---
title: "AIRR metadata prep"
editor: source
author: "Joe Boktor"
date: '2025-07-08'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

Environment setup

```{r setup, message=FALSE, warning=FALSE}
home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))

# Additional libraries
library(data.table)

# Plotting libraries
library(ggsci)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(gt)
library(gtsummary)

```


Loading metadata and TCRB data

```{r, eval=FALSE}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>%
  mutate(case_control_other_latest = 
    factor(case_control_other_latest, levels = c("Control", "Case"))) %>% 
  glimpse()

trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()

```



# Visualizing Analysis Metadata

How many samples are in the analysis? From which cohorts? Home many total donors? How many PD/Control?

```{r, eval=FALSE}
# 1. Unique participants by cohort and disease status (with totals)
participant_summary_with_totals <- metadata %>%
  dplyr::distinct(participant_id, study, case_control_other_latest) %>%
  dplyr::count(study, case_control_other_latest) %>%
  pivot_wider(
    names_from = case_control_other_latest,
    values_from = n,
    values_fill = 0
  ) %>%
  dplyr::rename(
    "Study" = study,
    "Control" = Control,
    "PD" = Case
  ) %>%
  dplyr::mutate(
    Total = Control + PD
  ) %>%
  gt() %>%
  tab_header(
    title = "Unique Participants by Cohort and Disease Status"
  ) %>%
  tab_style(
    style = cell_fill(color = "lightgray"),
    locations = cells_column_labels()
  ) %>%
  fmt_number(
    columns = c("Control", "PD", "Total"),
    decimals = 0
  )

# 2. Unique samples by cohort and disease status (with totals)
sample_summary_with_totals <- metadata %>%
  dplyr::count(study, case_control_other_latest) %>%
  pivot_wider(
    names_from = case_control_other_latest,
    values_from = n,
    values_fill = 0
  ) %>%
  dplyr::rename(
    "Study" = study,
    "Control" = Control,
    "PD" = Case
  ) %>%
  dplyr::mutate(
    Total = Control + PD
  ) %>%
  gt() %>%
  tab_header(
    title = "Unique Samples by Cohort and Disease Status"
  ) %>%
  tab_style(
    style = cell_fill(color = "lightgray"),
    locations = cells_column_labels()
  ) %>%
  fmt_number(
    columns = c("Control", "PD", "Total"),
    decimals = 0
  )

# # Display both tables
# participant_summary_with_totals
# sample_summary_with_totals

# Save both tables
participant_summary_with_totals %>%
  gtsave(filename = glue("{wkdir}/figures/tables/participant_summary_{Sys.Date()}.html"))

sample_summary_with_totals %>%
  gtsave(filename = glue("{wkdir}/figures/tables/sample_summary_{Sys.Date()}.html"))

```


```{r}
htmltools::includeHTML(glue("{wkdir}/figures/tables/participant_summary_2025-07-08.html"))

htmltools::includeHTML(glue("{wkdir}/figures/tables/sample_summary_2025-07-08.html"))

```


Visualzing demographic and clinical data for samples.

```{r, eval=FALSE}
# plot the distribution of total samples per participant
p_sample_counts <- metadata %>% 
    dplyr::summarise(n = n(), .by = "participant_id") %>% 
    left_join(metadata) %>% 
    ggplot(aes(x = n, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw(base_size = 16) +
    labs(x = "Samples per participant", 
         y = "Count",
         fill = "Disease Status")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/sample_counts_{Sys.Date()}.png"),
    plot = p_sample_counts,
    width = 4,
    height = 3.5
)

baseline_meta <- metadata %>% 
    select(participant_id, age_at_baseline, case_control_other_latest, sex, study) %>% 
    distinct()

p_age_density <- baseline_meta %>% 
    ggplot(aes(x = age_at_baseline, color = case_control_other_latest)) +
    geom_density(alpha = 0.5, size = 2) +
    scale_color_d3() +
    theme_bw(base_size = 16) +
    labs(x = "Age at Baseline",
         y = "Density",
         color = NULL) +
    theme(legend.position = "none")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/age_density_{Sys.Date()}.png"),
    plot = p_age_density,
    width = 4,
    height = 3.5
)

p_sex_dist <- baseline_meta %>%
    # Plot sex distribution
    ggplot(aes(x = sex, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw(base_size = 16) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Sex", 
         y = "Count",
         fill = "Disease Status")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/sex_dist_{Sys.Date()}.png"),
    plot = p_sex_dist,
    width = 3.5,
    height = 3.5
)

# Plot study distribution
p_study_dist <- baseline_meta %>%
    ggplot(aes(x = study, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw(base_size = 16) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Study", 
         y = "Count",
         fill = "Disease Status")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/study_dist_{Sys.Date()}.png"),
    plot = p_study_dist,
    width = 4,
    height = 3.5
)



fig2_panel <- (
  p_age_density +
  p_sex_dist   +
  p_study_dist +
  p_sample_counts
) +
  plot_layout(
    nrow = 1, guides = "collect",
    widths = c(1.3, 0.6, 1, 1.3)
  ) &
  theme(legend.position = "bottom")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/FIGURE2_top_panel_{Sys.Date()}.png"),
    plot = fig2_panel,
    width = 12,
    height = 4
)

```


Visualzing TCRB count distributions

```{r}
sid_counts <- trust4_reports_tcrb %>%
    group_by(sample_id, C) %>%
    dplyr::summarise(n = n()) %>%
    mutate(grouping = "sample") %>%
    dplyr::rename(unit = sample_id) %>%
    glimpse()
pid_counts <- trust4_reports_tcrb %>%
    group_by(participant_id, C) %>%
    dplyr::summarise(n = n()) %>%
    mutate(grouping = "donor") %>%
    dplyr::rename(unit = participant_id) %>%
    glimpse()
sid_study_pair <- metadata %>% select(sample_id, study) %>% distinct() %>% 
  dplyr::rename(unit = sample_id)
pid_study_pair <- metadata %>% select(participant_id, study) %>% distinct() %>% 
  dplyr::rename(unit = participant_id)

rec_counts_df <- bind_rows(sid_counts, pid_counts) %>%
    mutate(cell_type_origin = case_when(
        grepl("TR", C) ~ "TCR",
        # grepl("IG", C) ~ "BCR",
        TRUE ~ "ERROR"
    )) %>%
    left_join(bind_rows(sid_study_pair, pid_study_pair)) %>% 
    glimpse()


p_ecdf_TCRS <- rec_counts_df %>%
    filter(cell_type_origin == "TCR") %>% 
    mutate(grouping = factor(grouping, levels = c("sample", "donor"))) %>%
    ggplot(aes(n)) +
    stat_ecdf(aes(color = study), geom = "step", linewidth = 1, alpha = 0.8) +
    facet_grid(cell_type_origin ~ grouping, scales = "free") +
    scale_x_log10() +
    theme_bw() +
    labs(
        x = "Number of Receptors",
        y = "Cumulative Proportion",
        color = NULL
    ) +
    scale_color_nejm()
    # scale_color_manual(values = c(
    #     "TRAC" = "#a6cee3",
    #     "TRBC" = "#1f78b4",
    #     "TRDC" = "#fb9a99",
    #     "TRGC" = "#e31a1c"
    # ))

ggsave(
    filename = glue("{wkdir}/figures/AIRR/ecdf_TRUST4_TCRS_{Sys.Date()}.png"),
    plot = p_ecdf_TCRS,
    width = 5,
    height = 2.25
)


```



Visualize HLA genotype risk and protective status

```{r, eval=FALSE}

p_arcas_hla_risk <- metadata %>% 
  filter(DRB1_risk != "none") %>% 
  # Calculate counts for each group
  group_by(DRB1_grouped, DRB1_risk, case_control_other_latest) %>%
  summarise(count = n(), .groups = "drop") %>%
  # Create the plot
  ggplot(aes(x = fct_reorder(DRB1_grouped, count), y = count, fill = case_control_other_latest)) +
  geom_bar(stat = "identity") +
  # Add count labels on top of each bar
  # geom_text(aes(label = count),
  #           position = position_stack(vjust = 0.5),
  #           size = 3) +
  scale_fill_d3() +
  facet_grid(DRB1_risk ~ ., scales = "free", space = "free") +
  theme_bw() +
  labs(
    # title = "Distribution of Case/Control Status by DRB1 Genotype",
    x = NULL,
    y = "Count",
    fill = "Case/Control Status"
  ) +
  coord_flip() +
  theme(
    # axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    # Add left padding to prevent cutoff
    plot.margin = margin(t = 5, r = 5, b = 5, l = 40, unit = "pt"),
    strip.text.y   = element_text(angle = 0),
    strip.text.y.left  = element_text(angle = 0),
    strip.text.y.right = element_text(angle = 0)
  )

ggsave(
  glue("{wkdir}/figures/AIRR/arcasHLA/arcas_hla_risk_barplot_{Sys.Date()}.png"),
  p_arcas_hla_risk,
  width = 5,
  height = 3.5
)

```



Visualize Read depth vs clonotype count and clonoal volume

```{r}
metadata %>% glimpse()

clonotype_volume_df <- trust4_reports_tcrb %>% 
  group_by(sample_id) %>% 
  dplyr::summarize(tcrb_volume = sum(`#count`)) %>% 
  ungroup() %>% 
  glimpse()

clonotype_count_df <- trust4_reports_tcrb %>% 
  group_by(sample_id) %>% 
  dplyr::summarize(tcrb_count = n()) %>% 
  ungroup() %>% 
  glimpse()

count_volume_df <- clonotype_count_df %>% 
  left_join(clonotype_volume_df) %>% 
  left_join(metadata) %>% 
  glimpse()

lmm_model_count <- lme(aligned_reads ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = metadata,
    na.action = na.omit
)
broom.mixed::tidy(lmm_model_count)

lmm_model_tcellreads <- lme(est_tcell_reads_abis ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = metadata,
    na.action = na.omit
)
broom.mixed::tidy(lmm_model_tcellreads) %>% 
    DT::datatable()


p_aligned_reads_boxplot <- count_volume_df %>% 
    ggplot(aes(y = aligned_reads, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study, nrow = 1) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "Aligned Reads (Picard)", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

p_tcell_reads_boxplot <- count_volume_df %>% 
    ggplot(aes(y = est_tcell_reads_abis, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study, nrow = 1) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "T-Cell Estimated Reads (ABIS)", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

p_seq_depth_boxplots <- p_aligned_reads_boxplot + p_tcell_reads_boxplot
ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_boxplots_{Sys.Date()}.png"),
    plot = p_seq_depth_boxplots,
    width = 10,
    height = 4
)

# Interesting that aligned reads are consistent between cohorts and disease status but T-cell reads are significant in both conditions

# correlating Aligned reads with clonotype and clonal volume
p_seq_depth_tcell_abis <- count_volume_df %>% 
  ggplot(aes(x = est_tcell_reads_abis, y = tcrb_volume, color = study)) +
  geom_point(alpha = 0.5, size = 0.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  ggpubr::stat_cor(
    method = "pearson",
    label.x.npc = 0.7,
    label.y.npc = c(0.39, 0.37, 0.35, 0.33),
    aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),
    size = 3, # larger font
    digits = 2, # two decimal places
    p.accuracy = 0.001, # p-value accuracy
    show.legend = FALSE,
    parse = TRUE # pretty math notation
  ) +
  scale_color_jama() +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(text = NA))) +
  labs(x = "T-Cell Estimated Reads (ABIS)",
       y = "Clonotype Volume",
       color = "Study") +
  theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_tcell_abis_{Sys.Date()}.png"),
    plot = p_seq_depth_tcell_abis,
    width = 5,
    height = 4
)

p_seq_depth_aligned_reads <- count_volume_df %>% 
  ggplot(aes(x = aligned_reads, y = tcrb_volume, color = study)) +
  geom_point(alpha = 0.5, size = 0.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  ggpubr::stat_cor(
    method = "pearson",
    # label.x = 7.75, #label.y = 3.1,
    label.x.npc = 0.7,
    label.y.npc = c(0.39, 0.37, 0.35, 0.33),
    aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),
    size = 3, # larger font
    digits = 2, # two decimal places
    p.accuracy = 0.001, # p-value accuracy
    show.legend = FALSE,
    parse = TRUE # pretty math notation
  ) +
  scale_color_jama() +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(text = NA))) +
  labs(x = "Aligned Reads (Picard)",
       y = "Clonotype Volume",
       color = "Study") +
  theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_aligned_reads_{Sys.Date()}.png"),
    plot = p_seq_depth_aligned_reads,
    width = 5,
    height = 4
)


p_aligned_and_tcell_reads <- p_seq_depth_aligned_reads + p_seq_depth_tcell_abis

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_aligned_and_tcell_reads_{Sys.Date()}.png"),
    plot = p_aligned_and_tcell_reads,
    width = 10,
    height = 4
)


```



```{r}
sessionInfo()
```
