---
title: "ClusTCR Statistics"
editor: source
author: "Joe Boktor"
date: '2025-09-08'
eval: false
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

# ClusTCR Statistics

Environment setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(data.table)
library(seqinr)
library(progress)
library(immunarch)
library(nlme)
library(broom)
library(broom.mixed)
library(parallelDist)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(scico)
# library(glmmTMB)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))
source(paste0(src_dir, "/R_scripts/_airr_analyze-permanova.R"))

scratch_dir <- "/resnick/scratch/jbok/PDMBS"
gliph2_plots_dir <- glue("{wkdir}/figures/AIRR/gliph2/")
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
arcasHLA_interim <- glue("{wkdir}/data/interim/airr/arcasHLA")
gliph2_input_dir <- glue("{wkdir}/data/input/gliph2/param_files")
gliph2_res_dir <- glue("{wkdir}/data/interim/airr/gliph2/processing")
clustcr_dir <- glue("{wkdir}/data/interim/clusTCR")
clustcr_figs <- glue("{wkdir}/figures/clusTCR")
tcr_cdr3_riskscore_complied <- 
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")
cohort_subset_dir <- glue("{clustcr_dir}/results/cohort_subsets")

```

## Chi-squared test for PD enriched/depeleted clusters

Loading in compiled clusTCR results dataframe
```{r}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  # left_join(clonotype_volume_df, by = "sample_id") %>% 
  glimpse()

metadata_categories_df <- readRDS(
  glue("{wkdir}/data/interim/metadata/metadata_categories_dataframe_2025-07-21.rds")) %>% 
  glimpse()

clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds"))
clustcr_results_df %>% glimpse()

clustcr_results_df_filtered <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 9) %>% 
  glimpse()
clustcr_results_df_filtered %>% glimpse()
clustcr_results_df_filtered$cluster_id %>% unique() %>% length()
```


Executing Chi-squared test for PD enriched/depeleted clusters

```{r}
sample_groups_df <- clustcr_results_df_filtered %>%
    dplyr::select(participant_id, case_control_other_latest) %>%
    distinct() %>% glimpse()

grp_n <- sample_groups_df$case_control_other_latest %>% table()
case_n_base <- grp_n[["Case"]]
ctrl_n_base <- grp_n[["Control"]]
cidx_list <- unique(clustcr_results_df_filtered$cluster_id)

# 1.  unique donors per (cluster × group) in one pass 
per_cluster <- clustcr_results_df_filtered %>%
  dplyr::distinct(cluster_id, participant_id, case_control_other_latest) %>%
  dplyr::count(cluster_id, case_control_other_latest, name = "n") %>%
  pivot_wider(names_from = case_control_other_latest,
              values_from = n,
              values_fill = 0) %>%
  mutate(
    remCase = case_n_base - Case,
    remCtrl = ctrl_n_base - Control
  ) %>% 
  glimpse()

per_cluster %>% glimpse()

# per_cluster: data-frame with Case, Control, remCase, remCtrl columns
future::plan(future::multisession, workers = 4)
chi2_stats <- furrr::future_pmap_dfr(
  per_cluster %>% select(Case, Control, remCase, remCtrl),
  function(Case, Control, remCase, remCtrl) {
    mat <- matrix(c(Case, Control, remCase, remCtrl), nrow = 2)
    ## 1. χ² test (no Yates correction keeps it comparable to Fisher)
    ct <- suppressWarnings(chisq.test(mat, correct = FALSE))
    ## 2. Odds ratio for context
    or <- ((Case + 0.5)  * (remCtrl + 0.5)) / ((Control + 0.5) * (remCase + 0.5))
    list(
      chi2_p_value = ct$p.value,
      chi2_stat    = unname(ct$statistic),
      odds_ratio   = or,
      exp_case_cluster    = ct$expected[1],
      exp_control_cluster = ct$expected[2]
    )
  }
)
future::plan(sequential)

chi2_results_df <- bind_cols(per_cluster, chi2_stats) %>% 
    mutate(adj_chi2_p_value = stats::p.adjust(chi2_p_value, method = "fdr")) %>%
    arrange(chi2_p_value) %>% 
    glimpse()

chi2_results_df %>% 
    slice_head(n = 10, by = chi2_p_value) %>% 
    View()

saveRDS(chi2_results_df, file = 
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_{Sys.Date()}.rds"))


```


```{r}
chi2_results_df <- readRDS(
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_2025-09-08.rds")) %>% 
  glimpse()

cluster_sscore_filtered <- clustcr_results_df_filtered %>% 
  dplyr::select(cluster_id, Q_data, risk_score_p2_glrm_strat) %>% 
  group_by(cluster_id) %>% 
  dplyr::summarize(
    avg_risk_score = mean(risk_score_p2_glrm_strat, na.rm = TRUE),
    avg_selection_score = mean(log10(Q_data), na.rm = TRUE)) %>% 
  glimpse()

cluster_stats_filtered <- clustcr_results_df_filtered %>% 
  dplyr::select(cluster_id, motif, unique_donors_per_clust, unique_samples_per_clust, motif_length) %>% 
  distinct() %>% 
  left_join(cluster_sscore_filtered) %>% 
  glimpse()

# chi2_results_df %>% 
#   slice_min(chi2_p_value, n = 50) %>% 
#   View

chi2_results_df <- chi2_results_df %>% 
  left_join(cluster_stats_filtered) %>% 
  glimpse()

chi2_results_df %>% glimpse()

p_chi2_volcano <- chi2_results_df %>% 
  mutate(cluster_id = str_before_first(cluster_id, "_")) %>% 
  # Add a column to identify significant points
  mutate(
    significant = case_when(
      abs(log2(odds_ratio)) >= 0.5 & chi2_p_value <= 0.05 ~ case_when(
        odds_ratio > 1 ~ "Enriched in Cases",
        odds_ratio < 1 ~ "Enriched in Controls"
      ),
      TRUE ~ "Not Significant"
    ),
    # Create custom hover text
    hover_text = sprintf(
      "Cluster: %s\nP-value: %.2e\nOdds Ratio: %.2f\nUnique Donors: %d",
      cluster_id, chi2_p_value, odds_ratio, unique_donors_per_clust
    )
  ) %>%
  # Add a column to identify points to label
  mutate(
    label_text = case_when(
      # Top 15 most significant enriched in cases
      significant == "Enriched in Cases" & 
        chi2_p_value <= (arrange(filter(., significant == "Enriched in Cases"), chi2_p_value) %>% 
                         slice_head(n = 15) %>% 
                         pull(chi2_p_value) %>% 
                         max()) ~ cluster_id,
      # Top 8 most significant enriched in controls
      significant == "Enriched in Controls" & 
        chi2_p_value <= (arrange(filter(., significant == "Enriched in Controls"), chi2_p_value) %>% 
                         slice_head(n = 8) %>% 
                         pull(chi2_p_value) %>% 
                         max()) ~ cluster_id,
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(x = log2(odds_ratio), y = -log10(chi2_p_value), 
             color = significant, text = hover_text)) +
  geom_point(alpha = 0.5, aes(size = unique_donors_per_clust)) +
  # Add ggrepel text labels
  geom_label_repel(
    aes(label = label_text),
    size = 3,
      force = 100,
    max.overlaps = Inf,
    box.padding = 0.5,
    segment.color = "grey50",
    show.legend = FALSE,
    segment.curvature = 0.3,
    segment.ncp = 3,
    segment.angle = 10
  ) +
  scale_color_manual(
    values = c(
      "Enriched in Cases" = "#FF7F0F",
      "Enriched in Controls" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  labs(
    title = "Chi-squared test for PD enriched/depleted clusters",
    x = expression(log[2]~"(Odds Ratio)"),
    y = expression(-log[10]~"(P-value)"),
    size = "Unique Donors per Cluster",
    color = NULL

  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    # legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  ) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "darkgrey") +
  geom_vline(xintercept = c(-0.5, 0.5), 
             linetype = "dashed", 
             color = "darkgrey") +
  coord_cartesian(clip = "off")

ggsave(
  glue("{clustcr_figs}/clusTCR_chi2_volcano_odds_ratio_pvalue_{Sys.Date()}.png"),
  p_chi2_volcano,
  width = 9,
  height = 6,
  dpi = 600
)


# Examining relationship between odds ratio and selection score
chi2_results_df %>% 
  filter(chi2_p_value <= 0.05) %>% 
  ggplot(aes(x = log2(odds_ratio), y = avg_selection_score)) +
  geom_point(alpha = 0.5) +
  theme_bw()

chi2_results_df %>% 
  filter(chi2_p_value <= 0.05) %>% 
  ggplot(aes(x = log2(odds_ratio), y = avg_risk_score)) +
  geom_point(alpha = 0.5) +
  theme_bw()
# no obvious relationship between odds ratio and selection score or risk score


```

## Assessing Chi-squared test cluster generalizability

```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds")
  )
clustcr_results_df %>% glimpse()

```


```{r}
summary_paths <- list.files(
  glue("{clustcr_dir}/results/cohort_subsets"), pattern = "_summary.csv", 
  full.names = TRUE, recursive = TRUE
)

full_summary_path <- glue("{clustcr_dir}/results/clusTCR_summary_vgene_restricted_31cpus.csv")

summary_ls <- c(full_summary_path, summary_paths) %>% 
  purrr::set_names(basename(.)  %>% gsub("_summary.csv", "", .)) %>% 
  purrr::map(~ read_csv(.) %>% pull(motif))
names(summary_ls)[1] <- "full_dataset"
names(summary_ls)

comparison_df <- 
  combn(names(summary_ls), 2, simplify = FALSE) %>% 
  map_dfr(~ tibble(Var1 = .x[1], Var2 = .x[2])) %>%
  glimpse()

comparison_df <- comparison_df %>% 
  mutate(Var1_counts = map_int(Var1, ~ length(summary_ls[[.x]])),
         Var2_counts = map_int(Var2, ~ length(summary_ls[[.x]])),
         overlap = map2_int(Var1, Var2, ~ length(intersect(summary_ls[[.x]], summary_ls[[.y]]))),
         overlap_percent_var1 = (overlap / Var1_counts) * 100,
         overlap_percent_var2 = (overlap / Var2_counts) * 100
         ) %>% 
  glimpse()

# THis doesn't consider V-gene restriction -- not the best way to do this
View(comparison_df)


```


```{r}
# Saving pre-processed results ----
aggregate_clustcr_results <- function(clustcr_output_file, clustcr_summary_file, output_path) {
  clustcr_output <- read_csv(clustcr_output_file, col_select = -1)
  clustcr_summary <- read_csv(clustcr_summary_file) %>% dplyr::rename(cluster = ...1)
  clustcr_summary$cluster  <- unique(clustcr_output$cluster)

  clustcr_results_df <- clustcr_output %>%
    dplyr::left_join(clustcr_summary) %>%
    dplyr::left_join(tcrb_df, 
      by = c("junction_aa" = "CDR3b_aa", "v_call" = "TRBV"),
      relationship = "many-to-many") %>%
    dplyr::group_by(cluster) %>%
    dplyr::mutate(unique_donors_per_clust = dplyr::n_distinct(participant_id),
      unique_samples_per_clust = dplyr::n_distinct(sample_id)
    ) %>%
    dplyr::ungroup() %>% 
    dplyr::mutate(motif_length = motif %>% 
          gsub("\\[.*?\\]", "X", .) %>%  # Replace bracketed content with single X
          gsub("\\.", "", .) %>%         # Remove dots
          gsub("[^A-Z]", "", .) %>%      # Keep only uppercase letters
          nchar()
    ) %>% 
    dplyr::mutate(cluster_id = glue("{motif}_{v_call}")) %>% 
    # dplyr::mutate(cluster_id = glue("c{cluster}_{motif}")) %>% 
    glimpse()

  saveRDS(clustcr_results_df, output_path)
}

# Aggregating clustcr results for each cohort subset
for (spath in summary_paths) {
  sname <- spath %>% basename(.) %>% gsub("_summary.csv", "", .) %>% print()
  aggregate_clustcr_results(
    clustcr_output_file = spath %>% gsub("_summary", "_clusters", .),
    clustcr_summary_file = spath,
    output_path = glue("{cohort_subset_dir}/{sname}/{sname}_aggregated.rds")
  )
}

# Running Chi-squared test on each dataset
run_dataset_chi2 <- function(clustcr_results_path, outputdir = NULL) {
  sname <- clustcr_results_path %>% basename(.) %>% gsub("_aggregated.rds", "", .)
  message(glue("Running Chi-squared test on {sname}... \n"))
  output_path <- glue("{clustcr_dir}/results/cohort_subsets/{sname}/{sname}_chi2_results_df_{Sys.Date()}.rds")

  if (file.exists(output_path)) return()
  
  clustcr_results_df <- readRDS(clustcr_results_path)
  clustcr_results_df_filtered <- clustcr_results_df %>% 
    filter(unique_donors_per_clust >= 4) %>% 
    filter(motif_length >= 9)

  sample_groups_df <- clustcr_results_df_filtered %>%
      dplyr::select(participant_id, case_control_other_latest) %>%
      distinct()

  grp_n <- sample_groups_df$case_control_other_latest %>% table()
  case_n_base <- grp_n[["Case"]]
  ctrl_n_base <- grp_n[["Control"]]
  cidx_list <- unique(clustcr_results_df_filtered$cluster_id)

  # 1.  unique donors per (cluster × group) in one pass 
  per_cluster <- clustcr_results_df_filtered %>%
    dplyr::distinct(cluster_id, participant_id, case_control_other_latest) %>%
    dplyr::count(cluster_id, case_control_other_latest, name = "n") %>%
    pivot_wider(names_from = case_control_other_latest,
                values_from = n,
                values_fill = 0) %>%
    mutate(
      remCase = case_n_base - Case,
      remCtrl = ctrl_n_base - Control
    ) %>% 
    glimpse()

  # per_cluster: data-frame with Case, Control, remCase, remCtrl columns
  future::plan(future::multisession, workers = 15)
  chi2_stats <- furrr::future_pmap_dfr(
    per_cluster %>% dplyr::select(Case, Control, remCase, remCtrl),
    function(Case, Control, remCase, remCtrl) {
      mat <- matrix(c(Case, Control, remCase, remCtrl), nrow = 2)
      ## 1. χ² test (no Yates correction keeps it comparable to Fisher)
      ct <- suppressWarnings(chisq.test(mat, correct = FALSE))
      ## 2. Odds ratio for context
      or <- ((Case + 0.5)  * (remCtrl + 0.5)) / ((Control + 0.5) * (remCase + 0.5))
      list(
        chi2_p_value = ct$p.value,
        chi2_stat    = unname(ct$statistic),
        odds_ratio   = or,
        exp_case_cluster    = ct$expected[1],
        exp_control_cluster = ct$expected[2]
      )
    }
  )

  chi2_results_df <- dplyr::bind_cols(per_cluster, chi2_stats) %>% 
      dplyr::mutate(
        adj_chi2_p_value = stats::p.adjust(chi2_p_value, method = "fdr")) %>%
      dplyr::arrange(chi2_p_value) %>% 
      glimpse()

  if (is.null(outputdir)) {
    outputdir <- dirname(clustcr_results_path)
  }
  saveRDS(chi2_results_df, 
    glue("{outputdir}/{sname}_chi2_results_df_{Sys.Date()}.rds"))

}


# Separating out individual cohorts from the full clustering - aggregating - and then running Chi2 test
clustcr_results_df %>% glimpse()
cohort_from_full <- list()
for (cohort_name in unique(metadata$study)) {
  cohort_pids <- metadata %>% filter(study == cohort_name) %>% pull(participant_id) %>% unique()
  cohort_from_full[[cohort_name]] <- clustcr_results_df %>% 
    filter(participant_id %in% cohort_pids) %>% 
    dplyr::group_by(cluster) %>%
    dplyr::mutate(
      unique_donors_per_clust = dplyr::n_distinct(participant_id),
      unique_samples_per_clust = dplyr::n_distinct(sample_id)
    ) %>%
    dplyr::ungroup() %>% 
    filter(unique_donors_per_clust >= 4) %>% 
    glimpse()
  
  outdir <- glue("{cohort_subset_dir}/ONLY_{cohort_name}_from_full_clustering/")
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
  saveRDS(cohort_from_full[[cohort_name]], 
    glue("{outdir}/{cohort_name}_from_full_clustering_aggregated.rds")
      )
}


# Running Chi2 for each dataset
clustcr_subset_aggpaths <- list.files(
    cohort_subset_dir, full.names = TRUE, 
    recursive = TRUE, pattern = "aggregated.rds")

clustcr_subset_aggpaths %>% 
  purrr::walk(run_dataset_chi2)


# future::plan(sequential)
```


```{r}
clustcr_subset_chi2_paths <- list.files(
    cohort_subset_dir, full.names = TRUE, 
    recursive = TRUE, pattern = "_chi2_results_df_2025-08-25"
    )
clustcr_subset_chi2_paths_onlyfromfull <- list.files(
    cohort_subset_dir, full.names = TRUE, 
    recursive = TRUE, pattern = "_from_full_clustering_chi2_results_df_2025-08-25"
    )

# 

# clustcr_subset_chi2_paths
all_except_paths <- clustcr_subset_chi2_paths %>% keep(~ grepl("ALL_EXCEPT", .))
names(all_except_paths) <- all_except_paths %>% dirname() %>% basename()

only_paths <- clustcr_subset_chi2_paths_onlyfromfull
names(only_paths) <- only_paths %>% dirname() %>% basename()

# For each dataset at a time 
# study_names <- unique(metadata$study)
# study_name <- "HBS"
# all_except_df <- all_except_paths[[glue("ALL_EXCEPT_{study_name}")]] %>% 
#   readRDS() %>% 
#   dplyr::rename_at(vars(-cluster_id), ~ glue("{.}_ALL_EXCEPT")) %>% 
#   glimpse()

# only_df <- only_paths[[glue("ONLY_{study_name}_from_full_clustering")]] %>% 
#   readRDS() %>% 
#   dplyr::rename_at(vars(-cluster_id), ~ glue("{.}_ONLY")) %>% 
#   glimpse()

# full_df <- full_join(all_except_df, only_df) %>% 
#   glimpse()

# aset <- all_except_df$cluster_id %>%
#   unique()
# bset <- only_df$cluster_id %>% 
#   unique()
# intersect_set <- intersect(aset, bset)

# length(aset)
# length(bset)
# length(intersect_set)
# full_df %>% glimpse
# full_df %>% 
#   # filter(chi2_p_value_ALL_EXCEPT <= 0.05) %>% 
#   ggplot(aes(x = log2(odds_ratio_ALL_EXCEPT), y = log2(odds_ratio_ONLY))) +
#   geom_point() +
#   geom_abline(slope = 1, intercept = 0) +
#   theme_minimal()

# full dataset chi2 results
chi2_results_df <- readRDS(
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_2025-09-08.rds")) %>% 
  glimpse()

chi2_results_df$cluster_id %>% unique() %>% length()

# top 25 hits
top35full <- chi2_results_df %>% slice_min(chi2_p_value, n = 35) %>% 
  pull(cluster_id)

# Identifying clusters that are significant in at least two separate cohort_subset_dir
only_dfs <- only_paths %>% 
  purrr::map(readRDS) %>% 
  bind_rows(.id = "dataset") %>% 
  mutate(sig_dir = case_when(
    log2(odds_ratio) > 0.5 ~ "PD UP",
    log2(odds_ratio) < -0.5 ~ "PD DN",
    TRUE ~ "TOSS"
  )) %>% 
  glimpse()

cluster_hits <- only_dfs %>% 
  filter(chi2_p_value <= 0.05) %>% 
  group_by(cluster_id, sig_dir) %>% 
  summarise(n = n(), .groups = "drop") %>%
  filter(sig_dir != "TOSS") %>% 
  filter(n >= 2) %>% 
  pull(cluster_id) %>% 
  unique()
length(cluster_hits)


only_dfs %>% glimpse()

p_all_reprod_cids <- only_dfs %>% 
  filter(cluster_id %in% cluster_hits) %>% 
  mutate(color_var = case_when(
    (sig_dir == "PD UP" & chi2_p_value <= 0.05) ~ "Enriched in PD",
    (sig_dir == "PD DN" & chi2_p_value <= 0.05) ~ "Enriched in Control",
    TRUE ~ "Not Significant"
  )) %>% 
  mutate(cohort = gsub("ONLY_|_from_full_clustering", "", dataset)) %>%
  ggplot(aes(x = log2(odds_ratio), y = fct_reorder(cluster_id, log2(odds_ratio)))) +
  geom_bar(stat = "identity", aes(fill = color_var)) +
  facet_wrap(~ cohort, nrow = 1) +
  labs( x = expression(log[2]~"Odds Ratio"), y = "Cluster ID", fill = NULL) +
  scale_fill_manual(
    values = c(
      "Enriched in PD" = "#FF7F0F",
      "Enriched in Control" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  theme_bw() +
  theme(axis.text.y = element_blank(),
    panel.grid = element_blank()
    )

ggsave(
  glue("{clustcr_figs}/clusTCR_cohort_reproducible_clusters_ALL_{Sys.Date()}.png"),
  p_all_reprod_cids,
  width = 10,
  height = 4
)

p_top35_reprod_cids <- only_dfs %>% 
  filter(cluster_id %in% cluster_hits) %>% 
  filter(cluster_id %in% top35full) %>% 
  mutate(color_var = case_when(
    (sig_dir == "PD UP" & chi2_p_value <= 0.05) ~ "Enriched in PD",
    (sig_dir == "PD DN" & chi2_p_value <= 0.05) ~ "Enriched in Control",
    TRUE ~ "Not Significant"
  )) %>% 
  mutate(cohort = gsub("ONLY_|_from_full_clustering", "", dataset)) %>%
  ggplot(aes(x = log2(odds_ratio), y = fct_reorder(cluster_id, log2(odds_ratio)))) +
  geom_bar(stat = "identity", aes(fill = color_var), width = 0.5) +
  facet_wrap(~ cohort, nrow = 1) +
  labs( x = expression(log[2]~"Odds Ratio"), y = "Cluster ID", fill = NULL) +
  scale_fill_manual(
    values = c(
      "Enriched in PD" = "#FF7F0F",
      "Enriched in Control" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  theme_bw()

ggsave(
  glue("{clustcr_figs}/clusTCR_cohort_reproducible_clusters_TOP_{Sys.Date()}.png"),
  p_top35_reprod_cids,
  width = 10,
  height = 4
)

# only_dfs %>% glimpse()
# only_dfs %>% 
#   filter(cluster_id %in% cluster_hits) %>% View
#   filter(cluster_id %in% top35full) %>% 
#   View()
```



## Logistic Regression Models for Cluster Membership  ----

Now running a logistic regression model to test for PD enriched/depeleted clusters

```{r}
clustcr_results_df_filtered %>% glimpse()
clonotype_volume_df_pid <- trust4_reports_tcrb %>% 
  group_by(participant_id) %>% 
  dplyr::summarize(
      clonal_volume = sum(`#count`),
      clonal_count = n()
      ) %>%
  glimpse()


metadata %>% glimpse()
model_meta_df <- metadata %>% 
  # dplyr::select(participant_id, age_at_baseline, sex, study, est_tcell_reads_abis, case_control_other_latest) %>%
  dplyr::group_by(participant_id, age_at_baseline, sex, study, case_control_other_latest, MHC_all_PC1, MHC_all_PC2, MHC_all_PC3) %>%
  dplyr::summarize(est_tcell_reads_abis_pid = sum(est_tcell_reads_abis)) %>% 
  distinct() %>% 
  glimpse()

saveRDS(model_meta_df,
  glue("{clustcr_dir}/results/metadata_for_logreg_{Sys.Date()}.rds")
)

per_cluster_norm_pid <- clustcr_results_df_filtered %>%
  as.data.table() %>% 
  dplyr::group_by(cluster_id, case_control_other_latest, participant_id) %>%
  dplyr::summarize(count = sum(count)) %>%
  left_join(clonotype_volume_df_pid) %>% 
  mutate(norm_count = count / clonal_volume) %>% 
  left_join(model_meta_df) %>% 
  glimpse()

# per_cluster_norm_pid$cluster_id %>% unique() %>% length()
# cluster_df <- per_cluster_norm_pid %>% filter(cluster_id == "c0_CSARD.g.SGANVLTF") %>% 
#   glimpse()

run_cluster_logreg_analysis <- function(cluster_df) {
  require(magrittr)
  cid <- unique(cluster_df$cluster_id)
  clustcr_dir <- 
    "/resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/clusTCR"
  model_meta_df <- readRDS(
    glue::glue("{clustcr_dir}/results/metadata_for_logreg_2025-08-17.rds")
  )
  cluster_df <- cluster_df %>% 
    dplyr::full_join(model_meta_df) %>% 
    dplyr::mutate(norm_count = tidyr::replace_na(norm_count, 0))

  # Models ----
  null_model <- glm(
    # norm_count ~ est_tcell_reads_abis_pid + age_at_baseline + sex,
    # norm_count ~ age_at_baseline + sex,
    norm_count ~ MHC_all_PC1 + MHC_all_PC2 + MHC_all_PC3 + est_tcell_reads_abis_pid,
    data = cluster_df,
    family = binomial(link = "logit")
  )
  full_model <- glm(
    # norm_count ~ case_control_other_latest + est_tcell_reads_abis_pid + age_at_baseline + sex,
    # norm_count ~ case_control_other_latest + age_at_baseline + sex,
    norm_count ~ case_control_other_latest + MHC_all_PC1 + MHC_all_PC2 + MHC_all_PC3 + est_tcell_reads_abis_pid,
    data = cluster_df,
    family = binomial(link = "logit")
  )

  model_comp <- anova(full_model, null_model, test = "LRT")
  lrt_res <- model_comp %>% broom::tidy()

  logreg_anova_df <- lrt_res[2,] %>% 
    dplyr::mutate(full_model = lrt_res$term[1]) %>% 
    dplyr::mutate(cluster_id = cid)

  logreg_reslist <- list(
      "cluster_id" = cid,
      "logit_null" = null_model,
      "logit_full" = full_model,
      "comparison_res" = model_comp,
      "logit_anova_df" = logreg_anova_df
  )
  saveRDS(logreg_reslist, 
    glue::glue("{clustcr_dir}/results/logistic_regression_age-sex-covars/",
      "{cid}__{Sys.Date()}.rds")
      )
}

cluster_list <- per_cluster_norm_pid %>% 
  ungroup() %>% 
  dplyr::group_split(cluster_id) %>% 
  setNames(unique(per_cluster_norm_pid$cluster_id))
length(cluster_list)
names(cluster_list) %>% head()

# unrun_clusters <- names(cluster_list) %>% 
#   discard(~ file.exists(glue("{clustcr_dir}/results/logistic_regression/{.}__2025-08-17.rds")))
# unrun_clusters %>% length()

unrun_clusters <- names(cluster_list) %>% 
  discard(~ file.exists(glue("{clustcr_dir}/results/logistic_regression/{.}__2025-08-17.rds")))
unrun_clusters %>% length()

```

Executing Logistic Regression Models for Cluster Membership (WEIGHTED BY CLONAL VOLUME)

```{r}
# future::plan(future::multisession, workers = 15)
# options(future.globals.maxSize = 2 * 1024^3)  # 2GB per worker
# cluster_list %>% 
# # cluster_list[unrun_clusters] %>%
#   furrr::future_walk(run_cluster_logreg_analysis, .options = furrr_options(seed = TRUE))
# future::plan(sequential)

# Running on cluster in parallel
# configure registry ----
breg <- breg_setup(cluster_name = "clusTCR_logreg", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )
# Submit Jobs ----
jobs <- batchMap(
  fun = run_cluster_logreg_analysis,
  args = list(cluster_list),
  reg = breg
)
jobs[, chunk := chunk(job.id, chunk.size = 40)]
print(jobs[, .N, by = chunk])

# can probably do two per minute so x jobs / 2 = minutes
submitJobs(jobs,
  resources = list(
    walltime = "1:00:00",
    memory = "5GB",
    ncpus = 1,
    max.concurrent.jobs = 9999
  )
)

```



Aggregating and visualizing Logistic Regression results

```{r}
# Utility funcs
pull_logreg_anova_df <- function(res) {
  res$logit_anova_df
}

pull_logreg_full_model <- function(res) {
  res$logit_full %>% broom::tidy()
}
```


```{r}
future::plan(future::multisession, workers = 15)
options(future.globals.maxSize = 2 * 1024^3) 

logreg_reslist <- list.files(
  glue("{clustcr_dir}/results/logistic_regression"),
  full.names = TRUE)

logreg_anova_df <- logreg_reslist %>% 
  # purrr::map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  furrr::future_map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  bind_rows() %>% 
  glimpse()
logreg_anova_df %>% slice_min(p.value, n = 20)
saveRDS(
  logreg_anova_df,
  glue("{clustcr_dir}/results/logistic_regression_anova_df_{Sys.Date()}.rds")
)

full_models_df <- logreg_reslist %>% 
  purrr::set_names(basename(.)  %>% str_before_first("__")) %>% 
  purrr::map(~ readRDS(.) %>% pull_logreg_full_model()) %>% 
  bind_rows(.id = "cluster_id") %>% 
  glimpse()
saveRDS(
  full_models_df,
  glue("{clustcr_dir}/results/logistic_regression_full_models_df_{Sys.Date()}.rds")
)

#______________________________________________________________________________
# Age / sex covars only ----
logreg_reslist_agesex <- list.files(
  glue("{clustcr_dir}/results/logistic_regression_age-sex-covars"),
  full.names = TRUE)
length(logreg_reslist_agesex)

logreg_anova_df <- logreg_reslist_agesex %>% 
  # purrr::map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  furrr::future_map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  bind_rows() %>% 
  glimpse()
logreg_anova_df %>% slice_min(p.value, n = 20)
saveRDS(
  logreg_anova_df,
  glue("{clustcr_dir}/results/logistic_regression_anova_df_agesex_{Sys.Date()}.rds")
)

full_models_df <- logreg_reslist_agesex %>% 
  purrr::set_names(basename(.)  %>% str_before_first("__")) %>% 
  furrr::future_map(~ readRDS(.) %>% pull_logreg_full_model()) %>% 
  bind_rows(.id = "cluster_id") %>% 
  glimpse()
saveRDS(
  full_models_df,
  glue("{clustcr_dir}/results/logistic_regression_full_models_df_agesex_{Sys.Date()}.rds")
)

#______________________________________________________________________________
# MHC seqdpeth covars only ----
logreg_reslist_agesex <- list.files(
  glue("{clustcr_dir}/results/logistic_regression_age-sex-covars"),
  full.names = TRUE)
length(logreg_reslist_agesex)

logreg_anova_df <- logreg_reslist_agesex %>% 
  # purrr::map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  furrr::future_map(~ readRDS(.x) %>% pull_logreg_anova_df()) %>% 
  bind_rows() %>% 
  glimpse()
logreg_anova_df %>% slice_min(p.value, n = 20)
saveRDS(
  logreg_anova_df,
  glue("{clustcr_dir}/results/logistic_regression_anova_df_mhcseqdepth_{Sys.Date()}.rds")
)

full_models_df <- logreg_reslist_agesex %>% 
  purrr::set_names(basename(.)  %>% str_before_first("__")) %>% 
  furrr::future_map(~ readRDS(.) %>% pull_logreg_full_model()) %>% 
  bind_rows(.id = "cluster_id") %>% 
  glimpse()
saveRDS(
  full_models_df,
  glue("{clustcr_dir}/results/logistic_regression_full_models_df_mhcseqdepth_{Sys.Date()}.rds")
)

future::plan(sequential)
```

```{r}
full_models_df %>% glimpse()
full_models_df %>% 
  filter(term == "case_control_other_latestControl") %>% 
  slice_min(p.value, n = 20) %>% 
  View()

full_models_df %>% 
  filter(term == "case_control_other_latestCase") %>% 
  slice_min(p.value, n = 20) %>% 
  View()

```




## PERMANOVA screening for metadata explained variance in cluster membership

NOTE: This section uses data from the clusTCR distances notebook


Subsetting distance matrices into random subsamples of samples
```{r}
# Saving subsampled distance matrix and metadata
subsample_metadata_and_distmat <- function(iter, permanova_metadata, distance_matrix, outdir) {
    set.seed(iter)
    sid_subsample <- permanova_metadata %>% 
        rownames_to_column(var = "sample_id") %>%
        group_by(participant_id) %>%
        dplyr::slice_sample(n = 1) %>%
        ungroup() %>%
        pull(sample_id)
    
    common_samples <- intersect(rownames(distance_matrix), rownames(permanova_metadata))
    i_common_samples <- intersect(common_samples, sid_subsample)
    distance_matrix_filtered <- as.dist(distance_matrix[i_common_samples, i_common_samples])
    permanova_metadata_filtered <- permanova_metadata[i_common_samples, ]
    
    # Saving distance matrix and metadata
    saveRDS(distance_matrix_filtered,
        glue("{outdir}/distance_matrix_filtered_subsample_{iter}.rds")
    )
    saveRDS(permanova_metadata_filtered,
        glue("{outdir}/permanova_metadata_filtered_subsample_{iter}.rds")
    )
}

# full vs pd only
# binary and weighted matrices

clustcr_mat_w_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_SSxRS_2025-08-18.rds"))
clustcr_mat_b_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_SSxRS_2025-08-18.rds"))
clustcr_mat_w_eucdist_input <- clustcr_mat_w_eucdist %>% 
  as.matrix() %>% as.data.frame()
clustcr_mat_b_eucdist_input <- clustcr_mat_b_eucdist %>% 
  as.matrix() %>% as.data.frame()

clustcr_perm_outputs <- c(
  "weighted_SSxRS_PD_only",
  "binary_SSxRS_PD_only",
  "weighted_SSxRS_full",
  "binary_SSxRS_full"
)

target_metadata <- colnames(metadata) %>% 
    discard(. %in% c("sample_id"))

# Get sample metadata for PERMANOVA
permanova_metadata <- metadata %>%
    filter(sample_id %nin% low_vol_samples$sample_id) %>%
    column_to_rownames("sample_id") %>%
    select(all_of(target_metadata), participant_id) %>%
    glimpse()

clustcr_mat_b_eucdist_input[pd_sids, pd_sids] %>% glimpse()
clustcr_mat_w_eucdist_input[pd_sids, pd_sids] %>% glimpse()

for (output in clustcr_perm_outputs) {
  output_path <- glue("{clustcr_dir}/results/permanova/inputs/{output}")
  dir.create(output_path, showWarnings = FALSE, recursive = TRUE)

  if (grepl("weighted", output)) {
    distance_matrix <- clustcr_mat_w_eucdist_input
  } else {
    distance_matrix <- clustcr_mat_b_eucdist_input
  }
  if (grepl("PD_only", output)) {
    pd_sids <- metadata %>%
      filter(sample_id %nin% low_vol_samples$sample_id) %>%
      filter(case_control_other_latest == "Case") %>% 
      filter(sample_id %in% rownames(distance_matrix)) %>%
      pull(sample_id) %>% 
      unique()
    distance_matrix <- distance_matrix[pd_sids, pd_sids]
  }

  for (i in 1:50) {
    message(glue("Subsampling {output} for iteration {i}"))
    subsample_metadata_and_distmat(i, permanova_metadata, distance_matrix, output_path)
  }
}

```

Executing slurm batchtools for PERMANOVA

```{r}
perm_resdir <- glue("{clustcr_dir}/results/permanova/results")
clustcr_perm_outputs %>% purrr::map( ~ {
  dir.create(glue("{perm_resdir}/{.}"), showWarnings = FALSE, recursive = TRUE)
})

batch_params <- 
  expand.grid("var" = target_metadata, "i" = 1:50, "output" = clustcr_perm_outputs) %>% 
    mutate(
        n_perms = 9999,
        n_cores = 1,
        dmat_path = glue("{clustcr_dir}/results/permanova/inputs/{output}/distance_matrix_filtered_subsample_{i}.rds"),
        meta_path = glue("{clustcr_dir}/results/permanova/inputs/{output}/permanova_metadata_filtered_subsample_{i}.rds"),
        model_formula_string = glue("distance_matrix ~ {var}"),
        outpath = glue("{perm_resdir}/{output}/",
            "permanova_{var}_nperms_{n_perms}_subsample_{i}_{Sys.Date()}.rds")
    ) %>%
    filter(var != "participant_id") %>% 
    dplyr::select(-c(var, i, output)) %>% 
    glimpse()

# configure registry ----
breg <- breg_setup(cluster_name = "PERM_ClusTCR", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )

# Submit Jobs ----
jobs <- batchMap(
  fun = run_permanova,
  args = batch_params,
  reg = breg
)
nrow(batch_params)

# 221600 / 1000
# (120 * 15) / 60 - needs about 30 hours for 120 jobs with 4 cores each
jobs[, chunk := chunk(job.id, chunk.size = 120)]
print(jobs[, .N, by = chunk])

submitJobs(jobs,
  resources = list(
    walltime = "2-01:00:00",
    memory = "10GB",
    ncpus = 4,
    max.concurrent.jobs = 9999
  )
)

```


Aggregating PERMANOVA results

```{r}
perm_resdir <- glue("{clustcr_dir}/results/permanova/results")

res_files <- list.files(perm_resdir, full.names = TRUE, recursive = TRUE)
length(res_files)
# res_files[1] %>% readRDS() %>% broom::tidy()

# loading in metadata categories


cond <- "binary_SSxRS_PD_only"
# cond <- "weighted_SSxRS_PD_only"
perm_resdir_w_full <- glue("{clustcr_dir}/results/permanova/results/{cond}")
res_files <- fs::dir_ls(perm_resdir_w_full, recurse = TRUE, type = "file")
length(res_files)

future::plan("multisession", workers = 8)
permanova_results <- res_files %>%
    purrr::set_names(basename(.)) %>%
    furrr::future_map(~ readRDS(.) %>% broom::tidy()) %>%
    bind_rows(.id = "model_info") %>%
    janitor::clean_names() %>%
    mutate(r2_percent = round(r2 * 100, 3)) %>%
    arrange(p_value) %>%
    glimpse()

passing_meta <- permanova_results %>%   
    filter(term %nin% c("Residual", "Total")) %>% 
    filter(p_value <= 0.05) %>% 
    pull(term) %>% unique()

p_permanova <- permanova_results %>%
    filter(term %nin% c("Residual", "Total")) %>%
    filter(term %in% passing_meta) %>% 
    arrange(p_value) %>%
    mutate(
        significant = p_value < 0.05,
        p_value_formatted = format.pval(p_value, digits = 3)
    ) %>%
    ggplot(aes(y = fct_reorder(term, r2), x = r2 * 100)) +
    geom_point(aes(color = significant),
        size = 0.8,
        position = position_jitter(height = 0.2, width = 0), alpha = 0.85
    ) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.5) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    labs(
        title = "PERMANOVA Results: TCR Repertoire Similarity",
        x = "R² % (Effect Size)",
        y = NULL
    ) +
    theme(legend.position = "bottom")

# ggsave(
#     glue("{clustcr_figs}/permanova_results_{cond}_{Sys.Date()}.png"),
#     p_permanova,
#     width = 8,
#     height = 20
# )

# Summarize
perm_summary_df <- permanova_results %>%
  group_by(term) %>%
  summarise(
    mean_log10_p = mean(-log10(p_value), na.rm = TRUE),
    sd_log10_p = sd(-log10(p_value), na.rm = TRUE),
    mean_r2_percent = mean(r2 * 100, na.rm = TRUE),
    sd_r2_percent = sd(r2 * 100, na.rm = TRUE),
    mean_statistic = mean(statistic, na.rm = TRUE),
    sd_statistic = sd(statistic, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  filter(!grepl("guid|GUID", term)) %>% 
  left_join(metadata_categories_df, by = c("term" = "metadata")) %>% 
  glimpse()

# Flag terms above threshold
pval_thresh <- -log10(0.05)
hits <- perm_summary_df %>%
  filter(mean_log10_p > pval_thresh) %>% 
  glimpse()
# View(hits)

labeled_points <- hits %>%
  # filter(metadata_class %nin% c("Biomarkers Genetics", "Biomarkers Omics")) %>%
  filter(metadata_class %in% c("Motor Severity", "Clinical History", "Cognition")) %>%
  slice_max(mean_r2_percent, n = 30, with_ties = FALSE) %>% 
  filter(grepl("moca|mds_updrs|hoehn|levodopa|medication", term)) %>%
  glimpse()
# View(labeled_points)

metadata_class_color_var <- metadata_categories_df %>% 
  select(metadata_class, class_color) %>% 
  distinct() %>% 
  deframe()
metadata_subclass_color_var <- metadata_categories_df %>% 
  select(metadata_subclass, class_color) %>% 
  distinct() %>% 
  deframe()

# Plot
p_perm_base <- ggplot(perm_summary_df, aes(x = mean_log10_p, y = mean_r2_percent)) +
  geom_errorbar(aes(ymin = mean_r2_percent - sd_r2_percent,
                    ymax = mean_r2_percent + sd_r2_percent),
                width = 0.01, alpha = 0.1) +
  geom_errorbarh(aes(xmin = mean_log10_p - sd_log10_p,
                     xmax = mean_log10_p + sd_log10_p),
                 height = 0.01, alpha = 0.1) +
  geom_point(alpha = 0.95, aes(color = metadata_class)) +
    scale_y_log10() +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed") +
  labs(
    x = expression("Mean -log"[10]*"(p-value)"),
    y = "Mean R² (%) Explained Variance",
    # title = glue("PERMANOVA Summary by Term: {cond}"),
    color = "Metadata Class"
  ) +
  scale_color_manual(values = metadata_class_color_var) +
  scale_fill_manual(values = metadata_class_color_var, guide = "none") +
  theme_bw()
# ggsave(
#     glue("{clustcr_figs}/permanova_results_{cond}_summary_base_{Sys.Date()}.png"),
#     p_perm_base,
#     width = 10,
#     height = 6,
#     dpi = 600
# )

set.seed(4)
p_perm_annot <- p_perm_base +
  ggrepel::geom_label_repel(
    data = labeled_points,
    aes(label = term, fill = metadata_class),
    size = 2.5,
    max.overlaps = Inf,
    force = 50,
    force_pull = 0,
    nudge_x = ifelse(labeled_points$mean_log10_p > 0, 1, -1),
    nudge_y = 0.25,
    box.padding = 0.8,
    point.padding = 0.5,
    segment.color = "darkblue",
    segment.size = 0.2,
    segment.alpha = 0.75,
    alpha = 0.75,
    direction = "both",
    segment.curvature = 0.3,
    segment.ncp = 3,
    segment.angle = 10
  )

ggsave(
    glue("{clustcr_figs}/permanova_results_{cond}_summary_annot_{Sys.Date()}.png"),
    p_perm_annot,
    width = 7,
    height = 4,
    dpi = 600
)

```




Running Glmmer models to screen for metadata explained variance in cluster membership

```{r}
clustcr_results_df %>% glimpse()
cluster_stats %>% glimpse()

sid_count_summary_per_cluster <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 8) %>% 
  group_by(cluster, motif, v_call, sample_id, participant_id) %>% 
  dplyr::summarize(
    unique = n_distinct(junction_aa),
    total = sum(count)
  ) %>% 
  glimpse()

sid_count_summary_per_cluster_wide <- sid_count_summary_per_cluster %>% 
  ungroup() %>% 
  mutate(cluster = glue("clust_{cluster}_{v_call}")) %>% 
  select(-c(motif, v_call)) %>% 
  pivot_wider(
    # id_cols = c(cluster),
    names_from = cluster,
    values_from = c(unique, total),
    names_glue = "{cluster}__{.value}",
    values_fill = 0
  )

dim(sid_count_summary_per_cluster_wide)
sid_count_summary_per_cluster_wide %>% colnames() %>% head()

# 200158  total clusters being tested in this case
full_df <- sid_count_summary_per_cluster_wide %>% 
  left_join(metadata) %>% 
  glimpse()

# metadata %>% glimpse()

library(glmmTMB)
# full_df[["clust_0_TRBV20-1__unique"]]
top_clusts <- cluster_stats %>% 
  slice_max(unique_donors_per_clust, n = 10) %>% 
  pull(cluster)
sid_count_summary_per_cluster %>% glimpse()
sid_count_summary_per_cluster %>% 
  filter(cluster %in% top_clusts)  %>% 
  mutate(cluster = glue("clust_{cluster}_{v_call}")) %>% 
  pull(cluster) %>% 
  unique()

model_df <- full_df %>% 
  mutate(
    in_k_unique = .data[["clust_37759_TRBV7-3__unique"]],
    not_in_k_unique = clonal_count - in_k_unique,
    in_k_total = .data[["clust_37759_TRBV7-3__total"]],
    not_in_k_total = clonal_volume - in_k_total
    ) %>% 
  dplyr::select(
    in_k_unique, not_in_k_unique, in_k_total, not_in_k_total, not_in_k_total,
    est_tcell_reads_abis, case_control_other_latest,
    participant_id, sample_id) %>% 
  glimpse()

# full vs null -----------------------------------------------------
null_mod_unique <- glmmTMB::glmmTMB(
  cbind(in_k_unique, not_in_k_unique) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
full_mod_unique <- glmmTMB::glmmTMB(
  cbind(in_k_unique, not_in_k_unique) ~ est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
anova(null_mod_unique, full_mod_unique) %>% as.data.frame()


null_mod_total <- glmmTMB::glmmTMB(
  cbind(in_k_total, not_in_k_total) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
full_mod_total <- glmmTMB::glmmTMB(
  cbind(in_k_total, not_in_k_total) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
anova(null_mod_unique, full_mod_unique, test = "LRT")              # likelihood-ratio test



# library(glmmTMB)
# null_mod_unique <- glmmTMB(
#   cbind(in_k_unique, not_in_k_unique) ~ case_control_other_latest +
#                                          scale(est_tcell_reads_abis) +
#                                          (1|participant_id),
#   data = model_df,
#   family = betabinomial(link = "logit"),            # ← handles over-dispersion
#   control = glmmTMBControl(optimizer = optim,
#                            optArgs = list(method = "BFGS")) # more robust optimiser
# )
# summary(null_mod_unique)

library(DescTools)
DescTools::SeparationTest(in_k_unique ~ case_control_other_latest,
                          data = model_df)   # TRUE  = separation

```




Running Glmmer models to screen for metadata explained variance in cluster membership (WEIGHTED BY CLONAL VOLUME)

```{r}
clustcr_results_df %>% glimpse()
cluster_stats %>% glimpse()

sid_count_summary_per_cluster <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 8) %>% 
  group_by(cluster, motif, v_call, sample_id, participant_id) %>% 
  dplyr::summarize(
    unique = n_distinct(junction_aa),
    total = sum(count)
  ) %>% 
  glimpse()

sid_count_summary_per_cluster_wide <- sid_count_summary_per_cluster %>% 
  ungroup() %>% 
  mutate(cluster = glue("clust_{cluster}_{v_call}")) %>% 
  select(-c(motif, v_call)) %>% 
  pivot_wider(
    # id_cols = c(cluster),
    names_from = cluster,
    values_from = c(unique, total),
    names_glue = "{cluster}__{.value}",
    values_fill = 0
  )

dim(sid_count_summary_per_cluster_wide)
sid_count_summary_per_cluster_wide %>% colnames() %>% head()

# 200158  total clusters being tested in this case
full_df <- sid_count_summary_per_cluster_wide %>% 
  left_join(metadata) %>% 
  glimpse()

# metadata %>% glimpse()

library(glmmTMB)
# full_df[["clust_0_TRBV20-1__unique"]]
top_clusts <- cluster_stats %>% 
  slice_max(unique_donors_per_clust, n = 10) %>% 
  pull(cluster)
sid_count_summary_per_cluster %>% glimpse()
sid_count_summary_per_cluster %>% 
  filter(cluster %in% top_clusts)  %>% 
  mutate(cluster = glue("clust_{cluster}_{v_call}")) %>% 
  pull(cluster) %>% 
  unique()

model_df <- full_df %>% 
  mutate(
    in_k_unique = .data[["clust_37759_TRBV7-3__unique"]],
    not_in_k_unique = clonal_count - in_k_unique,
    in_k_total = .data[["clust_37759_TRBV7-3__total"]],
    not_in_k_total = clonal_volume - in_k_total
    ) %>% 
  dplyr::select(
    in_k_unique, not_in_k_unique, in_k_total, not_in_k_total, not_in_k_total,
    est_tcell_reads_abis, case_control_other_latest,
    participant_id, sample_id) %>% 
  glimpse()

# full vs null -----------------------------------------------------
null_mod_unique <- glmmTMB::glmmTMB(
  cbind(in_k_unique, not_in_k_unique) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
full_mod_unique <- glmmTMB::glmmTMB(
  cbind(in_k_unique, not_in_k_unique) ~ est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
anova(null_mod_unique, full_mod_unique) %>% as.data.frame()


null_mod_total <- glmmTMB::glmmTMB(
  cbind(in_k_total, not_in_k_total) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
full_mod_total <- glmmTMB::glmmTMB(
  cbind(in_k_total, not_in_k_total) ~ case_control_other_latest + est_tcell_reads_abis + (1|participant_id),
  data = model_df, family = binomial(link = "logit")
)
anova(null_mod_unique, full_mod_unique, test = "LRT")              # likelihood-ratio test



# library(glmmTMB)
# null_mod_unique <- glmmTMB(
#   cbind(in_k_unique, not_in_k_unique) ~ case_control_other_latest +
#                                          scale(est_tcell_reads_abis) +
#                                          (1|participant_id),
#   data = model_df,
#   family = betabinomial(link = "logit"),            # ← handles over-dispersion
#   control = glmmTMBControl(optimizer = optim,
#                            optArgs = list(method = "BFGS")) # more robust optimiser
# )
# summary(null_mod_unique)

library(DescTools)
DescTools::SeparationTest(in_k_unique ~ case_control_other_latest,
                          data = model_df)   # TRUE  = separation

```

Running Glmmer models to screen for metadata explained variance in cluster membership (UNWEIGHTED - BINARY)

First formatting the data for modeling
```{r}
clustcr_results_df %>% glimpse()
# model_meta_df <- readRDS(
#     glue("{clustcr_dir}/results/metadata_for_logreg_2025-08-17.rds")) %>% 
#     glimpse()
model_meta_df <- readRDS(
    glue("{clustcr_dir}/results/metadata_for_logreg_2025-08-17.rds")) %>% 
    glimpse()

sid_count_summary_per_cluster <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 8) %>% 
  dplyr::select(cluster_id, participant_id) %>% 
  ungroup() %>% 
  distinct() %>% 
  glimpse()

sid_count_summary_per_cluster_wide <- sid_count_summary_per_cluster %>% 
    mutate(value = 1) %>% 
    pivot_wider(
        names_from = cluster_id,
        values_from = value,
        values_fill = 0
    )

dim(sid_count_summary_per_cluster_wide)
sid_count_summary_per_cluster_wide %>% colnames() %>% head()
# 99,105  total clusters being tested in this case

# Joining with particpant level metadata
full_df <- sid_count_summary_per_cluster_wide %>% 
  left_join(model_meta_df)


# Saving the full dataframe
saveRDS(
  full_df,
  glue("{clustcr_dir}/results/logistic_regression_input_binary_dataframe_{Sys.Date()}.rds")
)
saveRDS(
  full_df_rich,
  glue("{clustcr_dir}/results/logistic_regression_input_binary_dataframe_RICH_{Sys.Date()}.rds")
)

```


Running a single cluster test and wrapping code into a function

```{r}
full_df <- readRDS(glue("{clustcr_dir}/results/logistic_regression_input_binary_dataframe_2025-09-08.rds"))
full_df[1:5, 1:5]
full_df %>% colnames() %>% tail(n=10)
# full_df[["CSARD..G.GYTF_TRBV20-1"]]

# Manually running a single cluster test ----
cid_list <- full_df[, -1] %>% colnames() %>% head(n=30)
cid <- "YASSSg.YEQYF_TRBV26"
null_formula <- glue("{cid} ~ est_tcell_reads_abis_pid + age_at_baseline + sex")
full_formula <- glue("{cid} ~ case_control_other_latest + est_tcell_reads_abis_pid + age_at_baseline + sex")

null_model <- glm(as.formula(null_formula), data = full_df, family = binomial(link = "logit"))
full_model <- glm(as.formula(full_formula), data = full_df, family = binomial(link = "logit"))

model_comp <- anova(full_model, null_model, test = "LRT")
lrt_res <- model_comp %>% broom::tidy()
logreg_anova_df <- lrt_res[2,] %>% 
    dplyr::mutate(full_model = lrt_res$term[1]) %>% 
    dplyr::mutate(cluster_id = cid)

logreg_reslist <- list(
    "cluster_id" = cid,
    "logit_null" = broom::tidy(null_model),
    "logit_full" = broom::tidy(full_model),
    "comparison_res" = model_comp,
    "logit_anova_df" = logreg_anova_df,
    "full_model_glance" = broom::glance(full_model),
    "null_model_glance" = broom::glance(null_model)
)

full_df %>% colnames() %>% tail(n=30)

# Wrapping code into a function ----
run_binary_cluster_logreg_analysis <- function(cid_list) {
    require(magrittr)
    require(glue)
    require(dplyr)
    require(tictoc)
    
    # reading the full dataframe
    clustcr_dir <- "/resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/clusTCR"
    full_df <- readRDS(glue("{clustcr_dir}/results/logistic_regression_input_binary_dataframe_2025-09-08.rds"))

    for (cid in cid_list) {
      tic("Fitting the models")
      null_formula <- glue("`{cid}` ~ est_tcell_reads_abis_pid + age_at_baseline + sex")
      full_formula <- glue("`{cid}` ~ case_control_other_latest + est_tcell_reads_abis_pid + age_at_baseline + sex")

      null_model <- glm(as.formula(null_formula), data = full_df, family = binomial(link = "logit"))
      full_model <- glm(as.formula(full_formula), data = full_df, family = binomial(link = "logit"))

      model_comp <- anova(full_model, null_model, test = "LRT")
      lrt_res <- model_comp %>% broom::tidy()

      logreg_anova_df <- lrt_res[2,] %>% 
          dplyr::mutate(full_model = lrt_res$term[1]) %>% 
          dplyr::mutate(cluster_id = cid)

      logreg_reslist <- list(
          "cluster_id" = cid,
          "logit_null" = broom::tidy(null_model),
          "logit_full" = broom::tidy(full_model),
          "comparison_res" = model_comp,
          "logit_anova_df" = logreg_anova_df,
          "full_model_glance" = broom::glance(full_model),
          "null_model_glance" = broom::glance(null_model)
      )

      saveRDS(logreg_reslist, 
        glue("/resnick/scratch/jbok/tmp_clusTCR_logreg/logistic_regression_binary/{cid}__2025-09-08.rds"))
      toc()
    }
}


# # test with list of cids
# run_binary_cluster_logreg_analysis(cid_list)

```

Executing log regression model at scale

Executing Logistic Regression Models for Cluster Membership (UNWEIGHTED - BINARY)

```{r}
cid_run_list <- unique(sid_count_summary_per_cluster$cluster_id) #%>%
  # discard(~ file.exists(glue("/resnick/scratch/jbok/tmp_clusTCR_logreg/logistic_regression_binary/{.}__2025-09-08.rds")))

length(unique(sid_count_summary_per_cluster$cluster_id))
length(cid_run_list)
cid_run_list[1:10]

# Running on cluster in parallel
breg <- breg_setup(cluster_name = "clusTCR_logreg_binary", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )
# Submit Jobs ----
jobs <- batchMap(
  fun = run_binary_cluster_logreg_analysis,
  args = list(chunk_func(cid_run_list, n = 20)),
  reg = breg
)

# jobs[, chunk := chunk(job.id, chunk.size = 40)]
# print(jobs[, .N, by = chunk])

# can probably do two per sqminute so x jobs / 2 = minutes
submitJobs(jobs,
  resources = list(
    walltime = "3:00:00",
    memory = "5GB",
    ncpus = 1,
    max.concurrent.jobs = 9999
  )
)

```


## Binary clusTCR logreg with longitudinal metadata averages (PD-only + study-adjusted)

Setup run layout and conventions

```{r}
target_metadata_classes <- c("Cognition", "Motor Severity")
target_case_group <- "Case"

logreg_metaavg_root <- glue("{clustcr_dir}/results/logistic_regression_binary_metaavg")
logreg_metaavg_dirs <- list(
  metadata = glue("{logreg_metaavg_root}/metadata"),
  input = glue("{logreg_metaavg_root}/input"),
  pairwise = glue("{logreg_metaavg_root}/pairwise"),
  summaries = glue("{logreg_metaavg_root}/summaries"),
  manifests = glue("{logreg_metaavg_root}/manifests")
)

purrr::walk(logreg_metaavg_dirs, ~ dir.create(.x, showWarnings = FALSE, recursive = TRUE))

model_spec_tag <- "pdonly_study_mhc_seqdepth_age_sex_metaavg"
run_id <- glue("{model_spec_tag}_{Sys.Date()}")

safe_file_token <- function(x) {
  x %>% gsub("[^A-Za-z0-9._-]", "_", .)
}

logreg_metaavg_dirs
run_id
```

Build participant-level averaged metadata table (numeric only, selected classes)

```{r}
metadata_long <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")
) %>%
  dplyr::filter(case_control_other_latest != "Other")

metadata_categories_df <- readRDS(
  glue("{wkdir}/data/interim/metadata/metadata_categories_dataframe_2025-07-21.rds")
)

candidate_metadata <- metadata_categories_df %>%
  dplyr::filter(metadata_class %in% target_metadata_classes) %>%
  dplyr::pull(metadata) %>%
  unique()

numeric_candidate_metadata <- candidate_metadata %>%
  intersect(colnames(metadata_long))

numeric_candidate_metadata <- numeric_candidate_metadata[
  purrr::map_lgl(metadata_long[, numeric_candidate_metadata, drop = FALSE], is.numeric)
]

metadata_pd_only <- metadata_long %>%
  dplyr::filter(case_control_other_latest == target_case_group)

model_meta_fixed_df <- metadata_pd_only %>%
  dplyr::group_by(
    participant_id, age_at_baseline, sex, study,
    MHC_all_PC1, MHC_all_PC2, MHC_all_PC3
  ) %>%
  dplyr::summarise(
    est_tcell_reads_abis_pid = sum(est_tcell_reads_abis, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::distinct()

model_meta_avg_df <- metadata_pd_only %>%
  dplyr::group_by(participant_id) %>%
  dplyr::summarise(
    dplyr::across(
      all_of(numeric_candidate_metadata),
      ~ mean(.x, na.rm = TRUE),
      .names = "avg__{.col}"
    ),
    .groups = "drop"
  )

model_meta_avg_df <- model_meta_fixed_df %>%
  dplyr::left_join(model_meta_avg_df, by = "participant_id") %>%
  glimpse()

metaavg_terms <- grep("^avg__", colnames(model_meta_avg_df), value = TRUE)
metaavg_terms <- metaavg_terms[
  purrr::map_lgl(metaavg_terms, ~ {
    x <- model_meta_avg_df[[.x]]
    sum(!is.na(x)) >= 20 && dplyr::n_distinct(x, na.rm = TRUE) > 1
  })
]

meta_manifest <- tibble::tibble(
  run_id = run_id,
  target_case_group = target_case_group,
  target_metadata_classes = paste(target_metadata_classes, collapse = ";"),
  n_candidate_metadata = length(candidate_metadata),
  n_numeric_metadata = length(numeric_candidate_metadata),
  n_metaavg_terms_used = length(metaavg_terms),
  n_participants = nrow(model_meta_avg_df)
)

saveRDS(
  model_meta_avg_df,
  glue("{logreg_metaavg_dirs$metadata}/metadata_for_logreg_metaavg_{run_id}.rds")
)
readr::write_csv(
  meta_manifest,
  glue("{logreg_metaavg_dirs$manifests}/metadata_manifest_{run_id}.csv")
)
```

Build unweighted binary cluster membership model input (PD-only)

```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds")
)

sid_count_summary_per_cluster <- clustcr_results_df %>%
  dplyr::filter(unique_donors_per_clust >= 10) %>%
  dplyr::filter(motif_length >= 8) %>%
  dplyr::filter(case_control_other_latest == target_case_group) %>%
  dplyr::select(cluster_id, participant_id) %>%
  dplyr::distinct()

sid_count_summary_per_cluster_wide <- sid_count_summary_per_cluster %>%
  dplyr::mutate(value = 1L) %>%
  tidyr::pivot_wider(
    names_from = cluster_id,
    values_from = value,
    values_fill = 0L
  )

model_input_binary_metaavg <- sid_count_summary_per_cluster_wide %>%
  dplyr::left_join(model_meta_avg_df, by = "participant_id") %>%
  glimpse()

cluster_cols <- setdiff(colnames(sid_count_summary_per_cluster_wide), "participant_id")

model_input_path <- glue(
  "{logreg_metaavg_dirs$input}/logreg_binary_input_pdonly_metaavg_{run_id}.rds"
)
saveRDS(model_input_binary_metaavg, model_input_path)

input_manifest <- tibble::tibble(
  run_id = run_id,
  n_participants = nrow(model_input_binary_metaavg),
  n_clusters = length(cluster_cols),
  n_metaavg_terms = length(metaavg_terms),
  input_rds = model_input_path
)
readr::write_csv(
  input_manifest,
  glue("{logreg_metaavg_dirs$manifests}/input_manifest_{run_id}.csv")
)

```

Define pairwise null/full logistic modeling function

```{r}
run_binary_cluster_logreg_metaavg <- function(job_df, input_rds, run_tag, model_spec_tag, out_dir) {
  require(glue)
  require(dplyr)
  require(broom)

  full_df <- readRDS(input_rds)
  required_covars <- c(
    "study",
    "MHC_all_PC1", "MHC_all_PC2", "MHC_all_PC3",
    "est_tcell_reads_abis_pid", "age_at_baseline", "sex"
  )

  for (i in seq_len(nrow(job_df))) {
    cid <- job_df$cluster_id[i]
    meta_term <- job_df$meta_term[i]
    outpath <- job_df$output_path[i]

    if (file.exists(outpath)) {
      next
    }

    result_obj <- tryCatch({
      if (!cid %in% colnames(full_df)) {
        stop(glue("Missing cluster column: {cid}"))
      }
      if (!meta_term %in% colnames(full_df)) {
        stop(glue("Missing metadata term column: {meta_term}"))
      }

      model_df <- full_df %>%
        dplyr::select(
          participant_id,
          outcome = all_of(cid),
          all_of(required_covars),
          meta_value = all_of(meta_term)
        ) %>%
        dplyr::filter(!is.na(meta_value)) %>%
        dplyr::filter(
          !is.na(study),
          !is.na(MHC_all_PC1), !is.na(MHC_all_PC2), !is.na(MHC_all_PC3),
          !is.na(est_tcell_reads_abis_pid), !is.na(age_at_baseline), !is.na(sex)
        )

      if (nrow(model_df) < 30) {
        stop("Insufficient rows after filtering (<30)")
      }
      if (dplyr::n_distinct(model_df$outcome, na.rm = TRUE) < 2) {
        stop("Outcome has no variance")
      }
      if (dplyr::n_distinct(model_df$meta_value, na.rm = TRUE) < 2) {
        stop("Metadata term has no variance")
      }

      null_model <- glm(
        outcome ~ study + MHC_all_PC1 + MHC_all_PC2 + MHC_all_PC3 +
          est_tcell_reads_abis_pid + age_at_baseline + sex,
        data = model_df,
        family = binomial(link = "logit")
      )
      full_model <- glm(
        outcome ~ study + MHC_all_PC1 + MHC_all_PC2 + MHC_all_PC3 +
          est_tcell_reads_abis_pid + age_at_baseline + sex + meta_value,
        data = model_df,
        family = binomial(link = "logit")
      )

      model_comp <- anova(full_model, null_model, test = "LRT")
      lrt_res <- broom::tidy(model_comp)
      logreg_anova_df <- lrt_res[2, ] %>%
        dplyr::mutate(
          cluster_id = cid,
          meta_term = meta_term,
          run_id = run_tag,
          model_spec_tag = model_spec_tag
        )

      list(
        status = "ok",
        cluster_id = cid,
        meta_term = meta_term,
        run_id = run_tag,
        model_spec_tag = model_spec_tag,
        n_obs = nrow(model_df),
        logit_null = broom::tidy(null_model),
        logit_full = broom::tidy(full_model),
        comparison_res = model_comp,
        logit_anova_df = logreg_anova_df,
        full_model_glance = broom::glance(full_model),
        null_model_glance = broom::glance(null_model)
      )
    }, error = function(e) {
      list(
        status = "error",
        cluster_id = cid,
        meta_term = meta_term,
        run_id = run_tag,
        model_spec_tag = model_spec_tag,
        error_message = as.character(e$message)
      )
    })

    saveRDS(result_obj, outpath)
  }
}

```

Submit pairwise models in Slurm-parallel with rerun-safe tracking

```{r}
cluster_keep <- model_input_binary_metaavg %>%
  dplyr::summarise(dplyr::across(all_of(cluster_cols), ~ sum(.x == 1, na.rm = TRUE))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "cluster_id", values_to = "n_pos") %>%
  dplyr::filter(n_pos >= 5) %>%
  dplyr::pull(cluster_id)

existing_files <- list.files(
  logreg_metaavg_dirs$pairwise,
  pattern = glue("__{run_id}\\.rds$"),
  full.names = FALSE
)
existing_key <- tibble::tibble(output_file = existing_files)

pairwise_df <- tidyr::crossing(
  cluster_id = cluster_keep,
  meta_term = metaavg_terms
) %>%
  dplyr::mutate(
    output_file = glue(
      "{safe_file_token(cluster_id)}__{safe_file_token(meta_term)}__{run_id}.rds"
    ),
    output_path = glue("{logreg_metaavg_dirs$pairwise}/{output_file}")
  )

pairwise_pending <- pairwise_df %>%
  dplyr::anti_join(existing_key, by = "output_file")

readr::write_csv(
  pairwise_df %>%
    dplyr::mutate(already_ran = output_file %in% existing_key$output_file) %>%
    dplyr::select(cluster_id, meta_term, output_file, already_ran),
  glue("{logreg_metaavg_dirs$manifests}/pairwise_manifest_{run_id}.csv")
)

if (nrow(pairwise_pending) > 0) {
  chunk_size <- 25
  pairwise_chunks <- split(
    pairwise_pending,
    ceiling(seq_len(nrow(pairwise_pending)) / chunk_size)
  )

  breg <- breg_setup(
    cluster_name = "clusTCR_logreg_binary_metaavg",
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
  )

  jobs <- batchMap(
    fun = run_binary_cluster_logreg_metaavg,
    args = list(
      job_df = pairwise_chunks,
      input_rds = rep(model_input_path, length(pairwise_chunks)),
      run_tag = rep(run_id, length(pairwise_chunks)),
      model_spec_tag = rep(model_spec_tag, length(pairwise_chunks)),
      out_dir = rep(logreg_metaavg_dirs$pairwise, length(pairwise_chunks))
    ),
    reg = breg
  )

  submitJobs(
    jobs,
    resources = list(
      walltime = "3:00:00",
      memory = "5GB",
      ncpus = 1,
      max.concurrent.jobs = 9999
    )
  )
} else {
  message("No pending pairwise jobs; all outputs already exist.")
}
```

Aggregate pairwise results and save run summaries

```{r}
pairwise_paths <- list.files(
  logreg_metaavg_dirs$pairwise,
  pattern = glue("__{run_id}\\.rds$"),
  full.names = TRUE
)

pairwise_res <- purrr::map(pairwise_paths, ~ tryCatch(readRDS(.x), error = function(e) NULL))
pairwise_res <- pairwise_res[!purrr::map_lgl(pairwise_res, is.null)]

summary_df <- purrr::map_dfr(pairwise_res, function(x) {
  if (identical(x$status, "ok")) {
    lrt_p <- x$logit_anova_df$p.value %>% as.numeric()
    full_meta <- x$logit_full %>% dplyr::filter(term == "meta_value")
    tibble::tibble(
      cluster_id = x$cluster_id,
      meta_term = x$meta_term,
      fit_ok = TRUE,
      p_value = lrt_p,
      meta_estimate = full_meta$estimate %>% as.numeric(),
      meta_term_p_value = full_meta$p.value %>% as.numeric(),
      n_obs = x$n_obs,
      delta_aic = x$null_model_glance$AIC - x$full_model_glance$AIC,
      delta_bic = x$null_model_glance$BIC - x$full_model_glance$BIC
    )
  } else {
    tibble::tibble(
      cluster_id = x$cluster_id,
      meta_term = x$meta_term,
      fit_ok = FALSE,
      p_value = NA_real_,
      meta_estimate = NA_real_,
      meta_term_p_value = NA_real_,
      n_obs = NA_real_,
      delta_aic = NA_real_,
      delta_bic = NA_real_,
      error_message = x$error_message
    )
  }
}) %>%
  dplyr::mutate(
    run_id = run_id,
    model_spec_tag = model_spec_tag,
    p_value_adj_bh = p.adjust(p_value, method = "BH")
  ) %>%
  dplyr::arrange(p_value)

run_manifest <- tibble::tibble(
  run_id = run_id,
  model_spec_tag = model_spec_tag,
  total_pairs_planned = nrow(pairwise_df),
  total_pairs_completed = nrow(summary_df),
  n_success = sum(summary_df$fit_ok, na.rm = TRUE),
  n_failed = sum(!summary_df$fit_ok, na.rm = TRUE)
)

saveRDS(
  summary_df,
  glue("{logreg_metaavg_dirs$summaries}/pairwise_logreg_summary_{run_id}.rds")
)
readr::write_csv(
  summary_df,
  glue("{logreg_metaavg_dirs$summaries}/pairwise_logreg_summary_{run_id}.csv")
)
readr::write_csv(
  run_manifest,
  glue("{logreg_metaavg_dirs$manifests}/run_manifest_{run_id}.csv")
)
```


Checking on the results

```{r}
res_paths <- list.files(
    glue("/resnick/scratch/jbok/tmp_clusTCR_logreg/logistic_regression_binary"),
    full.names = TRUE
)
length(res_paths)

future::plan(future::multisession, workers = 4)
# Contains p-values for the LRT comparing the null and full models with the addition of case/control status to the full model
full_logreg_binary_model_res <- res_paths %>% 
  purrr::set_names( ~ basename(.) %>% gsub("__2025-09-08.rds", "", .)) %>% 
  future_map(~ possibly(readRDS, NULL)(.))

saveRDS(full_logreg_binary_model_res, 
  glue("{clustcr_dir}/results/logistic_regression_binary_all_model_results_list_{Sys.Date()}.rds")
  )

anova_lrt_pval_df <- res_paths %>% 
  furrr::future_map(~ possibly(readRDS, NULL)(.)$logit_anova_df) %>% 
  bind_rows() %>% 
  janitor::clean_names() %>% 
  dplyr::select(cluster_id, p_value) %>% 
  arrange(p_value) %>% 
  glimpse()
anova_lrt_pval_df %>% glimpse()

saveRDS(anova_lrt_pval_df, 
  glue("{clustcr_dir}/results/logistic_regression_binary_anova_lrt_pval_df_{Sys.Date()}.rds")
  )

full_model_stats_df <- res_paths %>% 
  purrr::set_names( ~ basename(.) %>% gsub("__2025-09-08.rds", "", .)) %>% 
  future_map(~ possibly(readRDS, NULL)(.)$logit_full) %>% 
  bind_rows(.id = "cluster_id") %>% 
  janitor::clean_names() %>% 
  glimpse()

saveRDS(full_model_stats_df, 
  glue("{clustcr_dir}/results/logistic_regression_binary_full_model_stats_df_{Sys.Date()}.rds")
  )

```


```{r}
# Loading chi2 results
chi2_results_df <- readRDS(
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_2025-09-08.rds")) %>% 
  glimpse()
# Loading logreg results --
# ANOVA LRT p-values
anova_lrt_pval_df <- readRDS(
  glue("{clustcr_dir}/results/logistic_regression_binary_anova_lrt_pval_df_2025-09-10.rds")
  ) %>% 
  dplyr::rename(anova_lrt_pval = p_value)  %>% 
  mutate(anova_lrt_pval_adjusted = p.adjust(anova_lrt_pval, method = "BH")) %>% 
  arrange(anova_lrt_pval_adjusted) %>% 
  glimpse()

# Full model stats
full_model_stats_df <- readRDS(
  glue("{clustcr_dir}/results/logistic_regression_binary_full_model_stats_df_2025-09-10.rds")
  )
# # Loading model results  ( how to see full aggregated results)
# full_logreg_binary_model_res <- readRDS(
#   glue("{clustcr_dir}/results/logistic_regression_binary_all_model_results_list_2025-09-10.rds")
#   )

# # comparing chi2 and anova lrt p-values
# chi2_results_df %>% 
#   left_join(anova_lrt_pval_df, by = "cluster_id") %>% 
#   ggplot(aes(x = anova_lrt_pval, y = chi2_p_value)) +
#   geom_point() +
#   geom_abline(slope = 1, intercept = 0) +
#   theme_minimal()

p_model_pval_comparison_plot <- chi2_results_df %>% 
  left_join(anova_lrt_pval_df, by = "cluster_id") %>% 
  ggplot(aes(x = -log10(anova_lrt_pval), y = -log10(chi2_p_value))) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed") +
  labs(x = expression(-log[10](ANOVA~LRT~"P-value")), y = expression(-log[10]("Chi2 P-value"))) +
  coord_equal() +
  theme_bw()

ggsave(p_model_pval_comparison_plot, 
  filename = glue("{clustcr_figs}/Chi2_vs_LogReg_LRT_pval_comparisons_{Sys.Date()}.png"),
  width = 6,
  height = 6
)

# Double checking directionality is consistent between full logreg models and chi2
full_model_stats_df %>%
  filter(term == "case_control_other_latestControl") %>% 
  left_join(chi2_results_df, by = "cluster_id") %>% 
  left_join(anova_lrt_pval_df, by = "cluster_id") %>%
  filter(anova_lrt_pval < 0.05) %>% 
  ggplot(aes(x = -estimate, y = log2(odds_ratio))) +
  geom_point() +
  # geom_abline(slope = 1, intercept = 0) +
  theme_bw()

# Notes: looks like directionality is consistent; however chi2 has much more spread and a natural distribution while logreg is much more specific to a handful of clusters


# Now looking at a volcano plot of case/control estimates with the LRT p-value
full_model_stats_df %>%
  filter(term == "case_control_other_latestControl") %>% 
  left_join(anova_lrt_pval_df, by = "cluster_id") %>%
  # filter(anova_lrt_pval < 0.05) %>% 
  ggplot(aes(x = -estimate, y = -log10(p_value))) +
  geom_point() +
  theme_bw()
# Notes = this is pretty weird ..? Looks like is some strange structure to the estimate scores

full_model_stats_df %>%
  filter(term == "case_control_other_latestControl") %>% 
  left_join(anova_lrt_pval_df, by = "cluster_id") %>%
  filter(anova_lrt_pval < 0.05) %>%
  slice_max(order_by = abs(estimate), n = 100) %>%
  View



# cid <- "CASS.G..NSPLHF_TRBV12-3"
# full_mod <- readRDS(glue("/resnick/scratch/jbok/tmp_clusTCR_logreg/logistic_regression_binary/{cid}__2025-09-08.rds"))$logit_full %>% 
#   janitor::clean_names() %>% 
#   glimpse()
# # readRDS(glue("/resnick/scratch/jbok/tmp_clusTCR_logreg/logistic_regression_binary/{cid}__2025-09-08.rds"))$full_model_glance

# raw_binary_data <- full_df %>% 
#   select(detected = {{cid}}, case_control_other_latest, est_tcell_reads_abis_pid, age_at_baseline, sex, participant_id) %>% 
#   mutate(cid = {{cid}}) %>% 
#   glimpse()

# raw_binary_data %>% 
#   ggplot(aes(y = detected, x = case_control_other_latest)) +
#   geom_boxplot()
#   # geom_point() +
#   # geom_smooth(method = "glm", method.args = list(family = "binomial")) 

```

Plotting a volcano plot with the LRT p-value and odds ratio

```{r}
cluster_sscore_filtered <- clustcr_results_df_filtered %>% 
  dplyr::select(cluster_id, Q_data, risk_score_p2_glrm_strat) %>% 
  group_by(cluster_id) %>% 
  dplyr::summarize(
    avg_risk_score = mean(risk_score_p2_glrm_strat, na.rm = TRUE),
    avg_selection_score = mean(log10(Q_data), na.rm = TRUE)) %>% 
  glimpse()

cluster_stats_filtered <- clustcr_results_df_filtered %>% 
  dplyr::select(cluster_id, motif, unique_donors_per_clust, unique_samples_per_clust, motif_length) %>% 
  distinct() %>% 
  left_join(cluster_sscore_filtered) %>% 
  glimpse()


volcano_df <- chi2_results_df %>%
  left_join(anova_lrt_pval_df, by = "cluster_id") %>%
  left_join(cluster_stats_filtered) %>% 
  left_join(full_model_stats_df %>% filter(term == "case_control_other_latestControl"), 
    by = "cluster_id") %>%
    arrange(anova_lrt_pval) %>% 
  glimpse()

volcano_df %>% glimpse()
# volcano_df$unique_donors_per_clust %>% summary()

p_lrt_volcano <- volcano_df %>% 
  mutate(cluster_label = str_before_first(cluster_id, "_")) %>% 
  # Add a column to identify significant points
  mutate(
    significant = case_when(
      abs(log2(odds_ratio)) >= 0.5 & anova_lrt_pval <= 0.05 ~ case_when(
        odds_ratio > 1 ~ "Enriched in Cases",
        odds_ratio < 1 ~ "Enriched in Controls"
      ),
      TRUE ~ "Not Significant"
    ),
    # Create custom hover text
    hover_text = sprintf(
      "Cluster: %s\nLRT P-value: %.2e\nOdds Ratio: %.2f\nUnique Donors: %d",
      cluster_label, anova_lrt_pval, odds_ratio, unique_donors_per_clust
    )
  ) %>%
  # Add a column to identify points to label
  mutate(
    label_text = case_when(
      # Top 5 most significant enriched in cases
      significant == "Enriched in Cases" & 
        anova_lrt_pval <= (arrange(filter(., significant == "Enriched in Cases"), anova_lrt_pval) %>% 
                         slice_head(n = 5) %>% 
                         pull(anova_lrt_pval) %>% 
                         max()) ~ cluster_label,
      # Top 5 most significant enriched in controls
      significant == "Enriched in Controls" & 
        anova_lrt_pval <= (arrange(filter(., significant == "Enriched in Controls"), anova_lrt_pval) %>% 
                         slice_head(n = 5) %>% 
                         pull(anova_lrt_pval) %>% 
                         max()) ~ cluster_label,
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot(aes(
    x = log2(odds_ratio), 
    # x = odds_ratio, 
    y = -log10(anova_lrt_pval), 
             color = significant, 
             text = hover_text)) +
  geom_point(alpha = 0.6, aes(size = unique_donors_per_clust), shape = 21) +
  # Add ggrepel text labels
  geom_label_repel(
    aes(label = label_text),
    size = 3,
      force = 100,
    max.overlaps = Inf,
    box.padding = 0.5,
    segment.color = "grey50",
    show.legend = FALSE,
    segment.curvature = 0.3,
    segment.ncp = 3,
    segment.angle = 10
  ) +
  scale_color_manual(
    values = c(
      "Enriched in Cases" = "#FF7F0F",
      "Enriched in Controls" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  labs(
    # title = "Chi-squared test for PD enriched/depleted clusters",
    x = expression(log[2]~"(Odds Ratio)"),
    y = expression(-log[10]~"(P-value)"),
    size = "Unique Donors per Cluster",
    color = NULL

  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    # legend.title = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  ) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "darkgrey") +
  geom_vline(xintercept = c(-0.5, 0.5), 
             linetype = "dashed", 
             color = "darkgrey") +
  coord_cartesian(clip = "off")

ggsave(
  glue("{clustcr_figs}/clusTCR_VolcanoPlot_LogReg-LRT-pvalue_vs_odds_ratio_{Sys.Date()}.png"),
  p_lrt_volcano,
  width = 7,
  height = 4 #,dpi = 600
)

seq_converg_df <- volcano_df %>% 
  mutate(direction = case_when(
    log2(odds_ratio) > 1 ~ "Enriched in Cases",
    log2(odds_ratio) < 1 ~ "Enriched in Controls",
    TRUE ~ "Not Significant"
  )) %>% 
  filter(anova_lrt_pval < 0.05) %>%
  group_by(direction, motif) %>% 
  dplyr::summarize(n = n()) %>% 
  arrange(desc(n)) %>% 
  glimpse()

seq_converg_df %>% 
  filter(direction == "Enriched in Cases") %>% 
  slice_max(order_by = n, n = 10) %>% 
  DT::datatable()

seq_converg_df %>% 
  filter(direction == "Enriched in Controls") %>% 
  slice_max(order_by = n, n = 10) %>% 
  DT::datatable()




# # metadata$diagnosis_latest %>% table
# metadata_pdonly <- metadata %>% filter(case_control_other_latest == "Case")

# metadata_pdonly$code_upd2313_posture %>% table
# metadata_pdonly$code_upd2301_speech_problems %>% table
# metadata_pdonly$code_upd2hy_hoehn_and_yahr_stage %>% table

# metadata_pdonly$mds_updrs_part_ii_summary_score %>% table
# metadata_pdonly$mds_updrs_part_iii_summary_score %>% table

# metadata_pdonly$upd23a_medication_for_pd %>% table
# metadata_pdonly$use_of_pd_medication %>% table
# metadata_pdonly$moca_total_score %>% table
# metadata_pdonly$moca_delayed_recall_subscore %>% table


# table(metadata$diagnosis_latest, metadata$study)
# metadata %>% 
#   filter(case_control_other_latest != "Other") %>% 
#   ggplot(aes(x = on_levodopa, y = mds_updrs_part_iii_summary_score)) +
#   geom_boxplot() +
#   # geom_point() +
#   theme_bw()


```

Comparing logreg model with chi2 results

```{r}
chi2_results_df <- readRDS(
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_2025-09-08.rds")) %>% 
  glimpse()

```

