---
title: "Adaptive Immune Receptor Repertoire in Parkinson's Disease: Repertoire Distance Analysis"
editor: source
author: "Joe Boktor"
date: '2025-07-10'
format: 
  html:
    font-family: helvetica neue
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 4
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
---

Environment setup

```{r setup, message=FALSE, warning=FALSE}
home_dir <- "/central/groups/MazmanianLab/joeB"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/parkinsons-microbial-blood-signatures")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))

# Additional libraries
library(data.table)

# Plotting libraries
library(ggsci)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(gt)
library(gtsummary)
library(immunarch)
library(nlme)

metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
imm_plots_dir <- glue("{wkdir}/figures/AIRR/immunarch")
```

Loading metadata and TCRB data

```{r, eval=FALSE}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-17.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  glimpse()

key_meta_df <- metadata %>% 
  dplyr::select(
    sample_id, participant_id, case_control_other_latest, 
    age_at_baseline, sex, abis_t_cell, ethnicity, race, visit_month, study,
    aligned_reads, est_tcell_reads_abis, DRB1_grouped, DRB1_risk
  ) %>% 
  glimpse()

trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()
```



### Calculating Immunarch verlap metrics in parallel (Public Overlap, Jaccard, Tversky, Cosine, Morisita)

```{r}
# Function to calculate overlap for a single method
calculate_overlap <- function(method, immdata_path, wkdir, metrics_dir) {
    require(immunarch)
    require(glue)

    # Load the immdata object
    immdata <- readRDS(immdata_path)

    message(glue("Calculating {method} overlap metrics"))
    metric_data <- repOverlap(
        immdata$data,
        .method = method,
        .downsample = TRUE,
        .verbose = TRUE
    )
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-overlap-{method}_{Sys.Date()}.rds"
        )
    )
}

# Create parameter list for each method
overlap_methods = c(
#   "jaccard", "public", "overlap", "tversky", 
#     "cosine", 
    "morisita"
)

# Setup batchtools registry
cluster_run <- glue("{get_time()}_Repertoire_Overlap_2025-07-08")
message("\n\nRUNNING: ", cluster_run, "\n")
breg <- makeRegistry(
    file.dir = glue(
        "{wkdir}/.cluster_runs/",
        cluster_run
    ),
    seed = 42
)
breg$cluster.functions <- batchtools::makeClusterFunctionsSlurm(
    template = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
    scheduler.latency = 0.1,
    fs.latency = 1
)

# Create parameter list for each method
immdata_path <- glue(
    "{wkdir}/data/interim/airr/immunarch/",
    "immunarch-obj_TCRB_2025-07-08.rds")

batchtools_params <- data.frame(method = overlap_methods) %>% 
    mutate(
        immdata_path = immdata_path,
        wkdir = wkdir,
        metrics_dir = metrics_dir
    )

# Submit jobs
jobs <- batchMap(
    fun = calculate_overlap,
    args = batchtools_params,
    reg = breg
)

submitJobs(jobs,
    resources = list(
        walltime = "7-00:00:00",
        memory = "200GB",
        ncpus = 32,
        max.concurrent.jobs = 9999
    )
)

```

# Calculating IRTransport Distance

This approach is based on the IRTransport package which utilizes optimal transport of ..


First, split each sample_id into a properly formmatted tsv file for IRTransport input

```{r}
irt_input_dir <- glue("{wkdir}/data/interim/irtransport/irt_input")
irt_output_dir <- glue("{wkdir}/data/interim/irtransport/irt_output")
dir.create(irt_input_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(irt_output_dir, showWarnings = FALSE, recursive = TRUE)

# trust4_reports_tcrb %>% glimpse()
# trust4_reports_tcrb  %>% distinct() %>% glimpse()

fmt_for_irt <- trust4_reports_tcrb %>% 
    dplyr::select(sample_id, participant_id, cdr3b = CDR3aa, vb = v_gene, j_gene) %>% 
    distinct() %>% 
    glimpse()


# future::plan(future::multisession, workers = 30)
unique(trust4_reports_tcrb$sample_id) %>% 
    discard(~file.exists(glue("{irt_input_dir}/{.}.tsv"))) %>% 
    purrr::walk(
        .f = function(x) {
            df <- fmt_for_irt %>% 
                dplyr::filter(sample_id == x)
            write_tsv(df, file = glue("{irt_input_dir}/{x}.tsv"))
        }
    )

# fmt_for_irt



```
