---
title: "cdr3-QTL Analysis"
editor: source
author: "Joe Boktor"
date: '2025-08-07'
eval: false
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
    fig-height: 8
    fig-width: 14
---

This notebook will be following analysis defined in (Ishigaki et al., 2022) and implemented [here](https://github.com/immunogenomics/cdr3-QTL).

## Analysis Setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor) 
library(glue)
library(furrr)
library(tictoc)
library(progress)
library(entropy)
library(data.table)
library(seqinr)
library(progress)
library(nlme)
library(broom.mixed)
library(MVLM)
# plotting
library(ggsci)
library(scico)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)

home_dir <- "/central/groups/MazmanianLab/joeB"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/parkinsons-microbial-blood-signatures")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/compute_CDR3-risk-score.R"))
# source(paste0(src_dir, "/R_scripts/analyze_CDR3qtl.R"))
scratch_dir <- "/central/scratch/jbok/PDMBS"
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
entropy_dir <- glue("{wkdir}/data/interim/cdr3_entropy")
entropy_plots_dir <- glue("{wkdir}/figures/AIRR/cdr3_entropy")
git_dir <- glue("{home_dir}/git/cdr3-QTL")
hla_plots_dir <- glue("{wkdir}/figures/AIRR/arcasHLA")
hla_dir <- glue("{wkdir}/data/interim/airr/arcasHLA/analysis_cleaned")

c3_pheno_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_phenotypes")
gdata_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_hla_genotypes")
hla_4d_to_aa_dir <- glue("{wkdir}/data/input/cdr3qtl/hla_4d_to_aa")
tcr_cdr3_riskscore_dir <- glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores")
tcr_cdr3_riskscore_complied <- 
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")

# dir.create(tcr_cdr3_riskscore_dir, showWarnings = FALSE, recursive = TRUE)
# dir.create(tcr_cdr3_riskscore_complied, showWarnings = FALSE, recursive = TRUE)
# dir.create(gdata_dir, showWarnings = FALSE, recursive = TRUE)
# dir.create(c3_pheno_dir, showWarnings = FALSE, recursive = TRUE)
future::plan(multisession, workers = 16)
```


```{r}
cdr3_phenotypes <- c(
  "cdr3_p1_strat_filtered_cv1k", "cdr3_p1_unstrat_filtered_cv1k", 
  "cdr3_p2_strat_filtered_cv1k", "cdr3_p2_unstrat_filtered_cv1k")

all_imgt_pos <- c(
    "P104","P105","P106","P107","P108","P109","P110","P111",
    "P111.1","P112.2","P112.1",
    "P112","P113","P114","P115","P116","P117","P118")
imgt_pos_testable <- all_imgt_pos[4:16]

align_imgt <- function(CDR3){
    AA <- unlist(strsplit(CDR3, ""));
    N_AA_fow <- nchar(CDR3) %/% 2 + nchar(CDR3) %% 2;
    N_AA_na <- 18 - nchar(CDR3);
    AAmod <- c(
        AA[1:(N_AA_fow)],
        rep("NA",N_AA_na),
        AA[(N_AA_fow + 1):nchar(CDR3)]
    );
    names(AAmod) <- all_imgt_pos
    return(t(AAmod))
}

```


Plotting HLA risk score distribution

```{r}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  mutate(case_control_other_latest = factor(case_control_other_latest, levels = c("Control", "Case"))) %>% 
  glimpse()

p_hla_risk_score_dist <- metadata %>%
    dplyr::select(participant_id, case_control_other_latest, pd_hla_risk_score) %>% 
    distinct() %>% 
    ggplot(aes(x = case_control_other_latest, y = pd_hla_risk_score)) +
    # geom_violin() +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.3, height = 0, alpha = 0.3) +
    theme_bw() +
    # ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
    labs(x = NULL, y = "HLA Risk Score") +
    scale_x_discrete(labels = c("Control", "Case")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/hla_risk_score_dist.png"),
    plot = p_hla_risk_score_dist,
    width = 3,
    height = 4
)

```



Here is a single instance example of calculating CDR3 risk score beta-coefficients using the HLA risk score covariate.

```{r}
dlist <- generate_model_inputs_cdr3_riskscore(
    phenotype_type = "cdr3_p1_unstrat_filtered_cv1k",
    cdr3_pos = "P109",
    cdr3length = NA_character_
)

# M <- dlist$M %>% glimpse()
pdata <- dlist$pdata %>% glimpse()
covar_df <- dlist$covar_df %>% glimpse()

# running the LM for each CDR3 AA
mod_list <- unique(pdata$AA) %>%
    purrr::set_names(glue("PD-CDR3-Risk-Score_CDR3-{.}")) %>% 
    purrr::map(
        ~ run_lm_site_pos_pair_riskscore(
            pdata = pdata,
            covar_df = covar_df,
            cdr3_aa = .
        )
)
pdata$participant_id %>% unique() %>% length()

lm_results_df <- mod_list %>% 
    purrr::map(broom::tidy) %>% 
    bind_rows(.id = "model_name") %>% 
    mutate(cdr3_aa = strex::str_after_last(model_name, "CDR3-")) %>% 
    glimpse()

```

## Executing the HLA-risk score LM calculation for each CDR3 AA phenotype 

```{r}
# slurm batchtools parallelzed alternative
batch_params <- expand.grid(
    "phenotype_type" = cdr3_phenotypes, 
    "cdr3_pos" = imgt_pos_testable,
    "cdr3length" = c(12:18)
    ) %>% 
    as.data.frame() %>%
    mutate(
        phenotype_type = as.character(phenotype_type),
        cdr3_pos = as.character(cdr3_pos),
        results_dir = glue("{wkdir}/data/interim/cdr3_qtl/results_cdr3-risk-score/{phenotype_type}"),
        cdr3length = as.character(cdr3length),
        wkdir = wkdir,
        cdr3length = case_when(
            grepl("_unstrat_", phenotype_type) ~ NA_character_,
            TRUE ~ cdr3length
        )
    ) %>%
    dplyr::distinct() %>%
    filter(!file.exists(glue("{results_dir}/lm_model_results__",
        "{phenotype_type}__CDR3pos-{cdr3_pos}__CDR3len-{cdr3length}.csv"))) %>% 
    glimpse()

# View(batch_params)

calculate_cdr3_riskscore_betacoefs(
    results_dir = batch_params$results_dir[1],
    phenotype_type = batch_params$phenotype_type[1],
    cdr3_pos = batch_params$cdr3_pos[1],
    cdr3length = batch_params$cdr3length[1]
)

safe_calculate_cdr3_riskscore_betacoefs_detailed <- purrr::safely(
  calculate_cdr3_riskscore_betacoefs,
  otherwise = NULL,
  quiet = FALSE
)

# future::plan(multisession, workers = 8)
1:nrow(batch_params) %>% 
    purrr::walk(
        ~ safe_calculate_cdr3_riskscore_betacoefs_detailed(
            results_dir = batch_params$results_dir[.],
            phenotype_type = batch_params$phenotype_type[.],
            cdr3_pos = batch_params$cdr3_pos[.],
            cdr3length = batch_params$cdr3length[.]
        )
    )

# # Trouble-shooting
# calculate_cdr3_riskscore_betacoefs(
#     results_dir = glue("{wkdir}/data/interim/cdr3_qtl/results_cdr3-risk-score/cdr3_p1_strat_filtered_cv1k"),
#     phenotype_type = "cdr3_p1_strat_filtered_cv1k",
#     cdr3_pos = "P110",
#     cdr3length = 12
# )

# dlist <- generate_model_inputs_cdr3_riskscore(
#     phenotype_type = "cdr3_p1_strat_filtered_cv1k",
#     cdr3_pos = "P110",
#     cdr3length = 12
# )

# pdata <- dlist$pdata
# covar_df <- dlist$covar_df

# # running the LM for each combination
# mod_list <- unique(pdata$AA) %>%
#     purrr::set_names(glue("PD-CDR3-Risk-Score_CDR3-{.}")) %>% 
#     purrr::map(
#         ~ run_lm_site_pos_pair_riskscore(
#             pdata = pdata, 
#             covar_df = covar_df,
#             cdr3_aa = .
#         )
#     )

# lm_results_df <- mod_list %>% 
#     purrr::map(broom::tidy) %>% 
#     bind_rows(.id = "model_name") %>% 
#     mutate(cdr3_aa = strex::str_after_last(model_name, "CDR3-")) %>% 
#     glimpse()

```


Loading in the results

```{r}
resdir <- glue("{wkdir}/data/interim/cdr3_qtl/results_cdr3-risk-score")

cdr3_betacoef <- list.files(resdir, pattern = ".csv", full.names = TRUE, recursive = TRUE) %>% 
    purrr::set_names(basename(.)) %>% 
    purrr::map(read_csv) %>% 
    bind_rows(.id = "file_name") %>% 
    glimpse()

cdr3_betacoef %<>% 
    mutate(
          descriptor = basename(file_name) %>% gsub("lm_model_results__", "", .),
          pheno = str_before_first(descriptor, "__"),
          cdr3_pos = descriptor %>% str_after_first("CDR3pos-") %>% str_before_first("__"),
          cdr3_len = descriptor %>% str_after_first("CDR3len-") %>% str_before_first(".csv")
    ) %>% 
    janitor::clean_names() %>% 
    filter(!is.na(cdr3_aa)) %>% 
    glimpse()

# # viewing table of top hits across all phenotypes
# cdr3_betacoef %>% 
#     filter(term == "pd_hla_risk_score") %>% 
#     slice_min(p_value, n = 300) %>% 
#     View()

# Visualizing the distribution of p-values across all phenotypes
cdr3_betacoef %>% 
    ggplot(aes(p_value)) +
    facet_wrap(~pheno,ncol=2) +
    geom_histogram(bins = 100) +
    theme_bw() +
    labs(title = "CDR3 risk score beta-coefficients")

cdr3_betacoef_sig <- cdr3_betacoef %>% 
    group_by(pheno) %>% 
    filter(term == "pd_hla_risk_score") %>% 
    mutate(adj_p = p.adjust(p_value, method = "BH")) %>% 
    filter(adj_p < 0.05) %>% 
    arrange(adj_p) %>% 
    glimpse()

# Number of significant hits across all phenotypes
cdr3_betacoef_sig$pheno %>% table()

betacoef_list <- cdr3_betacoef_sig %>% split(.$pheno)

cdr3_betacoef %>% 
    filter(pheno == "cdr3_p2_strat_filtered_cv1k") %>% 
    glimpse()

# ----------------------------------------------------------------------------
# Creating sequence logo for the most significant CDR3 AA associations ----

library(ggseqlogo)
logo_df_full <- betacoef_list$cdr3_p2_strat_filtered_cv1k %>% 
    mutate(cdr3_len = as.numeric(cdr3_len)) %>% 
    mutate(cdr3_pos = factor(cdr3_pos, levels = all_imgt_pos)) %>% 
    glimpse()
# View(logo_df_full)

logo_df_summed <- logo_df_full %>% 
    group_by(cdr3_aa, cdr3_pos) %>% 
    summarize(summed_effect = sum(estimate), .groups = "drop") %>% 
    glimpse()
logo_df_summed_mat <- logo_df_summed %>% 
    arrange(cdr3_pos) %>% 
    pivot_wider(names_from = cdr3_pos, values_from = summed_effect, values_fill = 0) %>% 
    column_to_rownames(var = "cdr3_aa") %>% 
    as.matrix()

# Saving cumulative sequence logo
pseqlogo_summed <- ggseqlogo(logo_df_summed_mat, seq_type='aa',  method='custom') +
    ylab('Effect size') +
    xlab('CDR3 amino acid position') +
    scale_x_continuous(
        breaks = 1:ncol(logo_df_summed_mat),                 # positions to tick
        labels = colnames(logo_df_summed_mat)                # set custom labels
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_p2_strat_filtered_cv1k_logo_summed.png"),
    plot = pseqlogo_summed,
    width = 3.5,
    height = 3,
    dpi = 600
)
ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_p2_strat_filtered_cv1k_logo_summed_large.png"),
    plot = pseqlogo_summed,
    width = 10,
    height = 4
)

# To create sequence logo - sum the effect sizes of significant CDR3 AA assocations 

logo_df_full %>% glimpse
# l <- 15
for (l in unique(logo_df_full$cdr3_len)) {
    logo_df_summed <- logo_df_full %>% 
        filter(cdr3_len == l) %>% 
        group_by(cdr3_aa, cdr3_pos) %>% 
        summarize(summed_effect = sum(estimate), .groups = "drop") %>% 
        glimpse()
    logo_df_summed_mat <- logo_df_summed %>% 
        pivot_wider(names_from = cdr3_pos, values_from = summed_effect, values_fill = 0) %>% 
        column_to_rownames(var = "cdr3_aa") %>% 
        as.matrix()
    pseqlogo_summed_len <- ggseqlogo(logo_df_summed_mat, seq_type='aa',  method='custom') +
        ylab('Effect size') +
        xlab('CDR3 amino acid position') +
        ggtitle(glue("CDR3 length {l}")) +
        scale_x_continuous(
            breaks = 1:ncol(logo_df_summed_mat),                 # positions to tick
            labels = colnames(logo_df_summed_mat)                # set custom labels
        ) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
    ggsave(
        filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_p2_strat_filtered_cv1k_logo_summed_len{l}.png"),
        plot = pseqlogo_summed_len,
        width = 5,
        height = 4
    )

}





```


# Scoring TCRs with HLA risk score Beta-Coefficients to generate CDR3 risk scores

Here we develop a single case of scoring a test TCR

```{r}
trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()

trust4_reports_tcrb %>% glimpse()
betacoef_list %>% glimpse()


# Breaking down the function
x <- "CASSVGQGLGYTF"
# align_imgt_df()
# df <- data.frame(ptarget=unlist(strsplit(x, "")), pos=1:len_cdr3 )
# df$tag <- paste0(df$ptarget, ":", df$pos);

len_cdr3 <- nchar(x)
cdr3_aln <- align_imgt(x)
cdr3_data <- data.frame(ptarget = cdr3_aln[1,], pos = colnames(cdr3_aln)) %>% 
    mutate(tag = paste0(ptarget, ":", pos)) %>% 
    glimpse()
phenotype_type <- "cdr3_p2_strat_filtered_cv1k"
pos_beta <- subset(betacoef_list[[phenotype_type]], cdr3_len==len_cdr3 ) #len_cdr3=15
pos_beta$tag <- paste0(pos_beta$cdr3_aa, ":", pos_beta$cdr3_pos)
pos_beta <- pos_beta[,c("tag","estimate")]

cdr3_data <- merge( pos_beta, cdr3_data, by="tag" )
cdr3_data
if ( nrow(cdr3_data) == 0 ){
    score = 0
} else {
    score = sum(cdr3_data$estimate)
}
score

```


Executing TCR scoring at scale for all 6 + million TCRs

```{r}
# quick speed test
for (k in c(10, 100, 1000)) {
    tic()
    score_tcrs_with_betacoefs_list(
        trust4_reports_tcrb$CDR3aa[1:k],
        "cdr3_p2_strat_filtered_cv1k",
        betacoef_list
        )
    toc()
}
# 100k would proably take 1.5 minutes, 1M would take 15 minutes 6 million TCRs would take 1.5 hours * 4 iterations - is way too slow

future::plan(multisession, workers = 8)
p_score_list <- list()
for (p in names(betacoef_list)) {
    message(glue("{get_time()} Scoring {p}..."))
    p_score_list[[p]] <- trust4_reports_tcrb$CDR3aa %>% 
        purrr::set_names() %>% 
        furrr::future_map(
            ~ score_tcrs_with_betacoefs(x = ., 
                phenotype_type = p,
                betacoef_list)
        )
}


```


```{r}
# slurm batchtools parallelzed alternative
# lets put about 100k TCRs per chunk
length(trust4_reports_tcrb$CDR3aa)/100000
tcr_chunks <- chunk_func(trust4_reports_tcrb$CDR3aa, 67) # 67 chunks of 100k TCRs
names(tcr_chunks)

batch_params <- expand.grid(
    "x" = tcr_chunks,
    "phenotype_type" = names(betacoef_list)) %>% 
    as.data.frame() %>% 
    mutate(betacoef_list = list(betacoef_list)) %>% 
    mutate(filename = glue(
        "{tcr_cdr3_riskscore_dir}/chunk_{names(x)}__{phenotype_type}__padj05.rds")) %>% 
    glimpse()

# configure registry ----
breg <- breg_setup(cluster_name = "CDR3_QTL_RISK_SCORE", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )

# Submit Jobs ----
jobs <- batchMap(
  fun = score_tcrs_with_betacoefs_list,
  args = batch_params,
  reg = breg
)
nrow(batch_params)

# jobs[, chunk := chunk(job.id, chunk.size = 40)]
# print(jobs[, .N, by = chunk])

# can probably do two per minute so x jobs / 2 = minutes
submitJobs(jobs,
  resources = list(
    walltime = "1:00:00",
    memory = "5GB",
    ncpus = 1,
    max.concurrent.jobs = 9999
  )
)

```

Aggregating scored TCRs

```{r}
chunked_tcr_scores <- list.files(
    tcr_cdr3_riskscore_dir,
    pattern = "chunk_",
    full.names = TRUE,
    recursive = TRUE
) 

name_map <- c(
    "cdr3_p1_strat_filtered_cv1k" = "risk_score_p1_full_strat",
    "cdr3_p1_unstrat_filtered_cv1k" = "risk_score_p1_full_unstrat",
    "cdr3_p2_strat_filtered_cv1k" = "risk_score_p2_glrm_strat",
    "cdr3_p2_unstrat_filtered_cv1k" = "risk_score_p2_glrm_unstrat"
)

tmp_df <- list()
for (p in cdr3_phenotypes) {
    tmp_df[[p]] <- chunked_tcr_scores %>% 
        keep(grepl(p, .)) %>% 
        purrr::set_names(basename(.)) %>% 
        purrr::map(~ readRDS(.) %>% stack()) %>% 
        bind_rows(.id = "path") %>% 
        dplyr::rename(CDR3aa = ind, !!name_map[[p]] := values) %>% 
        mutate(CDR3aa = as.character(CDR3aa)) %>% 
        mutate(chunk_n = str_after_first(path, "chunk_") %>% str_before_first("__") %>% as.numeric()) %>% 
        arrange(chunk_n) %>% 
        glimpse()
}

# checking that CDR3aa are all aligned
cdr3_phenotypes %>% purrr::map(
    ~ all(tmp_df[[.]]$CDR3aa == tmp_df[["cdr3_p1_strat_filtered_cv1k"]]$CDR3aa)
)

# tmp_df[["cdr3_p1_strat_filtered_cv1k"]] %>% glimpse()
# tmp_df[["cdr3_p1_strat_filtered_cv1k"]]$path %>% unique()

# tst <- tmp_df[["cdr3_p1_strat_filtered_cv1k"]] %>% 
#     mutate(chunk_n = str_after_first(path, "chunk_") %>% str_before_first("__") %>% as.numeric()) %>% 
#     arrange(chunk_n) %>% 
#     glimpse()

tst$chunk_n %>% table()

cdr3_risk_score_df <- data.frame(
    CDR3aa = tmp_df[[cdr3_phenotypes[1]]]$CDR3aa,
    risk_score_p1_full_strat = tmp_df[["cdr3_p1_strat_filtered_cv1k"]]$risk_score_p1_full_strat,
    risk_score_p1_full_unstrat = tmp_df[["cdr3_p1_unstrat_filtered_cv1k"]]$risk_score_p1_full_unstrat,
    risk_score_p2_glrm_strat = tmp_df[["cdr3_p2_strat_filtered_cv1k"]]$risk_score_p2_glrm_strat,
    risk_score_p2_glrm_unstrat = tmp_df[["cdr3_p2_unstrat_filtered_cv1k"]]$risk_score_p2_glrm_unstrat
) %>% 
    glimpse()

saveRDS(cdr3_risk_score_df, 
    glue("{tcr_cdr3_riskscore_complied}/",
        "risk_scores_padj05_{Sys.Date()}.rds"))

```

# Visualizing CDR3 risk score results

```{r}
cdr3_risk_score_df <- readRDS(
    glue("{tcr_cdr3_riskscore_complied}/",
        "risk_scores_padj05_2025-08-10.rds")) %>% 
        dplyr::rename(ind = CDR3aa)
cdr3_risk_score_df %>% glimpse()

trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()

metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  mutate(case_control_other_latest = factor(case_control_other_latest, levels = c("Control", "Case"))) %>% 
  glimpse()

trust4_reports_tcrb_c3rc <- bind_cols(
    trust4_reports_tcrb,
    cdr3_risk_score_df) %>% 
    glimpse()

# Checking that CDR3aa are all aligned
trust4_reports_tcrb_c3rc$CDR3aa[1:20000] == trust4_reports_tcrb_c3rc$ind[1:20000]
all(trust4_reports_tcrb_c3rc$CDR3aa == trust4_reports_tcrb_c3rc$ind)

trust4_reports_tcrb_c3rc %>% glimpse()

avg_cdr3_risk_score_df <- trust4_reports_tcrb_c3rc %>% 
    group_by(participant_id, CDR3aa_length) %>% 
    # dplyr::summarize(
    #     avg_risk_score_p1_full_strat = mean(risk_score_p1_full_strat, na.rm = TRUE),
    #     avg_risk_score_p1_full_unstrat = mean(risk_score_p1_full_unstrat, na.rm = TRUE),
    #     avg_risk_score_p2_glrm_strat = mean(risk_score_p2_glrm_strat, na.rm = TRUE),
    #     avg_risk_score_p2_glrm_unstrat = mean(risk_score_p2_glrm_unstrat, na.rm = TRUE)
    # ) %>% 
    dplyr::summarize(
        across(starts_with("risk_score"), 
               list(
                   avg = ~mean(., na.rm = TRUE),
                   p75 = ~quantile(., 0.75, na.rm = TRUE),
                   p90 = ~quantile(., 0.90, na.rm = TRUE)
               ), 
               .names = "{.col}_{.fn}"),
    ) %>% 
    dplyr::left_join(
        metadata %>% select(participant_id, case_control_other_latest, pd_hla_risk_score),
        relationship = "many-to-many") %>% 
    dplyr::distinct() %>% 
    glimpse()

# avg_cdr3_risk_score_df %>% glimpse()
avg_cdr3_risk_score_df_long <- avg_cdr3_risk_score_df %>% 
    dplyr::select(-contains("_p75"), -contains("_p90")) %>% 
    pivot_longer(
        cols = ends_with("avg"),
        names_to = "risk_score_type",
        values_to = "risk_score_value"
    ) %>%
    dplyr::mutate(risk_score_type = case_when(
        risk_score_type == "risk_score_p1_full_strat_avg" ~ "P1 Full Stratified",
        risk_score_type == "risk_score_p1_full_unstrat_avg" ~ "P1 Full Unstratified",
        risk_score_type == "risk_score_p2_glrm_strat_avg" ~ "P2 GLRM Stratified",
        risk_score_type == "risk_score_p2_glrm_unstrat_avg" ~ "P2 GLRM Unstratified"
    )) %>% 
    glimpse()

p_cdr3vshla_risk_score <- avg_cdr3_risk_score_df_long %>% 
    filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    ggplot(aes(x = pd_hla_risk_score, y = risk_score_value, color = case_control_other_latest)) +
    geom_point(alpha = 0.8, shape = 21) +
    theme_bw() +
    geom_smooth(method = "lm") +
    scale_color_d3() +
    ggpubr::stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.9) +
    facet_grid(CDR3aa_length~risk_score_type, scales = "free") +
    labs(title="CDR3 risk score",x="HLA risk score",y="CDR3 risk score") +
    theme(legend.position = "bottom")

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_vshla_risk_score_all_methods.png"),
    plot = p_cdr3vshla_risk_score,
    width = 15,
    height = 15
)


p_cdr3vshla_risk_score_onlyp2strat <- avg_cdr3_risk_score_df_long %>% 
    filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    filter(risk_score_type == "P2 GLRM Stratified") %>% 
    ggplot(aes(x = pd_hla_risk_score, y = risk_score_value, color = case_control_other_latest)) +
    geom_point(alpha = 0.8, shape = 21) +
    theme_bw() +
    geom_smooth(method = "lm") +
    scale_color_d3() +
    ggpubr::stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.99) +
    facet_wrap(~CDR3aa_length, nrow = 2, scales = "free") +
    labs(title="CDR3 risk score vs HLA risk score",x="HLA risk score",y="CDR3 risk score (P2 GLRM Stratified)") +
    theme(legend.position = "bottom")

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_vshla_risk_score_p2strat.png"),
    plot = p_cdr3vshla_risk_score_onlyp2strat,
    width = 10,
    height = 8
)

p_cdr3vshla_risk_score_onlyp2strat_len15only <- avg_cdr3_risk_score_df_long %>% 
    filter(CDR3aa_length == 15) %>%
    filter(risk_score_type == "P2 GLRM Stratified") %>% 
    ggplot(aes(x = pd_hla_risk_score, y = risk_score_value, color = case_control_other_latest)) +
    geom_point(alpha = 0.5, shape = 21) +
    theme_bw() +
    geom_smooth(method = "lm") +
    scale_color_d3() +
    ggpubr::stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.99) +
    # facet_wrap(~CDR3aa_length, nrow = 2, scales = "free") +
    labs(x="HLA risk score", y="CDR3 risk score") +
    theme(legend.position = "none")

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_vshla_risk_score_p2strat_len15only.png"),
    plot = p_cdr3vshla_risk_score_onlyp2strat_len15only,
    width = 2.5,
    height = 2.5
)

p_cdr3vscasecontrol <- avg_cdr3_risk_score_df_long %>% 
    filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    filter(risk_score_type == "P2 GLRM Stratified") %>% 
    ggplot(aes(x = case_control_other_latest, y = risk_score_value)) +
    geom_point(alpha = 0.3, position = position_jitter(width = 0.2)) +
    geom_boxplot(outlier.shape = NA) +
    facet_wrap(~CDR3aa_length, nrow = 2, scales = "free_y") +
    theme_bw() +
    labs(title="Average CDR3 risk score",x="Case/Control",y="CDR3 risk score (P2 GLRM Stratified)")

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_vscasecontrol.png"),
    plot = p_cdr3vscasecontrol,
    width = 8,
    height = 5
)



# Looking at different points of the risk score distribution
avg_cdr3_risk_score_df %>% glimpse()

avg_cdr3_risk_score_df_long_p75 <- avg_cdr3_risk_score_df %>% 
    pivot_longer(
        cols = "risk_score_p2_glrm_strat_p75",
        names_to = "risk_score_type",
        values_to = "risk_score_value"
    ) %>% 
    # dplyr::mutate(risk_score_type = case_when(
    #     risk_score_type == "avg_risk_score_p1_full_strat" ~ "P1 Full Stratified",
    #     risk_score_type == "avg_risk_score_p1_full_unstrat" ~ "P1 Full Unstratified",
    #     risk_score_type == "avg_risk_score_p2_glrm_strat" ~ "P2 GLRM Stratified",
    #     risk_score_type == "avg_risk_score_p2_glrm_unstrat" ~ "P2 GLRM Unstratified"
    # )) %>% 
    glimpse()

p_cdr3vscasecontrol_p75 <- avg_cdr3_risk_score_df_long_p75 %>% 
    filter(CDR3aa_length == 15) %>% 
    # filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    # filter(risk_score_type == "P2 GLRM Stratified") %>% 
    ggplot(aes(x = case_control_other_latest, y = risk_score_value)) +
    geom_point(alpha = 0.3, position = position_jitter(width = 0.2)) +
    geom_boxplot(outlier.shape = NA) +
    # ggpubr::stat_compare_means(method = "wilcox.test", label = "p.signif") +
    facet_wrap(~CDR3aa_length, nrow = 2, scales = "free_y") +
    theme_bw() +
    labs(title="CDR3 risk score at \n75th percentile",x=NULL, y="CDR3 risk score") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_vs_casecontrol_p75.png"),
    plot = p_cdr3vscasecontrol_p75,
    width = 3,
    height = 5
)

```


What does this distribution look like within repertoires?

```{r}
# trust4_reports_tcrb_c3rc %>% 
#     filter(CDR3aa_length == 15) %>% glimpse()

# Create ggridgeline density plot for 25 randomly selected participants
# Load required package
library(ggridges)

# Set seed for reproducibility and select 25 random participants
set.seed(123)
random_participants <- trust4_reports_tcrb_c3rc$participant_id %>% 
    unique() %>% 
    sample(100)

# Create the ggridgeline density plot
ggridge_density_plot <- trust4_reports_tcrb_c3rc %>% 
    # filter(participant_id %in% random_participants) %>% 
    # filter(CDR3aa_length == 15) %>% 
    filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    ggplot(aes(
        x = risk_score_p2_glrm_strat, 
        y = case_control_other_latest,
        fill = case_control_other_latest
    )) +
    geom_density_ridges(
        alpha = 0.7,
        scale = 2,
        rel_min_height = 0.01
    ) +
    scale_fill_d3() +
    theme_ridges() +
    facet_wrap(~CDR3aa_length, ncol = 1, scales = "free_y") +
    labs(
        title = "Distribution of CDR3 Risk Scores (P2 GLRM Stratified)",
        subtitle = "25 Randomly Selected Participants - CDR3 Length 15",
        x = "Risk Score (P2 GLRM Stratified)",
        y = "Participant ID",
        fill = "Case/Control Status"
    ) +
    theme(
        legend.position = "bottom",
        axis.text.y = element_text(size = 8),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
    )

# Display the plot
ggridge_density_plot

# Save the plot
ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/ggridge_density_25participants_p2strat.png"),
    plot = ggridge_density_plot,
    width = 6,
    height = 12,
    dpi = 300
)
# Create the ggridgeline density plot
ggridge_density_plot <- trust4_reports_tcrb_c3rc %>% 
    # filter(participant_id %in% random_participants) %>% 
    # filter(CDR3aa_length == 15) %>% 
    filter(CDR3aa_length >= 12 & CDR3aa_length <= 18) %>% 
    ggplot(aes(
        x = risk_score_p2_glrm_strat, 
        y = case_control_other_latest,
        fill = case_control_other_latest
    )) +
    geom_density_ridges(
        alpha = 0.7,
        scale = 2,
        rel_min_height = 0.01
    ) +
    scale_fill_d3() +
    theme_ridges() +
    facet_wrap(~CDR3aa_length, ncol = 1, scales = "free_y") +
    labs(
        title = "Distribution of CDR3 Risk Scores (P2 GLRM Stratified)",
        subtitle = "25 Randomly Selected Participants - CDR3 Length 15",
        x = "Risk Score (P2 GLRM Stratified)",
        y = "Participant ID",
        fill = "Case/Control Status"
    ) +
    theme(
        legend.position = "bottom",
        axis.text.y = element_text(size = 8),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
    )

# Display the plot
ggridge_density_plot

# Save the plot
ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/ggridge_density_25participants_p2strat.png"),
    plot = ggridge_density_plot,
    width = 6,
    height = 12,
    dpi = 300
)

```


Attempt at looking at clusters of values

```{r}
library(dbscan)
my_vector <- c(1, 1.2, 1.1, 5, 5.2, 5.1, 10, 10.3, 10.2, 15, 15.1, 15.2, 20)
my_data <- as.matrix(my_vector) # Convert to matrix
kNNdistplot(my_data, k = 2) # k should be minPts - 1
# Choose suitable eps and minPts based on kNNdistplot and data characteristics
# For example:
eps_val <- 0.5 
minPts_val <- 2

dbscan_result <- dbscan(my_data, eps = eps_val, minPts = minPts_val)

```



```{r}
# Load GGally if not already loaded
library(GGally)

# Create a correlation plot for the four average risk score variables
correlation_plot <- avg_cdr3_risk_score_df %>%
  select(
    avg_risk_score_p1_full_strat,
    avg_risk_score_p1_full_unstrat, 
    avg_risk_score_p2_glrm_strat,
    avg_risk_score_p2_glrm_unstrat
  ) %>%
  # Rename columns for better readability
  rename(
    "P1 Full Stratified" = avg_risk_score_p1_full_strat,
    "P1 Full Unstratified" = avg_risk_score_p1_full_unstrat,
    "P2 GLRM Stratified" = avg_risk_score_p2_glrm_strat,
    "P2 GLRM Unstratified" = avg_risk_score_p2_glrm_unstrat
  ) %>%
  ggpairs(
    # Customize the appearance
    upper = list(
      continuous = wrap("cor", 
                       method = "pearson",
                       size = 4,
                       color = "black",
                       digits = 3)
    ),
    lower = list(
      continuous = wrap("points", 
                       alpha = 0.6,
                       size = 0.8,
                       color = "steelblue")
    ),
    diag = list(
      continuous = wrap("densityDiag", 
                       alpha = 0.7,
                       fill = "steelblue",
                       color = "darkblue")
    )
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10, face = "bold")
  ) +
  labs(
    title = "Correlation Matrix of CDR3 Risk Scores",
    subtitle = "Pearson correlations between different CDR3 risk score methods"
  )

# Display the plot
correlation_plot

# Save the plots
ggsave(
  filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_correlation_plot.png"),
  plot = correlation_plot,
  width = 12,
  height = 10
)

```



Visualizing an example result
```{r}
source(paste0(src_dir, "/R_scripts/analyze_CDR3qtl.R"))

dlist <- generate_model_inputs_cdr3_riskscore(
    phenotype_type = "cdr3_p2_strat_filtered_cv1k",
    cdr3_pos = "P110",
    cdr3length = 13
)

# M <- dlist$M %>% glimpse()
pdata <- dlist$pdata %>% glimpse()
covar_df <- dlist$covar_df %>% glimpse()
df <- merge(pdata,covar_df,by="participant_id") %>% 
    # filter(AA == "R") %>% 
    glimpse()

# df %>% ggplot(aes(x=pd_hla_risk_score,y=normrate)) +
#     geom_point(alpha = 0.3) +
#     theme_bw() +
#     geom_smooth(method = "lm") +
#     facet_wrap(~AA,ncol=7,scale="free")+ #
#     labs(title="CDR3-R risk score",x="HLA risk score",y="CDR3-R frequency")

df %>% ggplot(aes(x=pd_hla_risk_score,y=normrate)) +
    geom_point(alpha = 0.8, shape = 21) +
    theme_bw() +
    geom_smooth(method = "lm") +
    facet_wrap(~AA,ncol=7,scale="free")+ #
    ggside::geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5, fill = "steelblue") +
    ggside::scale_xsidey_continuous(expand = c(0, 0)) +
    ggside::theme_ggside_void() +
    theme(
        ggside.panel.scale = 0.3,  # Increase space for y-side panel (default is 0.2)
        ggside.axis.text = element_blank(),
        ggside.axis.ticks = element_blank()
    ) +
    labs(title="CDR3 risk score (P110)",x="HLA risk score",y="CDR3 P2-AA (13aa) frequency")



dlist <- generate_model_inputs_cdr3_riskscore(
    phenotype_type = "cdr3_p2_unstrat_filtered_cv1k",
    cdr3_pos = "P110",
    cdr3length = NA_character_
)

# M <- dlist$M %>% glimpse()
pdata <- dlist$pdata %>% glimpse()
covar_df <- dlist$covar_df %>% glimpse()
df <- merge(pdata,covar_df,by="participant_id") %>% 
    # filter(AA == "R") %>% 
    glimpse()

p_example <-df %>% ggplot(aes(x=pd_hla_risk_score,y=normrate)) +
    geom_point(alpha = 0.8, shape = 21) +
    theme_bw() +
    geom_smooth(method = "lm") +
    facet_wrap(~AA,ncol=7,scale="free")+ #
    ggside::geom_ysidedensity(aes(x = after_stat(density)), alpha = 0.5, fill = "steelblue") +
    ggside::scale_xsidey_continuous(expand = c(0, 0)) +
    ggside::theme_ggside_void() +
    theme(
        ggside.panel.scale = 0.3,  # Increase space for y-side panel (default is 0.2)
        ggside.axis.text = element_blank(),
        ggside.axis.ticks = element_blank()
    ) +
    labs(title="CDR3 risk score (P110)",x="HLA risk score",y="CDR3 P2-AA frequency")

ggsave(
    filename = glue("{wkdir}/figures/CDR3-qtl/cdr3-risk-score/cdr3_risk_score_example_p2unstrat_p110-13.png"),
    plot = p_example,
    width = 13,
    height = 6
)

```
