---
title: "Clustering TCR repertoires"
editor: source
author: "Joe Boktor"
date: '2025-03-24'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---


Environment setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(glue)
library(listenv)
library(data.table)
library(seqinr)
library(immunarch)
library(nlme)
library(broom)
library(broom.mixed)
library(parallelDist)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(scico)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
dint <- paste0(wkdir, "/data/interim")

source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))
source(paste0(src_dir, "/R_scripts/_airr_analyze-permanova.R"))

scratch_dir <- "/resnick/scratch/jbok/PDMBS"
gliph2_plots_dir <- glue("{wkdir}/figures/AIRR/gliph2/")
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
arcasHLA_interim <- glue("{wkdir}/data/interim/airr/arcasHLA")
gliph2_input_dir <- glue("{wkdir}/data/input/gliph2/param_files")
gliph2_res_dir <- glue("{wkdir}/data/interim/airr/gliph2/processing")
clustcr_dir <- glue("{wkdir}/data/interim/clusTCR")
clustcr_figs <- glue("{wkdir}/figures/clusTCR")
tcr_cdr3_riskscore_complied <- 
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")
cohort_subset_dir <- glue("{clustcr_dir}/results/cohort_subsets")

# dir.create(clustcr_figs, showWarnings = FALSE, recursive = TRUE)
# dir.create(glue("{clustcr_dir}/input"), showWarnings = FALSE, recursive = TRUE)
# dir.create(glue("{clustcr_dir}/results"), showWarnings = FALSE, recursive = TRUE)

```


Loading metadata and TCRB data

```{r, eval=FALSE}
trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()

trust4_reports_tcrb %>% glimpse()
# trust4_reports_tcrb$sample_id %>% unique() %>% length()
# trust4_reports_tcrb$participant_id %>% unique() %>% length()
metadata %>% select(participant_id, case_control_other_latest) %>% distinct() %>%
  select(case_control_other_latest) %>% table()

clonotype_volume_df <- trust4_reports_tcrb %>% 
  group_by(sample_id) %>% 
  # group_by(participant_id) %>% 
  dplyr::summarize(
      clonal_volume = sum(`#count`),
      clonal_count = n()
      ) %>%
  glimpse()

clonotype_volume_df$clonal_count %>% summary()
clonotype_volume_df$clonal_volume %>% summary()

clonotype_volume_df$clonal_volume %>% summary()
low_vol_samples <- clonotype_volume_df %>% 
  filter(clonal_volume < 200) %>% 
  glimpse()

metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  left_join(clonotype_volume_df, by = "sample_id") %>% 
  glimpse()

metadata_categories_df <- readRDS(
  glue("{wkdir}/data/interim/metadata/metadata_categories_dataframe_2025-07-21.rds")) %>% 
  glimpse()

```


## Analysis Setup - Prepping data for GLIPH2

Prepping TRUST4 data for GLIPH2

```{r}
trust4_reports_tcrb %>% glimpse()

gliph2_cdr3b_df <- trust4_reports_tcrb %>%
    mutate("subject_condition" = paste0(sample_id, ":", case_control_other_latest)) %>%
    mutate(TRBV = str_before_first(V, "\\*")) %>%
    mutate(TRBJ = str_before_first(J, "\\*")) %>%
    mutate(CDR3a = "NA") %>% 
    dplyr::select(
        "#CDR3b" = CDR3aa, TRBV, TRBJ, CDR3a, 
        "subject:condition" = subject_condition, count = "#count"
    ) %>%
    glimpse()

gliph2_cdr3b_df %>% glimpse()

write.table(
  gliph2_cdr3b_df,
  file = glue("{gliph2_input_dir}/amppd_tcrb_{Sys.Date()}.txt"),
  quote = FALSE,
  sep = "\t",
  col.names = FALSE,
  row.names = FALSE
)

trust4_reports_tcrb %>% glimpse()

clustcr_cdr3b_df <- trust4_reports_tcrb %>%
    mutate(TRBV = str_before_first(V, "\\*")) %>%
    mutate(TRBJ = str_before_first(J, "\\*")) %>%
    dplyr::select(
        "CDR3b_aa" = CDR3aa, TRBV, TRBJ, sample_id, participant_id,
        condition = case_control_other_latest, count = "#count"
    ) %>%
    glimpse()

readr::write_tsv(clustcr_cdr3b_df, file = glue("{clustcr_dir}/input/amppd_tcrb_{Sys.Date()}.tsv"))

# latest file:
# amppd_tcrb_2025-08-04

# Saaving cohort subssets for clustTCR analysis 
full_amppd_tcrb <- trust4_reports_tcrb %>%
  mutate(TRBV = str_before_first(V, "\\*")) %>%
  mutate(TRBJ = str_before_first(J, "\\*")) %>%
  dplyr::select(
      "CDR3b_aa" = CDR3aa, TRBV, TRBJ, sample_id, participant_id,
      condition = case_control_other_latest, count = "#count", study
  ) %>%
  glimpse()

# smalltest <- full_amppd_tcrb %>% sample_n(10000)
# readr::write_tsv(smalltest, file = glue("{clustcr_dir}/input/amppd_tcrb_small_test_{Sys.Date()}.tsv"))

for (cohort in unique(full_amppd_tcrb$study)) {
  message(glue("Writing {cohort} data"))
  one_cohort_df <- full_amppd_tcrb %>% filter(study == cohort)
  readr::write_tsv(one_cohort_df, file = glue("{clustcr_dir}/input/ONLY_{cohort}_{Sys.Date()}.tsv"))
  all_except_one_cohort_df <- full_amppd_tcrb %>% filter(study != cohort)
  readr::write_tsv(all_except_one_cohort_df, file = glue("{clustcr_dir}/input/ALL_EXCEPT_{cohort}_{Sys.Date()}.tsv"))
}

```


Saving HLA file for GLIPH2

```{r}
genotypes_cleaned <- readRDS(
    glue("{arcasHLA_interim}/genotypes_cleaned_2025-07-06.rds")
)

# alleles_to_remove <- c("E", "F", "G", "H", "J", "K", "L")
# alleles_to_remove <- c(
#     paste0(alleles_to_remove, "1"), 
#     paste0(alleles_to_remove, "2")
#     )
mhc1_alleles <- c("A", "B", "C")
mhc2_alleles <- c("DPA1", "DPB1", "DQA1", "DQB1", "DRB1")
major_hla_alleles <- c(mhc1_alleles, mhc2_alleles)

gliph2_hla_df <- genotypes_cleaned %>%
    left_join(metadata %>% select(participant_id, sample_id, case_control_other_latest),
        by = c("participant_id" = "participant_id")
    ) %>%
    select(-c(participant_id, case_control_other_latest)) %>%
    select(sample_id, all_of(paste0(rep(major_hla_alleles, each = 2), 1:2))) %>%
    # select(-all_of(alleles_to_remove)) %>%
    glimpse()

write.table(
  gliph2_hla_df,
  file = glue("{gliph2_input_dir}/amppd_hla_{Sys.Date()}.txt"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

# latest file:
# amppd_hla_2025-07-08
```


### Running GLIPH2

Gliph2 executable for Linux server Centos 7  was downloaded from [here](http://50.255.35.37:8080/) including [Human v2.0](http://50.255.35.37:8080/tools) reference data.


The command below was executed in a slurm script found in `notebooks/shell_scripts/slurm_run_gliph2.sh`


```{r}
# # Running test
# cd /resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/airr/gliph2/tcrb_results_test

# /resnick/groups/MazmanianLab/jboktor/software/gliph2/irtools.centos -c /resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/input/gliph2/param_files/gliph2_tcrb_TEST.cfg

# exec_path <- glue("{home_dir}/software/gliph2/irtools.centos")
# gliph2_out_dir <- glue("{wkdir}/data/interim/airr/gliph2/tcrb_results")
# dir.create(gliph2_out_dir, showWarnings = FALSE, recursive = TRUE)

# glue(
#   "cd {gliph2_out_dir} && ",
#   "{exec_path} -c {gliph2_input_dir}/gliph2_tcrb.cfg"
#   )
```


# Running ClusTCR

```{r}
# cd /resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/notebooks/python_scripts
# mamba activate clustcr
# python run_clusTCR.py

```


Running the generalized clusTCR script

```{r}
# # Testing generalized script
# shell_do(
#   glue(
#     "mamba activate clustcr && ",
#     "python {src_dir}/python_scripts/run_clusTCR_general.py",
#       " --tcrb {clustcr_dir}/input/amppd_tcrb_small_test_2025-08-22.tsv",
#       " --outdir {clustcr_dir}/results/NEWRESULTSTESTME",
#       " --outname TESTNEWSCRIPT",
#       " --include-vgene",
#       " --cdr3-col CDR3b_aa",
#       " --vgene-col TRBV",
#       " --n-cpus all"
#   )
# )

tcrb_input_paths <- list.files(glue("{clustcr_dir}/input"), pattern = "ONLY_|ALL_EXCEPT_")
names(tcrb_input_paths) <- str_before_last(tcrb_input_paths, "_")

for (dataset in names(tcrb_input_paths)) {
  message(glue("Running {dataset}"))
  shell_do(
    glue(
      "sbatch --job-name=clustcr_test_{dataset} ",
      "--cpus-per-task=32 ",
      "--time=1-00:00:00 ",
      "--mem=300G ",
      "--output={wkdir}/.cluster_runs/{get_time()}_clustcr_test_{dataset}_%j.out ",
      "--error={wkdir}/.cluster_runs/{get_time()}_clustcr_test_{dataset}_%j.err ",
      "--wrap=\"",
      "source ~/.bashrc && mamba activate clustcr && ",
      "python {src_dir}/python_scripts/run_clusTCR_general.py",
        " --tcrb {clustcr_dir}/input/{tcrb_input_paths[dataset]}",
        " --outdir {clustcr_dir}/results/cohort_subsets/{dataset}",
        " --outname {dataset}",
        " --include-vgene",
        " --cdr3-col CDR3b_aa",
        " --vgene-col TRBV",
        " --n-cpus 32",
      "\""
    )
  )
}

# Running levenshtein script
for (dataset in names(tcrb_input_paths)) {
  message(glue("Running {dataset}"))
  shell_do(
    glue(
      "sbatch --job-name=clustcr_test_{dataset} ",
      "--cpus-per-task=32 ",
      "--time=1-00:00:00 ",
      "--mem=300G ",
      "--output={wkdir}/.cluster_runs/{get_time()}_clustcr_test_{dataset}_%j.out ",
      "--error={wkdir}/.cluster_runs/{get_time()}_clustcr_test_{dataset}_%j.err ",
      "--wrap=\"",
      "source ~/.bashrc && mamba activate clustcr && ",
      "python {src_dir}/python_scripts/run_clusTCR_general_levenshtein.py",
        " --tcrb {clustcr_dir}/input/{tcrb_input_paths[dataset]}",
        " --outdir {clustcr_dir}/results/cohort_subsets/{dataset}_levenshtein",
        " --outname {dataset}",
        " --include-vgene",
        " --cdr3-col CDR3b_aa",
        " --vgene-col TRBV",
        " --n-cpus 32",
      "\""
    )
  )
}

# Run entire dataset with levenshtein distance
shell_do(
  glue(
    "sbatch --job-name=clustcr_test_levenshtein_full ",
    "--cpus-per-task=32 ",
    "--time=2-00:00:00 ",
    "--mem=600G ",
    "--output={wkdir}/.cluster_runs/{get_time()}_clustcr_test_levenshtein_full_%j.out ",
    "--error={wkdir}/.cluster_runs/{get_time()}_clustcr_test_levenshtein_full_%j.err ",
    "--wrap=\"",
    "source ~/.bashrc && mamba activate clustcr && ",
    "python {src_dir}/python_scripts/run_clusTCR_general_levenshtein.py",
      " --tcrb {clustcr_dir}/input/amppd_tcrb_2025-08-04.tsv",
      " --outdir {clustcr_dir}/results/levenshtein_full",
      " --outname levenshtein_full",
      " --include-vgene",
      " --cdr3-col CDR3b_aa",
      " --vgene-col TRBV",
      " --n-cpus 32",
      "\""
  )
)

```



## Interpreting & Wrangling ClusTCR results

```{r}
# loading sample - TCRb data
clustcr_cdr3b_df <- readr::read_tsv(
  glue("{clustcr_dir}/input/amppd_tcrb_2025-08-04.tsv")) %>% 
  dplyr::rename(case_control_other_latest = condition) %>% 
  glimpse()
clustcr_cdr3b_df %>% glimpse()

# Reading in CDR3 risk score data
cdr3_risk_score_df <- readRDS(glue("{tcr_cdr3_riskscore_complied}/",
        "risk_scores_padj05_2025-08-10.rds")) %>% 
    distinct() %>% 
    dplyr::rename(CDR3b_aa = CDR3aa) %>% 
    glimpse()
cdr3_risk_score_df %>% glimpse()

# olga/sonnia probabilities 
genprob_case <- read_csv(
  glue("{dint}/olga_sonnia/results/tcrb_case_results/tcrb_case_results.csv")) %>% 
  mutate(case_control_other_latest = "Case") %>% 
  dplyr::rename(CDR3b_aa = amino_acid, TRBV = v_gene, TRBJ = j_gene)
genprob_ctrl <- read_csv(
  glue("{dint}/olga_sonnia/results/tcrb_ctrl_results/tcrb_ctrl_results.csv")) %>% 
  mutate(case_control_other_latest = "Control") %>% 
  dplyr::rename(CDR3b_aa = amino_acid, TRBV = v_gene, TRBJ = j_gene)
genprob <- bind_rows(genprob_case, genprob_ctrl) %>% 
  glimpse()

# Joining CDR3 risk score data and olga/sonnia probabilities to input data
tcrb_df <- clustcr_cdr3b_df %>% 
  left_join(cdr3_risk_score_df) %>% 
  left_join(genprob) %>% 
  glimpse()
tcrb_df %>% glimpse()

```


```{r}
# Loading ClusTCR output ----
clustcr_output <- read_csv(
  glue("{clustcr_dir}/results/clusTCR_output_vgene_restricted_31cpus.csv"),
  col_select = -1) %>% 
  glimpse()
clustcr_summary <- read_csv(
  glue("{clustcr_dir}/results/clusTCR_summary_vgene_restricted_31cpus.csv")) %>% 
  dplyr::rename(cluster = ...1) %>% 
  glimpse()
clustcr_summary$cluster  <- unique(clustcr_output$cluster)
clustcr_output %>% glimpse()
clustcr_summary %>% glimpse()


# Merging ClusTCR output with metadata ----
clustcr_results_df <- clustcr_output %>%
  left_join(clustcr_summary) %>%
  left_join(tcrb_df, 
    by = c("junction_aa" = "CDR3b_aa", "v_call" = "TRBV"),
    relationship = "many-to-many") %>%
  group_by(cluster) %>%
  mutate(
    unique_donors_per_clust = n_distinct(participant_id),
    unique_samples_per_clust = n_distinct(sample_id)
  ) %>%
  ungroup() %>% 
  mutate(motif_length = motif %>% 
         gsub("\\[.*?\\]", "X", .) %>%  # Replace bracketed content with single X
         gsub("\\.", "", .) %>%         # Remove dots
         gsub("[^A-Z]", "", .) %>%      # Keep only uppercase letters
         nchar()
  ) %>% 
  # mutate(cluster_id = glue("c{cluster}_{motif}")) %>% 
  mutate(cluster_id = glue("{motif}_{v_call}")) %>% 
  glimpse()

saveRDS(clustcr_results_df, 
  glue(
    "{clustcr_dir}/results/",
    "compiled_clustcr_results_df_{Sys.Date()}.rds")
  )

```

Visualziing trends between Selection score and CDR3 risk score

```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds")
  )  %>% glimpse()
clustcr_results_df %>% glimpse()

cluster_sscore <- clustcr_results_df %>% 
  dplyr::select(cluster, Q_data, risk_score_p2_glrm_strat) %>% 
  group_by(cluster) %>% 
  dplyr::summarize(
    avg_risk_score = mean(risk_score_p2_glrm_strat, na.rm = TRUE),
    avg_selection_score = mean(log10(Q_data), na.rm = TRUE)) %>% 
  glimpse()
cluster_stats <- clustcr_results_df %>% 
  dplyr::select(cluster, motif, unique_donors_per_clust, unique_samples_per_clust, motif_length) %>% 
  distinct() %>% 
  left_join(cluster_sscore) %>% 
  glimpse()

# visualziing top clusters by selection score and risk
cluster_stats %>% glimpse()
cluster_stats %>%
  filter(avg_risk_score != 0) %>%
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 9) %>% 
  select(cluster, avg_selection_score, avg_risk_score) %>% 
  distinct() %>% 
  ggplot(aes(x = avg_selection_score, y = avg_risk_score)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  ggside::geom_xsidedensity(aes(y = after_stat(density)), 
                           position = "stack", 
                           fill = "steelblue", 
                           alpha = 0.7) +
  ggside::geom_ysidedensity(aes(x = after_stat(density)), 
                           position = "stack", 
                           fill = "coral", 
                           alpha = 0.7) +
  ggside::scale_xsidey_continuous(breaks = NULL) +
  ggside::scale_ysidex_continuous(breaks = NULL) +
  theme_minimal() +
  theme(ggside.panel.scale = 0.3)

```


**Notes**
- Out of 6,679,301 non-unique CDR3b sequences, 2,977,430 get clustered into 210,314 groups
- We will filter our clusters to contain at least 3 unique donors (participants)


### Generating cluster membership matrices

```{r}
clustcr_filter_summarized <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 8) %>% 
  dplyr::group_by(sample_id, cluster_id) %>%
  dplyr::summarize(count = sum(count)) %>% 
  glimpse()

sid_count_sum <- clustcr_filter_summarized %>%
  dplyr::group_by(sample_id) %>% 
  dplyr::summarize(sid_total_count = sum(count)) %>% 
  glimpse()

clust_mat_weighted <- clustcr_filter_summarized %>% 
  left_join(sid_count_sum, by = "sample_id") %>% 
  mutate(norm_count = count / sid_total_count) %>% 
  dplyr::select(sample_id, cluster_id, norm_count) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = norm_count,
    values_fill = 0
  )
clust_mat_binary <- clustcr_filter_summarized %>% 
  mutate(present = 1) %>% 
  dplyr::select(sample_id, cluster_id, present) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = present,
    values_fill = 0
  )

# Saving data filtered for extreme Selection Scores or CDR3 risk scores
saveRDS(
  clust_mat_weighted, 
  glue("{clustcr_dir}/results/clust_mat_weighted_{Sys.Date()}.rds")
)
saveRDS(
  clust_mat_binary, 
  glue("{clustcr_dir}/results/clust_mat_binary_{Sys.Date()}.rds")
)

```



# Interpreting & Wrangling GLIPH2 results

Reading in GLIPH2 results

```{r}
gliph_dir <- glue("{wkdir}/data/interim/airr/gliph2/tcrb_results_try3")
cluster_df <- readr::read_csv(
    glue("{gliph_dir}/AMPPD_TCRB_cluster.csv")) %>%
    glimpse()

cluster_df %>% glimpse()
# cluster_df$final_score %>% log10() %>% {-(.)} %>% hist(1000)

sid_pid_key <- key_meta_df %>% 
    dplyr::select(sample_id, participant_id) %>% 
    glimpse()

cluster_df_meta <- cluster_df %>% 
    tidyr::separate(Sample, into = c("sample_id", "case_control_other_latest"), sep = ":", remove = FALSE) %>% 
    left_join(sid_pid_key) %>% 
    glimpse()

cluster_df_meta %>% glimpse()

saveRDS(cluster_df_meta, file = glue("{gliph2_res_dir}/cluster_df_meta_{Sys.Date()}.rds"))

```


Filtering clusters for Case/Control Fisher's test 

- Minimum 5 unique donors per cluster

- Motif length > 3 fixed positions

```{r}
cluster_df_meta <- readRDS(glue("{gliph2_res_dir}/cluster_df_meta_2025-05-23.rds"))

cluster_pid_comp <- cluster_df_meta %>% 
    dplyr::select(index, participant_id) %>% 
    distinct() %>% 
    group_by(index) %>% 
    dplyr::summarise(n_participants = n()) %>% 
    ungroup()

unique_pid_thresh_inx <- cluster_pid_comp %>% 
    filter(n_participants >= 5) %>% 
    pull(index)

cluster_df_meta_f1 <- cluster_df_meta %>% 
    filter(index %in% unique_pid_thresh_inx) %>% 
    glimpse()

cluster_df_meta_f1 %>% glimpse()

# Second round of filtering for clusters with Fisher score <= 0.05
#' This round of filtering remmoves motifs that are likely 
#' influenced by germline genes and are common in a naive 
#' inexperience human TCR repertoire population filtering for 
#' significant values increases the chance that the motif is 
#' associated with an antigenic peptide ..

cluster_df_meta_f2 <- cluster_df_meta_f1 %>%
    filter(Fisher_score <= 0.05) %>%
    glimpse()
cluster_df_meta_f2 %>% glimpse()

# cluster_df_meta_f2$pattern %>% unique
# cluster_df_meta_f2$pattern %>% 
#     gsub("\\%", "", .) %>% nchar()

# Filter 3 clusters with pattern length > 3 fixed pos AAs
cluster_df_meta_f3 <- cluster_df_meta_f2 %>% 
    mutate(pattern_len = gsub("\\%", "", pattern) %>% nchar()) %>% 
    filter(pattern_len > 3) %>% 
    glimpse()

```


```{r}
# Case / control total DONORS
sample_groups_df <- cluster_df_meta %>%
    dplyr::select(participant_id, case_control_other_latest) %>%
    distinct()
grp_n <- sample_groups_df$case_control_other_latest %>% table()
case_n_base <- grp_n[["Case"]]
ctrl_n_base <- grp_n[["Control"]]
cidx_list <- unique(cluster_df_meta_f3$index)

fisher_results_df <- tibble(
    index = numeric(),
    type = character(),
    cases_in_cluster = numeric(),
    controls_in_cluster = numeric(),
    cases_not_in_cluster = numeric(),
    controls_not_in_cluster = numeric(),
    fisher_p_value = numeric(),
    odds_ratio = numeric(),
    conf_low = numeric(),
    conf_high = numeric()
)
pb <- progress_bar$new(
    format = "Processing clusters [:bar] :percent eta: :eta",
    total = length(cidx_list),
    clear = FALSE,
    width = 80
)

# For each cluster, calculate the values and perform Fisher's test
for (cidx in cidx_list) {
    pb$tick()
    
    cidx_df <- cluster_df_meta_f3 %>%
        filter(index == cidx)
    
    # Get counts
    case_n_cluster <- cidx_df %>%
        filter(case_control_other_latest == "Case") %>%
        pull(participant_id) %>%
        n_distinct()
    ctrl_n_cluster <- cidx_df %>%
        filter(case_control_other_latest == "Control") %>%
        pull(participant_id) %>%
        n_distinct()
    
    # Calculate remaining counts
    case_n_remaining <- case_n_base - case_n_cluster
    ctrl_n_remaining <- ctrl_n_base - ctrl_n_cluster
    
    # Perform Fisher's test
    fisher_test <- fisher.test(matrix(
        c(
            case_n_cluster, ctrl_n_cluster,
            case_n_remaining, ctrl_n_remaining
        ),
        nrow = 2
    ))
    
    # Add results to dataframe
    fisher_results_df <- fisher_results_df %>%
        add_row(
            index = cidx,
            type = unique(cidx_df$type),
            cases_in_cluster = case_n_cluster,
            controls_in_cluster = ctrl_n_cluster,
            cases_not_in_cluster = case_n_remaining,
            controls_not_in_cluster = ctrl_n_remaining,
            fisher_p_value = fisher_test$p.value,
            odds_ratio = fisher_test$estimate[1],
            conf_low = fisher_test$conf.int[1],
            conf_high = fisher_test$conf.int[2]
        )
}

# Sort by p-value and display results
fisher_results_df <- fisher_results_df %>%
    mutate(adj_fisher_p_value = p.adjust(fisher_p_value, method = "BH")) %>%
    arrange(fisher_p_value) %>% 
    glimpse()

# View the first few rows
fisher_results_df %>% View()

```


## TESTING A FASTER VERSION OF THE FISHER'S TEST

```{r}
# To retain as many clusters as possible
cluster_df_meta_f3 <- cluster_df_meta_f1 %>% 
    mutate(pattern_len = gsub("\\%", "", pattern) %>% nchar()) %>% 
    filter(pattern_len > 3) %>% 
    glimpse()

cluster_df_meta_f3 %>% glimpse()

sample_groups_df <- cluster_df_meta_f3 %>%
    dplyr::select(participant_id, case_control_other_latest) %>%
    distinct()

grp_n <- sample_groups_df$case_control_other_latest %>% table()
case_n_base <- grp_n[["Case"]]
ctrl_n_base <- grp_n[["Control"]]
cidx_list <- unique(cluster_df_meta_f3$index)


# 1.  unique donors per (cluster × group) in one pass 
per_cluster <- cluster_df_meta_f3 %>%
  dplyr::distinct(index, type, participant_id, case_control_other_latest) %>%
  dplyr::count(index, type, case_control_other_latest, name = "n") %>%
  pivot_wider(names_from = case_control_other_latest,
              values_from = n,
              values_fill = 0)                                          # columns: Case, Control

per_cluster %>% glimpse()

# 2.  add the 'remaining' cells once 
per_cluster <- per_cluster %>%
  mutate(
    remCase = case_n_base - Case,
    remCtrl = ctrl_n_base - Control
  ) %>% 
  glimpse()

# # 3.  run Fisher’s test row-wise (vectorised over rows) 
# future::plan(future::multisession, workers = 7)
# fisher_stats <- furrr::future_pmap_dfr(
#   # supply the four cell counts as parallel vectors
#   per_cluster %>% dplyr::select(Case, Control, remCase, remCtrl),
#   function(Case, Control, remCase, remCtrl) {
#     ft <- fisher.test(matrix(c(Case+1, Control+1, remCase+1, remCtrl+1), nrow = 2))
#     list(
#       fisher_p_value = ft$p.value,
#       odds_ratio     = unname(ft$estimate),
#       conf_low       = ft$conf.int[1],
#       conf_high      = ft$conf.int[2]
#     )
#   }
# )

# # 4.  final result 
# fisher_results_df <- bind_cols(per_cluster, fisher_stats) %>% 
#     glimpse()

# fisher_results_df$adj_fisher_p_value %>% sort() %>% head
# fisher_results_df %>% 
#     slice_head(n = 10, by = fisher_p_value) %>% 
#     View()

# per_cluster: data-frame with Case, Control, remCase, remCtrl columns
future::plan(future::multisession, workers = 15)
chi2_stats <- furrr::future_pmap_dfr(
  per_cluster %>% select(Case, Control, remCase, remCtrl),
  function(Case, Control, remCase, remCtrl) {
    mat <- matrix(c(Case, Control, remCase, remCtrl), nrow = 2)
    ## 1. χ² test (no Yates correction keeps it comparable to Fisher)
    ct <- suppressWarnings(chisq.test(mat, correct = FALSE))
    ## 2. Odds ratio for context
    or <- (Case * remCtrl) / (Control * remCase)
    list(
      chi2_p_value = ct$p.value,
      chi2_stat    = unname(ct$statistic),
      odds_ratio   = or,
      exp_case_cluster    = ct$expected[1],
      exp_control_cluster = ct$expected[2]
    )
  }
)

chi2_results_df <- bind_cols(per_cluster, chi2_stats) %>% 
    mutate(adj_chi2_p_value = stats::p.adjust(chi2_p_value, method = "fdr")) %>%
    arrange(chi2_p_value) %>% 
    glimpse()

chi2_results_df %>% 
    slice_head(n = 10, by = chi2_p_value) %>% 
    View()

saveRDS(chi2_results_df, file = 
  glue("{gliph2_res_dir}/chi2_results_df_{Sys.Date()}.rds"))

```


TEMP FITTING LMs over time x disease metric to classify clusters as worsening, improving, or stagnant

```{r}
library(broom)

chi2_results_df <- readRDS(
  glue("{gliph2_res_dir}/chi2_results_df_2025-05-26.rds"))

full_meta_df %>% glimpse()
# For each participant and each disease metric, calculate slopes and quality of fit
# For each participant, fit a linear model and extract the slope and R²

# code_upd2hy_hoehn_and_yahr_stage
# mds_updrs_part_iii_summary_score
# mds_updrs_part_iv_summary_score
# upsit_total_score
# rbd_info_source

slopes_df <- full_meta_df %>%
  filter(!is.na(mds_updrs_part_iii_summary_score), !is.na(visit_month)) %>%
  group_by(participant_id) %>%
  do({
    fit = lm(mds_updrs_part_iii_summary_score ~ visit_month, data = .)
    tidy_fit = broom::tidy(fit)
    glance_fit = broom::glance(fit)
    data.frame(
      slope = tidy_fit$estimate[tidy_fit$term == "visit_month"],
      intercept = tidy_fit$estimate[tidy_fit$term == "(Intercept)"],
      r.squared = glance_fit$r.squared,
      n = nrow(.)
    )
  }) %>%
  ungroup()

# categorize each metric sample as reliably increasing, decreasing, highly variable, or not stagnant
metric_type <- slopes_df %>% 
    mutate(metric_type = case_when(
        slope > 0.2 & r.squared > 0.3 ~ "increasing",
        slope < -0.2 & r.squared > 0.3 ~ "decreasing",
        r.squared < 0.3 ~ "highly variable",
        TRUE ~ "stagnant"
    )) %>% 
    mutate(metric_type = factor(metric_type, levels = c("increasing", "decreasing", "highly variable", "stagnant"))) %>% 
    select(participant_id, metric_type) %>% 
    glimpse()

# table(full_meta_df$case_control_other_latest, full_meta_df$upsit_total_score)
# table(full_meta_df$study, full_meta_df$mds_updrs_part_ii_summary_score)

full_meta_df %>% 
    left_join(metric_type, by = "participant_id") %>% 
    filter(!is.na(metric_type)) %>%
    ggplot(aes(x = visit_month, y = mds_updrs_part_iii_summary_score, color = metric_type)) +
    facet_grid(metric_type~case_control_other_latest) +
    scale_color_aaas() +
    geom_line(aes(group = participant_id), alpha = 0.3)
    # geom_smooth(aes(group = participant_id), alpha = 0.1, method = "lm", se = FALSE, linewidth = 0.5)


# For each sample_id calculate the fraction of membership in TCR meta-clusters that are PD enriched / PD depleted
cluster_id_annotation <- chi2_results_df %>% 
    mutate(
        significant = case_when(
            abs(log2(odds_ratio)) >= 0.5 & chi2_p_value <= 0.05 ~ case_when(
                odds_ratio > 1 ~ "Enriched in Cases",
                odds_ratio < 1 ~ "Enriched in Controls"
            ),
            TRUE ~ "Not Significant"
        )
    )  %>% 
    select(index, type, significant) %>% 
    distinct() %>% 
    glimpse()

# cluster_df_meta_f3 <- cluster_df_meta_f1 %>% 
#     mutate(pattern_len = gsub("\\%", "", pattern) %>% nchar()) %>% 
#     filter(pattern_len > 3) %>% 
#     glimpse()

# full_meta_df %>% 
#     left_join(metric_type, by = "participant_id") %>% 
#     filter(!is.na(metric_type)) 

sid_cluster_enrichment <- cluster_df_meta_f3 %>% 
  dplyr::select(index, type, sample_id, participant_id, Freq, case_control_other_latest) %>%
  left_join(cluster_id_annotation, by = c("index", "type")) %>% 
  distinct() %>% 
  group_by(participant_id, significant, case_control_other_latest) %>%
  dplyr::summarise(total_freq = sum(Freq)) %>%
  group_by(participant_id) %>%
  dplyr::mutate(frac_prop = total_freq / sum(total_freq)) %>%
  # left_join(full_meta_df, by = "participant_id") %>% 
  left_join(metric_type, by = "participant_id") %>% 
  # filter(!is.na(metric_type)) %>%
  glimpse()


# Optionally, order participants by fraction of "Enriched in Cases" for better visualization
participant_order <- sid_cluster_enrichment %>%
  filter(significant == "Enriched in Cases") %>%
  arrange(desc(frac_prop)) %>%
  pull(participant_id)

# Set factor levels for consistent color mapping and ordering
sid_cluster_enrichment$significant <- factor(
  sid_cluster_enrichment$significant,
  levels = c("Enriched in Cases", "Enriched in Controls", "Not Significant")
)
sid_cluster_enrichment$participant_id <- factor(
  sid_cluster_enrichment$participant_id,
  levels = participant_order
)


sid_cluster_enrichment %>% glimpse()

# Plot
p_metacluster_enrichment_disease <- ggplot(sid_cluster_enrichment, aes(
  y = participant_id,
  x = frac_prop,
  fill = significant
)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(
    values = c(
      "Enriched in Cases" = "#FF7F0F",
      "Enriched in Controls" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  labs(
    x = "Fraction of repertoire",
    y = "Participant ID",
    fill = "Cluster Enrichment",
    title = "Fraction of GLIPH2 Meta-Cluster Composition by Participant"
  ) +
  theme_bw() +
  facet_wrap(~case_control_other_latest, ncol = 1, scales = "free_y") +
  theme(
    axis.text.y = element_blank(),
    legend.position = "right"
  )

p_metacluster_enrichment_disease_severity_group <- sid_cluster_enrichment %>% 
  filter(case_control_other_latest == "Case") %>% 
  filter(!is.na(metric_type)) %>% 
  ggplot(aes(
    y = participant_id,
    x = frac_prop,
    fill = significant
    )) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(
    values = c(
      "Enriched in Cases" = "#FF7F0F",
      "Enriched in Controls" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  labs(
    x = "Fraction of repertoire",
    y = "Participant ID",
    fill = "Cluster Enrichment",
    title = "Fraction of GLIPH2 Meta-Cluster Composition by Participant"
  ) +
  theme_bw() +
  facet_wrap(~metric_type, ncol = 1, scales = "free_y") +
  theme(
    axis.text.y = element_blank(),
    legend.position = "right"
  )

ggsave(
  glue("{gliph2_plots_dir}/tcr_cluster_enrichment_disease_cond_{Sys.Date()}.png"),
  plot = p_metacluster_enrichment_disease,
  width = 8,
  height = 6
)

ggsave(
  glue("{gliph2_plots_dir}/tcr_cluster_enrichment_disease_severity_group_{Sys.Date()}.png"),
  plot = p_metacluster_enrichment_disease_severity_group,
  width = 8,
  height = 6
)


p_metacluster_enrichment_disease_severity_group_contrank <- sid_cluster_enrichment %>% 
  filter(case_control_other_latest == "Case") %>% 
  filter(!is.na(metric_type)) %>% 
  ggplot(aes(
    y = participant_id,
    x = frac_prop,
    fill = significant
    )) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  scale_fill_manual(
    values = c(
      "Enriched in Cases" = "#FF7F0F",
      "Enriched in Controls" = "#1F77B4",
      "Not Significant" = "grey50"
    )
  ) +
  labs(
    x = "Fraction of repertoire",
    y = "Participant ID",
    fill = "Cluster Enrichment",
    title = "Fraction of GLIPH2 Meta-Cluster Composition by Participant"
  ) +
  theme_bw() +
  facet_wrap(~metric_type, ncol = 1, scales = "free_y") +
  theme(
    axis.text.y = element_blank(),
    legend.position = "right"
  )

```



```{r}
p_fishers_volc <- chi2_results_df %>% 
    # Add a column to identify significant points
    mutate(
        significant = case_when(
            abs(log2(odds_ratio)) >= 0.5 & chi2_p_value <= 0.05 ~ case_when(
                odds_ratio > 1 ~ "Enriched in Cases",
                odds_ratio < 1 ~ "Enriched in Controls"
            ),
            TRUE ~ "Not Significant"
        ),
        # Create custom hover text
        hover_text = sprintf(
            "Type: %s\nP-value: %.2e\nOdds Ratio: %.2f",
            type, chi2_p_value, odds_ratio
        )
    ) %>%
    # Add a column to identify points to label
    mutate(
        label_text = case_when(
            # Top 8 most significant enriched in cases
            significant == "Enriched in Cases" & 
                chi2_p_value <= (arrange(filter(., significant == "Enriched in Cases"), chi2_p_value) %>% 
                                 slice_head(n = 15) %>% 
                                 pull(chi2_p_value) %>% 
                                 max()) ~ type,
            # Top 8 most significant enriched in controls
            significant == "Enriched in Controls" & 
                chi2_p_value <= (arrange(filter(., significant == "Enriched in Controls"), chi2_p_value) %>% 
                                 slice_head(n = 8) %>% 
                                 pull(chi2_p_value) %>% 
                                 max()) ~ type,
            TRUE ~ NA_character_
        )
    ) %>%
    ggplot(aes(x = log2(odds_ratio), 
               y = -log10(chi2_p_value),
               color = significant,
               text = hover_text)) +
    geom_point(alpha = 0.8, size = 1) +
    # Add ggrepel text labels
    geom_label_repel(
        aes(label = label_text),
        size = 3,
        max.overlaps = Inf,
        box.padding = 0.5,
        segment.color = "grey50",
        show.legend = FALSE,
        segment.curvature = 0.3, # Controls how curved the lines are (0 = straight)
        segment.ncp = 3, # Number of control points (higher = smoother curves)
        segment.angle = 10 # Angle of the curve
    ) +
    scale_color_manual(
        values = c(
            "Enriched in Cases" = "#FF7F0F",
            "Enriched in Controls" = "#1F77B4",
            "Not Significant" = "grey50"
        )
    ) +
    theme_bw() +
    theme(
        legend.position = "right",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 12)
    ) +
    geom_hline(yintercept = -log10(0.05), 
               linetype = "dashed", 
               color = "darkgrey") +
    geom_vline(xintercept = c(-0.5, 0.5), 
               linetype = "dashed", 
               color = "darkgrey") +
    labs(
        x = expression(log[2]~"(Odds Ratio)"),
        y = expression(-log[10]~"(p-value)"),
        title = "TCR Cluster Enrichment Analysis"
    ) +
    coord_cartesian(clip = "off")

# p_fishers_volc
ggsave(
    glue("{gliph2_plots_dir}/tcr_cluster_volcano_{Sys.Date()}.png"),
    plot = p_fishers_volc,
    width = 8,
    height = 6
)


# # Convert to plotly with custom hover template
# p_fishers_volc_plotly <- ggplotly(p_fishers_volc, tooltip = "text") %>%
#     layout(hoverlabel = list(bgcolor = "white"))

# # Save the plotly object as an interactive HTML
# htmlwidgets::saveWidget(
#     p_fishers_volc_plotly, 
#     file = glue("{gliph2_plots_dir}/tcr_cluster_volcano_interactive.html"),
#     selfcontained = TRUE  # This ensures all dependencies are included in the file
# )


```



```{r}
sample_groups_df <- cluster_df_meta_f3 %>%
    dplyr::select(participant_id, case_control_other_latest) %>%
    distinct()
grp_n <- sample_groups_df$case_control_other_latest %>% table()
case_n_base <- grp_n[["Case"]]
ctrl_n_base <- grp_n[["Control"]]
cidx_list <- unique(cluster_df_meta_f3$index)


# 1.  unique donors per (cluster × group) in one pass 
per_cluster <- cluster_df_meta_f3 %>%
  dplyr::distinct(index, type, participant_id, case_control_other_latest) %>%
  dplyr::count(index, type, case_control_other_latest, name = "n") %>%
  pivot_wider(names_from = case_control_other_latest,
              values_from = n,
              values_fill = 0)                                          # columns: Case, Control

# 2.  add the 'remaining' cells once 
per_cluster <- per_cluster %>%
  mutate(
    remCase = case_n_base - Case,
    remCtrl = ctrl_n_base - Control
  ) %>% 
  glimpse()

# 3.  run Fisher’s test row-wise (vectorised over rows) 
future::plan(future::multisession, workers = 7)
fisher_stats <- furrr::future_pmap_dfr(
  # supply the four cell counts as parallel vectors
  per_cluster %>% dplyr::select(Case, Control, remCase, remCtrl),
  function(Case, Control, remCase, remCtrl) {
    ft <- fisher.test(matrix(c(Case+1, Control+1, remCase+1, remCtrl+1), nrow = 2))
    list(
      fisher_p_value = ft$p.value,
      odds_ratio     = unname(ft$estimate),
      conf_low       = ft$conf.int[1],
      conf_high      = ft$conf.int[2]
    )
  }
)

# 4.  final result 
fisher_results_df <- bind_cols(per_cluster, fisher_stats) %>% 
    glimpse()

# fisher_results_df$adj_fisher_p_value %>% sort() %>% head
fisher_results_df %>% 
    slice_head(n = 10, by = fisher_p_value) %>% 
    View()

# per_cluster: data-frame with Case, Control, remCase, remCtrl columns
chi2_stats <- furrr::future_pmap_dfr(
  per_cluster %>% select(Case, Control, remCase, remCtrl),
  function(Case, Control, remCase, remCtrl) {
    mat <- matrix(c(Case, Control, remCase, remCtrl), nrow = 2)
    ## 1. χ² test (no Yates correction keeps it comparable to Fisher)
    ct <- suppressWarnings(chisq.test(mat, correct = FALSE))
    ## 2. Odds ratio for context
    or <- (Case * remCtrl) / (Control * remCase)
    list(
      chi2_p_value = ct$p.value,
      chi2_stat    = unname(ct$statistic),
      odds_ratio   = or,
      exp_case_cluster    = ct$expected[1],
      exp_control_cluster = ct$expected[2]
    )
  }
)

chi2_results_df <- bind_cols(per_cluster, fisher_stats, chi2_stats) %>% 
    mutate(adj_chi2_p_value = stats::p.adjust(chi2_p_value, method = "fdr")) %>%
    arrange(chi2_p_value) %>% 
    glimpse()

chi2_results_df %>% 
    slice_head(n = 10, by = chi2_p_value) %>% 
    View()

```




## Building scaled ML models for GLIPH2 cluster based prediction of clinical outcomes

First, let's try a bunch of Chi-squared models for categorical outcomes.
```{r}
# cluster_df <- cluster_df_meta_f3
# meta_df <- full_meta_df
# meta_col <- "DRB1_risk"
# seed <- 1

cluster_meta_association <- function(cluster_df, meta_df, meta_col, seed) {

    # Filtering out sample_id with missing values from random sample selection process
    new_meta <- meta_df %>%
        select(sample_id, all_of(meta_col)) %>% 
        filter(!is.na(!!sym(meta_col))) %>% 
        filter(!!sym(meta_col) != "") %>% 
        mutate(!!sym(meta_col) := stringr::str_replace_all(!!sym(meta_col), "[^A-Za-z0-9]", "_")) %>%
        distinct() %>% 
        glimpse()
    
    if (nrow(new_meta) == 0) return(NULL)
    
    new_meta_order <- new_meta[[meta_col]] %>%
        table() %>% 
        sort(decreasing = TRUE) %>% 
        names()

    set.seed(seed)
    rand_sids <- cluster_df  %>% 
        inner_join(new_meta, by = "sample_id") %>% 
        select(sample_id, participant_id) %>% 
        distinct() %>% 
        group_by(participant_id) %>%
        slice_sample(n = 1, ) %>% 
        pull(sample_id)

  # Join metadata
  df <- cluster_df %>%
    filter(sample_id %in% rand_sids) %>% 
    select(type, sample_id) %>%
    distinct() %>%
    left_join(new_meta, by = "sample_id")
  
  # Check type of metadata column
  meta_vec <- df[[meta_col]]
  is_cat <- is.factor(meta_vec) || is.character(meta_vec)
  
  if (is_cat) {
    # Categorical: For each cluster, run chi-squared test
    per_cluster <- df %>%
      group_by(type, !!sym(meta_col)) %>%
      summarise(n = n_distinct(sample_id), .groups = "drop") %>%
      mutate(!!sym(meta_col) := factor(!!sym(meta_col), levels = new_meta_order)) %>%
      arrange(!!sym(meta_col)) %>% 
      pivot_wider(names_from = !!sym(meta_col), values_from = n, values_fill = 0)
    
    # Get group sizes for "remaining" counts
    group_sizes <- new_meta %>%
      filter(sample_id %in% rand_sids) %>% 
      filter(!is.na(!!sym(meta_col))) %>%
      mutate(!!sym(meta_col) := factor(!!sym(meta_col), levels = new_meta_order)) %>% 
      group_by(!!sym(meta_col)) %>%
      summarize(total_n = n()) %>% 
      arrange(!!sym(meta_col))
    
    # CHECK THAT THE ORDER OF THE GROUPS ALIGN
    group_sizes_order <- group_sizes %>% pull(!!sym(meta_col))
    actual_n_order <- colnames(per_cluster)[-1]
    order_check <- group_sizes_order == actual_n_order
    if (!all(order_check)) {
      stop("The order of the groups does not align")
    }

    # Add remaining counts for each group
    for (grp in group_sizes[[meta_col]]) {
      per_cluster[[paste0("rem_", grp)]] <- group_sizes$total_n[group_sizes[[meta_col]] == grp] - per_cluster[[as.character(grp)]]
    }

    per_cluster %<>% column_to_rownames(var = "type")
    
    # # FOR TESTING
    # vals_df <- per_cluster
    # vals <- as.list(vals_df[1,])

    # Run chi-squared for each cluster
    chi2_stats <- purrr::pmap_dfr(
      per_cluster,
      function(...) {
        vals <- list(...)
        n_groups <- length(vals) / 2
        mat <- rbind(
          as.numeric(vals[1:n_groups]), # categories in cluster
          as.numeric(vals[(n_groups+1):(2*n_groups)]) # categories in remaining
        )
        ct <- suppressWarnings(chisq.test(mat, correct = FALSE))
        tibble(
          chi2_p_value = ct$p.value,
          chi2_stat = unname(ct$statistic)
        )
      }
    )
    result <- bind_cols(per_cluster, chi2_stats)
    result <- result %>%
      mutate(adj_chi2_p_value = p.adjust(chi2_p_value, method = "fdr")) %>%
      arrange(chi2_p_value)
    message(glue("COMPLETED: {meta_col} has {nrow(result)} rows"))
    return(result)
  }

  stop("meta_col must be either categorical (factor/character)")
}



full_meta_df %>% glimpse()
# Select only categorical metadata
categorical_meta <- full_meta_df %>% 
  mutate_if(is.factor, as.character) %>% 
  select_if(is.character) %>% 
  colnames()

# filtering out metadata that has more than 20 unique values
testable_categorical_meta <- categorical_meta %>% 
  purrr::map_df(function(x) {
    tibble(
      meta_col = x,
      n_unique = length(unique(full_meta_df[[x]]))
    )
  }) %>% 
  arrange(desc(n_unique)) %>% 
  filter(n_unique <= 20 & n_unique > 1) %>% 
  pull(meta_col)

#' Running with different seeds to shuffle random selection of samples per participant


chi2_param_grid <- expand_grid(
  # meta_col = "DRB1_risk", # for testing
  meta_col = testable_categorical_meta,
  seed = 1:10) %>% 
  glimpse()

# #Single example run
# result_cat <- cluster_meta_association(
#   cluster_df = cluster_df_meta_f3,
#   meta_df = full_meta_df,
#   meta_col = "upd129_gait"
# )


# with 15 workers this will take about 20 minutes
tic()
# future::plan(future::sequential)
future::plan(future::multisession, workers = 15)
1:nrow(chi2_param_grid) %>% 
  # discard(~ file.exists(glue(
  #   "{gliph2_res_dir}/chi2_association_screening/",
  #   "{chi2_param_grid$meta_col[.]}_ITER-{chi2_param_grid$seed[.]}_2025-05-23.rds"))) %>%   
  furrr::future_walk(function(x) {
    gliph2_res_dir <- "/resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/airr/gliph2/processing"   
    result_cat <- cluster_meta_association(
      cluster_df = cluster_df_meta_f3,
      meta_df = full_meta_df,
      meta_col = chi2_param_grid$meta_col[x],
      seed = chi2_param_grid$seed[x]
    )
    saveRDS(result_cat, file = glue(
      "{gliph2_res_dir}/chi2_association_screening/",
      "{chi2_param_grid$meta_col[x]}_ITER-{chi2_param_grid$seed[x]}_{Sys.Date()}.rds"))
  })
toc()

```



```{r}
chi2_screen_res_dir <- glue("{gliph2_res_dir}/chi2_association_screening")
chi2_screen_res_paths <- list.files(chi2_screen_res_dir, pattern = "ITER", full.names = TRUE)

chi2_res_df <- chi2_screen_res_paths %>% 
  purrr::set_names(basename(str_before_last(., "_"))) %>% 
  purrr::map(function(x) {
    df <- readRDS(x)
    if (is.null(df)) return(NULL)
    df %>% 
      tibble::rownames_to_column(var = "type") %>% 
      dplyr::select(type, chi2_p_value, chi2_stat, adj_chi2_p_value)
  }) %>%
  bind_rows(.id = "meta_iter") %>% 
  glimpse()

chi2_res_df %>% glimpse()

# result_cat %>% filter(adj_chi2_p_value <= 0.05) %>% glimpse
meta_cats_df <- readRDS(glue("{wkdir}/",
  "data/interim/metadata/2023-06-06_metadata_categories_dataframe.rds")) %>% 
  glimpse()
```


```{r}
chi2_res_df %<>% 
  mutate(meta = str_before_first(meta_iter, "_ITER-1")) %>% 
  as.data.table() %>% 
  glimpse()

chi2_summary_stats <- chi2_res_df %>%
  group_by(meta) %>%
  summarise(
    across(
      c(chi2_p_value, chi2_stat, adj_chi2_p_value),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  arrange(adj_chi2_p_value_mean) %>%
  glimpse()


# chi2_summary_stats %>% View
chi2_summary_stats %>% glimpse()
p <- chi2_summary_stats %>% 
  ggplot(aes(x=-log10(adj_chi2_p_value_mean), y=chi2_stat_mean)) +
  geom_point() +
  theme_bw() +
  labs(
    x = expression(-log[10]~"(p-value)"),
    y = "Chi-squared Statistic"
  )

ggsave(
  glue("{gliph2_plots_dir}/chi2_summary_stats_{Sys.Date()}.png"),
  plot = p,
  width = 8,
  height = 6
)


chi2_summary_stats %>% filter(adj_chi2_p_value_mean <= 0.05) %>% View
chi2_summary_stats %>% View

chi2_summary_stats$chi2_p_value_mean %>% hist()

full_meta_df$genetic_status_enrollment %>% table()
full_meta_df$genetic_status_enrollment %>% unique()
full_meta_df$upd2106_dopamine_dysregulation_syndrome_features %>% table()
full_meta_df$upd2317c_rest_tremor_amplitude_right_lower_extremity %>% table()
full_meta_df$DRB1_grouped %>% table()
full_meta_df$DRB1_risk %>% table()



chi2_res_df_DRB1grpd <- chi2_screen_res_paths %>% 
  keep(grepl("DRB1_risk", .)) %>% 
  purrr::set_names(basename(str_before_last(., "_"))) %>% 
  purrr::map(function(x) {
    df <- readRDS(x)
    if (is.null(df)) return(NULL)
    df %>% tibble::rownames_to_column(var = "type")
  }) %>%
  bind_rows(.id = "meta_iter") %>% 
  glimpse()

View(chi2_res_df_DRB1grpd)



```


## TRYING TO SETUP DATA FOR ML MODELS

```{r}
chi2_results_df <- readRDS(
  glue("{gliph2_res_dir}/chi2_results_df_2025-05-26.rds"))
cluster_id_annotation <- chi2_results_df %>% 
    mutate(
        significant = case_when(
            abs(log2(odds_ratio)) >= 0.5 & chi2_p_value <= 0.05 ~ case_when(
                odds_ratio > 1 ~ "Enriched in Cases",
                odds_ratio < 1 ~ "Enriched in Controls"
            ),
            TRUE ~ "Not Significant"
        )
    )  %>% 
    select(index, type, significant) %>% 
    distinct() %>% 
    glimpse()

useful_cluster_ids <- cluster_id_annotation %>% 
  filter(significant != "Not Significant") %>% 
  pull(type)

cluster_pid_comp <- cluster_df_meta %>% 
    dplyr::select(index, participant_id) %>% 
    distinct() %>% 
    group_by(index) %>% 
    dplyr::summarise(n_participants = n()) %>% 
    ungroup()

unique_pid_thresh_inx <- cluster_pid_comp %>% 
    filter(n_participants >= 10) %>% 
    pull(index)

cluster_df_meta_f1 <- cluster_df_meta %>% 
    filter(index %in% unique_pid_thresh_inx) %>% 
    glimpse()
# cluster_df_meta_f1 %>% glimpse()

# Filter 3 clusters with pattern length > 3 fixed pos AAs
cluster_df_meta_f3 <- cluster_df_meta_f1 %>% 
    mutate(pattern_len = gsub("\\%", "", pattern) %>% nchar()) %>% 
    filter(pattern_len > 3) %>% 
    glimpse()
```


```{r}
library(tidyverse)
library(tidymodels)

cluster_df_meta_f3 %>% glimpse()
full_meta_df %>% glimpse()

# random sample_id selection per participant_id
set.seed(42)
rand_sids <- cluster_df_meta_f3  %>% 
    select(sample_id, participant_id) %>% 
    distinct() %>% 
    group_by(participant_id) %>%
    slice_sample(n = 1) %>% 
    pull(sample_id)

# Aggregate features per sample_id (example: sum Freq by sample_id and type)
features_df <- cluster_df_meta_f3 %>%
    filter(sample_id %in% rand_sids) %>% 
      # FILTERING WITH CHI2 results (this time not using 2nd filter round)
    filter(type %in% useful_cluster_ids) %>%  
    dplyr::group_by(sample_id, type) %>% 
    dplyr::summarise(count = sum(Freq, na.rm = TRUE), .groups = "drop") %>% 
    # sample_n(1000) %>% 
    pivot_wider(names_from = type, values_from = count, values_fill = 0)

feat_cols <- colnames(features_df)[-1]
length(feat_cols)
# Find columns with >0 in at least 5 samples
cols_to_keep <- feat_cols[
  colSums(features_df[feat_cols] > 0, na.rm = TRUE) >= 5
]
length(cols_to_keep)

filtered_features_df <- features_df %>%
  select(sample_id, all_of(cols_to_keep))
filtered_features_df %>% dim()

# Join with case/control status
meta_status <- full_meta_df %>%
  filter(case_control_other_latest %in% c("Case", "Control")) %>% 
  select(sample_id, predict_var = mds_updrs_part_iii_summary_score) %>% 
  drop_na(predict_var) %>% 
  glimpse()

model_df <- filtered_features_df %>%
  inner_join(meta_status, by = "sample_id")

model_df %>% dim()
```


Spliting the data into training and testing sets.

```{r}
set.seed(123)
data_split <- initial_split(model_df, strata = predict_var)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Building a recipe
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(sample_id, new_role = "id")
#   step_zv(all_predictors()) %>%
#   step_normalize(all_predictors())

# Building a model specification
lasso_spec <- logistic_reg(
  penalty = tune(),   # LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

lasso_spec <- logistic_reg(
  penalty = tune(),   # LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

# Building a workflow
wf <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(rec)

# Tuning the grid of lambda values
lambda_grid <- grid_regular(penalty(), levels = 10)  # 10 values from min to max
set.seed(123)
folds <- vfold_cv(train_data, v = 5, strata = predict_var)
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = lambda_grid,
  metrics = metric_set(roc_auc, accuracy)
)

# Selecting the best models and evaluating performance
best_lambda <- select_best(tune_res, "roc_auc")
final_wf <- finalize_workflow(wf, best_lambda)
final_fit <- fit(final_wf, data = train_data)
# Predict on test set
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# Evaluate
roc_auc(test_pred, truth = predict_var, .pred_Case)
accuracy(test_pred, truth = predict_var, 
.pred_class = ifelse(.pred_Case > 0.5, "Case", "Control"))


```


Trying with Random Forests

```{r}
# future::plan(future::multisession, workers = 7)

# Building a recipe (keep this the same)
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(sample_id, new_role = "id")

# Random Forest Model - no tuning
rf_spec <- rand_forest(
  trees = 500,  # reasonable default
  min_n = 5     # reasonable default
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Building a workflow
wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rec)

# Fit the model directly (no tuning)
set.seed(123)
final_fit <- fit(wf, data = train_data)

# Predict on test set
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# Evaluate
roc_auc(test_pred, truth = predict_var, .pred_Case)
accuracy(test_pred, truth = predict_var, 
         .pred_class = ifelse(.pred_Case > 0.5, "Case", "Control"))

# Get feature importance
rf_fit <- extract_fit_parsnip(final_fit)
var_imp <- vip::vip(rf_fit, num_features = 20)
print(var_imp)
```



## GIANA Clustering 

Setting up working directory

```{r}
giana_src <- "/resnick/groups/MazmanianLab/jboktor/git/GIANA"
giana_wkdir <- glue("{wkdir}/data/interim/giana")
giana_input_dir <- glue("{wkdir}/data/interim/giana/input")
giana_output_dir <- glue("{wkdir}/data/interim/giana/results")
dir.create(giana_input_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(giana_output_dir, recursive = TRUE, showWarnings = FALSE)

```

Preparing data for GIANA 


Format will only requires a txt or tsv file input where the first column is the CDR3b amino acid sequence - we will also add a sample_id column and count column. 

```{r}
trust4_reports_tcrb %>% glimpse()
giana_input_df <- trust4_reports_tcrb %>% 
  dplyr::select(CDR3aa, v_gene, 
  #count = `#count`, 
  sample_id) %>% 
  glimpse()

head(giana_input_df)
# write to file
write_tsv(giana_input_df, file = glue("{giana_input_dir}/full_tcrb_{Sys.Date()}.tsv"))


```

Executing GIANA.


Below is a python command to run in the terminal - make sure you are in the giana mamba environment with the necessary packages installed for execution.

```{r}
# execute using slurm script
shell_do(
  glue("sbatch {wkdir}/notebooks/shell_scripts/slurm_run_GIANA.sh")
)

# python /resnick/groups/MazmanianLab/jboktor/git/GIANA/GIANA.4.1.py \
#                 -f /resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/giana/input/full_tcrb_2025-07-10.tsv \
#                 -S 3         \     # Smith-Waterman score cutoff (↑ stricter) default value is 3.6 - lowered to allow for a variety of disease relavent TCR clusters
#                 -t 10          \   # isometric distance cutoff (smaller → faster) default value is 10
#                 -N 32          \   # threads for FAISS step
#                 -v                 # exclude v_gene from clustering criteria
#                 -o /resnick/groups/MazmanianLab/jboktor/PDMBS/pdairr/data/interim/giana/results/



```

## GIANA Results Analysis

```{r}

```





```{r}
sessionInfo()
```

