---
title: "cdr3-QTL Single-Cell Localization"
editor: source
author: "Joe Boktor"
date: '2026-04-12'
eval: false
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
    fig-height: 8
    fig-width: 14
---

This notebook localizes CDR3 risk scores (Ishigaki et al., 2022; [GitHub](https://github.com/immunogenomics/cdr3-QTL)) to specific T cell subtypes using external single-cell RNA-seq + TCR-seq datasets from Parkinson's disease studies:

- **Wang et al. 2021** — Blood + CSF T cells (10x 5' GEX + V(D)J); 121K T cells, 21 clusters; [Cell Discovery](https://doi.org/10.1038/s41421-021-00280-3)
- **Moquin-Beaudry et al. 2025** — PBMC scRNA-seq + TCR-seq; 78,876 cells, 38 immune subtypes; 14 PD + 6 atypical + 10 HC; [Brain](https://doi.org/10.1093/brain/awaf066)
- **Ma et al. 2025** — Post-mortem brain (SN + cortex) snRNA-seq + TCR-seq; 207K nuclei; [Nature Communications](https://doi.org/10.1038/s41467-025-62478-3)


## Analysis Setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(glue)
library(furrr)
library(tictoc)
library(progress)
library(entropy)
library(data.table)
library(seqinr)
library(nlme)
library(broom.mixed)
library(Seurat)
library(readxl)
# plotting
library(ggsci)
library(scico)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(ggridges)
library(dunn.test)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/compute_CDR3-risk-score.R"))
scratch_dir <- "/resnick/scratch/jbok/PDMBS"
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
entropy_dir <- glue("{wkdir}/data/interim/cdr3_entropy")
entropy_plots_dir <- glue("{wkdir}/figures/AIRR/cdr3_entropy")
git_dir <- glue("{home_dir}/git/cdr3-QTL")
hla_plots_dir <- glue("{wkdir}/figures/AIRR/arcasHLA")
hla_dir <- glue("{wkdir}/data/interim/airr/arcasHLA/analysis_cleaned")

c3_pheno_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_phenotypes")
gdata_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_hla_genotypes")
hla_4d_to_aa_dir <- glue("{wkdir}/data/input/cdr3qtl/hla_4d_to_aa")
tcr_cdr3_riskscore_dir <- glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores")
tcr_cdr3_riskscore_complied <-
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")

# External single-cell data directories
ext_sc_dir <- glue("{wkdir}/data/input/external_singlecell")
wang_dir <- glue("{ext_sc_dir}/wang_2021")
moquin_dir <- glue("{ext_sc_dir}/moquin_beaudry_2025")
ma_dir <- glue("{ext_sc_dir}/ma_2025")
sc_results_dir <- glue("{wkdir}/data/interim/cdr3_qtl/singlecell_localization")
sc_figures_dir <- glue("{wkdir}/figures/CDR3-qtl/singlecell_localization")

dir.create(wang_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(moquin_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(ma_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(sc_results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(sc_figures_dir, showWarnings = FALSE, recursive = TRUE)

future::plan(multisession, workers = 8)
```


# Study 1: Wang et al. 2021 — Blood & CSF T Cells

## Download Wang et al. 2021 Data

Data deposited on [Zenodo record 3993994](https://zenodo.org/record/3993994):

- `counts.rds` (382MB) — raw count matrix
- `normalized.rds` (417MB) — normalized expression matrix
- `metadata.txt` (16MB) — cell annotations with T cell cluster labels + TCR info
- `README.txt`

CSF TCR data originally from Gate et al. (GEO: [GSE134578](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE134578)).

```{r download-wang}
zenodo_base <- "https://zenodo.org/record/3993994/files"
wang_files <- c("counts.rds", "normalized.rds", "metadata.txt", "README.txt")

for (f in wang_files) {
    dest <- glue("{wang_dir}/{f}")
    if (!file.exists(dest)) {
        message(glue("Downloading {f} via curl..."))
        exit_code <- system2("curl",
            args = c("-L", "-o", shQuote(dest),
                     "--retry", "3",
                     "--retry-delay", "5",
                     "-C", "-",
                     shQuote(glue("{zenodo_base}/{f}"))),
            stdout = "", stderr = ""
        )
        if (exit_code != 0) {
            warning(glue("curl failed for {f} (exit code {exit_code})"))
            if (file.exists(dest)) file.remove(dest)
        }
    } else {
        message(glue("{f} already exists, skipping."))
    }
}

# Verify downloads
wang_downloaded <- file.info(list.files(wang_dir, full.names = TRUE))
wang_downloaded$file <- basename(rownames(wang_downloaded))
wang_downloaded %>% dplyr::select(file, size) %>% print()

```


## Load & Inspect Wang et al. 2021 Metadata

```{r load-wang-metadata}
wang_meta <- read.delim(glue("{wang_dir}/metadata.txt"), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    janitor::clean_names()

message(glue("Wang et al. 2021: {nrow(wang_meta)} cells"))
wang_meta %>% glimpse()

# Inspect available columns — looking for CDR3, cell type, donor, condition
wang_meta %>% colnames() %>% sort() %>% print()
```


## Extract CDR3-Beta Sequences with Cell Type Annotations

```{r wang-extract-cdr3b}
# Inspect TCR-related columns
wang_meta %>%
    dplyr::select(matches("cdr3|CDR3|trb|TRB|tcr|TCR|chain|clono", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

# Inspect cell type / cluster columns
wang_meta %>%
    dplyr::select(matches("cluster|cell.?type|subtype|annotation|ident", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

# Inspect condition/donor columns
wang_meta %>%
    dplyr::select(matches("condition|disease|status|group|donor|patient|sample|subject", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

```


```{r wang-clean-cdr3b}

wang_cdr3b <- wang_meta %>%
    mutate(condition = strex::str_before_first(sample_type, "\\(")) %>% 
    mutate(condition = gsub("MCI", "PD", condition)) %>% 
    mutate(condition = gsub("Parkinson", "PD", condition)) %>% 
    mutate(tissue = case_when(
        grepl("CSF", sample_type) ~ "CSF",
        grepl("Blood", sample_type) ~ "Blood",
        TRUE ~ "ERROR"
    )) %>% 
    rename(
        cdr3b_aa = beta_cdr3_aa,
        t_cell_subtype = cell_type2,
        donor_id = sample
    ) %>%
    filter(!is.na(cdr3b_aa) & cdr3b_aa != "") %>%
    filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
    mutate(
        study = "Wang_2021",
        compartment = tissue
    ) %>%
    glimpse()

# CDR3 length distribution
wang_cdr3b %>%
    mutate(cdr3_len = nchar(cdr3b_aa)) %>%
    dplyr::count(cdr3_len) %>%
    print()

# T cell subtype breakdown
wang_cdr3b %>%
    dplyr::count(t_cell_subtype, sort = TRUE) %>%
    print()

# Condition breakdown
wang_cdr3b %>%
    dplyr::count(condition) %>%
    print()

```



# Study 2: Moquin-Beaudry et al. 2025 — PBMC Immune Atlas

scRNA-seq + TCR-seq of sorted PBMCs from 14 PD + 6 atypical Parkinsonism + 10 HC donors. 78,876 cells across 38 immune subtypes. Shared TCR clones identified across PD patients. Published in [Brain](https://doi.org/10.1093/brain/awaf066); preprint: [bioRxiv 10.1101/2024.07.26.605020](https://doi.org/10.1101/2024.07.26.605020).

## Download Moquin-Beaudry et al. 2025 Data

GEO accession not yet publicly available (PMC embargo until 2026-05-26). Data may need to be requested from the authors (CHU Sainte-Justine / Univ. Montreal; corresponding author: Martine Tétreault). Check supplementary materials on [Brain (Oxford Academic)](https://academic.oup.com/brain/advance-article-abstract/doi/10.1093/brain/awaf066/8127334).

```{r download-moquin}
# TODO: Update with GEO accession or direct download URL when available
# GEO accession: GSE?????? (check Brain article Data Availability section)
#
# Expected data:
#   - scRNA-seq count matrix / Seurat object
#   - TCR-seq: filtered_contig_annotations.csv or clonotype files
#   - Cell type annotations (38 subtypes)
#   - Donor metadata (PD, atypical Parkinsonism, HC)
#
# When GEO accession becomes available:
# geo_acc <- "GSE??????"
# moquin_supp_url <- glue("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE???nnn/{geo_acc}/suppl/")
# system2("curl", args = c("-L", "-o", shQuote(glue("{moquin_dir}/data.tar.gz")),
#     "--retry", "3", "-C", "-", shQuote(glue("{moquin_supp_url}/{geo_acc}_RAW.tar"))))

message("Moquin-Beaudry et al. 2025: GEO accession pending. Check paper for data availability.")
```


## Load & Process Moquin-Beaudry et al. 2025 Data

```{r load-moquin}
# TODO: Uncomment and adjust once data is downloaded

# Load cell metadata with TCR info
# moquin_meta <- read_csv(glue("{moquin_dir}/metadata.csv")) %>%
#     janitor::clean_names() %>%
#     glimpse()

# Inspect columns for CDR3, cell type, donor, condition
# moquin_meta %>% colnames() %>% sort() %>% print()
# moquin_meta %>%
#     dplyr::select(matches("cdr3|CDR3|trb|TRB|tcr|TCR|chain|clono", ignore.case = TRUE)) %>%
#     colnames() %>% print()
```


```{r moquin-extract-cdr3b}
# TODO: Uncomment and adjust column names once data structure is known
# Expected: 38 immune subtypes including T cell subsets with paired TCR

# moquin_cdr3b <- moquin_meta %>%
#     rename(
#         cdr3b_aa = <CDR3_BETA_COLUMN>,
#         t_cell_subtype = <CELLTYPE_COLUMN>,
#         donor_id = <DONOR_COLUMN>,
#         condition = <CONDITION_COLUMN>
#     ) %>%
#     filter(!is.na(cdr3b_aa) & cdr3b_aa != "") %>%
#     filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
#     mutate(
#         study = "MoquinBeaudry_2025",
#         tissue = "Blood",
#         compartment = "peripheral"
#     ) %>%
#     glimpse()

# moquin_cdr3b %>% dplyr::count(t_cell_subtype, sort = TRUE) %>% print()
# moquin_cdr3b %>% dplyr::count(condition) %>% print()
```


# Study 4: Ma et al. 2025 — Brain T Cells (Substantia Nigra)

## Download Ma et al. 2025 Data

Processed data and supplementary materials deposited on Zenodo:

- [Processed data](https://zenodo.org/records/19454950) — `data.zip` (6.7GB): snRNA-seq processed objects
- [Supplementary data](https://zenodo.org/records/19457732) — `SupplementaryData.zip` (105MB): supplementary tables
- [GitHub](https://github.com/dalhoomist/T-cell_and_glial_pathology_in_PD): analysis code

```{r download-ma}
# -- Processed data (6.7 GB) --
ma_data_zip <- glue("{ma_dir}/data.zip")
if (!file.exists(ma_data_zip)) {
    message("Downloading Ma et al. 2025 processed data (6.7 GB)...")
    system2("curl",
        args = c("-L", "-o", shQuote(ma_data_zip),
                 "--retry", "3", "--retry-delay", "5", "-C", "-",
                 shQuote("https://zenodo.org/records/19454950/files/data.zip")),
        stdout = "", stderr = ""
    )
} else {
    message("data.zip already exists, skipping.")
}

# -- Supplementary data (105 MB) --
ma_supp_zip <- glue("{ma_dir}/SupplementaryData.zip")
if (!file.exists(ma_supp_zip)) {
    message("Downloading Ma et al. 2025 supplementary data (105 MB)...")
    system2("curl",
        args = c("-L", "-o", shQuote(ma_supp_zip),
                 "--retry", "3", "--retry-delay", "5", "-C", "-",
                 shQuote("https://zenodo.org/records/19457732/files/SupplementaryData.zip")),
        stdout = "", stderr = ""
    )
} else {
    message("SupplementaryData.zip already exists, skipping.")
}

# Verify downloads
file.info(c(ma_data_zip, ma_supp_zip)) %>%
    mutate(file = basename(rownames(.))) %>%
    dplyr::select(file, size) %>%
    print()
```


## Extract & Inspect Ma et al. 2025 Archives

```{r extract-ma}
# Extract processed data (use system unzip — R's unzip() fails on large/complex zips)
ma_data_dir <- glue("{ma_dir}/data")
if (!dir.exists(ma_data_dir)) {
    message("Extracting data.zip...")
    system2("unzip", args = c("-o", shQuote(ma_data_zip), "-d", shQuote(ma_dir)))
}

# Extract supplementary data
ma_supp_dir <- glue("{ma_dir}/SupplementaryData")
if (!dir.exists(ma_supp_dir)) {
    message("Extracting SupplementaryData.zip...")
    system2("unzip", args = c("-o", shQuote(ma_supp_zip), "-d", shQuote(ma_dir)))
}

# List contents
message("=== Processed data contents ===")
list.files(ma_data_dir, recursive = TRUE, full.names = FALSE) %>% print()

message("\n=== Supplementary data contents ===")
list.files(ma_supp_dir, recursive = TRUE, full.names = FALSE) %>% print()
```


## Load Ma et al. 2025 Single-Cell Data

```{r load-ma-data}
# Identify RDS/h5ad/h5seurat files in the extracted data
ma_rds_files <- list.files(ma_data_dir, pattern = "\\.rds$|\\.RDS$",
                           recursive = TRUE, full.names = TRUE)
ma_h5_files <- list.files(ma_data_dir, pattern = "\\.h5ad$|\\.h5seurat$|\\.h5$",
                          recursive = TRUE, full.names = TRUE)

message("RDS files found:")
ma_rds_files %>% print()
message("H5 files found:")
ma_h5_files %>% print()

# Load the primary object — adjust path based on extracted contents
# should be processed_tcell_obj.rds
ma_obj <- readRDS(ma_rds_files[6])
ma_obj

head(ma_obj@meta.data) %>% glimpse()
colnames(ma_obj@meta.data) %>% sort() %>% print()
```


## Load Patient Metadata (Supplementary Table 3, Sheet A)

```{r ma-patient-metadata}
ma_supp_path <- glue("{ma_dir}/Supplementary Data")

# Patient metadata — Sheet A
ma_patient_meta <- read_excel(
    glue("{ma_supp_path}/41467_2025_62478_MOESM3_ESM.xlsx"),
    sheet = "A") %>%
    janitor::clean_names() %>%
    filter(tcr_seq == "Y") %>% 
    glimpse()

# Key columns: tcr_seq, sn_rna_seq, st, tb_number, sequencing_id,
#   brain_region, condition, neuropath_diagnosis, age, sex, path_braak_nft,
#   path_braak_asyn, path_cerad
ma_patient_meta %>% dplyr::count(condition) %>% print()
ma_patient_meta %>% dplyr::count(brain_region) %>% print()

```


## Extract CDR3-Beta Sequences from Ma et al. 2025

TCR clonotype data from Supplementary Table 5:

- **Sheet A**: Full alpha (TRA) repertoire (155K sequences) — columns: Clonotype, copy, Sample, Region, Condition
- **Sheet B**: Curated paired chains (19K sequences, ~12K TRB) — columns: Clonotype, copy, Sample, Chain, Condition

Clonotype format: `{CDR3_AA}_{V_gene}_{J_gene}` (e.g., `ASSAYRLETQY_hTRBV2*03_hTRBJ2-5*01`)

```{r ma-load-tcr}
# Sheet A is 100% alpha (TRA) — no beta sequences
# Sheet B has paired chains: 6,971 TRA + 12,121 TRB with explicit Chain column
ma_tcr_sheet_b <- read_excel(
    glue("{ma_supp_path}/41467_2025_62478_MOESM5_ESM.xlsx"),
    sheet = "B") %>%
    janitor::clean_names() %>%
    rename(clonotype_id = 1)

message(glue("Sheet B: {nrow(ma_tcr_sheet_b)} clonotypes"))
ma_tcr_sheet_b %>% dplyr::count(chain) %>% print()
```


```{r ma-extract-cdr3b}
# Parse CDR3 AA, V gene, J gene from Clonotype column
# Format: {CDR3_AA}_{V_gene}_{J_gene}
parse_clonotype <- function(df) {
    df %>%
        mutate(
            cdr3_aa = str_extract(clonotype, "^[^_]+"),
            v_gene = str_extract(clonotype, "(?<=_)[^_]+(?=_)"),
            j_gene = str_extract(clonotype, "[^_]+$")
        )
}

# Extract TRB from Sheet B and parse clonotype fields
ma_trb <- ma_tcr_sheet_b %>%
    filter(chain == "hTRB") %>%
    parse_clonotype() %>%
    rename(donor_id = sample)

message(glue("TRB clonotypes: {nrow(ma_trb)}"))

# Join brain region from patient metadata (Sheet B lacks Region column)
ma_trb <- ma_trb %>%
    left_join(
        ma_patient_meta %>%
            dplyr::select(tb_number, brain_region) %>% distinct(),
        by = c("donor_id" = "tb_number")
    ) %>%
    rename(tissue = brain_region)

message(glue("Tissue NAs after join: {sum(is.na(ma_trb$tissue))}"))

# Deduplicate by donor + CDR3 sequence
ma_trb_all <- ma_trb %>%
    distinct(donor_id, cdr3_aa, .keep_all = TRUE)

message(glue("Unique TRB clonotypes: {nrow(ma_trb_all)}"))
ma_trb_all %>% dplyr::count(condition) %>% print()
ma_trb_all %>% dplyr::count(tissue) %>% print()
```


```{r ma-clean-cdr3b}
# Filter to valid CDR3-beta lengths for IMGT alignment (12-18 AA)
ma_cdr3b <- ma_trb_all %>%
    rename(cdr3b_aa = cdr3_aa) %>%
    filter(!is.na(cdr3b_aa) & cdr3b_aa != "") %>%
    filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
    mutate(
        study = "Ma_2025",
        compartment = "CNS",
        # No single-cell subtype annotations — bulk TCR from snRNA-seq
        t_cell_subtype = "T_cell_brain"
    ) %>%
    glimpse()

# CDR3 length distribution
ma_cdr3b %>%
    mutate(cdr3_len = nchar(cdr3b_aa)) %>%
    dplyr::count(cdr3_len) %>%
    print()

# Condition breakdown
ma_cdr3b %>% dplyr::count(condition) %>% print()
ma_cdr3b %>% dplyr::count(tissue) %>% print()

message(glue("Ma et al. 2025: {nrow(ma_cdr3b)} TRB CDR3 sequences (length 12-18)"))

```


# Harmonize CDR3-Beta Data Across Studies

```{r harmonize-cdr3b}
# Standardize columns across studies:
#   donor_id, condition, tissue, t_cell_subtype, cdr3b_aa, study, compartment

# Select harmonized columns from each study
wang_harmonized <- wang_cdr3b %>%
    dplyr::select(donor_id, condition, tissue, t_cell_subtype, cell,
                  cdr3b_aa, study, compartment)

# Moquin-Beaudry — uncomment when data is available
# moquin_harmonized <- moquin_cdr3b %>%
#     dplyr::select(donor_id, condition, tissue, t_cell_subtype,
#                   cdr3b_aa, study, compartment)

ma_harmonized <- ma_cdr3b %>%
    dplyr::select(donor_id, condition, tissue, t_cell_subtype,
                  cdr3b_aa, study, compartment)

# Combine all studies (add moquin_harmonized when available)
all_cdr3b <- bind_rows(wang_harmonized, ma_harmonized) %>%
    mutate(
        # Standardize condition labels
        condition = case_when(
            grepl("PD|parkinson|case", condition, ignore.case = TRUE) ~ "PD",
            grepl("HC|healthy|control|ctrl", condition, ignore.case = TRUE) ~ "HC",
            TRUE ~ condition
        ),
        cdr3_len = nchar(cdr3b_aa)
    ) %>%
    glimpse()

# Summary
message(glue("Total CDR3-beta entries: {nrow(all_cdr3b)}"))
all_cdr3b %>% dplyr::count(study, condition) %>% print()
all_cdr3b %>% dplyr::count(study, compartment) %>% print()
all_cdr3b %>% dplyr::count(study, t_cell_subtype, sort = TRUE) %>% print()

# CDR3 length distribution by study
p_cdr3_len <- all_cdr3b %>%
    ggplot(aes(x = cdr3_len, fill = study)) +
    geom_bar(position = "dodge") +
    theme_bw() +
    labs(x = "CDR3-beta length (AA)", y = "Count", fill = "Study") +
    scale_fill_npg()
p_cdr3_len

ggsave(
    filename = glue("{sc_figures_dir}/cdr3b_length_distribution.png"),
    plot = p_cdr3_len,
    width = 8, height = 5, dpi = 300
)

# Save harmonized data
saveRDS(all_cdr3b, glue("{sc_results_dir}/harmonized_cdr3b_singlecell.rds"))
```


# Score CDR3-Beta Sequences with CDR3 Risk Score

Applying the CDR3-QTL HLA risk score beta-coefficients (Ishigaki et al. 2022) to single-cell CDR3-beta sequences, using the same scoring function from [notebook 04](AIRR_cdr3-QTL_04-cdr3riskscore.qmd).

## Load Pre-Computed Beta-Coefficients

```{r load-betacoefs}
resdir <- glue("{wkdir}/data/interim/cdr3_qtl/results_cdr3-risk-score")

all_imgt_pos <- c(
    "P104","P105","P106","P107","P108","P109","P110","P111",
    "P111.1","P112.2","P112.1",
    "P112","P113","P114","P115","P116","P117","P118")

cdr3_betacoef <- list.files(resdir, pattern = ".csv", full.names = TRUE, recursive = TRUE) %>%
    purrr::set_names(basename(.)) %>%
    purrr::map(read_csv) %>%
    bind_rows(.id = "file_name") %>%
    mutate(
        descriptor = basename(file_name) %>% gsub("lm_model_results__", "", .),
        pheno = str_before_first(descriptor, "__"),
        cdr3_pos = descriptor %>% str_after_first("CDR3pos-") %>% str_before_first("__"),
        cdr3_len = descriptor %>% str_after_first("CDR3len-") %>% str_before_first(".csv")
    ) %>%
    janitor::clean_names() %>%
    filter(!is.na(cdr3_aa))

# Filter to significant associations
cdr3_betacoef_sig <- cdr3_betacoef %>%
    group_by(pheno) %>%
    filter(term == "pd_hla_risk_score") %>%
    mutate(adj_p = p.adjust(p_value, method = "BH")) %>%
    filter(adj_p < 0.05) %>%
    arrange(adj_p)

betacoef_list <- cdr3_betacoef_sig %>% split(.$pheno)

message(glue("Loaded {nrow(cdr3_betacoef_sig)} significant beta-coefficients across {length(betacoef_list)} phenotypes"))
names(betacoef_list) %>% print()
```


## Score Single-Cell CDR3-Beta Sequences

```{r score-cdr3b}
all_cdr3b <- readRDS(glue("{sc_results_dir}/harmonized_cdr3b_singlecell.rds"))

# Get unique CDR3-beta sequences to avoid redundant scoring
unique_cdr3b <- unique(all_cdr3b$cdr3b_aa)
message(glue("Scoring {length(unique_cdr3b)} unique CDR3-beta sequences"))

# Score across all phenotype types
tic("CDR3 risk scoring")
score_results <- list()
for (p in names(betacoef_list)) {
    message(glue("Scoring phenotype: {p}"))
    score_results[[p]] <- unique_cdr3b %>%
        purrr::set_names() %>%
        furrr::future_map_dbl(
            ~ score_tcrs_with_betacoefs(
                x = .,
                phenotype_type = p,
                betacoef_list = betacoef_list
            )
        )
}
toc()

# Convert to data frame — one column per phenotype
score_df <- score_results %>%
    purrr::imap(~ tibble(cdr3b_aa = names(.x), !!.y := .x)) %>%
    purrr::reduce(left_join, by = "cdr3b_aa")

message(glue("Scored {nrow(score_df)} unique CDR3-beta sequences"))
score_df %>% glimpse()
```


## Join Scores to Cell-Level Metadata

```{r join-scores}
scored_cells <- all_cdr3b %>%
    left_join(score_df, by = "cdr3b_aa")

message(glue("Scored entries: {nrow(scored_cells)}"))
scored_cells %>% glimpse()

# Save scored data
saveRDS(scored_cells, glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))
```


# Analyze CDR3 Risk Scores by T Cell Subtype

## PD vs HC Comparison Within Each Subtype

```{r pd-vs-hc-subtype}
scored_cells <- readRDS(glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))
# scored_cells$tissue %>% table()
primary_pheno <- "cdr3_p2_strat_filtered_cv1k"
scored_cells %>% glimpse()

# Statistical testing: PD vs HC within each subtype
subtype_tests <- scored_cells %>%
    # filter(study == "Wang_2021") %>%
    filter(condition != "Alzheimer") %>%
  filter(t_cell_subtype != "Gamma delta T cells")  %>% 
    # group_by(t_cell_subtype) %>%
    group_by(study, t_cell_subtype, tissue) %>%
    filter(n_distinct(condition) == 2) %>%
    summarize(
        n_pd = sum(condition == "PD"),
        n_hc = sum(condition == "HC"),
        median_pd = median(.data[[primary_pheno]][condition == "PD"]),
        median_hc = median(.data[[primary_pheno]][condition == "HC"]),
        wilcox_p = tryCatch(
            wilcox.test(
                .data[[primary_pheno]][condition == "PD"],
                .data[[primary_pheno]][condition == "HC"]
            )$p.value,
            error = function(e) NA_real_
        ),
        .groups = "drop"
    ) %>%
    mutate(
        delta_median = median_pd - median_hc,
        adj_p = p.adjust(wilcox_p, method = "BH")
    ) %>%
    arrange(adj_p)

subtype_tests %>% print(n = 50)
# subtype_tests %>% View()
saveRDS(subtype_tests, glue("{sc_results_dir}/subtype_wilcox_tests.rds"))

```

## CDR3 Risk Score Distributions by Subtype

```{r}
wang_data <- scored_cells %>%
  filter(condition != "Alzheimer", study == "Wang_2021") %>% 
  filter(t_cell_subtype != "Gamma delta T cells") 

# ── One-vs-rest Wilcoxon tests, stratified by tissue ──
subtype_vs_rest <- wang_data %>%
  distinct(tissue, t_cell_subtype) %>%
  pmap_dfr(function(tissue, t_cell_subtype) {
    tissue_data <- wang_data %>% filter(tissue == !!tissue)
    in_grp  <- tissue_data %>% filter(t_cell_subtype == !!t_cell_subtype) %>% pull(.data[[primary_pheno]])
    out_grp <- tissue_data %>% filter(t_cell_subtype != !!t_cell_subtype) %>% pull(.data[[primary_pheno]])
    
    # Guard against subtypes with too few cells in a tissue
    if (length(in_grp) < 5 || length(out_grp) < 5) {
      return(tibble(
        tissue = tissue, t_cell_subtype = t_cell_subtype,
        n = length(in_grp), median_subtype = median(in_grp),
        median_rest = median(out_grp), wilcox_p = NA_real_,
        delta_median = median(in_grp) - median(out_grp)
      ))
    }
    
    wt <- wilcox.test(in_grp, out_grp)
    tibble(
      tissue = tissue,
      t_cell_subtype = t_cell_subtype,
      n = length(in_grp),
      median_subtype = median(in_grp),
      median_rest = median(out_grp),
      wilcox_p = wt$p.value,
      delta_median = median(in_grp) - median(out_grp)
    )
  }) %>%
  # Adjust p-values within each tissue separately
  group_by(tissue) %>%
  mutate(
    p_adj = p.adjust(wilcox_p, method = "BH"),
    direction = case_when(
      is.na(p_adj)                      ~ "NS",
      p_adj < 0.05 & delta_median > 0   ~ "Reduced Protection",
      p_adj < 0.05 & delta_median < 0   ~ "Increased Protection",
      TRUE                              ~ "NS"
    ),
    p_label = case_when(
      is.na(p_adj)  ~ NA_character_,
      p_adj < 0.001 ~ "p < 0.001",
      p_adj < 0.05  ~ paste0("p = ", formatC(p_adj, format = "f", digits = 3)),
      TRUE          ~ NA_character_
    )
  ) %>%
  ungroup()

# ── Merge direction back into plot data ──
plot_data <- wang_data %>%
  left_join(
    subtype_vs_rest %>% select(tissue, t_cell_subtype, direction, p_label, delta_median),
    by = c("tissue", "t_cell_subtype")
  ) %>%
  mutate(t_cell_subtype = reorder(t_cell_subtype, .data[[primary_pheno]], FUN = median))

# ── Annotation df for p-value labels (significant only) ──
sig_labels <- subtype_vs_rest %>%
  filter(!is.na(p_label)) %>%
  mutate(t_cell_subtype = factor(t_cell_subtype, levels = levels(plot_data$t_cell_subtype)))

# ── Kruskal-Wallis per tissue ──
kw_labels <- wang_data %>%
  group_by(tissue) %>%
  summarise(
    kw_stat = kruskal.test(.data[[primary_pheno]] ~ t_cell_subtype)$statistic,
    kw_p    = kruskal.test(.data[[primary_pheno]] ~ t_cell_subtype)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    kw_label = glue("Kruskal-Wallis \u03C7\u00B2 = {round(kw_stat, 1)}, p < 2.2e-16")
  )

# ── Plot ──
p_subtype_enrichment <- ggplot(plot_data, aes(x = .data[[primary_pheno]],
                                               y = t_cell_subtype,
                                               fill = direction)) +
  geom_density_ridges(alpha = 0.7, scale = 1.1, linewidth = 0.4,
                      rel_min_height = 0.01) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey40") +
  # P-values for significant subtypes
  geom_text(
    data = sig_labels,
    aes(x = 1.8, y = t_cell_subtype, label = p_label),
    inherit.aes = FALSE,
    hjust = 0, size = 2.5, fontface = "italic", color = "grey30"
  ) +
  # Kruskal-Wallis annotation per tissue panel
  geom_text(
    data = kw_labels,
    aes(x = 2.8, y = Inf, label = kw_label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1.5, size = 2.5, fontface = "italic", color = "grey30"
  ) +
  facet_grid(. ~ tissue) +
  scale_fill_manual(
    values = c(
      "Reduced Protection"  = "#E8853B",
      "Increased Protection" = "#5B7E9E",
      "NS"                   = "#E8E8E8"
    ),
    name = "vs. all other subtypes"
  ) +
  coord_cartesian(xlim = c(-1.75, 2.5), clip = "off") +
  theme_minimal(base_size = 12) +
  labs(
    x = "CDR3 Risk Score",
    y = NULL,
    title = "CDR3 risk score distributions vary across T cell subtypes",
    subtitle = "Color indicates subtypes significantly enriched or depleted relative to all others within each tissue (BH-adjusted p < 0.05)"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "grey40"),
    plot.margin = margin(5.5, 50, 5.5, 5.5),
    axis.text.y = element_text(size = 9),
    strip.text = element_text(face = "bold", size = 11)
  )

ggsave(
  filename = 
  glue("{sc_figures_dir}/cdr3_riskscore_by_subtype_enrichment_by_tissue.png"),
  plot = p_subtype_enrichment, width = 12, height = 8, dpi = 300
)

```


```{r}
scored_cells <- readRDS(glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))
primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

# ── Data ──
wang_data <- scored_cells %>%
  filter(condition != "Alzheimer", study == "Wang_2021") %>% 
  filter(t_cell_subtype != "Gamma delta T cells") 

# ── Kruskal-Wallis per tissue ──
kw_per_tissue <- wang_data %>%
  group_by(tissue) %>%
  summarise(
    kw_stat = kruskal.test(.data[[primary_pheno]] ~ t_cell_subtype)$statistic,
    kw_p    = kruskal.test(.data[[primary_pheno]] ~ t_cell_subtype)$p.value,
    .groups = "drop"
  ) %>%
  mutate(kw_label = glue("Kruskal-Wallis \u03C7\u00B2 = {round(kw_stat, 1)}, p = {formatC(kw_p, format = 'e', digits = 1)}"))

# ── Per-tissue, per-subtype Wilcoxon: PD vs HC ──
subtype_tests_tissue <- wang_data %>%
  distinct(tissue, t_cell_subtype) %>%
  pmap_dfr(function(tissue, t_cell_subtype) {
    tissue_subtype_data <- wang_data %>%
      filter(tissue == !!tissue, t_cell_subtype == !!t_cell_subtype)
    
    pd_vals <- tissue_subtype_data %>% filter(condition == "PD") %>% pull(.data[[primary_pheno]])
    hc_vals <- tissue_subtype_data %>% filter(condition != "PD") %>% pull(.data[[primary_pheno]])
    
    if (length(pd_vals) < 5 || length(hc_vals) < 5) {
      return(tibble(
        tissue = tissue, t_cell_subtype = t_cell_subtype,
        n_pd = length(pd_vals), n_hc = length(hc_vals),
        median_pd = median(pd_vals), median_hc = median(hc_vals),
        wilcox_p = NA_real_, delta_median = median(pd_vals) - median(hc_vals)
      ))
    }
    
    wt <- wilcox.test(pd_vals, hc_vals)
    tibble(
      tissue = tissue,
      t_cell_subtype = t_cell_subtype,
      n_pd = length(pd_vals),
      n_hc = length(hc_vals),
      median_pd = median(pd_vals),
      median_hc = median(hc_vals),
      wilcox_p = wt$p.value,
      delta_median = median(pd_vals) - median(hc_vals)
    )
  }) %>%
  group_by(tissue) %>%
  mutate(p_adj = p.adjust(wilcox_p, method = "BH")) %>%
  ungroup()

# ── Significance labels ──
sig_labels <- subtype_tests_tissue %>%
  filter(!is.na(p_adj), p_adj < 0.05) %>%
  mutate(
    p_label = case_when(
      p_adj < 0.001 ~ formatC(p_adj, format = "e", digits = 1),
      TRUE ~ formatC(p_adj, format = "f", digits = 3)
    ),
    sig_star = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    ),
    display_label = paste0(sig_star, "  p = ", p_label)
  )

# ── Build plot data with consistent subtype ordering ──
plot_data <- wang_data %>%
  mutate(t_cell_subtype = reorder(t_cell_subtype, .data[[primary_pheno]], FUN = median))

# Match factor levels for annotation layers
sig_labels <- sig_labels %>%
  mutate(t_cell_subtype = factor(t_cell_subtype, levels = levels(plot_data$t_cell_subtype)))

kw_per_tissue_plot <- kw_per_tissue  # already has tissue column for faceting

# ── Plot ──
p_subtype_scores_groups <- ggplot(plot_data, aes(x = .data[[primary_pheno]],
                                                  y = t_cell_subtype,
                                                  fill = condition)) +
  ggridges::geom_density_ridges(alpha = 0.6, scale = 1.2, linewidth = 0.5,
                                rel_min_height = 0.01) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  # P-value labels for significant subtypes
  geom_text(
    data = sig_labels,
    aes(x = 1.85, y = t_cell_subtype, label = display_label),
    inherit.aes = FALSE,
    hjust = 0, size = 2.5, fontface = "italic", color = "grey30"
  ) +
  # Kruskal-Wallis per tissue panel
  geom_text(
    data = kw_per_tissue_plot,
    aes(x = 3, y = Inf, label = kw_label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1.5, size = 2.5, fontface = "italic", color = "grey30"
  ) +
  facet_grid(. ~ tissue) +
  scale_fill_manual(values = c("darkgrey", "darkred")) +
  theme_minimal(base_size = 12) +
  labs(
    x = "CDR3 Risk Score",
    y = NULL,
    fill = "Condition",
    title = "CDR3 risk score distributions shift in PD across T cell subtypes",
    subtitle = "PD vs. HC Wilcoxon tests per subtype within each tissue (BH-adjusted, significant shown)"
  ) +
  coord_cartesian(xlim = c(-2, 2.8), clip = "off") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 9, color = "grey40"),
    plot.margin = margin(5.5, 50, 5.5, 5.5),
    axis.text.y = element_text(size = 9),
    strip.text = element_text(face = "bold", size = 11)
  )

ggsave(
  filename = glue("{sc_figures_dir}/cdr3_riskscore_by_subtype_groups_by_tissue.png"),
  plot = p_subtype_scores_groups,
  width = 12, height = 8, dpi = 300
)

```



## Summary Heatmap: Mean CDR3 Risk Score by Subtype x Condition

```{r summary-heatmap}
heatmap_data <- scored_cells %>%
  filter(t_cell_subtype != "Gamma delta T cells") %>% 
    group_by(study, t_cell_subtype, tissue, condition) %>%
    summarize(
        mean_score = mean(.data[[primary_pheno]]),
        n_cells = n(),
        .groups = "drop"
    )

p_heatmap <- heatmap_data %>%
    filter(study == "Wang_2021") %>%
    filter(condition != "Alzheimer") %>%
    filter(t_cell_subtype != "Gamma delta T cells") %>% 
    mutate(
        t_cell_subtype = factor(
            t_cell_subtype,
            levels = scored_cells %>%
                filter(condition != "Alzheimer", study == "Wang_2021") %>%
                group_by(t_cell_subtype) %>%
                summarize(median_score = median(.data[[primary_pheno]]), .groups = "drop") %>%
                arrange(median_score) %>%
                pull(t_cell_subtype)
        )
    ) %>%
    ggplot(aes(x = condition, y = t_cell_subtype, fill = mean_score)) +
    geom_tile() +
    geom_text(aes(label = n_cells), size = 2.5) +
    facet_wrap( ~ tissue) +
    scale_fill_scico(palette = "vik", midpoint = 0) +
    theme_bw() +
    labs(
        x = "Condition", y = "T Cell Subtype",
        fill = "Mean CDR3\nRisk Score",
        title = "CDR3 Risk Score Heatmap"
    )

ggsave(
    filename = glue("{sc_figures_dir}/cdr3_riskscore_heatmap.png"),
    plot = p_heatmap,
    width = 6, height = 7, dpi = 300
)
p_heatmap

```


## Closely examining the gamma delta T cell subtype

```{r}
wang_meta <- read.delim(glue("{wang_dir}/metadata.txt"), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    janitor::clean_names()

message(glue("Wang et al. 2021: {nrow(wang_meta)} cells"))
wang_count_mat <- readRDS(glue("{wang_dir}/counts.rds"))

# Match metadata to count matrix columns
common_cells <- intersect(colnames(wang_count_mat), wang_meta$cell)
wang_meta <- wang_meta %>%
    filter(cell %in% common_cells) %>%
    mutate(cell = factor(cell, levels = common_cells)) %>%
    arrange(cell) %>%
    mutate(cell = as.character(cell))
wang_count_mat <- wang_count_mat[, common_cells]

message(glue("Matched cells: {length(common_cells)}"))

# Blood only for marker sanity check
wang_blood_meta <- wang_meta %>%
    filter(grepl("Blood", sample_type))
wang_blood_counts <- wang_count_mat[, wang_blood_meta$cell]

message(glue("Blood cells: {ncol(wang_blood_counts)}"))
wang_blood_meta %>% count(cell_type2, sort = TRUE) %>% print(n = 30)

# Canonical/commonly used gamma-delta T-cell markers
gd_markers <- c("TRDC", "TRGC1", "TRGC2", "TRDV2", "TRGV9", "TRGV10", "TRGC3")
gd_markers_found <- intersect(gd_markers, rownames(wang_blood_counts))
gd_markers_missing <- setdiff(gd_markers, gd_markers_found)

message(glue("Markers found: {length(gd_markers_found)} / {length(gd_markers)}"))
if (length(gd_markers_missing) > 0) {
    message(glue("Missing markers: {paste(gd_markers_missing, collapse = ', ')}"))
}

if (length(gd_markers_found) == 0) {
    stop("None of the requested gamma-delta markers were found in wang_blood_counts.")
}

# Log-normalize counts to counts-per-10k
lib_size <- Matrix::colSums(wang_blood_counts)
norm_counts <- Matrix::t(Matrix::t(wang_blood_counts) / lib_size * 1e4)
log_norm_counts <- log1p(norm_counts)

# Build marker summary by cell subtype
marker_long <- purrr::map_dfr(gd_markers_found, function(g) {
    vals <- as.numeric(log_norm_counts[g, ])
    tibble(
        marker = g,
        cell = colnames(log_norm_counts),
        expr = vals
    )
}) %>%
    left_join(
        wang_blood_meta %>% select(cell, cell_type2),
        by = "cell"
    ) %>%
    group_by(cell_type2, marker) %>%
    summarize(
        avg_expr = mean(expr, na.rm = TRUE),
        pct_expr = mean(expr > 0, na.rm = TRUE) * 100,
        .groups = "drop"
    )

subtype_order <- marker_long %>%
    filter(marker == "TRDC") %>%
    arrange(desc(avg_expr)) %>%
    pull(cell_type2)

p_gd_markers <- marker_long %>%
    mutate(
        cell_type2 = factor(cell_type2, levels = subtype_order),
        marker = factor(marker, levels = gd_markers_found)
    ) %>%
    ggplot(aes(x = marker, y = cell_type2)) +
    geom_point(aes(size = pct_expr, color = avg_expr)) +
    scale_color_scico(palette = "vik", midpoint = 0) +
    scale_size_continuous(range = c(1, 8), limits = c(0, 100)) +
    theme_bw() +
    labs(
        x = "Gamma-delta marker",
        y = "Wang blood cell subtype",
        color = "Avg log1p(CP10k)",
        size = "% cells > 0",
        title = "Gamma-delta T-cell marker expression across Wang blood subtypes"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold")
    )

ggsave(
    filename = glue("{sc_figures_dir}/wang_blood_gamma_delta_marker_check.png"),
    plot = p_gd_markers,
    width = 8,
    height = 7,
    dpi = 300
)
# p_gd_markers
# it looks like these cells do express Gamma / Delta TCR receptors 

```


## Wang-focused mechanistic follow-up analyses

```{r wang-followup-setup}
primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

wang_meta <- read.delim(glue("{wang_dir}/metadata.txt"), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    janitor::clean_names()
wang_count_mat <- readRDS(glue("{wang_dir}/counts.rds"))
scored_cells <- readRDS(glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))
scored_cells %>% glimpse()

# Harmonize Wang metadata and align with count matrix columns
common_cells <- intersect(colnames(wang_count_mat), wang_meta$cell)
wang_meta <- wang_meta %>%
    filter(cell %in% common_cells) %>%
    mutate(
        condition = strex::str_before_first(sample_type, "\\("),
        condition = gsub("MCI", "PD", condition),
        condition = gsub("Parkinson", "PD", condition),
        condition = case_when(
            grepl("PD|parkinson|case", condition, ignore.case = TRUE) ~ "PD",
            grepl("HC|healthy|control|ctrl", condition, ignore.case = TRUE) ~ "HC",
            TRUE ~ condition
        ),
        tissue = case_when(
            grepl("CSF", sample_type) ~ "CSF",
            grepl("Blood", sample_type) ~ "Blood",
            TRUE ~ "Unknown"
        ),
        donor_id = sample,
        t_cell_subtype = cell_type2,
        cdr3b_aa = beta_cdr3_aa,
        study = "Wang_2021",
        compartment = tissue
    ) %>%
    arrange(match(cell, common_cells))

wang_count_mat <- wang_count_mat[, wang_meta$cell]
stopifnot(all(colnames(wang_count_mat) == wang_meta$cell))

wang_seurat <- CreateSeuratObject(
    counts = wang_count_mat,
    meta.data = wang_meta %>% column_to_rownames("cell")
)

# Attach CDR3 score primarily by cell barcode, then fallback by donor+CDR3
wang_scores <- scored_cells %>%
    filter(study == "Wang_2021") %>%
    mutate(cell = paste0(donor_id, "-", cell))

meta_tbl <- wang_seurat@meta.data %>%
    rownames_to_column("cell") %>%
    left_join(
        wang_scores %>% select(cell, all_of(primary_pheno)),
        by = "cell"
    )

missing_score <- is.na(meta_tbl[[primary_pheno]])
if (any(missing_score)) {
    fallback_tbl <- wang_scores %>%
        select(donor_id, cdr3b_aa, all_of(primary_pheno)) %>%
        distinct()
    meta_tbl <- meta_tbl %>%
        left_join(fallback_tbl, by = c("donor_id", "cdr3b_aa"), suffix = c("", "_fallback")) %>%
        mutate(
            !!primary_pheno := if_else(
                is.na(.data[[primary_pheno]]),
                .data[[paste0(primary_pheno, "_fallback")]],
                .data[[primary_pheno]]
            )
        ) %>%
        select(-all_of(paste0(primary_pheno, "_fallback")))
}

wang_seurat <- AddMetaData(
    object = wang_seurat,
    metadata = meta_tbl %>% column_to_rownames("cell")
)

wang_followup <- wang_seurat@meta.data %>%
    rownames_to_column("cell") %>%
    filter(tissue == "Blood", condition %in% c("PD", "HC")) %>%
    filter(!is.na(.data[[primary_pheno]])) %>%
    group_by(donor_id, clonotype) %>%
    mutate(clone_size = n()) %>%
    ungroup()

message(glue("Follow-up cells retained: {nrow(wang_followup)}"))
message(glue("Missing CDR3 score fraction: {round(mean(is.na(meta_tbl[[primary_pheno]])), 4)}"))
wang_followup %>% count(t_cell_subtype, sort = TRUE) %>% print(n = 30)

saveRDS(wang_followup, glue("{sc_results_dir}/wang_followup_blood_meta.rds"))
```

```{r wang-analysis-4-donor-coupling}
subtype_map <- list(
    cytotoxic_cd4 = "Cytotoxic CD4+ T cells",
    central_memory_cd8 = "Central memory CD8+ T cells"
)

missing_subtypes <- setdiff(unlist(subtype_map), unique(wang_followup$t_cell_subtype))
if (length(missing_subtypes) > 0) {
    stop(glue("Missing required subtypes: {paste(missing_subtypes, collapse = ', ')}"))
}

donor_pair <- wang_followup %>%
    filter(t_cell_subtype %in% unlist(subtype_map)) %>%
    mutate(subtype_key = case_when(
        t_cell_subtype == subtype_map$cytotoxic_cd4 ~ "cytotoxic_cd4",
        t_cell_subtype == subtype_map$central_memory_cd8 ~ "central_memory_cd8",
        TRUE ~ "other"
    )) %>%
    filter(subtype_key != "other") %>%
    group_by(condition, donor_id, subtype_key) %>%
    summarize(mean_score = mean(.data[[primary_pheno]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = subtype_key, values_from = mean_score) %>%
    drop_na(cytotoxic_cd4, central_memory_cd8)

cor_stats <- donor_pair %>%
    group_by(condition) %>%
    summarize(
        n_donors = n(),
        spearman_rho = cor(cytotoxic_cd4, central_memory_cd8, method = "spearman"),
        spearman_p = suppressWarnings(cor.test(cytotoxic_cd4, central_memory_cd8, method = "spearman")$p.value),
        pearson_r = cor(cytotoxic_cd4, central_memory_cd8, method = "pearson"),
        pearson_p = suppressWarnings(cor.test(cytotoxic_cd4, central_memory_cd8, method = "pearson")$p.value),
        .groups = "drop"
    )

p_donor_coupling <- donor_pair %>%
    ggplot(aes(x = central_memory_cd8, y = cytotoxic_cd4, color = condition, label = donor_id)) +
    geom_point(size = 3, alpha = 0.9) +
    ggrepel::geom_text_repel(size = 3, max.overlaps = 50) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
    theme_bw() +
    labs(
        x = "Mean CDR3 score (Central memory CD8)",
        y = "Mean CDR3 score (Cytotoxic CD4)",
        title = "Within-donor coupling of key Wang blood populations"
    )

ggsave(
    filename = glue("{sc_figures_dir}/wang_donor_coupling_cytotoxicCD4_vs_cmCD8.png"),
    plot = p_donor_coupling, width = 7, height = 6, dpi = 300
)

write_csv(cor_stats, glue("{sc_results_dir}/wang_donor_coupling_stats.csv"))
saveRDS(donor_pair, glue("{sc_results_dir}/wang_donor_coupling_pairs.rds"))

cor_stats %>% print()
p_donor_coupling
```

```{r wang-analysis-1-clone-size}
clone_size_df <- wang_followup %>%
    filter(t_cell_subtype %in% c("Cytotoxic CD4+ T cells", "Central memory CD8+ T cells")) %>%
    mutate(
        clone_bin = case_when(
            clone_size == 1 ~ "singleton",
            clone_size >= 2 & clone_size <= 4 ~ "small_2_4",
            clone_size >= 5 & clone_size <= 19 ~ "medium_5_19",
            clone_size >= 20 ~ "large_20plus",
            TRUE ~ "other"
        ),
        clone_bin = factor(clone_bin, levels = c("singleton", "small_2_4", "medium_5_19", "large_20plus"))
    ) %>%
    filter(!is.na(clone_bin))

clone_models <- clone_size_df %>%
    group_split(t_cell_subtype) %>%
    set_names(unique(clone_size_df$t_cell_subtype)) %>%
    purrr::map(~ lm(reformulate(c("log1p(clone_size)", "condition", "log1p(clone_size):condition"), response = primary_pheno), data = .x))

clone_model_stats <- purrr::imap_dfr(
    clone_models, ~ broom::tidy(.x, conf.int = TRUE) %>% mutate(t_cell_subtype = .y))
clone_counts <- clone_size_df %>% dplyr::count(t_cell_subtype, condition, clone_bin)

p_clone_bins <- clone_size_df %>%
    ggplot(aes(x = clone_bin, y = .data[[primary_pheno]], fill = condition)) +
    geom_violin(scale = "width", alpha = 0.6, trim = FALSE) +
    geom_boxplot(width = 0.2, outlier.size = 0.2, position = position_dodge(width = 0.9)) +
    facet_wrap(~ t_cell_subtype, scales = "free_y") +
    theme_bw() +
    labs(
        x = "Clone size bin",
        y = "CDR3 risk score",
        title = "CDR3 score stratified by clone size in key blood subtypes"
    )

ggsave(
    filename = glue("{sc_figures_dir}/wang_clone_size_stratified_cdr3_score.png"),
    plot = p_clone_bins, width = 10, height = 6, dpi = 300
)

write_csv(clone_model_stats, glue("{sc_results_dir}/wang_clone_size_lm_stats.csv"))
write_csv(clone_counts, glue("{sc_results_dir}/wang_clone_size_bin_counts.csv"))
saveRDS(clone_size_df, glue("{sc_results_dir}/wang_clone_size_input_df.rds"))

clone_model_stats %>% print(n = 50)
p_clone_bins
```

```{r wang-analysis-3-transitioning-retained}
cd8_transition_df <- wang_followup %>%
    filter(grepl("CD8", t_cell_subtype)) %>%
    mutate(
        cd8_state = case_when(
            grepl("Central memory", t_cell_subtype, ignore.case = TRUE) ~ "central_memory",
            grepl("Transition", t_cell_subtype, ignore.case = TRUE) ~ "transitional",
            grepl("Terminal|Effector", t_cell_subtype, ignore.case = TRUE) ~ "terminal_effector",
            TRUE ~ NA_character_
        )
    ) %>%
    filter(!is.na(cd8_state), !is.na(clonotype), clonotype != "")

clonotype_labels <- cd8_transition_df %>%
    group_by(condition, donor_id, clonotype) %>%
    summarize(
        has_central = any(cd8_state == "central_memory"),
        has_downstream = any(cd8_state %in% c("transitional", "terminal_effector")),
        .groups = "drop"
    ) %>%
    filter(has_central) %>%
    mutate(transition_status = if_else(has_downstream, "transitioning", "retained"))

clonotype_scores <- cd8_transition_df %>%
    inner_join(clonotype_labels, by = c("condition", "donor_id", "clonotype")) %>%
    group_by(condition, donor_id, clonotype, transition_status) %>%
    summarize(mean_score = mean(.data[[primary_pheno]], na.rm = TRUE), .groups = "drop")

transition_tests <- clonotype_scores %>%
    group_by(condition) %>%
    summarize(
        n_transitioning = sum(transition_status == "transitioning"),
        n_retained = sum(transition_status == "retained"),
        p_value = tryCatch(
            wilcox.test(mean_score ~ transition_status)$p.value,
            error = function(e) NA_real_
        ),
        delta_median = median(mean_score[transition_status == "transitioning"], na.rm = TRUE) -
            median(mean_score[transition_status == "retained"], na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"))

p_transition <- clonotype_scores %>%
    ggplot(aes(x = transition_status, y = mean_score, fill = transition_status)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_boxplot(width = 0.2, outlier.size = 0.4) +
    facet_wrap(~ condition) +
    theme_bw() +
    labs(
        x = "CD8 clonotype class",
        y = "Mean CDR3 risk score",
        title = "Transitioning vs retained central-memory CD8 clonotypes"
    ) +
    theme(legend.position = "none")

ggsave(
    filename = glue("{sc_figures_dir}/wang_cd8_transitioning_vs_retained_cdr3_score.png"),
    plot = p_transition, width = 8, height = 5, dpi = 300
)

write_csv(transition_tests, glue("{sc_results_dir}/wang_cd8_transitioning_vs_retained_stats.csv"))
saveRDS(clonotype_labels, glue("{sc_results_dir}/wang_cd8_transitioning_labels.rds"))
saveRDS(clonotype_scores, glue("{sc_results_dir}/wang_cd8_transitioning_scores.rds"))

transition_tests %>% print()
p_transition
```

```{r wang-analysis-2-trajectory}
trajectory_df <- wang_followup %>%
    filter(grepl("CD8", t_cell_subtype)) %>%
    mutate(
        cd8_state = case_when(
            grepl("Central memory", t_cell_subtype, ignore.case = TRUE) ~ "central_memory",
            grepl("Transition", t_cell_subtype, ignore.case = TRUE) ~ "transitional",
            grepl("Terminal|Effector", t_cell_subtype, ignore.case = TRUE) ~ "terminal_effector",
            TRUE ~ NA_character_
        ),
        pseudo_stage = case_when(
            cd8_state == "central_memory" ~ 1,
            cd8_state == "transitional" ~ 2,
            cd8_state == "terminal_effector" ~ 3,
            TRUE ~ NA_real_
        )
    ) %>%
    filter(!is.na(pseudo_stage))

trajectory_model <- lm(
    reformulate(c("pseudo_stage", "condition", "pseudo_stage:condition"), response = primary_pheno),
    data = trajectory_df
)

trajectory_stats <- broom::tidy(trajectory_model, conf.int = TRUE)

p_trajectory <- trajectory_df %>%
    ggplot(aes(x = pseudo_stage, y = .data[[primary_pheno]], color = condition)) +
    geom_jitter(width = 0.08, alpha = 0.1, size = 0.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
    scale_x_continuous(
        breaks = c(1, 2, 3),
        labels = c("Central memory", "Transitional", "Terminal/effector")
    ) +
    theme_bw() +
    labs(
        x = "CD8 trajectory proxy stage",
        y = "CDR3 risk score",
        title = "CDR3 score trend across CD8 differentiation trajectory"
    )

ggsave(
    filename = glue("{sc_figures_dir}/wang_cd8_trajectory_cdr3_score_trend.png"),
    plot = p_trajectory, width = 9, height = 5, dpi = 300
)

write_csv(trajectory_stats, glue("{sc_results_dir}/wang_cd8_trajectory_lm_stats.csv"))
saveRDS(trajectory_df, glue("{sc_results_dir}/wang_cd8_trajectory_input_df.rds"))

trajectory_stats %>% print(n = 20)
p_trajectory
```

```{r wang-analysis-5-de}
# ────────────────────────────────────────────────────────────────────────────
# Primary DE: MAST hurdle model with donor random intercept.
#   ~ score_group + (1 | donor_id)   method = "glmer"
#
# Partitioning rules:
#   - Drop score==0 cells (CDR3 has no match in any significant HLA-risk tag;
#     "uninformative", not "low").
#   - Median-split the remaining non-zero score distribution, per subtype.
#     Every non-zero cell contributes to high or low — no middle is thrown
#     away, no clone collapse.
#   - All features kept (TCR V/J/C genes retained — they're informative about
#     cell-type composition, not just artifacts).
#
# Pseudobulk edgeR is kept below as a sensitivity analysis, not primary.
# ────────────────────────────────────────────────────────────────────────────

suppressPackageStartupMessages({
    library(MAST); library(SummarizedExperiment); library(edgeR)
})

de_subtypes <- c("Cytotoxic CD4+ T cells", "Central memory CD8+ T cells")

# --- 1. Partition: drop score==0, median-split within subtype ---
de_input <- wang_followup %>%
    filter(condition == "PD") %>%
    filter(t_cell_subtype %in% de_subtypes) %>%
    filter(.data[[primary_pheno]] != 0) %>%
    group_by(t_cell_subtype) %>%
    mutate(
        .med = median(.data[[primary_pheno]]),
        score_group = if_else(.data[[primary_pheno]] > .med, "high", "low")
    ) %>%
    ungroup() %>%
    select(-.med)

message("Cells per (subtype, score_group) after zero-drop + median split:")
de_input %>% dplyr::count(t_cell_subtype, score_group) %>% print()

message("Donor × score_group cell counts (no clone collapse):")
de_input %>% dplyr::count(t_cell_subtype, donor_id, score_group) %>%
    tidyr::pivot_wider(names_from = score_group, values_from = n, values_fill = 0) %>%
    arrange(t_cell_subtype, donor_id) %>% print(n = 50)

# --- 2. MAST glmer with donor random intercept (PRIMARY) ---
# FindMarkers does not expose random-effect formula syntax; call MAST::zlm directly.
# Recipe: zlm(~ score_group + (1|donor_id), method='glmer', ebayes=FALSE,
#             strictConvergence=FALSE). Uses lme4 under the hood.
run_mast_glmer <- function(df_sub, seurat_obj,
                           min_pct = 0.05, n_cores = 8,
                           checkpoint_dir = sc_results_dir) {
    subtype_name <- df_sub$t_cell_subtype[1]
    subtype_slug <- gsub("[^A-Za-z0-9]+", "_", subtype_name)
    lrt_path <- glue("{checkpoint_dir}/_mast_glmer_lrt__{subtype_slug}.rds")
    pct_path <- glue("{checkpoint_dir}/_mast_glmer_pct__{subtype_slug}.rds")

    if (file.exists(lrt_path) && file.exists(pct_path)) {
        message(glue("[{subtype_name}] resuming from checkpoint: {basename(lrt_path)}"))
        lrt <- readRDS(lrt_path)
        pct_info <- readRDS(pct_path)
    } else {
        cells_use <- df_sub$cell
        obj_sub <- subset(seurat_obj, cells = cells_use)
        obj_sub$score_group <- factor(
            df_sub$score_group[match(colnames(obj_sub), df_sub$cell)],
            levels = c("low", "high")
        )
        obj_sub$donor_id <- factor(df_sub$donor_id[match(colnames(obj_sub), df_sub$cell)])
        obj_sub <- NormalizeData(obj_sub, verbose = FALSE)

        norm_mat <- GetAssayData(obj_sub, layer = "data")
        det_low  <- Matrix::rowMeans(norm_mat[, obj_sub$score_group == "low"]  > 0)
        det_high <- Matrix::rowMeans(norm_mat[, obj_sub$score_group == "high"] > 0)
        keep_genes <- names(det_low)[pmax(det_low, det_high) >= min_pct]
        obj_sub <- subset(obj_sub, features = keep_genes)

        message(glue("[{subtype_name}] cells: high={sum(obj_sub$score_group=='high')} low={sum(obj_sub$score_group=='low')}; genes after {min_pct*100}% filter: {length(keep_genes)}; donors: {nlevels(obj_sub$donor_id)}"))

        expr_mat <- as.matrix(GetAssayData(obj_sub, layer = "data"))
        cdat <- data.frame(
            wellKey     = colnames(obj_sub),
            score_group = obj_sub$score_group,
            donor_id    = obj_sub$donor_id,
            row.names   = colnames(obj_sub)
        )
        fdat <- data.frame(primerid = rownames(obj_sub), row.names = rownames(obj_sub))
        sca <- MAST::FromMatrix(
            exprsArray   = expr_mat,
            cData        = cdat,
            fData        = fdat,
            check_sanity = FALSE
        )

        options(mc.cores = n_cores)
        tic(glue("MAST glmer fit {subtype_name}"))
        zlm_fit <- MAST::zlm(
            formula           = ~ score_group + (1 | donor_id),
            sca               = sca,
            method            = "glmer",
            ebayes            = FALSE,
            strictConvergence = FALSE,
            parallel          = TRUE
        )
        toc()

        tic(glue("MAST LRT {subtype_name}"))
        lrt <- MAST::summary(zlm_fit, doLRT = "score_grouphigh")$datatable
        toc()

        pct_info <- list(
            genes    = rownames(obj_sub),
            det_low  = det_low[rownames(obj_sub)],
            det_high = det_high[rownames(obj_sub)]
        )
        saveRDS(lrt,      lrt_path)
        saveRDS(pct_info, pct_path)
        message(glue("[{subtype_name}] checkpoint saved: {basename(lrt_path)}"))
    }

    # Extract from the summary table (use dplyr::rename explicitly — S4Vectors masks it)
    p_tbl <- lrt[contrast == "score_grouphigh" & component == "H",
                 .(primerid, `Pr(>Chisq)`)] %>%
        as_tibble() %>%
        dplyr::rename(gene = primerid, p_val = `Pr(>Chisq)`)

    fc_tbl <- lrt[contrast == "score_grouphigh" & component == "logFC",
                  .(primerid, coef, ci.hi, ci.lo)] %>%
        as_tibble() %>%
        dplyr::rename(gene        = primerid,
                      avg_log2FC  = coef,
                      logFC_ci_hi = ci.hi,
                      logFC_ci_lo = ci.lo) %>%
        mutate(
            avg_log2FC  = avg_log2FC  / log(2),
            logFC_ci_hi = logFC_ci_hi / log(2),
            logFC_ci_lo = logFC_ci_lo / log(2)
        )

    pct_tbl <- tibble(
        gene  = pct_info$genes,
        pct.1 = pct_info$det_high,   # "high" is ident.1
        pct.2 = pct_info$det_low
    )

    out <- p_tbl %>%
        left_join(fc_tbl,  by = "gene") %>%
        left_join(pct_tbl, by = "gene") %>%
        mutate(
            p_val_adj      = p.adjust(p_val, method = "BH"),
            t_cell_subtype = subtype_name
        )

    n_na <- sum(is.na(out$p_val))
    message(glue("[{subtype_name}] MAST glmer: {nrow(out)} genes tested; {n_na} failed convergence (p=NA)."))
    out
}

mast_glmer_results <- de_input %>%
    group_split(t_cell_subtype) %>%
    purrr::map_dfr(~ run_mast_glmer(.x, seurat_obj = wang_seurat))

# --- 3. Primary volcano from MAST glmer ---
de_results <- mast_glmer_results %>%
    mutate(
        p_val_adj_plot = pmax(p_val_adj, .Machine$double.xmin, na.rm = TRUE),
        neglog10_padj  = -log10(p_val_adj_plot),
        sig            = !is.na(p_val_adj) & p_val_adj < 0.05,
        method         = "MAST_glmer_donor_RE"
    )

de_labels <- de_results %>%
    filter(sig, is.finite(avg_log2FC), is.finite(neglog10_padj)) %>%
    group_by(t_cell_subtype) %>%
    group_modify(~ bind_rows(
        .x %>% filter(avg_log2FC > 0) %>% arrange(p_val_adj, desc(avg_log2FC)) %>% slice_head(n = 15),
        .x %>% filter(avg_log2FC < 0) %>% arrange(p_val_adj, avg_log2FC)       %>% slice_head(n = 15)
    ) %>% distinct(gene, .keep_all = TRUE)) %>%
    ungroup()

p_de_volcano <- de_results %>%
    filter(!is.na(p_val_adj)) %>%
    ggplot(aes(x = avg_log2FC, y = neglog10_padj, color = sig)) +
    geom_point(alpha = 0.6, size = 1) +
    ggrepel::geom_text_repel(
        data = de_labels, aes(label = gene),
        size = 2.6, max.overlaps = Inf, box.padding = 0.3, point.padding = 0.1,
        min.segment.length = 0, show.legend = FALSE
    ) +
    facet_wrap(~ t_cell_subtype, scales = "free_y") +
    theme_bw() +
    labs(
        x = "log2FC (high CDR3 score vs low; MAST glmer, (1|donor_id))",
        y = "-log10(FDR)",
        title = "PD blood DE by CDR3-score median split — MAST with donor random intercept"
    ) +
    scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "darkred")) +
    theme(legend.position = "none")

ggsave(
    filename = glue("{sc_figures_dir}/wang_pd_de_high_vs_low_cdr3score_volcano.png"),
    plot = p_de_volcano, width = 10, height = 5, dpi = 300
)

write_csv(de_results, glue("{sc_results_dir}/wang_pd_de_high_vs_low_cdr3score_MAST_glmer.csv"))
saveRDS(de_input, glue("{sc_results_dir}/wang_pd_de_input_groups.rds"))

cat("\n=== MAST glmer: top 15 per subtype (primary) ===\n")
de_results %>%
    filter(!is.na(p_val_adj)) %>%
    arrange(p_val_adj) %>%
    group_by(t_cell_subtype) %>%
    slice_head(n = 15) %>%
    print(n = 30)

# ────────────────────────────────────────────────────────────────────────────
# Pseudobulk edgeR — SENSITIVITY ANALYSIS ONLY. Not primary.
# Kept for method cross-check. Reuses the same de_input (no clone collapse,
# no TCR filter). Drops donors that don't appear in both groups for this
# paired-donor design (a necessity for edgeR's blocking factor — not for MAST).
# ────────────────────────────────────────────────────────────────────────────
counts_mat <- GetAssayData(wang_seurat, assay = "RNA", layer = "counts")

pseudobulk_de <- function(df_sub, min_cells = 10) {
    subtype_name <- df_sub$t_cell_subtype[1]
    df_sub <- df_sub %>% mutate(pb_sample = paste(donor_id, score_group, sep = "__"))
    sample_key <- df_sub %>% distinct(pb_sample, donor_id, score_group)
    cell_counts <- df_sub %>% dplyr::count(pb_sample)
    keep_samples <- cell_counts %>% filter(n >= min_cells) %>% pull(pb_sample)
    if (length(keep_samples) < 4) {
        warning(glue("{subtype_name}: only {length(keep_samples)} pseudobulk samples with >={min_cells} cells; skipping edgeR."))
        return(tibble())
    }
    sample_key <- sample_key %>% filter(pb_sample %in% keep_samples)
    if (dplyr::n_distinct(sample_key$score_group) < 2) {
        warning(glue("{subtype_name}: only one group passes cell-count threshold; skipping edgeR."))
        return(tibble())
    }
    cells_per_sample <- split(df_sub$cell[df_sub$pb_sample %in% keep_samples],
                              df_sub$pb_sample[df_sub$pb_sample %in% keep_samples])
    pb_counts <- sapply(cells_per_sample[keep_samples], function(cc) {
        Matrix::rowSums(counts_mat[, cc, drop = FALSE])
    })
    storage.mode(pb_counts) <- "integer"
    meta <- sample_key[match(colnames(pb_counts), sample_key$pb_sample), ]
    meta$score_group <- factor(meta$score_group, levels = c("low", "high"))
    meta$donor_id    <- factor(meta$donor_id)

    keep_donors <- meta %>% as_tibble() %>%
        group_by(donor_id) %>% filter(dplyr::n_distinct(score_group) == 2) %>%
        pull(donor_id) %>% as.character() %>% unique()
    if (length(keep_donors) < 2) {
        warning(glue("{subtype_name}: <2 donors in both groups; skipping edgeR."))
        return(tibble())
    }
    keep_cols <- meta$donor_id %in% keep_donors
    pb_counts <- pb_counts[, keep_cols]
    meta      <- droplevels(meta[keep_cols, ])

    y <- DGEList(counts = pb_counts, samples = meta)
    keep_genes <- filterByExpr(y, group = meta$score_group)
    y <- y[keep_genes, , keep.lib.sizes = FALSE]
    y <- calcNormFactors(y)
    design <- model.matrix(~ donor_id + score_group, data = meta)
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design)
    qlf <- glmQLFTest(fit, coef = "score_grouphigh")
    topTags(qlf, n = Inf, sort.by = "none")$table %>%
        as_tibble(rownames = "gene") %>%
        mutate(t_cell_subtype = subtype_name,
               n_donors       = length(keep_donors),
               n_samples      = ncol(pb_counts))
}

pb_results <- de_input %>%
    group_split(t_cell_subtype) %>%
    purrr::map_dfr(pseudobulk_de)

pb_out <- pb_results %>%
    dplyr::rename(p_val = PValue, p_val_adj = FDR, avg_log2FC = logFC) %>%
    mutate(method = "pseudobulk_edgeR_sensitivity")

write_csv(pb_out, glue("{sc_results_dir}/wang_pd_de_high_vs_low_cdr3score_pseudobulk_edgeR_sensitivity.csv"))

cat("\n=== Pseudobulk edgeR: top 15 per subtype (SENSITIVITY) ===\n")
pb_out %>% arrange(p_val_adj) %>% group_by(t_cell_subtype) %>% slice_head(n = 15) %>% print(n = 30)

p_de_volcano
```



## Session Info

```{r session-info}
sessionInfo()
```
