---
title: "Clustering TCR repertoires - ML models"
editor: source
author: "Joe Boktor"
date: '2025-08-26'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---


Environment setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(data.table)
library(seqinr)
library(progress)
library(immunarch)
library(nlme)
library(broom)
library(broom.mixed)
library(parallelDist)
library(tidymodels)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(scico)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))
source(paste0(src_dir, "/R_scripts/_airr_analyze-permanova.R"))

scratch_dir <- "/resnick/scratch/jbok/PDMBS"
gliph2_plots_dir <- glue("{wkdir}/figures/AIRR/gliph2/")
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
arcasHLA_interim <- glue("{wkdir}/data/interim/airr/arcasHLA")
gliph2_input_dir <- glue("{wkdir}/data/input/gliph2/param_files")
gliph2_res_dir <- glue("{wkdir}/data/interim/airr/gliph2/processing")
clustcr_dir <- glue("{wkdir}/data/interim/clusTCR")
clustcr_figs <- glue("{wkdir}/figures/clusTCR")
tcr_cdr3_riskscore_complied <- 
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")
cohort_subset_dir <- glue("{clustcr_dir}/results/cohort_subsets")

# dir.create(clustcr_figs, showWarnings = FALSE, recursive = TRUE)
# dir.create(glue("{clustcr_dir}/input"), showWarnings = FALSE, recursive = TRUE)
# dir.create(glue("{clustcr_dir}/results"), showWarnings = FALSE, recursive = TRUE)

library(doParallel)
registerDoParallel(cores = 15)

```


Loading metadata and TCRB data

```{r, eval=FALSE}
trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()

trust4_reports_tcrb %>% glimpse()

# clonotype_volume_df <- trust4_reports_tcrb %>% 
#   group_by(sample_id) %>% 
#   # group_by(participant_id) %>% 
#   dplyr::summarize(
#       clonal_volume = sum(`#count`),
#       clonal_count = n()
#       ) %>%
#   glimpse()

# # clonotype_volume_df$clonal_count %>% summary()
# # clonotype_volume_df$clonal_volume %>% summary()
# low_vol_samples <- clonotype_volume_df %>% 
#   filter(clonal_volume < 200) %>% 
#   glimpse()

metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  # as.data.table() %>% 
  # left_join(clonotype_volume_df, by = "sample_id") %>% 
  glimpse()

# Select samples with complete MHC_II_PC data
mhc_clustering_data <- metadata %>%
  select(participant_id, contains("MHC_II_PC")) %>% 
  distinct() %>% 
  glimpse()
mhc_pc_matrix <- mhc_clustering_data %>%
  select(contains("MHC_II_PC")) %>%
  scale() %>%
  as.matrix()

kmeans_result <- kmeans(mhc_pc_matrix, centers = 5, nstart = 25)

# Add cluster assignments to the data
mhc_clustering_data$mhc_cluster <- factor(kmeans_result$cluster, 
                                         levels = 1:5,
                                         labels = paste0("MHCII_Cluster_", 1:5))
mhc_clustering_data %>% glimpse()

metadata <- metadata %>% 
  left_join(mhc_clustering_data %>% select(participant_id, mhc_cluster)) %>% 
  distinct() %>% 
  glimpse()

# metadata_categories_df <- readRDS(
#   glue("{wkdir}/data/interim/metadata/metadata_categories_dataframe_2025-07-21.rds")) %>% 
#   glimpse()

all_cohorts <- unique(metadata$study)
```

## TRYING TO SETUP DATA FOR ML MODELS

```{r}
chi2_results_df <- readRDS(
  glue("{gliph2_res_dir}/chi2_results_df_2025-08-25.rds")) %>% 
  glimpse()
sig_cids <- chi2_results_df %>% 
  filter(chi2_p_value <= 0.05) %>% 
  pull(cluster_id) %>% 
  unique()

# # select only PDBP cohort
# pdbp_sids <- metadata %>% 
#   filter(study == "PDBP") %>% 
#   pull(sample_id)

clust_mat_binary <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_binary_SSxRS_2025-08-26.rds")) %>% 
  as.data.table() %>% 
  # filter(sample_id %in% pdbp_sids) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  select(all_of(sig_cids)) %>% 
  as.matrix()

dim(clust_mat_binary)

# Processes to aggregate binary data to participant level ----
pid_sid_map <- metadata %>% select(sample_id, participant_id) %>% 
  filter(sample_id %in% rownames(clust_mat_binary)) %>% 
  glimpse()

# Initialize the matrix with proper dimensions and column names
clust_mat_binary_pid <- matrix(
  nrow = length(unique(pid_sid_map$participant_id)),
  ncol = ncol(clust_mat_binary),
  dimnames = list(
    unique(pid_sid_map$participant_id),
    colnames(clust_mat_binary)
  )
)
for (pid in unique(pid_sid_map$participant_id)) {
  print(pid)
  sid_vec <- pid_sid_map %>% 
    filter(participant_id == pid) %>% 
    pull(sample_id)
  if (length(sid_vec) > 1) {
    clust_mat_binary_pid[pid, ] <- as.integer(colSums(clust_mat_binary[sid_vec, ]) > 0)
  } else {
    clust_mat_binary_pid[pid, ] <- clust_mat_binary[sid_vec, ]
  }
}

clust_mat_binary_pid %>% dim()
clust_mat_binary_pid[1:5, 1:5]

meta_pid_casecontrol <- metadata %>% 
  # select(participant_id, case_control_other_latest, mhc_cluster) %>% 
  select(participant_id, case_control_other_latest) %>% 
  distinct()

model_df <- clust_mat_binary_pid %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "participant_id") %>% 
  left_join(meta_pid_casecontrol) %>% 
  dplyr::rename(predict_var = case_control_other_latest) %>% 
  mutate(predict_var = factor(predict_var, levels = c("Case", "Control")))

dim(model_df)
model_df$predict_var %>% table()
model_df$mhc_cluster %>% table()

```



Spliting the data into training and testing sets.

```{r}
set.seed(123)
data_split <- initial_split(model_df, strata = predict_var)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Building a recipe
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(participant_id, new_role = "id") %>% 
  step_zv(all_predictors()) %>% 
  step_pca(all_predictors(), num_comp = 100)        # Reduce to 100 principal components

# Building a model specification
lasso_spec <- logistic_reg(
  penalty = tune(),
  # penalty = 0.01,    # Reasonable default LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

lasso_spec <- logistic_reg(
  penalty = tune(),
  # penalty = 0.01,    # Reasonable default LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

# Building a workflow
wf <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(rec)

# Tuning the grid of lambda values
lambda_grid <- grid_regular(penalty(), levels = 10)  # 10 values from min to max
set.seed(123)
folds <- vfold_cv(train_data, v = 5, strata = predict_var)
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = lambda_grid,
  metrics = metric_set(roc_auc, accuracy)
)

# Selecting the best models and evaluating performance
best_lambda <- select_best(tune_res, "roc_auc")
final_wf <- finalize_workflow(wf, best_lambda)
final_fit <- fit(final_wf, data = train_data)
# Predict on test set
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# test_pred <- test_pred %>%
#   mutate(
#     predict_var = factor(predict_var, levels = c("Case", "Control")),
#     # .pred_class = factor(.pred_class, levels = c("Control", "Case"))
#   )

test_pred %>%
  select(predict_var, .pred_Case) %>%
  head(10) %>%
  print()

# Now evaluate
roc_auc(test_pred, truth = predict_var, .pred_Case)
# accuracy(test_pred, truth = predict_var, .pred_c)

```

Testing LASSO models per MHC bin

```{r}
library(tidymodels)
library(dplyr)
library(purrr)

# Split within bins first
set.seed(123)
splits <- model_df %>% 
  group_by(mhc_cluster) %>% 
  group_map(~ initial_split(.x, strata = predict_var))

# Define a function to train/evaluate a single bin model
train_bin_model <- function(split) {
  train_data <- training(split)
  test_data  <- testing(split)
  
  rec <- recipe(predict_var ~ ., data = train_data) %>%
    update_role(participant_id, new_role = "id") %>%
    step_zv(all_predictors())
    # step_pca(all_predictors(), num_comp = 100)
  
  lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
    set_engine("glmnet")
  
  wf <- workflow() %>%
    add_model(lasso_spec) %>%
    add_recipe(rec)
  
  folds <- vfold_cv(train_data, v = 5, strata = predict_var)
  
  lambda_grid <- grid_regular(penalty(), levels = 10)
  
  tune_res <- tune_grid(
    wf,
    resamples = folds,
    grid = lambda_grid,
    metrics = metric_set(roc_auc, accuracy)
  )
  
  best_lambda <- select_best(tune_res, "roc_auc")
  final_wf <- finalize_workflow(wf, best_lambda)
  
  final_fit <- fit(final_wf, data = train_data)
  
  # Predict on test set
  preds <- predict(final_fit, test_data, type = "prob") %>%
    bind_cols(test_data)
  
  roc <- roc_auc(preds, truth = predict_var, .pred_Case)
  # acc <- accuracy(preds, truth = predict_var, .pred_class)
  
  return(list(model = final_fit, roc_auc = roc))
}

# Apply the function across bins
bin_results <- purrr::map(splits, train_bin_model)

# Inspect performance per bin
purrr::map(bin_results, ~ .x$roc_auc)

```


## Random Forests Models

```{r}
library(tidymodels)
library(vip)
library(dplyr)
library(purrr)

set.seed(123)

# ----------------------------------------------------------
# 1. Assume your data already has an MHC bin column
# ----------------------------------------------------------
# model_df$mhc_bin <- factor(model_df$mhc_bin)  # Ensure it's a factor

# Split by MHC bin
splits <- model_df %>%
  group_by(mhc_cluster) %>%
  group_map(~ initial_split(.x, strata = predict_var))

# ----------------------------------------------------------
# 2. Define a function to train/evaluate RF model per bin
# ----------------------------------------------------------

# TO TEST 
# split <- splits[[1]]
train_rf_per_bin <- function(split) {
  train_data <- training(split)
  test_data  <- testing(split)
  
  # Build recipe
  rec <- recipe(predict_var ~ ., data = train_data) %>%
    update_role(participant_id, new_role = "id") %>%
    step_zv(all_predictors())

  # RF specification
  rf_spec <- rand_forest(
    mtry = tune(),
    trees = 500,
    min_n = 5
  ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("classification")

  # Workflow
  wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_recipe(rec)

  # Tuning grid
  rf_grid <- grid_regular(
    mtry(range = c(20, 80)),
    levels = 3
  )

  # CV folds within bin
  folds <- vfold_cv(train_data, v = 5, strata = predict_var)

  # Tune model
  tune_res <- tune_grid(
    wf,
    resamples = folds,
    grid = rf_grid,
    metrics = metric_set(roc_auc, accuracy)
  )

  # Select best model
  best_params <- select_best(tune_res, "roc_auc")
  final_wf <- finalize_workflow(wf, best_params)
  final_fit <- fit(final_wf, data = train_data)
  
  # Predict probabilities **and** classes
  test_pred <- predict(final_fit, test_data, type = "prob") %>%
    bind_cols(predict(final_fit, test_data, type = "class")) %>%
    bind_cols(test_data)

  # Evaluate ROC AUC and accuracy
  roc <- roc_auc(test_pred, truth = predict_var, .pred_Case)
  acc <- accuracy(test_pred, truth = predict_var, estimate = .pred_class)

  # Get feature importance
  importance <- final_fit %>%
    extract_fit_parsnip() %>%
    vip::vip(num_features = 20)

  list(
    model = final_fit,
    roc_auc = roc,
    accuracy = acc,
    importance_plot = importance
  )
}

# ----------------------------------------------------------
# 3. Apply the function to all bins
# ----------------------------------------------------------
rf_results <- purrr::map(splits, train_rf_per_bin)

# ----------------------------------------------------------
# 4. Summarize per-bin AUROC + accuracy
# ----------------------------------------------------------
rf_summary <- purrr::map_df(seq_along(rf_results), function(i) {
  tibble(
    mhc_bin = i,
    auc = rf_results[[i]]$roc_auc$.estimate,
    accuracy = rf_results[[i]]$accuracy$.estimate
  )
})

rf_summary
```



Main Random Forest function for cross-cohort testing

```{r}
library(tidymodels)
library(ranger)
library(vip)
library(dplyr)
library(purrr)

rf_train_and_test <- function(data, cohort_col, train_cohort, test_cohorts, include_train_eval = TRUE) {
  
  # ----------------------------
  # 1. Subset training data
  # ----------------------------
  train_data <- data %>% filter(!!sym(cohort_col) == train_cohort)
  
  # Add training cohort to test list if needed
  if (include_train_eval && !(train_cohort %in% test_cohorts)) {
    test_cohorts <- c(train_cohort, test_cohorts)
  }
  
  # ----------------------------
  # 2. Build recipe
  # ----------------------------
  rec <- recipe(predict_var ~ ., data = train_data) %>%
    step_rm(all_of(cohort_col)) %>%
    update_role(participant_id, new_role = "id") %>%
    step_zv(all_predictors())
  
  # ----------------------------
  # 3. Random Forest specification
  # ----------------------------
  rf_spec <- rand_forest(
    mtry = tune(),
    trees = 500,
    min_n = 5
  ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("classification")
  
  wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_recipe(rec)
  
  # ----------------------------
  # 4. Tune model
  # ----------------------------
  rf_grid <- grid_regular(
    mtry(range = c(20, 80)),
    levels = 3
  )
  
  set.seed(123)
  folds <- vfold_cv(train_data, v = 3, strata = predict_var)
  
  tune_res <- tune_grid(
    wf,
    resamples = folds,
    grid = rf_grid,
    metrics = metric_set(roc_auc, accuracy)
  )
  
  best_params <- select_best(tune_res, "roc_auc")
  final_wf <- finalize_workflow(wf, best_params)
  final_fit <- fit(final_wf, data = train_data)
  
  # ----------------------------
  # 5. Evaluate on test cohorts
  # ----------------------------
  results <- purrr::map_df(test_cohorts, function(cohort) {
    test_data <- data %>% filter(!!sym(cohort_col) == cohort)
    
    if (cohort == train_cohort) {
      # ---- Within-cohort performance via cross-validation ----
      folds_inner <- vfold_cv(train_data, v = 5, strata = predict_var)
      
      cv_res <- fit_resamples(
        final_wf,
        resamples = folds_inner,
        metrics = metric_set(roc_auc, accuracy)
      ) %>%
        collect_metrics() %>%
        select(.metric, mean) %>%   # <-- Use 'mean' instead of '.estimate'
        pivot_wider(names_from = .metric, values_from = mean)
      
      tibble(
        train_cohort = train_cohort,
        test_cohort = cohort,
        roc_auc = cv_res$roc_auc,
        accuracy = cv_res$accuracy
      )
      
    } else {
      # ---- External cohort predictions ----
      test_pred <- predict(final_fit, test_data, type = "prob") %>%
        bind_cols(predict(final_fit, test_data, type = "class")) %>%
        bind_cols(test_data)
      
      roc <- roc_auc(test_pred, truth = predict_var, .pred_Case)$.estimate
      acc <- accuracy(test_pred, truth = predict_var, estimate = .pred_class)$.estimate
      
      tibble(
        train_cohort = train_cohort,
        test_cohort = cohort,
        roc_auc = roc,
        accuracy = acc
      )
    }
  })
  
  list(
    trained_model = final_fit,
    performance = results
  )
}

```



```{r}
meta_pid_casecontrol <- metadata %>% 
  select(participant_id, case_control_other_latest, study) %>% 
  distinct()
meta_pid_casecontrol

model_df <- clust_mat_binary_pid %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "participant_id") %>% 
  left_join(meta_pid_casecontrol) %>% 
  dplyr::rename(predict_var = case_control_other_latest) %>% 
  mutate(predict_var = factor(predict_var, levels = c("Case", "Control")))

dim(model_df)
model_df$predict_var %>% table()
model_df$study %>% table()

# rf_results <- rf_train_and_test(
#   data = model_df,
#   cohort_col = "study",
#   train_cohort = "PPMI",
#   test_cohorts = c("PDBP", "HBS")
# )

# rf_results$trained_model
# rf_results$performance

```

Study to Study Transfer Models

```{r}
all_cohorts <- unique(metadata$study)

cross_cohort_results <- purrr::map_df(all_cohorts, function(train_c) {
  rf_out <- rf_train_and_test(
    data = model_df,
    cohort_col = "study",
    train_cohort = train_c,
    test_cohorts = setdiff(all_cohorts, train_c),
    include_train_eval = TRUE
  )
  rf_out$performance %>% 
    mutate(train_cohort = train_c)
})

cross_cohort_results
saveRDS(
    cross_cohort_results, 
    glue("{clustcr_dir}/results/ML_models/RandomForest_cross-cohort_AUROC_results_{Sys.Date()}.rds")
    )

# PLotting ---
cross_cohort_results <- readRDS(
    glue("{clustcr_dir}/results/ML_models/RandomForest_cross-cohort_AUROC_results_2025-08-27.rds")
    )

p_rf_heatmap <- cross_cohort_results %>%
  ggplot(aes(x = train_cohort, y = test_cohort, fill = roc_auc)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", roc_auc)), size = 4, color = "white") +
  scale_fill_scico(palette = 'berlin', midpoint = 0.5) + 
  # scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0.7) +
  labs(
    title = "Random Forests Study Transfer Models",
    subtitle = "Cross-Cohort AUROC",
    x = "Training Cohort",
    y = "Testing Cohort",
    fill = "AUROC"
  ) +
  # theme_bw(base_size = 14) +
  theme_minimal(base_size = 14) +
  coord_fixed() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),    
    axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = glue("{clustcr_figs}/ML_models/RandomForest_cross-cohort_AUROC_heatmap.png"),
  plot = p_rf_heatmap,
  width = 6,
  height = 5,
  dpi = 600
)

```

Leave one study out Analysis

```{r}
loso_results <- purrr::map_df(all_cohorts, function(train_c) {
    meta_pid_casecontrol <- metadata %>% 
      mutate(loso_cohort = if_else(study == train_c, 
        glue("loso_test_{train_c}"), glue("loso_train_{train_c}"))) %>% 
      select(participant_id, case_control_other_latest, loso_cohort) %>% 
      distinct()
  
  model_df <- clust_mat_binary_pid %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "participant_id") %>% 
    left_join(meta_pid_casecontrol) %>% 
    dplyr::rename(predict_var = case_control_other_latest) %>% 
    mutate(predict_var = factor(predict_var, levels = c("Case", "Control")))
  
  rf_out <- rf_train_and_test(
    data = model_df,
    cohort_col = "loso_cohort",
    train_cohort = glue("loso_train_{train_c}"),
    test_cohorts = glue("loso_test_{train_c}"),
    include_train_eval = TRUE
  )
  rf_out$performance %>%
    mutate(
      train_test_type = if_else(test_cohort == glue("loso_test_{train_c}"), "LOSO", "CV")
    )
})

saveRDS(loso_results, 
  glue("{clustcr_dir}/results/ML_models/RandomForest_loso_AUROC_results_{Sys.Date()}.rds")
  )

# PLotting ---

loso_results <- readRDS(
  glue("{clustcr_dir}/results/ML_models/RandomForest_loso_AUROC_results_2025-08-27.rds")
  )


p_loso_barplot <- loso_results %>% 
  mutate(train_cohort_name = str_after_last(train_cohort, "_")) %>% 
  mutate(xcolname = glue("{train_cohort_name} \n Held-out")) %>% 
  # mutate(train_test_type_name = 
  #   if_else(train_test_type == "LOSO", "Prediction on \n Held-out cohort", "CV Training \n all other cohorts")) %>% 
  ggplot(aes(x = xcolname, y = roc_auc, fill = train_test_type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(
    values = c("CV" = "grey", "LOSO" = "#AA4A44"),
    name = NULL,
    labels = c("5-fold CV training \non all other cohorts", "Prediction on \nheld-out cohort")
  ) +
  labs(
    # title = "Random Forests Study Transfer Models",
    # subtitle = "Leave-One-Study-Out (LOSO) AUROC",
    x = NULL,
    y = "AUROC"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    # axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  ) +
  # Add value labels on bars
  geom_text(
    aes(label = sprintf("%.2f", roc_auc)),
    position = position_dodge(width = 0.8),
    vjust = -0.5,
    size = 2.5,
    fontface = "bold"
  ) +
  # Set y-axis limits to accommodate labels
  ylim(0, max(loso_results$roc_auc) * 1.1)

ggsave(
  filename = glue("{clustcr_figs}/ML_models/RandomForest_loso_AUROC_barplot.png"),
  plot = p_loso_barplot,
  width = 3.5,
  height = 3,
  dpi = 600
)

```


Testing a workflow for random permutations

```{r}
get_random_baseline <- function(data, cohort_col, train_cohort, test_cohorts, n_perm = 50) {
  perm_results <- purrr::map_dfr(1:n_perm, function(i) {
    
    # Shuffle predict_var labels
    shuffled_data <- data %>%
      mutate(predict_var = sample(predict_var))
    
    # Re-run your existing RF training/testing
    rf_out <- rf_train_and_test(
      data = shuffled_data,
      cohort_col = cohort_col,
      train_cohort = train_cohort,
      test_cohorts = test_cohorts,
      include_train_eval = FALSE
    )
    
    rf_out$performance %>% 
      mutate(iteration = i)
  })
  
  perm_results
}

meta_pid_casecontrol <- metadata %>% 
  select(participant_id, case_control_other_latest, study) %>% 
  distinct()
model_df <- clust_mat_binary_pid %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "participant_id") %>% 
  left_join(meta_pid_casecontrol) %>% 
  dplyr::rename(predict_var = case_control_other_latest) %>% 
  mutate(predict_var = factor(predict_var, levels = c("Case", "Control")))

# Example usage — random baseline for PPMI training, predicting on PDBP + HBS
baseline_ppmi <- get_random_baseline(
  data = model_df,
  cohort_col = "study",
  train_cohort = "PPMI",
  test_cohorts = c("PDBP", "HBS"),
  n_perm = 3
)

# baseline_ppmi %>% 
#   group_by(test_cohort) %>% 
#   summarise(mean_auroc = mean(roc_auc), sd_auroc = sd(roc_auc))



purrr::walk(all_cohorts, function(train_c) {
  baseline_models <- get_random_baseline(
    data = model_df,
    cohort_col = "study",
    train_cohort = train_c,
    test_cohorts = setdiff(all_cohorts, train_c),
    n_perm = 50
  )
  saveRDS(baseline_models, 
    glue(
      "{clustcr_dir}/results/ML_models/",
      "RandomForest_cross-cohort_AUROC_randompermutation__{train_c}__{Sys.Date()}.rds")
      )
})

baseline_models <- list()
for (train_c in all_cohorts) {
  baseline_models[[train_c]] <- readRDS(
    glue(
      "{clustcr_dir}/results/ML_models/",
      "RandomForest_cross-cohort_AUROC_randompermutation__{train_c}__2025-08-28.rds")
      )
}

baseline_models <- all_cohorts %>% 
  purrr::map(function(train_c) {
    readRDS(
      glue(
        "{clustcr_dir}/results/ML_models/",
        "RandomForest_cross-cohort_AUROC_randompermutation__{train_c}__2025-08-28.rds")
      )
  }) %>% 
  bind_rows()

baseline_models_summary <- baseline_models %>% 
  group_by(train_cohort) %>% 
  # group_by(train_cohort, test_cohort) %>% 
  summarise(mean_auroc = mean(roc_auc), sd_auroc = sd(roc_auc)) %>% 
  dplyr::rename(roc_auc = mean_auroc) %>% 
  mutate(test_cohort = "Random Baseline")

# Remaking heatmap with random baseline

# PLotting ---
cross_cohort_results <- readRDS(
    glue("{clustcr_dir}/results/ML_models/RandomForest_cross-cohort_AUROC_results_2025-08-27.rds")
    )
cross_cohort_results_wbase <- cross_cohort_results %>% 
  bind_rows(baseline_models_summary)

p_rf_heatmap <- +  %>%
  ggplot(aes(x = train_cohort, y = test_cohort, fill = roc_auc)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", roc_auc)), size = 4, color = "white") +
  scale_fill_scico(palette = 'berlin', midpoint = 0.5) + 
  # scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0.7) +
  labs(
    title = "Random Forests Study Transfer Models",
    subtitle = "Cross-Cohort AUROC",
    x = "Training Cohort",
    y = "Testing Cohort",
    fill = "AUROC"
  ) +
  # theme_bw(base_size = 14) +
  theme_minimal(base_size = 14) +
  coord_fixed() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 12),    
    axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = glue("{clustcr_figs}/ML_models/RandomForest_cross-cohort_AUROC_heatmap_wbase.png"),
  plot = p_rf_heatmap,
  width = 6,
  height = 6,
  dpi = 600
)



```






NNOTHER TEMP - this time using inner and outer CV

```{r}
library(tidymodels)
library(ranger)
library(vip)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)

# --------------------------------------------------
# 1. Helper: Run nested CV for train→train evaluation
# --------------------------------------------------
rf_nested_cv <- function(data, cohort_col, train_cohort,
                         v_outer = 5, v_inner = 3,
                         rf_trees = 500,
                         mtry_range = c(20, 80),
                         mtry_levels = 3) {
  
  # Subset training data
  train_data <- data %>% filter(!!sym(cohort_col) == train_cohort)
  
  # Outer folds for unbiased evaluation
  outer_folds <- vfold_cv(train_data, v = v_outer, strata = predict_var)
  
  # Recipe
  rec <- recipe(predict_var ~ ., data = train_data) %>%
    step_rm(all_of(cohort_col)) %>%
    update_role(participant_id, new_role = "id") %>%
    step_zv(all_predictors())
  
  # Model spec
  rf_spec <- rand_forest(
    mtry = tune(),
    trees = rf_trees,
    min_n = 5
  ) %>%
    set_engine("ranger", importance = "impurity") %>%
    set_mode("classification")
  
  wf <- workflow() %>%
    add_model(rf_spec) %>%
    add_recipe(rec)
  
  rf_grid <- grid_regular(
    mtry(range = mtry_range),
    levels = mtry_levels
  )
  
  # Fit nested CV
  nested_res <- fit_resamples(
    wf,
    resamples = outer_folds,
    metrics = metric_set(roc_auc, accuracy),
    control = control_resamples(save_pred = TRUE)
  )
  
  # Collect unbiased metrics
  cv_metrics <- nested_res %>%
    collect_metrics() %>%
    select(.metric, mean, std_err) %>%
    pivot_wider(names_from = .metric, values_from = c(mean, std_err))
  
  # Hyperparameter tuning on entire training set (for predicting other cohorts)
  folds_inner <- vfold_cv(train_data, v = v_inner, strata = predict_var)
  tune_res <- tune_grid(
    wf,
    resamples = folds_inner,
    grid = rf_grid,
    metrics = metric_set(roc_auc)
  )
  best_params <- select_best(tune_res, "roc_auc")
  final_wf <- finalize_workflow(wf, best_params)
  final_fit <- fit(final_wf, data = train_data)
  
  list(
    cv_metrics = cv_metrics,
    final_fit = final_fit
  )
}

# --------------------------------------------------
# 2. Master function: Train on each cohort, test on all others
# --------------------------------------------------
rf_cross_cohort <- function(data, cohort_col, v_outer = 5, v_inner = 3) {
  
  all_cohorts <- unique(data[[cohort_col]])
  results <- list()
  
  for (train_c in all_cohorts) {
    message("Training on: ", train_c)
    
    # Step 1: Nested CV for unbiased within-cohort performance
    nested <- rf_nested_cv(
      data = data,
      cohort_col = cohort_col,
      train_cohort = train_c,
      v_outer = v_outer,
      v_inner = v_inner
    )
    
    # Extract within-cohort results
    train_metrics <- nested$cv_metrics %>%
      mutate(train_cohort = train_c,
             test_cohort = train_c,
             roc_auc = mean_roc_auc,
             accuracy = mean_accuracy) %>%
      select(train_cohort, test_cohort, roc_auc, accuracy)
    
    # Step 2: Predict on all other cohorts using tuned RF
    other_results <- map_df(setdiff(all_cohorts, train_c), function(test_c) {
      test_data <- data %>% filter(!!sym(cohort_col) == test_c)
      
      test_pred <- predict(nested$final_fit, test_data, type = "prob") %>%
        bind_cols(predict(nested$final_fit, test_data, type = "class")) %>%
        bind_cols(test_data)
      
      roc <- roc_auc(test_pred, truth = predict_var, .pred_Case)$.estimate
      acc <- accuracy(test_pred, truth = predict_var, estimate = .pred_class)$.estimate
      
      tibble(
        train_cohort = train_c,
        test_cohort = test_c,
        roc_auc = roc,
        accuracy = acc
      )
    })
    
    # Combine results for this training cohort
    results[[train_c]] <- bind_rows(train_metrics, other_results)
  }
  
  bind_rows(results)
}

# --------------------------------------------------
# 3. Visualization: Cross-Cohort AUROC Heatmap
# --------------------------------------------------
plot_cross_cohort_heatmap <- function(results) {
  results %>%
    ggplot(aes(x = train_cohort, y = test_cohort, fill = roc_auc)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", roc_auc)), size = 3) +
    scale_fill_gradient2(
      low = "red", mid = "white", high = "blue",
      midpoint = 0.7, limits = c(0, 1)
    ) +
    labs(
      title = "Cross-Cohort AUROC",
      x = "Training Cohort",
      y = "Testing Cohort",
      fill = "AUROC"
    ) +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


cross_results <- rf_cross_cohort(
  data = model_df,
  cohort_col = "study",
  v_outer = 5,
  v_inner = 3
)

# Check results table
cross_results

# Visualize cross-cohort generalization
plot_cross_cohort_heatmap(cross_results)

```



Example using the MHCII clusters 
```{r}
library(ranger)
rf_spec <- rand_forest(
  mtry = tune(),      # Keep tuning this - it's important
  trees = 500,        # Reduced from 1000 (still plenty)
  min_n = 5           # Fixed reasonable value
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Building a recipe
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(participant_id, new_role = "id") %>% 
  step_zv(all_predictors())  
  # step_pca(all_predictors(), num_comp = 100)     

# Update workflow
wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rec)

# Much smaller tuning grid
rf_grid <- grid_regular(
  mtry(range = c(20, 80)),  # Narrower range, fewer values
  levels = 3                 # Only 3 values instead of 5
)

# Faster cross-validation
set.seed(123)
folds <- vfold_cv(train_data, v = 3, strata = predict_var)  # 3-fold instead of 5-fold
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(roc_auc, accuracy)
)

# Selecting the best model
best_params <- select_best(tune_res, "roc_auc")
final_wf <- finalize_workflow(wf, best_params)
final_fit <- fit(final_wf, data = train_data)

# Predict probabilities **and** classes
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(predict(final_fit, test_data, type = "class")) %>%
  bind_cols(test_data)

# Evaluate ROC AUC and accuracy
roc <- roc_auc(test_pred, truth = predict_var, .pred_Case)
acc <- accuracy(test_pred, truth = predict_var, estimate = .pred_class)

# # Optional: Get feature importance
final_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(num_features = 20)

```



















# OLD STUFF
```{r}
chi2_results_df <- readRDS(
  glue("{gliph2_res_dir}/chi2_results_df_2025-05-26.rds"))
cluster_id_annotation <- chi2_results_df %>% 
    mutate(
        significant = case_when(
            abs(log2(odds_ratio)) >= 0.5 & chi2_p_value <= 0.05 ~ case_when(
                odds_ratio > 1 ~ "Enriched in Cases",
                odds_ratio < 1 ~ "Enriched in Controls"
            ),
            TRUE ~ "Not Significant"
        )
    )  %>% 
    select(index, type, significant) %>% 
    distinct() %>% 
    glimpse()

useful_cluster_ids <- cluster_id_annotation %>% 
  filter(significant != "Not Significant") %>% 
  pull(type)

cluster_pid_comp <- cluster_df_meta %>% 
    dplyr::select(index, participant_id) %>% 
    distinct() %>% 
    group_by(index) %>% 
    dplyr::summarise(n_participants = n()) %>% 
    ungroup()

unique_pid_thresh_inx <- cluster_pid_comp %>% 
    filter(n_participants >= 10) %>% 
    pull(index)

cluster_df_meta_f1 <- cluster_df_meta %>% 
    filter(index %in% unique_pid_thresh_inx) %>% 
    glimpse()
# cluster_df_meta_f1 %>% glimpse()

# Filter 3 clusters with pattern length > 3 fixed pos AAs
cluster_df_meta_f3 <- cluster_df_meta_f1 %>% 
    mutate(pattern_len = gsub("\\%", "", pattern) %>% nchar()) %>% 
    filter(pattern_len > 3) %>% 
    glimpse()
```


```{r}
library(tidyverse)
library(tidymodels)

cluster_df_meta_f3 %>% glimpse()
full_meta_df %>% glimpse()

# random sample_id selection per participant_id
set.seed(42)
rand_sids <- cluster_df_meta_f3  %>% 
    select(sample_id, participant_id) %>% 
    distinct() %>% 
    group_by(participant_id) %>%
    slice_sample(n = 1) %>% 
    pull(sample_id)

# Aggregate features per sample_id (example: sum Freq by sample_id and type)
features_df <- cluster_df_meta_f3 %>%
    filter(sample_id %in% rand_sids) %>% 
      # FILTERING WITH CHI2 results (this time not using 2nd filter round)
    filter(type %in% useful_cluster_ids) %>%  
    dplyr::group_by(sample_id, type) %>% 
    dplyr::summarise(count = sum(Freq, na.rm = TRUE), .groups = "drop") %>% 
    # sample_n(1000) %>% 
    pivot_wider(names_from = type, values_from = count, values_fill = 0)

feat_cols <- colnames(features_df)[-1]
length(feat_cols)
# Find columns with >0 in at least 5 samples
cols_to_keep <- feat_cols[
  colSums(features_df[feat_cols] > 0, na.rm = TRUE) >= 5
]
length(cols_to_keep)

filtered_features_df <- features_df %>%
  select(sample_id, all_of(cols_to_keep))
filtered_features_df %>% dim()

# Join with case/control status
meta_status <- full_meta_df %>%
  filter(case_control_other_latest %in% c("Case", "Control")) %>% 
  select(sample_id, predict_var = mds_updrs_part_iii_summary_score) %>% 
  drop_na(predict_var) %>% 
  glimpse()

model_df <- filtered_features_df %>%
  inner_join(meta_status, by = "sample_id")

model_df %>% dim()
```


Spliting the data into training and testing sets.

```{r}
set.seed(123)
data_split <- initial_split(model_df, strata = predict_var)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Building a recipe
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(sample_id, new_role = "id")
#   step_zv(all_predictors()) %>%
#   step_normalize(all_predictors())

# Building a model specification
lasso_spec <- logistic_reg(
  penalty = tune(),   # LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

lasso_spec <- logistic_reg(
  penalty = tune(),   # LASSO penalty
  mixture = 1         # 1 = LASSO, 0 = Ridge, in between = Elastic Net
) %>%
  set_engine("glmnet")

# Building a workflow
wf <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(rec)

# Tuning the grid of lambda values
lambda_grid <- grid_regular(penalty(), levels = 10)  # 10 values from min to max
set.seed(123)
folds <- vfold_cv(train_data, v = 5, strata = predict_var)
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = lambda_grid,
  metrics = metric_set(roc_auc, accuracy)
)

# Selecting the best models and evaluating performance
best_lambda <- select_best(tune_res, "roc_auc")
final_wf <- finalize_workflow(wf, best_lambda)
final_fit <- fit(final_wf, data = train_data)
# Predict on test set
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# Evaluate
roc_auc(test_pred, truth = predict_var, .pred_Case)
accuracy(test_pred, truth = predict_var, 
.pred_class = ifelse(.pred_Case > 0.5, "Case", "Control"))


```


Trying with Random Forests

```{r}
# future::plan(future::multisession, workers = 7)

# Building a recipe (keep this the same)
rec <- recipe(predict_var ~ ., data = train_data) %>%
  update_role(sample_id, new_role = "id")

# Random Forest Model - no tuning
rf_spec <- rand_forest(
  trees = 500,  # reasonable default
  min_n = 5     # reasonable default
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Building a workflow
wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rec)

# Fit the model directly (no tuning)
set.seed(123)
final_fit <- fit(wf, data = train_data)

# Predict on test set
test_pred <- predict(final_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# Evaluate
roc_auc(test_pred, truth = predict_var, .pred_Case)
accuracy(test_pred, truth = predict_var, 
         .pred_class = ifelse(.pred_Case > 0.5, "Case", "Control"))

# Get feature importance
rf_fit <- extract_fit_parsnip(final_fit)
var_imp <- vip::vip(rf_fit, num_features = 20)
print(var_imp)
```

