---
title: "Downloading AIRR-data"
editor: source
author: "Joe Boktor"
date: '2023-12-04'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    theme: cosmo
    column: page
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

```{r}
#| warning: false
#| message: false

library(googleCloudStorageR)
library(gargle)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(tidyverse)
library(plyr)
library(magrittr)
library(data.table)
# setDTthreads(16)

home_dir <- "/central/groups/MazmanianLab/joeB"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/parkinsons-microbial-blood-signatures")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
load_dot_env(glue("{wkdir}/.env"))
scratch_dir <- "/central/scratch/jbok/PDMBS"
t4_bucket = Sys.getenv('GCS_TRUST4')
```

```{r, eval=FALSE}
## Fetch token. See: https://developers.google.com/identity/protocols/oauth2/scopes
# scope <-c("https://www.googleapis.com/auth/cloud-platform")
# token <- gargle::token_fetch(scopes = scope)
token <- token_fetch()
googleCloudStorageR::gcs_auth(token = token)

```


```{r, eval=FALSE}
# Set the storage bucket location
gcs_get_bucket(Sys.getenv('GCS_TRUST4'))
# List the files in the bucket
bucket_files <- gcs_list_objects(Sys.getenv('GCS_TRUST4'))

trust4_file_paths <- list()
trust4_file_paths[["cdr3"]] <- bucket_files %>% 
  filter(grepl("_cdr3.out", name))
trust4_file_paths[["final_assemblies"]] <- bucket_files %>% 
  filter(grepl("_final.out", name))
trust4_file_paths[["final_reports"]] <- bucket_files %>% 
  filter(grepl("_report.tsv", name))

saveRDS(
  trust4_file_paths, 
  file = glue("{wkdir}/data/interim/airr/trust4_file-paths.rds")
)

# For HBS cohort
hbs_t4_buckets <- paste0("GCS_TRUST4_HBS", 4:5)

# bucket <- "GCS_TRUST4_HBS3"
for (bucket in hbs_t4_buckets) {
  message(glue("Downloading from {bucket}"))
  trust4_file_paths_hbs <- 
    shell_do(
      glue("gsutil ls {Sys.getenv(bucket)}/TRUST4workflow/**/*_{{report.tsv,final.out,cdr3.out}}"),
      stdout_path = TRUE
    )
  
  saveRDS(
    trust4_file_paths_hbs,
    glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_TRUST4-HBS_{bucket}.rds")
  )
}



# # bucket <- "GCS_TRUST4_HBS3"
# for (bucket in hbs_t4_buckets) {
#   message(glue("Downloading from {bucket}"))
#   cmd <- glue(
#     "gsutil ls '{Sys.getenv('bucket')}/TRUST4workflow/**/*_report.tsv' ",
#     "'{Sys.getenv('bucket')}/TRUST4workflow/**/*_final.out' ",
#     "'{Sys.getenv('bucket')}/TRUST4workflow/**/*_cdr3.out'"
#   )
#   trust4_file_paths_hbs <- shell_do(cmd, stdout_path = TRUE)
  
#   saveRDS(
#     trust4_file_paths_hbs,
#     glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_TRUST4-HBS_{bucket}.rds")
#   )
# }

```

```{r}
#' Function to download and parse data from Google Storage directly
#' into an R object
download_trust4_data <- function(file_path, 
                                bucket_name = Sys.getenv('GCS_TRUST4'), 
                                path_ext = "_report.tsv") {
  sample_name <- basename(file_path) %>% str_remove(path_ext)
  gcs_get_object(bucket = bucket_name, file_path) %>% 
    mutate(sample_id = sample_name)
}

```

```{r, eval=FALSE}
trust4_file_paths <- readRDS(
  file = glue("{wkdir}/data/interim/airr/trust4_file-paths.rds")
)

tic()
airr_reports <-
  trust4_file_paths$final_reports$name[1:2000] %>%
  set_names() %>%
  purrr::map( ~download_trust4_data(.))
toc()

saveRDS(
  airr_reports, 
  file = glue("{wkdir}/data/interim/airr/trust4_reports_list.rds")
)

airr_reports_df <- bind_rows(airr_reports)

saveRDS(
  airr_reports_df, 
  file = glue("{wkdir}/data/interim/airr/trust4_reports.rds")
)


```

```{r, eval=FALSE}
trust4_file_paths <- readRDS(
  file = glue("{wkdir}/data/interim/airr/trust4_file-paths.rds")
)

tic()
airr_reports <-
  trust4_file_paths$final_reports$name[2001:8436] %>%
  set_names() %>%
  purrr::map( ~download_trust4_data(.))
toc()

saveRDS(
  airr_reports, 
  file = glue("{wkdir}/data/interim/airr/trust4_reports_list_2001-8436.rds")
)

airr_reports_df <- bind_rows(airr_reports) %>% as.data.table()
saveRDS(
  airr_reports_df, 
  file = glue("{wkdir}/data/interim/airr/trust4_reports_2001-8436.rds")
)

```

```{r, eval=FALSE}
airr_reports_df_2000 <- readRDS(
  file = glue("{wkdir}/data/interim/airr/trust4_reports.rds")
)
airr_reports_df_2000 %<>% as.data.table()

saveRDS(
  bind_rows(
    airr_reports_df_2000,
    airr_reports_df
  ), 
  file = glue("{wkdir}/data/interim/airr/trust4_reports_8436.rds")
)

```

```{r}
airr_reports_df <- 
  readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_8436.rds")
)

write.csv(airr_reports_df, 
          file = 
            "{wkdir}/data/interim/airr/2023-05-04_TRUST4-reports-all.csv")

```

Adding HBS cohort and all remaining samples

```{r}
trust4_reports <- data.table::fread(
    glue("{wkdir}/data/interim/airr/2023-05-04_TRUST4-reports-all.csv"), 
    header = TRUE) %>%
    dplyr::select(-V1) %>% 
    glimpse()


hbs_and_final_reports <- list.files(glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/TRUST4_HBS/reports"),
  full.names = TRUE)

tic()
remaining_reports <- hbs_and_final_reports %>%
  purrr::set_names(basename(.) %>% str_remove("_report.tsv")) %>% 
  purrr::map( ~ read_tsv(.)) %>% 
  bind_rows(.id = "sample_id")
toc()

trust4_reports$sample_id %>% unique %>% length +
remaining_reports$sample_id %>% unique %>% length

# note that there are 641 redundant samples in the second analysis  - removing them below
c(trust4_reports$sample_id %>% unique(), remaining_reports$sample_id %>% unique()) %>% length
c(trust4_reports$sample_id %>% unique(), remaining_reports$sample_id %>% unique()) %>% unique %>% length

redundant_samples <- unique(remaining_reports$sample_id)[
  unique(remaining_reports$sample_id) %in% unique(trust4_reports$sample_id)]


remaining_reports_filt <- remaining_reports %>% 
  filter(sample_id %nin% redundant_samples)

trust4_reports_10579 <- bind_rows(
    trust4_reports,
    remaining_reports_filt
  ) %>% 
  glimpse()

# saveRDS(
#   airr_reports_df_x, 
#   file = glue("{wkdir}/data/interim/airr/trust4_reports_8436.rds")
# )

# From 10,579 samples - we assemble 27,985,042 TCR/ BCR receptors
write.csv(trust4_reports_10579, 
          glue("{wkdir}/data/interim/airr/2025-07-05_TRUST4-reports-all-10579.csv")
          )

```


#### Quality filtering TCRB sequences

```{r}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-17.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  glimpse()

key_meta_df <- metadata %>% 
  dplyr::select(sample_id, participant_id, case_control_other_latest, visit_month,
    age_at_baseline, sex, study, ethnicity, race, est_tcell_reads_abis) %>% 
  glimpse()

trust4_reports <- data.table::fread(
    glue("{wkdir}/data/interim/airr/2025-07-05_TRUST4-reports-all-10579.csv"), 
    header = TRUE) %>%
    dplyr::select(-V1) %>% 
    glimpse()

trust4_reports %>% glimpse()

# selecting for quality receptors for all data available for key samples of interest 
valid_aa <- "^[ACDEFGHIKLMNPQRSTVWY]+$"
trust4_reports_clean_df <- trust4_reports %>% 
  filter(sample_id %in% metadata$sample_id) %>% 
  filter(CDR3aa != "out_of_frame") %>%
  filter(!grepl("_", CDR3aa)) %>%
  filter(grepl(valid_aa, CDR3aa)) %>%
  filter(C != ".") %>% 
  glimpse()

trust4_reports_tcrb <- trust4_reports_clean_df %>% 
    filter(C == "TRBC") %>% 
    filter(V != "." & J != ".") %>%
    mutate(CDR3aa_length = nchar(CDR3aa)) %>%
    filter(CDR3aa_length >= 6) %>%
    filter(CDR3aa_length <= 26) %>%
    mutate(v_gene = str_before_first(V, "\\*")) %>%
    mutate(j_gene = str_before_first(J, "\\*")) %>% 
    left_join(key_meta_df) %>%
    glimpse()

saveRDS(
  trust4_reports_tcrb,
  file = glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_{Sys.Date()}.rds"),
  compress = FALSE
)

# Note that 30 samples failed to run TRUST4 - and are not included in the analysis 
# 30/9,955 (only 0.3% of samples)
metadata$sample_id %>% unique() %>% length() # 10,609 (originally - 9,955 after removing Other category)
trust4_reports$sample_id %>% unique() %>% length() # 10,579
trust4_reports_tcrb$sample_id %>% unique() %>% length() # 9928 

# trust4_reports_tcrb_df$sample_id %>% unique() %>% length()

```


Examining our TCRB data:


- From 10,609 samples - we assemble 27,985,042 TCR/ BCR receptors. 


- After filtering for only PD and Control samples (n = 10,579), annotated constant regions, complete and valid CDR3 sequences
 we get 19,355,603 TCR/ BCR receptors.


- After filtering for TCRB sequences with complete V and J gene annotations and CDR3 lengths between 6 and 26 amino acids - we get 6,679,301 TCRB receptors.


### Downloading CDR3 data

```{r, eval=FALSE}
# Viewing filepath df structure
cdr3_res <- list.files(data_out)
dups <- trust4_file_paths$cdr3$name %>%
  basename() %>%
  duplicated()

trust4_file_paths$cdr3 %>% glimpse
trust4_file_paths$final_assemblies %>% glimpse
trust4_file_paths$final_reports %>% glimpse
trust4_file_paths$cdr3$name %>% keep(grepl("attempt-", .))

# file size in bytes
trust4_file_paths$cdr3$size %>%
  strex::str_after_first(" ") %>%
  unique()
trust4_file_paths$final_assemblies$size %>%
  strex::str_after_last(" ") %>%
  unique()
trust4_file_paths$final_reports$size %>%
  strex::str_after_first(" ") %>%
  unique()
```

```{r, eval=FALSE}
# note may want to double check that the right attempt file was downloaded -- 
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_download_TRUST4_CDR3")
data_out <- glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/TRUST4_cdr3")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)

# filtering multiple attempts for largest output size
cdr3_gc_stats <- trust4_file_paths$cdr3 %>%
  mutate(size_numeric = case_when(
    grepl(" Kb", size) ~ gsub(" Kb", "", size) %>% as.numeric() * 1000,
    grepl(" Mb", size) ~ gsub(" Mb", "", size) %>% as.numeric() * 1000000
  )) %>%
  mutate(filename = basename(name)) %>%
    group_by(filename) %>%
    slice_max(order_by = size_numeric, n = 1, with_ties = FALSE) %>%
    glimpse

for (url in cdr3_gc_stats$name) {
  if (file.exists(file.path(data_out, basename(url)))) {
    next
  }
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu=200MB",
      " --time=10:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {t4_bucket}/{url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.02)
}

```

### Downloading Full Length Assemblies

```{r, eval=FALSE}
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_download_TRUST4_final_assemblies")
data_out <- glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/TRUST4_final_assemblies")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)

final_assemblies_gc_stats <- trust4_file_paths$final_assemblies %>%
  mutate(size_numeric = case_when(
    grepl(" Mb", size) ~ gsub(" Mb", "", size) %>% as.numeric() * 1000000
  )) %>%
  mutate(filename = basename(name)) %>%
    group_by(filename) %>%
    slice_max(order_by = size_numeric, n = 1, with_ties = FALSE) %>%
    glimpse

for (url in final_assemblies_gc_stats$name) {
  if (file.exists(file.path(data_out, basename(url)))) {
    next
  }
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu=500MB",
      " --time=10:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {t4_bucket}/{url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.01)
}

```


### Downloading HBS TRUST4 Data (Reports, CDR3, Final Assemblies)

```{r}
# slurm_out <- glue("{wkdir}/.cluster_runs/gcs_download_TRUST4_final_assemblies")
# data_out <- glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/TRUST4_final_assemblies")
# dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
# dir.create(data_out, recursive = TRUE, showWarnings = FALSE)

# hbs_t4_buckets <- paste0("GCS_TRUST4_HBS", 1:3)
hbs_t4_buckets <- paste0("GCS_TRUST4_HBS", 4:5)

hbs_t4_buckets_paths <- hbs_t4_buckets %>% 
  purrr::set_names() %>% 
  purrr::map(
    ~readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2025-07-03_URLs_TRUST4-HBS_{.}.rds")
      # glue("{wkdir}/data/interim/gcs_locations/2025-06-29_URLs_TRUST4-HBS_{.}.rds")
    ) %>% 
      data.frame(path = .)
  ) %>% 
  bind_rows(.id = "bucket") %>% 
  mutate(sample_id = basename(path) %>% str_before_last("_")) %>% 
  mutate(filetype = path %>% str_after_last("_")) %>%
  mutate(bucket_n = str_sub(bucket, -1) %>% as.numeric()) %>% 
  mutate(attempt_n = case_when(
    grepl("attempt-", path) ~ str_extract(path, "attempt-\\d+") %>% 
      gsub("attempt-", "", .) %>% as.numeric(),
    TRUE ~ 1
  )) %>% 
  glimpse()

# Filter runs to only include the latest bucket and highest attempt
# should have 1869, 278, 1

hbs_t4_buckets_paths %>% glimpse()
batch_params_temp <- hbs_t4_buckets_paths %>%
  # select sample run in the latest bucket
  group_by(sample_id, filetype) %>%
  slice_max(order_by = bucket_n, n = 1, with_ties = FALSE) %>% 
  # select highest attempt within a bucket
  slice_max(order_by = attempt_n, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  glimpse()

batch_params_temp %>% group_by(filetype, bucket) %>%
  dplyr::summarise(n = n())

# Assembling batch params
hbs_airr_data_dir <- glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/TRUST4_HBS")
batch_params <- batch_params_temp %>% 
  mutate(wkdir = wkdir, 
         data_out = case_when(
            filetype == "report.tsv" ~ glue("{hbs_airr_data_dir}/reports"),
            filetype == "final.out" ~ glue("{hbs_airr_data_dir}/final_assemblies"),
            filetype == "cdr3.out" ~ glue("{hbs_airr_data_dir}/cdr3")
         )) %>% 
        dplyr::select(url = path, data_out, wkdir) %>% 
  glimpse()


# function to download data from GCS using gsutil
download_trust4_shell_do <- function(url, data_out, wkdir) {
  require(glue)
  source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))
  shell_do(
    glue(
      "{wkdir}/notebooks/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
}

# configure registry ----
breg <- breg_setup(cluster_name = "TRUST4_HBS_GCS_DOWNLOAD", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )

# Submit Jobs ----
jobs <- batchMap(
  fun = download_trust4_shell_do,
  args = batch_params,
  reg = breg
)

jobs[, chunk := chunk(job.id, chunk.size = 10)]
print(jobs[, .N, by = chunk])

submitJobs(jobs,
  resources = list(
    walltime = "40:00",
    memory = "2GB",
    ncpus = 1,
    max.concurrent.jobs = 9999
  )
)

```

### Downloading Flagstat Data

```{r}
tic()
arcasHLA_urls <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_HLA4')}/arcasHLAworkflow/*/call-arcasHLAanalysis/*.json"),
  stdout_path = TRUE
)
toc()

flagstat_cram_urls <- flagstat_urls %>%
  discard(grepl("unmapped-f4_flagstat.tsv", .))
flagstat_bam_urls <- flagstat_urls %>%
  keep(grepl("unmapped-f4_flagstat.tsv", .))

```

### Downloading ArcasHLA data

```{r}
hla_buckets <- paste0("GCS_HLA", 5:6)

for (bucket in hla_buckets) {
  message(glue("Downloading from {bucket}"))
  arcasHLA_urls <- 
    shell_do(
      glue("gsutil ls {Sys.getenv(bucket)}/arcasHLAworkflow/*/call-arcasHLAanalysis/*.json"),
      stdout_path = TRUE
    )
  
  saveRDS(
    arcasHLA_urls,
    glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_arcasHLA-jsons_{bucket}.rds")
  )
}

arcasHLA_urls_GCS_HLA4 <- 
  shell_do(
    glue("gsutil ls {Sys.getenv('GCS_HLA4')}/arcasHLAworkflow/*/call-arcasHLAanalysis/cacheCopy/*.json"),
  stdout_path = TRUE
  )

saveRDS(
  arcasHLA_urls_GCS_HLA4,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_arcasHLA-jsons_GCS_HLA4.rds")
)

# For HBS cohort
hla_buckets <- paste0("GCS_HLA_HBS", 1:3)
for (bucket in hla_buckets) {
  message(glue("Downloading from {bucket}"))
  arcasHLA_urls <- 
    shell_do(
      glue("gsutil ls {Sys.getenv(bucket)}/arcasHLAworkflow/*/call-arcasHLAanalysis/*.json"),
      stdout_path = TRUE
    )
  
  saveRDS(
    arcasHLA_urls,
    glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_arcasHLA-jsons_{bucket}.rds")
  )
}

```


```{r}
hla_buckets <- paste0("GCS_HLA", 1:3)
hla_urls_list <- hla_buckets %>% 
  purrr::set_names() %>% 
  purrr::map(
    ~readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2024-12-15_URLs_arcasHLA-jsons_{.}.rds")
    )
  )

hla_urls_list[["GCS_HLA4"]] <- readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2024-12-13_URLs_arcasHLA-jsons_GCS_HLA4.rds")
    )
hla_urls_list[["GCS_HLA5"]] <- readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2024-12-18_URLs_arcasHLA-jsons_GCS_HLA5.rds")
    )
hla_urls_list[["GCS_HLA6"]] <- readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2024-12-18_URLs_arcasHLA-jsons_GCS_HLA6.rds")
    )

hla_urls_list %>% purrr::map_int(~length(.) / 2)
# bucket 3 has 951 when it should have 932 ?

```


For HBS cohort - replace chunk above with this chunk

```{r}
hla_buckets <- paste0("GCS_HLA_HBS", 1:3)

hla_urls_list <- hla_buckets %>% 
  purrr::set_names() %>% 
  purrr::map(
    ~readRDS(
      glue("{wkdir}/data/interim/gcs_locations/2025-06-27_URLs_arcasHLA-jsons_{.}.rds")
    )
  )

hla_urls_list %>% purrr::map_int(~length(.) / 2)
# GCS_HLA_HBS1 GCS_HLA_HBS2 GCS_HLA_HBS3 
#          611          159            1 
# Some duplicates in GCS_HLA_HBS1 since there should be 589 successful runs

```


```{r}
hla_urls_df <- hla_urls_list %>% 
  purrr::map(
    ~as_tibble(.) %>% 
      dplyr::rename(path = value) %>% 
      mutate(
        filename = basename(path),
        sample_id = str_extract(filename, "^[^.]+"),
        file_type = case_when(
          str_detect(filename, "\\.genotype\\.json$") ~ "genotype",
          str_detect(filename, "\\.genes\\.json$") ~ "genes",
          TRUE ~ NA_character_
        )
      )
  ) %>% 
  bind_rows(.id = "bucket") %>% 
  glimpse()

hla_urls_df %>% 
  janitor::get_dupes(sample_id) %>% 
  filter(dupe_count > 2)

hla_urls_df %>% 
  janitor::get_dupes(sample_id, bucket) %>% 
  filter(dupe_count > 2)

# duplicates are only present across buckets but not within buckets
# 16 samples are duplicated across buckets -- 

# Here we select samples run in the latest bucket run (re-runs)
hla_urls_df <- hla_urls_df %>% 
  mutate(bucket_num = str_extract(bucket, "\\d+")) %>% 
  group_by(sample_id, file_type) %>% 
  slice_max(order_by = bucket_num, n = 1, with_ties = FALSE) %>% 
  glimpse()

hla_urls_df %>% 
  group_by(bucket, file_type) %>% 
  dplyr::summarise(n = n())

# This is correct now
# A tibble: 6 × 3
# # Groups:   bucket [3]
#   bucket       file_type     n
#   <chr>        <chr>     <int>
# 1 GCS_HLA_HBS1 genes       589
# 2 GCS_HLA_HBS1 genotype    589
# 3 GCS_HLA_HBS2 genes       159
# 4 GCS_HLA_HBS2 genotype    159
# 5 GCS_HLA_HBS3 genes         1
# 6 GCS_HLA_HBS3 genotype      1

```


```{r}
slurm_out <- glue("{wkdir}/.cluster_runs/{get_time()}_gcs_download_arcasHLA")
data_out <- glue("{pdmbs_dir}/workflow/RNASEQ/AIRR/arcasHLA")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

gcs_paths_to_download <- hla_urls_df$path %>% 
  discard( ~ file.exists(file.path(data_out, basename(.))))

for (url in gcs_paths_to_download) {
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu=1GB",
      " --time=10:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  # Sys.sleep(0.1)
}

```

