title: "Quantifying Generation Probability and Selection Factors of CDR3β Sequences using IGoR/OLGA/SoNNia"
editor: source
author: "Joe Boktor"
date: '2025-05-04'
eval: false
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
    fig-height: 8
    fig-width: 14
---

This notebook performs OLGA and SoNNia analyses to quantify the probability of CDR3β generation and produce $P_{gen}$ and $P_{post}$ statistics for each receptor. Then we perform distribution testing across samples as a pairwise distance metric.

## Analysis Setup

Environment setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(data.table)
library(fastICA)
library(parallelDist)
library(mclust)
library(seqinr)
library(progress)
library(transport)
library(immunarch)
library(nlme)
library(broom)
library(broom.mixed)
library(philentropy)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))

scratch_dir <- "/resnick/scratch/jbok/PDMBS"
alice_res_dir <- glue("{wkdir}/data/interim/ALICE/olga_parallel")
alicegit <- "/resnick/groups/MazmanianLab/jboktor/git/ALICE"
source(glue("{alicegit}/ALICE.R"))

```


Reading in metadata

```{r}
full_meta_df <- readRDS(
    glue("{wkdir}/data/interim/airr/full_meta_df_2025-06-07.rds")) %>% 
    glimpse()

mds_tests <- full_meta_df %>%
    dplyr::select(matches("mds_updrs.*_summary_score")) %>%
    colnames()

key_meta_df <- full_meta_df %>%
    dplyr::select(
        sample_id, case_control_other_latest,
        ethnicity, race,
        visit_month, age_at_baseline, sex, study, participant_id,
        raw_total_reads, raw_total_reads_binned, DRB1_grouped, DRB1_risk,
        all_of(mds_tests)
    ) %>%
    glimpse()

```


Analysis was run using R script and slurm (R-script needs to be executed in ALICE Git repo)

Reading in test results

```{r}
alice_res <- readRDS(
    glue("{wkdir}/data/interim/ALICE/tcrb_iter-100_nrec-1e6_2025-06-07.rds")
)

sapply(alice_res, nrow)
```


Results - pretty underwhelming. Probably don't have enough matched V/J seqs within a samples for the network analysis to pick up signatures.


     BF_1001_SVM0_5T1      BF_1029_SVM0_5T1      BF_1135_SVM0_5T1 
                    4                     4                     4 
 PD_PDBP166AHF_BLM0T1  PD_PDBP166AHF_SVM6T1  PD_PDBZ725PY8_BLM0T1 
                    6                     5                     5 
 PD_PDDB923WW2_BLM0T1 PD_PDDB923WW2_SVM12T1 PD_PDDB923WW2_SVM18T1 
                    6                     7                     5 
 PD_PDFU123GRM_BLM0T1  PD_PDKH605DLQ_SVM6T1  PD_PDKU416WUZ_BLM0T1 
                    4                     5                     2 
 PD_PDLH751UML_BLM0T1  PD_PDMZ472HZ5_BLM0T1 PD_PDMZ472HZ5_SVM12T1 
                    6                     2                     5 
PD_PDMZ472HZ5_SVM18T1 PD_PDMZ472HZ5_SVM24T1  PD_PDMZ472HZ5_SVM6T1 
                    4                     4                     6 
 PD_PDPK734UZM_BLM0T1 PD_PDPX797VDT_SVM18T1 PD_PDRM716VPC_SVM18T1 
                    2                     4                     6 
 PD_PDXD535AE5_BLM0T1 PD_PDXD535AE5_SVM12T1  PD_PDXZ620KYP_BLM0T1 
                    4                     5                     4 
 PD_PDZJ953DRP_BLM0T1       PP_15761_SVM6T1       PP_3086_SVM36T1 
                    4                     5                     1 
      PP_3112_SVM12T1        PP_3112_SVM6T1       PP_3173_SVM24T1 
                    1                     4                     5 
       PP_3180_SVM6T1       PP_3355_SVM12T1       PP_3380_SVM36T1 
                    2                     1                     6 
      PP_3432_SVM24T1        PP_3475_SVM6T1        PP_3481_SVM6T1 
                    4                     4                     4 
      PP_3521_SVM36T1       PP_3552_SVM24T1        PP_3564_BLM0T1 
                    2                     4                     2 
      PP_3564_SVM24T1       PP_3603_SVM12T1       PP_3603_SVM24T1 
                    2                     7                     6 
      PP_3603_SVM36T1        PP_3603_SVM6T1       PP_3613_SVM36T1 
                    6                     6                     2 
       PP_3635_BLM0T1       PP_3635_SVM24T1       PP_3759_SVM36T1 
                    1                     4                     4 
       PP_3787_BLM0T1       PP_3787_SVM12T1       PP_40004_BLM0T1 
                    1                     7                     4 
     PP_40586_SVM12T1      PP_40586_SVM24T1      PP_40754_SVM12T1 
                    5                     5                     1 
     PP_40754_SVM24T1       PP_40769_BLM0T1       PP_40769_SVM6T1 
                    5                     1                     4 
      PP_41281_BLM0T1       PP_41281_SVM6T1       PP_41285_BLM0T1 
                    4                     1                     4 
      PP_41296_BLM0T1       PP_41296_SVM6T1       PP_41564_BLM0T1 
                    4                     7                     7 
      PP_42126_BLM0T1       PP_42131_BLM0T1       PP_51594_BLM0T1 
                    4                     8                     2 
      PP_52644_BLM0T1       PP_54072_BLM0T1 
                    4                     4 

