---
title: "Adaptive Immune Receptor Repertoire in Parkinson's Disease: Optimal Transport and Probabilistic Modeling Approaches for Latent Space Embeddings"
editor: source
author: "Joe Boktor"
date: '2023-12-07'
bibliography: references.bib
format: 
  html:
    font-family: helvetica neue
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 4
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
    grid:
      sidebar-width: 300px
      body-width: 900px
      margin-width: 300px
      gutter-width: 1.5rem
---

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(data.table)
library(fastICA)
library(parallelDist)
library(mclust)
library(seqinr)
library(progress)
library(transport)
library(immunarch)
library(nlme)
library(broom.mixed)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(scico)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))
source(paste0(src_dir, "/R_scripts/_airr_analyze-permanova.R"))
scratch_dir <- "/resnick/scratch/jbok/PDMBS"
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
imm_plots_dir <- glue("{wkdir}/figures/AIRR/immunarch")
hla_dir <- glue("{wkdir}/data/interim/airr/arcasHLA/analysis_cleaned")
hla_pca_dir <- glue("{hla_dir}/pca_data")
deconv_outdir <- glue("{wkdir}/data/interim/celltype_deconvolution")

```


# Exploratory Analysis
Reading in metadata

```{r}
full_meta_df <- readRDS(
    glue("{wkdir}/data/interim/airr/full_meta_df_2025-06-09.rds")) %>% 
    glimpse()

mds_tests <- full_meta_df %>%
    dplyr::select(matches("mds_updrs.*_summary_score")) %>%
    colnames()

essential_meta <- c(
    "study", "raw_total_reads", "raw_total_reads_binned",
    "case_control_other_latest",
    "DRB1_grouped", "DRB1_risk",
    "age_at_baseline", "sex", mds_tests
)

key_meta_df <- full_meta_df %>% 
    dplyr::select(
        sample_id, case_control_other_latest,
        ethnicity, race,
        visit_month, age_at_baseline, sex, study, participant_id,
        raw_total_reads, raw_total_reads_binned, DRB1_grouped, DRB1_risk,
        all_of(mds_tests),
        contains("x_cell"), contains("epic"), contains("abis")
    ) %>% 
    glimpse()

```


Is there any difference in sequencing depth between disease groups / study cohorts?

```{r}
lmm_model_count <- lme(raw_total_reads ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = full_meta_df %>% filter(case_control_other_latest != "Other"),
    na.action = na.omit
)

summary(lmm_model_count)
broom.mixed::tidy(lmm_model_count) %>% 
    DT::datatable()

# Yes it does look there are sequencing depth differences between disease groups / study cohorts that need to be accounted for.
lmm_model_tcellreads_xcell <- nlme::lme(
    tcell_reads_x_cell ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = full_meta_df %>% filter(case_control_other_latest != "Other"),
    na.action = na.omit
)
lmm_model_tcellreads_epic <- nlme::lme(
    tcell_reads_epic ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = full_meta_df %>% filter(case_control_other_latest != "Other"),
    na.action = na.omit
)
lmm_model_tcellreads_abis <- nlme::lme(
    tcell_reads_abis ~ case_control_other_latest + study,
    random = ~ 1 | participant_id,
    data = full_meta_df %>% filter(case_control_other_latest != "Other"),
    na.action = na.omit
)

summary(lmm_model_tcellreads_xcell)
summary(lmm_model_tcellreads_epic)
summary(lmm_model_tcellreads_abis)

# It also looks like there are differences in T cell reads between disease groups / study cohorts that need to be accounted for.
p_seq_depth_dist <- full_meta_df %>% 
    filter(case_control_other_latest != "Other") %>% 
    ggplot(aes(y = raw_total_reads, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "Bulk RNA-Seq reads", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_dist_{Sys.Date()}.png"),
    plot = p_seq_depth_dist,
    width = 5,
    height = 4
)

p_seq_depth_tcell_xcell <- full_meta_df %>% 
    filter(case_control_other_latest != "Other") %>% 
    ggplot(aes(y = tcell_reads_x_cell, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "Estimated T-cell reads (xCell)", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_tcell_xcell_{Sys.Date()}.png"),
    plot = p_seq_depth_tcell_xcell,
    width = 5,
    height = 4
)

p_seq_depth_tcell_abis <- full_meta_df %>% 
    filter(case_control_other_latest != "Other") %>% 
    ggplot(aes(y = tcell_reads_abis, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "Estimated T-cell reads (ABIS)", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_tcell_abis_{Sys.Date()}.png"),
    plot = p_seq_depth_tcell_abis,
    width = 5,
    height = 4
)

p_seq_depth_tcell_epic <- full_meta_df %>% 
    filter(case_control_other_latest != "Other") %>% 
    ggplot(aes(y = tcell_reads_epic, x = case_control_other_latest)) +
    geom_point(size = 0.2, alpha = 0.75, 
        position = position_jitter(width = 0.4)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.75, width = 0.5) +
    facet_wrap(~study) +
    theme_bw() +
    scale_y_log10() +
    labs(y = "Estimated T-cell reads (EPIC)", x = NULL, color = "Quantile") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(legend.position = "top")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/seq_depth_tcell_epic_{Sys.Date()}.png"),
    plot = p_seq_depth_tcell_epic,
    width = 5,
    height = 4
)

```


## Sample Summary Plots

Visualzing demographic and clinical data for samples.

```{r}
# plot the distribution of total samples per participant
p_sample_counts <- full_meta_df %>% 
    group_by(participant_id) %>% 
    dplyr::summarise(n = n()) %>% 
    left_join(key_meta_df, by = c("participant_id")) %>% 
    ggplot(aes(x = n, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw() +
    labs(x = "Samples per participant", 
         y = "Count",
         fill = "Disease Status")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/sample_counts_{Sys.Date()}.png"),
    plot = p_sample_counts,
    width = 4,
    height = 3.5
)

baseline_meta <- key_meta_df %>% 
    select(participant_id, age_at_baseline, case_control_other_latest, sex, study) %>% 
    distinct()

p_age_density <- baseline_meta %>% 
    ggplot(aes(x = age_at_baseline, color = case_control_other_latest)) +
    geom_density(alpha = 0.5, size = 2) +
    scale_color_d3() +
    theme_bw() +
    labs(x = "Age at Baseline",
         y = "Density",
         fill = "Disease Status") +
    theme(legend.position = "none")

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/age_density_{Sys.Date()}.png"),
    plot = p_age_density,
    width = 4,
    height = 3.5
)

baseline_meta %>%
    # Plot sex distribution
    ggplot(aes(x = sex, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw() +
    labs(x = "Sex", 
         y = "Count",
         fill = "Disease Status") -> p_sex_dist

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/sex_dist_{Sys.Date()}.png"),
    plot = p_sex_dist,
    width = 3.5,
    height = 3.5
)

# Plot study distribution
baseline_meta %>%
    ggplot(aes(x = study, fill = case_control_other_latest)) +
    geom_bar(position = "stack", width = 0.6) +
    scale_fill_d3() +
    theme_bw() +
    labs(x = "Study", 
         y = "Count",
         fill = "Disease Status") -> p_study_dist

ggsave(
    filename = glue("{wkdir}/figures/AIRR/sample_summary/study_dist_{Sys.Date()}.png"),
    plot = p_study_dist,
    width = 4,
    height = 3.5
)

```


Visualizing samples x timepoints included in the TRUST4 analysis.

```{r}
time_point_meta <- full_meta_df %>% 
    select(participant_id, sample_id, visit_month, study, case_control_other_latest) %>% 
    distinct()

p_repertoire_samples <- time_point_meta %>% 
    ggplot(aes(x = visit_month,
        y = fct_reorder2(participant_id, visit_month, case_control_other_latest)
    )) +
    geom_line(aes(color = case_control_other_latest, group = participant_id)) +
    geom_point() +
    facet_grid(study ~ ., scales = "free_y", space = "free") +
    scale_color_d3() +
    theme_bw() +
    labs(x = "Visit Month", y = "Participant ID") +
    theme(axis.text.y = element_blank(), legend.position = "bottom")

ggsave(
    glue("{wkdir}/figures/AIRR/",
    "TRUST4-repertoire-samples_{Sys.Date()}.png"),
    p_repertoire_samples,
    width = 9, height = 11
)

```


## TRUST4 Data Processing

Loading TRUST4 data and filtering TCRB receptors for full length, in-frame, annotated V/J genes.

```{r}
trust4_reports <- fread(
    glue("{wkdir}/data/interim/airr/2023-05-04_TRUST4-reports-all.csv"), 
    header = TRUE) %>%
    dplyr::select(-V1) %>% 
    glimpse()

trust4_reports_tcrb <- trust4_reports %>% 
    filter(C == "TRBC") %>% 
    filter(V != "." & J != ".") %>% 
    filter(CDR3aa != "out_of_frame") %>% 
    filter(!grepl("_", CDR3aa)) %>%  
    glimpse()

trust4_reports_tcrb %>% glimpse()
trust4_reports_tcrb$cid_full_length %>% table()

# trust4_reports_tcrb$CDR3aa %>% unique %>% length
# trust4_reports_tcrb$CDR3aa %>% length
```

Resulting dataset contains 5,463,968 TCRB receptors 


Visualizing the distribution of TCR and BCR receptor counts per sample and donor


```{r}
trust4_reports_clean_df <- trust4_reports %>%
    filter(CDR3aa != "out_of_frame") %>%
    filter(!grepl("_", CDR3aa)) %>%
    filter(C != ".") %>%
    left_join(key_meta_df, by = c("sample_id")) %>%
    glimpse()

sid_counts <- trust4_reports_clean_df %>%
    group_by(sample_id, C) %>%
    dplyr::summarise(n = n()) %>%
    mutate(grouping = "sample") %>%
    dplyr::rename(unit = sample_id) %>%
    glimpse()
pid_counts <- trust4_reports_clean_df %>%
    group_by(participant_id, C) %>%
    dplyr::summarise(n = n()) %>%
    mutate(grouping = "donor") %>%
    dplyr::rename(unit = participant_id) %>%
    glimpse()

rec_counts_df <- bind_rows(sid_counts, pid_counts) %>%
    mutate(cell_type_origin = case_when(
        grepl("TR", C) ~ "TCR",
        grepl("IG", C) ~ "BCR",
        TRUE ~ "ERROR"
    )) %>%
    glimpse()

rec_counts_df$C %>% table()

p_ecdf <- rec_counts_df %>%
    mutate(cell_type_origin = factor(cell_type_origin, levels = c("TCR", "BCR"))) %>%
    mutate(grouping = factor(grouping, levels = c("sample", "donor"))) %>%
    ggplot(aes(n)) +
    stat_ecdf(aes(color = C), geom = "step", linewidth = 1, alpha = 0.8) +
    facet_grid(cell_type_origin ~ grouping, scales = "free") +
    scale_x_log10() +
    theme_bw() +
    labs(
        x = "Number of Receptors",
        y = "Cumulative Proportion",
        color = NULL
    ) +
    scale_color_manual(values = c(
        "IGHA1" = "#99d8c9",
        "IGHA2" = "#2ca25f",
        "IGKC" = "#e31a1c",
        "IGLC" = "#ff7f00",
        "IGHD" = "#b15928",
        "IGHE" = "#fed976",
        "IGHG1" = "#bfd3e6",
        "IGHG2" = "#8c96c6",
        "IGHG3" = "#8856a7",
        "IGHG4" = "#810f7c",
        "IGHM" = "#fb9a99",
        "TRAC" = "#a6cee3",
        "TRBC" = "#1f78b4",
        "TRDC" = "#fb9a99",
        "TRGC" = "#e31a1c"
    ))

ggsave(
    filename = glue("{wkdir}/figures/AIRR/ecdf_TRUST4_{Sys.Date()}.png"),
    plot = p_ecdf,
    width = 7,
    height = 4.5
)

p_ecdf_TCRS <- rec_counts_df %>%
    filter(cell_type_origin == "TCR") %>% 
    mutate(grouping = factor(grouping, levels = c("sample", "donor"))) %>%
    ggplot(aes(n)) +
    stat_ecdf(aes(color = C), geom = "step", linewidth = 1, alpha = 0.8) +
    facet_grid(cell_type_origin ~ grouping, scales = "free") +
    scale_x_log10() +
    theme_bw() +
    labs(
        x = "Number of Receptors",
        y = "Cumulative Proportion",
        color = NULL
    ) +
    scale_color_manual(values = c(
        "TRAC" = "#a6cee3",
        "TRBC" = "#1f78b4",
        "TRDC" = "#fb9a99",
        "TRGC" = "#e31a1c"
    ))

ggsave(
    filename = glue("{wkdir}/figures/AIRR/ecdf_TRUST4_TCRS_{Sys.Date()}.png"),
    plot = p_ecdf_TCRS,
    width = 7,
    height = 3
)

```



Visualizing clonotypes and clonal volume

```{r}
trust4_reports_tcrb %>% glimpse()
key_meta_df %>% glimpse()

trust4_reports_tcrb_df <- trust4_reports_tcrb %>%
    left_join(key_meta_df) %>%
    filter(case_control_other_latest != "Other") %>%
    glimpse()

clonotype_counts_df <- trust4_reports_tcrb_df %>%
    distinct() %>%
    group_by(sample_id) %>%
    dplyr::summarise(clonotype_count = n()) %>%
    left_join(key_meta_df) %>%
    glimpse()
clonotype_volume_df <- trust4_reports_tcrb_df %>%
    group_by(sample_id) %>%
    dplyr::summarise(clonotype_volume = sum(`#count`)) %>%
    left_join(key_meta_df) %>%
    glimpse()

# LMM model stats
lmm_model_count <- lme(clonotype_count ~ case_control_other_latest + age_at_baseline + sex + tcell_reads_epic,
    random = ~ 1 | participant_id,
    data = clonotype_counts_df
)
lmm_model_volume <- lme(clonotype_volume ~ case_control_other_latest + age_at_baseline + sex + tcell_reads_epic,
    random = ~ 1 | participant_id,
    data = clonotype_volume_df
)
summary(lmm_model_count)
summary(lmm_model_volume)
count_res_df <- broom.mixed::tidy(lmm_model_count) %>% print()
volume_res_df <- broom.mixed::tidy(lmm_model_volume) %>% print()

count_pval <- count_res_df %>%
    filter(term == "case_control_other_latestCase") %>%
    pull(p.value) %>%
    signif(2)
volume_pval <- volume_res_df %>%
    filter(term == "case_control_other_latestCase") %>%
    pull(p.value) %>%
    signif(2)

p_clonotype_count <- clonotype_counts_df %>%
    ggplot(aes(x = case_control_other_latest, y = clonotype_count)) +
    geom_point(aes(color = case_control_other_latest),
        size = 0.3,
        alpha = 0.2, position = position_jitter(width = 0.3)
    ) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    theme_bw() +
    scale_color_d3() +
    labs(x = NULL, y = "Clonal Count") +
    theme(legend.position = "none")
    # # Add p-value annotation using ggsignif package
    # ggsignif::geom_signif(
    #     comparisons = list(c("Control", "Case")),
    #     annotations = glue("*\np = {count_pval}"),
    #     y_position = max(clonotype_counts_df$clonotype_count) * 0.8,
    #     tip_length = 0,
    #     vjust = 0.6
    # )
p_clonotype_volume <- clonotype_volume_df %>%
    ggplot(aes(x = case_control_other_latest, y = clonotype_volume)) +
    geom_point(aes(color = case_control_other_latest),
        size = 0.3,
        alpha = 0.2, position = position_jitter(width = 0.3)
    ) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    theme_bw() +
    scale_color_d3() +
    labs(x = NULL, y = "Clonal Volume") +
    theme(legend.position = "none")
    # # Add p-value annotation using ggsignif package
    # ggsignif::geom_signif(
    #     comparisons = list(c("Control", "Case")),
    #     annotations = glue("ns\np = {volume_pval}"),
    #     y_position = max(clonotype_volume_df$clonotype_volume) * 0.8,
    #     tip_length = 0,
    #     vjust = 0.6
    # )

p_clonotype_count_volume <- p_clonotype_count + p_clonotype_volume

ggsave(
    filename = glue("{wkdir}/figures/AIRR/clonotype_count_volume_{Sys.Date()}.png"),
    plot = p_clonotype_count_volume,
    width = 3.5,
    height = 3.5
)

# ggsave(
#     filename = glue("{wkdir}/figures/AIRR/clonotype_count_{Sys.Date()}.png"),
#     plot = p_clonotype_count,
#     width = 2,
#     height = 3.5
# )
# ggsave(
#     filename = glue("{wkdir}/figures/AIRR/clonotype_volume_{Sys.Date()}.png"),
#     plot = p_clonotype_volume,
#     width = 2,
#     height = 3.5
# )

clonotype_volume_df %>% glimpse()
clonotype_counts_df %>% glimpse()

p_clonotype_count_raw_reads <- clonotype_counts_df %>%
    ggplot(aes(x = clonotype_count, y = raw_total_reads, color = study)) +
    geom_point(alpha = 0.1, size = 0.4) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_nejm() +
    ggpubr::stat_regline_equation(
        label.x.npc = "left",
        label.y.npc = "top"
    ) +
    ggside::geom_xsidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::geom_ysidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::scale_xsidey_continuous(breaks = NULL) +
    ggside::scale_ysidex_continuous(breaks = NULL) +
    coord_fixed() +
    labs(x = "Clonal Count", y = "Bulk RNAseq Reads") +
    theme(legend.position = "none")

p_clonotype_volume_raw_reads <- clonotype_volume_df %>%
    ggplot(aes(x = clonotype_volume, y = raw_total_reads, color = study)) +
    geom_point(alpha = 0.1, size = 0.4) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_nejm() +
    ggpubr::stat_regline_equation(
        label.x.npc = "left",
        label.y.npc = "top"
    ) +
    ggside::geom_xsidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::geom_ysidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::scale_xsidey_continuous(breaks = NULL) +
    ggside::scale_ysidex_continuous(breaks = NULL) +
    coord_fixed() +
    labs(x = "Clonal Volume", y = NULL)

p_clonotype_volumecount_raw_reads <- p_clonotype_count_raw_reads + p_clonotype_volume_raw_reads

ggsave(
    filename = glue("{wkdir}/figures/AIRR/clonotype_volume_count_raw_reads_{Sys.Date()}.png"),
    plot = p_clonotype_volumecount_raw_reads,
    width = 10,
    height = 3.5
)

p_clonotype_count_tcell_reads <- clonotype_counts_df %>%
    ggplot(aes(x = clonotype_count, y = tcell_reads_abis, color = study)) +
    geom_point(alpha = 0.1, size = 0.4) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_nejm() +
    ggpubr::stat_regline_equation(
        label.x.npc = "left",
        label.y.npc = "top"
    ) +
    ggside::geom_xsidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::geom_ysidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::scale_xsidey_continuous(breaks = NULL) +
    ggside::scale_ysidex_continuous(breaks = NULL) +
    coord_fixed() +
    labs(x = "Clonal Count", y = "Estimated T-Cell Reads (ABIS)") +
    theme(legend.position = "none")

p_clonotype_volume_tcell_reads <- clonotype_volume_df %>%
    ggplot(aes(x = clonotype_volume, y = tcell_reads_abis, color = study)) +
    geom_point(alpha = 0.1, size = 0.4) +
    geom_smooth(method = "lm", se = TRUE) +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    scale_color_nejm() +
    ggpubr::stat_regline_equation(
        label.x.npc = "left",
        label.y.npc = "top"
    ) +
    ggside::geom_xsidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::geom_ysidedensity(alpha = 0.5, show.legend = FALSE) +
    ggside::scale_xsidey_continuous(breaks = NULL) +
    ggside::scale_ysidex_continuous(breaks = NULL) +
    coord_fixed() +
    labs(x = "Clonal Volume", y = NULL)

p_clonotype_volumecount_tcell_reads <- p_clonotype_count_tcell_reads + p_clonotype_volume_tcell_reads


ggsave(
    filename = glue("{wkdir}/figures/AIRR/clonotype_volume_count_tcell_reads_{Sys.Date()}.png"),
    plot = p_clonotype_volumecount_tcell_reads,
    width = 10,
    height = 3.5
)


# Patchwork version
seq_depth_plots <- p_clonotype_volumecount_raw_reads /p_clonotype_volumecount_tcell_reads 
ggsave(
    filename = glue("{wkdir}/figures/AIRR/clonotype_volume_count_raw_reads_&_tcell_reads_{Sys.Date()}.png"),
    plot = seq_depth_plots,
    width = 12,
    height = 8
)

# # aplot version
# p_clonotype_volumecount_tcell_reads_aplot <- p_clonotype_count_tcell_reads  %>%  aplot::insert_right(p_clonotype_volume_tcell_reads)
# p_clonotype_volumecount_raw_reads_aplot <- p_clonotype_count_raw_reads  %>%  aplot::insert_right(p_clonotype_volume_raw_reads)
# seq_depth_plots <- p_clonotype_volumecount_raw_reads_aplot  %>%  aplot::insert_bottom(p_clonotype_volumecount_tcell_reads_aplot)

# ggsave(
#     filename = glue("{wkdir}/figures/AIRR/clonotype_volume_count_raw_reads_&_tcell_reads_{Sys.Date()}.png"),
#     plot = seq_depth_plots,
#     width = 10,
#     height = 8
# )

```



### Formatting TRUST4 data for Immunarch
```{r}
immunarch_formatted <- trust4_reports %>%
  dplyr::rename(
    Clones = `#count`,
    Proportion = frequency,
    CDR3.nt = CDR3nt,
    CDR3.aa = CDR3aa,
    V.name = V,
    D.name = D,
    J.name = J
  ) %>%
  dplyr::mutate(
    VJ.ins = NA_character_,
    VD.ins = NA_character_,
    DJ.ins = NA_character_,
    V.end = NA_integer_,
    D.start = NA_integer_,
    D.end = NA_integer_,
    J.start = NA_integer_,
    Sequence = NA_character_
  ) %>%
  glimpse()

# Split data into immunarch format and metadata
immunarch_data <- immunarch_formatted %>%
  dplyr::select(Clones, Proportion, CDR3.nt, CDR3.aa, V.name, D.name, J.name,
                VJ.ins, VD.ins, DJ.ins, Sequence)

metadata <- immunarch_formatted %>%
  dplyr::select(C, cid, cid_full_length, sample_id) %>% 
  glimpse()

# Adding in sample metadata

rna_metadata <- readRDS(
    glue("{wkdir}/data/interim/metadata/2023-07-14_phyloseq-metadata.rds")
    )$RNASEQ %>%
    glimpse()

metadata %<>% 
    left_join(rna_metadata, by = c("sample_id")) %>%
    glimpse()

# Create list structure expected by immunarch
immdata <- list(
  data = split(immunarch_data, immunarch_formatted$sample_id),
  meta = metadata
)

saveRDS(
    immdata,
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_{Sys.Date()}.rds")
)

```


### 

```{r}
immdata <- readRDS(
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_2024-11-21.rds")
)

#' Computes basic statistics, 
#' including clones/distributions of lengths and counts.
#' https://immunarch.com/articles/web_only/v3_basic_analysis.html
clonal_metrics = c("volume", "count", "len", "clones")
for (m in clonal_metrics) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repExplore(immdata$data, .method = m, .col = c("aa")) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "rep-explore-{m}_{Sys.Date()}.rds"
            )
    )
}

# Exploring Clonality
# https://immunarch.com/reference/repClonality.html
clone_mets <- c("clonal.prop", "homeo", "top", "rare")
for (m in clone_mets) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repClonality(immdata$data, .method = m) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "rep-clonality-{m}_{Sys.Date()}.rds"
            )
    )
}

# Gene Usage Analysis
# https://immunarch.com/reference/geneUsage.html
gene_metrics <- c("hs.trbv", "HomoSapiens.TRBJ", "macmul.IGHV")
for (g in gene_metrics) {
    message(glue("Calculating {g} gene usage metrics"))
    metric_data <- geneUsage(immdata$data, g) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "gene-usage-{g}_{Sys.Date()}.rds"
            )
    )
}

# Analyzing Repertoire Diversity
# https://immunarch.com/articles/web_only/v6_diversity.html
div_metrics <- c(
    "chao1", "hill", "div", "gini.simp", "inv.simp", 
    "gini", "raref", "d50", 
    "dxx"
)
for (d in div_metrics) {
    message(glue("Calculating {d} diversity metrics"))
    metric_data <- repDiversity(immdata$data, .method = d) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "rep-diversity-{d}_{Sys.Date()}.rds"
            )
    )
}


# TAKES FOREVER !
# Considering Repertoire Overlap (Metric of Similarity)
# https://immunarch.com/reference/repOverlap.html
overlap_methods = c(
  "jaccard"
    # "cosine", "morisita", "inc+public", "inc+morisita",
    # "public", "overlap", "jaccard", "tversky"
)
for (m in overlap_methods) {
    message(glue("Calculating {m} overlap metrics"))
    metric_data <- repOverlap(immdata$data,
        .method = m,
        .downsample = TRUE,
        .verbose = TRUE
    ) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "rep-overlap-{m}_{Sys.Date()}.rds"
            )
    )
}

# # TAKES FOREVER !
# # Considering Repertoire Overlap (Metric of Similarity)
# # https://immunarch.com/reference/repOverlap.html
# overlap_methods = c(
#   "jaccard"
#     # "cosine", "morisita", "inc+public", "inc+morisita",
#     # "public", "overlap", "jaccard", "tversky"
# )
# for (m in overlap_methods) {
#     message(glue("Calculating {m} overlap metrics"))
#     metric_data <- repOverlap(immdata$data,
#         .method = m,
#         .downsample = TRUE,
#         .verbose = TRUE
#     ) %>% 
#         glimpse()
#     saveRDS(
#         metric_data,
#         glue("{wkdir}/data/interim/airr/immunarch/metrics/",
#             "rep-overlap-{m}_{Sys.Date()}.rds"
#             )
#     )
# }

```

### Calculating overlap metrics in parallel

```{r}
# Function to calculate overlap for a single method
calculate_overlap <- function(method, immdata_path, wkdir) {
    require(immunarch)
    require(glue)

    # Load the immdata object
    immdata <- readRDS(immdata_path)

    message(glue("Calculating {method} overlap metrics"))
    metric_data <- repOverlap(
        immdata$data,
        .method = method,
        .downsample = TRUE,
        .verbose = TRUE
    )
    saveRDS(
        metric_data,
        glue("{wkdir}/data/interim/airr/immunarch/metrics/",
            "rep-overlap-{method}_{Sys.Date()}.rds"
        )
    )
}

# Create parameter list for each method
overlap_methods = c(
    "cosine", "morisita"
#   "jaccard", 
    # "public", "overlap", "jaccard", "tversky"
)

# Setup batchtools registry
cluster_run <- glue("{get_time()}_Repertoire_Overlap")
message("\n\nRUNNING: ", cluster_run, "\n")
breg <- makeRegistry(
    file.dir = glue(
        "{wkdir}/.cluster_runs/",
        cluster_run
    ),
    seed = 42
)
breg$cluster.functions <- batchtools::makeClusterFunctionsSlurm(
    template = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
    scheduler.latency = 0.1,
    fs.latency = 1
)

# Create parameter list for each method
immdata_path <- glue("{wkdir}/data/interim/airr/immunarch/",
    "immunarch-obj_2024-11-21.rds")

batchtools_params <- data.frame(method = overlap_methods) %>% 
    mutate(
        immdata_path = immdata_path,
        wkdir = wkdir
    )

# Submit jobs
jobs <- batchMap(
    fun = calculate_overlap,
    args = batchtools_params,
    reg = breg
)

submitJobs(jobs,
    resources = list(
        walltime = "7-00:00:00",
        memory = "300GB",
        ncpus = 8,
        max.concurrent.jobs = 9999
    )
)

```



### Formatting TRUST4 TCRBdata for Immunarch



```{r}
immunarch_formatted <- trust4_reports_tcrb %>%
  dplyr::rename(
    Clones = `#count`,
    Proportion = frequency,
    CDR3.nt = CDR3nt,
    CDR3.aa = CDR3aa,
    V.name = V,
    D.name = D,
    J.name = J
  ) %>%
  dplyr::mutate(
    VJ.ins = NA_character_,
    VD.ins = NA_character_,
    DJ.ins = NA_character_,
    V.end = NA_integer_,
    D.start = NA_integer_,
    D.end = NA_integer_,
    J.start = NA_integer_,
    Sequence = NA_character_
  ) %>%
  glimpse()

# Split data into immunarch format and metadata
immunarch_data <- immunarch_formatted %>%
  dplyr::select(Clones, Proportion, CDR3.nt, CDR3.aa, V.name, D.name, J.name,
                VJ.ins, VD.ins, DJ.ins, Sequence)

metadata <- immunarch_formatted %>%
  dplyr::select(C, cid, cid_full_length, sample_id) %>% 
  glimpse()

# Adding in sample metadata
rna_metadata <- readRDS(
    glue("{wkdir}/data/interim/metadata/2023-07-14_phyloseq-metadata.rds")
    )$RNASEQ %>%
    glimpse()

metadata %<>% 
    left_join(rna_metadata, by = c("sample_id")) %>%
    glimpse()

# Create list structure expected by immunarch
immdata <- list(
  data = split(immunarch_data, immunarch_formatted$sample_id),
  meta = metadata
)

saveRDS(
    immdata,
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_TCRB_{Sys.Date()}.rds")
)

```


### 

```{r}
immdata <- readRDS(
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_TCRB_2024-12-18.rds")
)
immdata %>% glimpse()

#' Computes basic statistics, 
#' including clones/distributions of lengths and counts.
#' https://immunarch.com/articles/web_only/v3_basic_analysis.html
clonal_metrics = c("volume", "count", "len", "clones")
for (m in clonal_metrics) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repExplore(immdata$data, .method = m, .col = c("aa")) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-explore-{m}_{Sys.Date()}.rds")
    )
}

# Exploring Clonality
# https://immunarch.com/reference/repClonality.html
clone_mets <- c("clonal.prop", "homeo", "top", "rare")
for (m in clone_mets) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repClonality(immdata$data, .method = m) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-clonality-{m}_{Sys.Date()}.rds"
            )
    )
}

# Gene Usage Analysis
# https://immunarch.com/reference/geneUsage.html
gene_metrics <- c("hs.trbv", "HomoSapiens.TRBJ")
for (g in gene_metrics) {
    message(glue("Calculating {g} gene usage metrics"))
    metric_data <- geneUsage(immdata$data, g) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/gene-usage-{g}_{Sys.Date()}.rds"
            )
    )
}

# Analyzing Repertoire Diversity
# https://immunarch.com/articles/web_only/v6_diversity.html
div_metrics <- c(
    # "chao1", "hill", 
    "div", "gini.simp", "inv.simp", 
    "gini", "raref", "d50", 
    "dxx"
)
for (d in div_metrics) {
    message(glue("Calculating {d} diversity metrics"))
    metric_data <- repDiversity(immdata$data, .method = d) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-diversity-{d}_{Sys.Date()}.rds"
            )
    )
}

```

### Calculating overlap metrics in parallel

```{r}
# Function to calculate overlap for a single method
calculate_overlap <- function(method, immdata_path, wkdir, metrics_dir) {
    require(immunarch)
    require(glue)

    # Load the immdata object
    immdata <- readRDS(immdata_path)

    message(glue("Calculating {method} overlap metrics"))
    metric_data <- repOverlap(
        immdata$data,
        .method = method,
        .downsample = TRUE,
        .verbose = TRUE
    )
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-overlap-{method}_{Sys.Date()}.rds"
        )
    )
}

# Create parameter list for each method
overlap_methods = c(
#   "jaccard", "public", "overlap", "tversky", 
    "cosine", "morisita"
)

# Setup batchtools registry
cluster_run <- glue("{get_time()}_Repertoire_Overlap")
message("\n\nRUNNING: ", cluster_run, "\n")
breg <- makeRegistry(
    file.dir = glue(
        "{wkdir}/.cluster_runs/",
        cluster_run
    ),
    seed = 42
)
breg$cluster.functions <- batchtools::makeClusterFunctionsSlurm(
    template = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
    scheduler.latency = 0.1,
    fs.latency = 1
)

# Create parameter list for each method
immdata_path <- glue(
    "{wkdir}/data/interim/airr/immunarch/",
    "immunarch-obj_TCRB_2024-12-18.rds")

batchtools_params <- data.frame(method = overlap_methods) %>% 
    mutate(
        immdata_path = immdata_path,
        wkdir = wkdir,
        metrics_dir = metrics_dir
    )

# Submit jobs
jobs <- batchMap(
    fun = calculate_overlap,
    args = batchtools_params,
    reg = breg
)

submitJobs(jobs,
    resources = list(
        walltime = "12:00:00",
        memory = "300GB",
        ncpus = 8,
        max.concurrent.jobs = 9999
    )
)

```


# Diversity Analysis of Beta Chain Repertoires

V-gene usage analysis

```{r}
# Visualizing v-gene usage
v_gene_usage <- readRDS(
    glue("{metrics_dir}/gene-usage-hs.trbv_2024-12-19.rds")
)
v_gene_usage_long <- v_gene_usage %>%
    pivot_longer(-Names, names_to = "sample_id", values_to = "counts") %>%
    mutate(counts = ifelse(is.na(counts), 0, counts)) %>%
    left_join(key_meta_df, by = c("sample_id")) %>%
    filter(case_control_other_latest != "Other") %>%
    group_by(sample_id) %>%
    mutate(vgenefreq = counts / sum(counts) * 100) %>%
    glimpse()

p_v_gene_usage <- v_gene_usage_long %>%
    ggplot(aes(
        x = fct_reorder(Names, vgenefreq, .desc = TRUE),
        y = vgenefreq
    )) +
    geom_point(aes(color = case_control_other_latest),
        position = position_jitterdodge(), alpha = 0.1, size = 0.05) +
    geom_boxplot(
        aes(group = interaction(Names, case_control_other_latest)),
        outlier.alpha = 0, # fill.alpha = 0.2,
        position = position_dodge(width = 0.75), alpha = 0.3, color = "black"
    ) +
    scale_color_d3() +
    scale_fill_d3() +
    theme_bw() +
    labs(
        x = "V-Gene", y = "V-Gene Frequency (%)",
        fill = "Disease Status",
        color = "Disease Status"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top"
    )

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_{Sys.Date()}.png"),
    plot = p_v_gene_usage,
    width = 14,
    height = 6
)

# Grouping by higher level gene id
v_gene_usage_long %>% glimpse()

v_gene_usage_long_broad <- v_gene_usage_long %>%
    mutate(vgenebroad = str_extract(Names, "^[A-Z0-9]+")) %>%
    group_by(sample_id, vgenebroad) %>%
    dplyr::summarise(counts = sum(counts)) %>%
    group_by(sample_id) %>%
    mutate(vgenebroadfreq = counts / sum(counts) * 100) %>%
    left_join(key_meta_df, by = c("sample_id")) %>%
    filter(case_control_other_latest != "Other") %>%
    mutate(case_control_other_latest = factor(case_control_other_latest, 
        levels = c("Control", "Case"))) %>%
    glimpse()

vgenebroad_order <- v_gene_usage_long_broad %>% 
    group_by(vgenebroad) %>% 
    dplyr::summarise(med_freq = median(vgenebroadfreq)) %>% 
    arrange(desc(med_freq)) %>% 
    pull(vgenebroad)

p_v_gene_usage_broad <- v_gene_usage_long_broad %>%
    filter(case_control_other_latest != "Other") %>%
    mutate(vgenebroad = factor(vgenebroad, levels = vgenebroad_order)) %>%
    ggplot(aes(
        x = vgenebroad,
        y = vgenebroadfreq, color = case_control_other_latest
    )) +
    geom_point(aes(color = case_control_other_latest),
        position = position_jitterdodge(), alpha = 0.1, size = 0.05) +
    geom_boxplot(
        aes(group = interaction(vgenebroad, case_control_other_latest)),
        outlier.alpha = 0, # fill.alpha = 0.2,
        position = position_dodge(width = 0.75), alpha = 0.3, color = "black"
    ) +
    scale_color_d3() +
    scale_fill_d3() +
    theme_bw() +
    labs(
        x = "V-Gene", y = "V-Gene Frequency (%)",
        fill = "Disease Status",
        color = "Disease Status"
    ) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_broad_{Sys.Date()}.png"),
    plot = p_v_gene_usage_broad,
    width = 14,
    height = 6
)

vgene_data %>% glimpse()
unique_vgenes <- unique(v_gene_usage_long_broad$vgenebroad)
lmm_results <- list()
# lmm_anova_res <- list()

# Iterate through each vgenebroad
for (vgene in unique_vgenes) {
    # Subset data for current vgene
    vgene_data <- v_gene_usage_long_broad %>%
        filter(vgenebroad == vgene)

    # Skip if not enough data
    if (nrow(vgene_data) < 10) {
        message(glue("Insufficient data for {vgene}"))
        lmm_results[[vgene]] <- NULL
        next
    }

    # Run linear mixed model
    tryCatch(
        {
            lmm_model <- nlme::lme(vgenebroadfreq ~ case_control_other_latest + age_at_baseline + sex + abis_t_cell ,
                random = ~ 1 | participant_id,
                data = vgene_data,
                control = lmeControl(opt = "optim")
            )

            # Store model summary
            lmm_results[[vgene]] <- lmm_model

            # Print basic results
            cat("Results for", vgene, ":\n")
            print(anova(lmm_model))
            cat("\n")
        },
        error = function(e) {
            cat("Error in model for", vgene, ":", e$message, "\n\n")
            lmm_results[[vgene]] <- NULL
        }
    )
}

vgene_lmm_res <- lmm_results %>%
    purrr::set_names(names(.)) %>%
    purrr::map(~ broom.mixed::tidy(.)) %>%
    bind_rows(.id = "vgene") %>%
    glimpse()

vgene_lmm_res_better <- tibble()
for (v in names(lmm_results)) {
    d <- lmm_results[[v]] %>% anova() %>% 
        rownames_to_column(var = "term") %>% 
        mutate(vgene = v)
    vgene_lmm_res_better %<>% bind_rows(d)
}
vgene_lmm_res_better %>% glimpse()

vgene_lmm_res_better <- vgene_lmm_res_better %>%
    dplyr::rename(p.value = `p-value`, f.value = `F-value`) %>% 
    filter(term == "case_control_other_latest") %>%
    glimpse()

# vgene_lmm_res_better %>% View()

# Create a plot of LMM results
vgene_lmm_plot <- vgene_lmm_res_better %>%
    # filter(term == "case_control_other_latestCase") %>%
    # Match the order of vgenes from the previous plot
    mutate(vgene = factor(vgene, levels = vgenebroad_order)) %>%
    # Add significance flag
    mutate(is_significant = p.value <= 0.05) %>%
    ggplot(aes(x = vgene, y = f.value)) +
    geom_point(aes(color = is_significant)) +
    # geom_pointrange(aes(
    #     ymin = estimate - std.error, 
    #     ymax = estimate + std.error,
    #     color = is_significant
    # ), size = 0.2) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_bw() +
    labs(x = NULL, y = "F-Statistic",
         title = "Linear Mixed Model Results for V Gene Usage") +
    theme(axis.text.x = element_blank(), legend.position = "top") +
    guides(color = "none")

# Stack the plots vertically using aplot
p_v_gene_usage_broad_lmm <- p_v_gene_usage_broad %>%
    aplot::insert_top(vgene_lmm_plot, 
                     height = 0.3)

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_broad_lmm_{Sys.Date()}.png"),
    plot = p_v_gene_usage_broad_lmm,
    width = 12,
    height = 6
)

# vgene_lmm_res %>%
#     filter(term == "case_control_other_latestCase") %>% 
#     DT::datatable()

lmm_results[["TRBV27"]] %>% anova()
lmm_results[["TRBV27"]] %>% broom.mixed::tidy() %>% print
# lmm_results[["TRBV27"]] %>% broom.mixed::tidy() %>% 
#     filter(term == "case_control_other_latestCase") %>% print()

```

Since data is not log-scaled and is in percentages, the interpretation of this result is that only one V-gene has a significant difference in frequency between control and case. However, the effect size is about a 0.11% difference in frequency between control and case - which is unlikely to be biologically meaningful.


```{r}

# gliph2_dir <- "/resnick/groups/MazmanianLab/jboktor/software/gliph2"

# out_prefix=Prefix #entry after this pound sign will be ignored
# cdr3_file=cdr3_file.txt
# hla_file =hla_file.txt
# refer_file=ref_file.txt
# v_usage_freq_file=ref_V.txt
# cdr3_length_freq_file=ref_L.txt
# local_min_pvalue=0.001
# p_depth = 1000
# global_convergence_cutoff = 1
# simulation_depth=1000
# kmer_min_depth=3
# local_min_OVE=10
# algorithm=GLIPH2
# all_aa_interchangeable=1


```


Shannon entropy and clonality can be calculated as follows:

$$
e = - \sum_{i = 1}^N p_i \log p_i
$$

and

$$
\text{clonality} = 1 - \frac{e}{\log N}
$$


Reading in and adding q0 to the hill diversity data

```{r, eval = FALSE}
immdata <- readRDS(
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_2024-11-21.rds")
)

hill_q0 <- names(immdata$data) %>% 
    purrr::set_names() %>% 
    purrr::map(~immdata$data[[.]] %>% nrow()) %>% 
    bind_rows(.id = "sample_id") %>% 
    pivot_longer(cols = everything(), 
        names_to = "Sample", values_to = "Value") %>% 
        mutate(Q = 0) %>% 
    glimpse()

hill_diversity <- readRDS(
    glue("{metrics_dir}/rep-diversity-hill_2024-12-19.rds")) %>% 
    bind_rows(hill_q0) %>% 
    glimpse()

saveRDS(hill_diversity, 
    glue("{metrics_dir}/rep-diversity-hill_2024-12-19_plus_q0.rds")
)

```


```{r}
hill_diversity <- readRDS(
    glue("{metrics_dir}/rep-diversity-hill_2024-12-19_plus_q0.rds")
)

hill_diversity_df <- hill_diversity %>%
    left_join(key_meta_df, by = c("Sample" = "sample_id")) %>%
    filter(!is.na(case_control_other_latest)) %>%
    glimpse()

hill_diversity_df %>% glimpse()

hill_lme_mods <- list()
for (q in unique(hill_diversity_df$Q)) {
    clonotype_counts_df <- hill_diversity_df %>% filter(Q == q) %>% glimpse()
    q_name <- as.character(q)
    hill_lme_mods[[q_name]] <- nlme::lme(
        Value ~ case_control_other_latest + age_at_baseline + sex + study + tcell_reads_abis,
        random = ~ 1 | participant_id,
        data = clonotype_counts_df,
        na.action = na.omit
    )
    print(summary(hill_lme_mods[[q_name]]))
}

hill_lme_mods_df <- names(hill_lme_mods) %>% 
    purrr::set_names() %>% 
    purrr::map(~hill_lme_mods[[.]] %>% broom.mixed::tidy()) %>% 
    bind_rows(.id = "Q") %>% 
    glimpse()

hill_lme_mods_df %>% 
    filter(term == "case_control_other_latestCase")  %>% 
    print()


p_hill_div_detailed <- hill_diversity_df %>%
    filter(Q > 0) %>% 
    ggplot(aes(
        x = Q, y = Value,
        color = case_control_other_latest, fill = case_control_other_latest
    )) +
    geom_line(aes(group = interaction(case_control_other_latest, Sample)),
        alpha = 0.1, size = 0.1) +
    geom_smooth(method = "loess", se = TRUE) +
    labs(x = "Q Value", y = "Hill Diversity") +
    scale_color_d3() +
    scale_fill_d3() +
    scale_y_log10() +
    theme_bw() +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/hill_diversity_beta_chain_detailed_{Sys.Date()}.png"),
    p_hill_div_detailed,
    width = 5,
    height = 4
)



# Plotting Hill Diversity
p_hill_div <- hill_diversity_df %>%
    filter(Q > 0) %>% 
    ggplot(aes(
        x = Q, y = Value,
        color = case_control_other_latest, fill = case_control_other_latest
    )) +
    geom_smooth(method = "loess", se = TRUE) +
    labs(x = "Q Value", y = "Hill Diversity") +
    scale_color_d3() +
    scale_fill_d3() +
    # scale_y_log10() +
    theme_bw() +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/hill_diversity_beta_chain_{Sys.Date()}.png"),
    p_hill_div,
    width = 5,
    height = 4
)

p_hill_div_bx <- hill_diversity_df %>%
    ggplot(aes(x = Q, y = Value)) +
    geom_point(
        aes(color = case_control_other_latest),
        position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
        alpha = 0.3, size = 0.8
    ) +
    geom_boxplot(
        aes(group = interaction(case_control_other_latest, Q)),
        outlier.alpha = 0, position = position_dodge(width = 0.8), alpha = 0.5) +
    labs(x = "Q Value", y = "Hill Diversity") +
    scale_y_log10() +
    scale_color_d3() +
    scale_fill_d3() +
    theme_bw() +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/hill_diversity_beta_chain_boxplot_{Sys.Date()}.png"),
    p_hill_div_bx,
    width = 6.5,
    height = 4
)



hill_diversity_df %>% glimpse()
p_seqdepth_vs_hill_div <- hill_diversity_df %>%
    ggplot(aes(x = Value, y = raw_total_reads)) +
    geom_point(size = 0.3, alpha = 0.1) +
    geom_smooth(method = "lm") +
    labs(x = "Hill Diversity", y = "Total Reads") +
    theme_bw() +
    facet_wrap(~Q,
        scales = "free",
        labeller = labeller(Q = function(x) paste("Q: ", x))
    ) +
    scale_y_log10() +
    theme(legend.position = "top") +
    stat_regline_equation(label.x.npc = 0.2, label.y.npc = 0.9) +
    stat_cor(method = "pearson", label.x.npc = 0.2, label.y.npc = 0.8)

ggsave(
    glue("{imm_plots_dir}/seqdepth_vs_hill_div_{Sys.Date()}.png"),
    p_seqdepth_vs_hill_div,
    width = 10,
    height = 8
)

p_mds4_vs_hill <- hill_diversity_df %>%
    ggplot(aes(x = Value, y = mds_updrs_part_iv_summary_score)) +
    geom_point(size = 0.3, alpha = 0.1) +
    geom_smooth(method = "lm") +
    labs(x = "Hill Diversity", y = "MDS-UPDRS Part IV") +
    theme_bw() +
    facet_wrap(~Q,
        scales = "free",
        labeller = labeller(Q = function(x) paste("Q: ", x))
    ) +
    theme(legend.position = "top") +
    stat_regline_equation(label.x.npc = 0.2, label.y.npc = 0.9) +
    stat_cor(method = "pearson", label.x.npc = 0.2, label.y.npc = 0.8)

ggsave(
    glue("{imm_plots_dir}/mds4_vs_hill_div_{Sys.Date()}.png"),
    p_mds4_vs_hill,
    width = 10,
    height = 8
)

p_age_vs_hill <- hill_diversity_df %>%
    ggplot(aes(x = Value, y = age_at_baseline)) +
    geom_point(size = 0.3, alpha = 0.1) +
    geom_smooth(method = "loess") +
    labs(x = "Hill Diversity", y = "Age at Baseline") +
    theme_bw() +
    facet_wrap(~Q,
        scales = "free",
        labeller = labeller(Q = function(x) paste("Q: ", x))
    ) +
    theme(legend.position = "top") +
    stat_regline_equation(label.x.npc = 0.2, label.y.npc = 0.9) +
    stat_cor(method = "pearson", label.x.npc = 0.2, label.y.npc = 0.8)

ggsave(
    glue("{imm_plots_dir}/age_vs_hill_div_{Sys.Date()}.png"),
    p_age_vs_hill,
    width = 10,
    height = 8
)

```




```{r}
# Create a named list where each element contains Values arranged by Q
hill_diversity_df %>% 
    select(Sample, Q, Value) %>% 
    glimpse()

# Create a named list where names are Sample IDs and values are vectors of diversity values by Q
hill_diversity_list <- hill_diversity_df %>%
    select(Sample, Q, Value) %>%
    group_by(Sample) %>%
    arrange(Sample, Q) %>%
    group_split() %>%
    setNames(unique(hill_diversity_df$Sample))

# Convert each group to a named vector where names are Q values
hill_diversity_list <- lapply(hill_diversity_list, function(df) {
    values <- df$Value
    names(values) <- df$Q
    return(values)
})

# Check the structure of the first few elements
str(hill_diversity_list[1:3])

# Create all pairwise combinations of samples
sample_names <- names(hill_diversity_list)
sample_pairs <- expand.grid(sample1 = sample_names, sample2 = sample_names, stringsAsFactors = FALSE) %>%
  filter(sample1 != sample2)  # Exclude self-comparisons

# Function to calculate correlation between two samples
calculate_correlation <- function(sample1, sample2) {
  # Extract diversity values for both samples
  values1 <- hill_diversity_list[[sample1]]
  values2 <- hill_diversity_list[[sample2]]
  
  # Ensure both vectors have the same Q values (names)
  common_qs <- intersect(names(values1), names(values2))
  
  # Calculate correlation if there are common Q values
  if (length(common_qs) > 1) {
    cor_result <- stats::cor(values1[common_qs], values2[common_qs], method = "pearson")
    return(tibble(sample1 = sample1, sample2 = sample2, correlation = cor_result))
  } else {
    return(tibble(sample1 = sample1, sample2 = sample2, correlation = NA_real_))
  }
}

# Set up parallel processing
plan(multisession, workers = availableCores() - 1)

# Run correlations in parallel
correlation_results <- furrr::future_map2_dfr(
  sample_pairs$sample1,
  sample_pairs$sample2,
  ~calculate_correlation(.x, .y),
  .options = furrr_options(seed = TRUE)
)

saveRDS(correlation_results,
    glue("{metrics_dir}/hill_diversity_pearson_correlations_{Sys.Date()}.rds")
)

```


```{r}
correlation_results <- readRDS(
    glue("{metrics_dir}/hill_diversity_pearson_correlations_2025-03-17.rds")
)

# Convert to a matrix format if needed
correlation_matrix <- correlation_results %>%
    pivot_wider(
        id_cols = sample1,
        names_from = sample2,
        values_from = correlation,
        values_fill = 1 # Should only be same pairs
    ) %>%
    column_to_rownames("sample1") %>%
    as.matrix()

# Preview results
head(correlation_results)
head(correlation_matrix)

# # Seriate the matrix
# sid_order_hillcor <- seriate_matrix_rows(correlation_matrix, seriate_method = "OLO")

# saveRDS(
#     sid_order_hillcor,
#     glue("{metrics_dir}/repertoire_stats/hill_diversity_pearson_cor_seriation_{Sys.Date()}.rds")
# )

sid_order_hillcor <- readRDS(
    glue("{metrics_dir}/repertoire_stats/hill_diversity_pearson_cor_seriation_2025-03-24.rds")
)

# Create a custom theme for heatmaps
theme_massive_heatmap <- function() {
  theme_void() +
    theme(
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        panel.border = element_blank()
    )
}


p_hill_cor_heatmap <- correlation_results %>%
    mutate(sample1 = factor(sample1, levels = sid_order_hillcor)) %>%
    mutate(sample2 = factor(sample2, levels = sid_order_hillcor)) %>%
    ggplot(aes(
        x = fct_relevel(sample1, sid_order_hillcor), 
        y = fct_relevel(sample2, sid_order_hillcor), 
        fill = correlation
    )) +
    geom_tile() +
    scale_fill_viridis_c(option = "mako") +
    labs(x = NULL, y = NULL, fill = "Hill Diversity \nPearson Correlation") +
    theme_void() +
    theme(
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        panel.border = element_blank()
    )

ggsave(
    glue("{imm_plots_dir}/hill_diversity_pearson_cor_heatmap_{Sys.Date()}.png"),
    p_hill_cor_heatmap,
    width = 10,
    height = 8
)

correlation_results_meta <- correlation_results %>%
    left_join(
        full_meta_df %>%
            select(sample_id, participant_id, case_control_other_latest, study) %>%
            rename_with(~ paste0(., "1"), everything()),
        by = c("sample1" = "sample_id1")
    ) %>%
    # Join metadata for sample2
    left_join(
        full_meta_df %>%
            select(sample_id, participant_id, case_control_other_latest, study) %>%
            rename_with(~ paste0(., "2"), everything()),
        by = c("sample2" = "sample_id2")
    ) %>%
    # Add comparison columns
    mutate(
        disease_grouping = case_when(
            (case_control_other_latest1 == "Case" & case_control_other_latest2 == "Case") ~ "Within \nCase",
            (case_control_other_latest1 == "Control" & case_control_other_latest2 == "Control") ~ "Within \nControl",
            (case_control_other_latest1 == "Case" & case_control_other_latest2 == "Control") |
                (case_control_other_latest1 == "Control" & case_control_other_latest2 == "Case") ~ "Between \nCase and Control",
            TRUE ~ "other"
        ),
        participant_grouping = if_else(
            participant_id1 == participant_id2,
            "Within Participant",
            "Between Participants"
        ),
        study_grouping = case_when(
            study1 == study2 ~ paste0("Within \n", study1),
            TRUE ~ paste0("Between \n", study1, " and ", study2)
        )
    ) %>%
    glimpse()

p_hill_div_vs_disease <- correlation_results_meta %>%
    ggplot(aes(x = disease_grouping, y = correlation)) +
    geom_boxplot() +
    # geom_violin() +
    theme_bw()

ggsave(
    glue("{imm_plots_dir}/hill_diversity_vs_disease_{Sys.Date()}.png"),
    p_hill_div_vs_disease,
    width = 6,
    height = 4
)

```

PERMANOVA 

```{r}
# Convert correlation matrix to distance matrix (1 - correlation)
distance_matrix <- 1 - correlation_matrix

# Define metadata columns for PERMANOVA
key_meta_df %>% glimpse()
full_meta_df %>% glimpse()
target_metadata <- c(
    "study", "raw_total_reads", "raw_total_reads_binned",
    "case_control_other_latest", # "participant_id",
    "DRB1_grouped", "DRB1_risk",
    "age_at_baseline", "sex", mds_tests
)

# Get sample metadata for PERMANOVA
permanova_metadata <- full_meta_df %>%
    filter(sample_id %in% rownames(distance_matrix)) %>%
    column_to_rownames("sample_id") %>%
    select(all_of(target_metadata), participant_id) %>%
    glimpse()

permanova_metadata %>% glimpse()
common_samples <- intersect(rownames(distance_matrix), rownames(permanova_metadata))

permanova_resdir <- glue("{metrics_dir}/permanova_hill_div_corr")
dir.create(permanova_resdir, showWarnings = FALSE, recursive = TRUE)

subsample_resdir <- glue(
    "{metrics_dir}/permanova_hill_div_corr_sid_subsample")
subsample_resdir_input <- glue(
    "{subsample_resdir}/model_input")
perm_resdir <- glue("{subsample_resdir}/results")

dir.create(subsample_resdir_input, showWarnings = FALSE, recursive = TRUE)
dir.create(perm_resdir, showWarnings = FALSE, recursive = TRUE)

# Saving subsampled distance matrix and metadata
for (i in 1:50) {
    set.seed(i)
    sid_subsample <- permanova_metadata %>% 
        rownames_to_column(var = "sample_id") %>%
        group_by(participant_id) %>%
        dplyr::slice_sample(n = 1) %>%
        ungroup() %>%
        glimpse() %>% 
        pull(sample_id)

    i_common_samples <- intersect(common_samples, sid_subsample)
    distance_matrix_filtered <- as.dist(distance_matrix[i_common_samples, i_common_samples])
    permanova_metadata_filtered <- permanova_metadata[i_common_samples, ] %>% glimpse()
    
    # Saving distance matrix and metadata
    saveRDS(
        distance_matrix_filtered,
        glue("{subsample_resdir_input}/",
        "distance_matrix_filtered_subsample_{i}.rds")
    )
    saveRDS(
        permanova_metadata_filtered,
        glue("{subsample_resdir_input}/",
        "permanova_metadata_filtered_subsample_{i}.rds")
    )
}

# Run PERMANOVA for each metadata variable
slurm_logs <- glue("{wkdir}/.cluster_runs/{Sys.Date()}_permanova_hill_div_corr_sid_subsample")
dir.create(slurm_logs, showWarnings = FALSE, recursive = TRUE)

nperms <- 9999
nthreads <- 1

for (i in 26:50) {
    for (var in target_metadata) {
        message("Running PERMANOVA for: ", var)

        # Important that distance_matrix is used here below
        form <- glue("distance_matrix ~ {var}")
        breg <- easy_batch(
            cluster_name = glue("pm_{i}_{var}"),
            fun = run_permanova,
            params = data.frame(
                "dmat_path" = glue("{subsample_resdir_input}/", 
                    "distance_matrix_filtered_subsample_{i}.rds"),
                "meta_path" = glue("{subsample_resdir_input}/",
                    "permanova_metadata_filtered_subsample_{i}.rds"),
                "model_formula_string" = form,
                "outpath" = glue("{perm_resdir}/permanova_{var}_nperms_{nperms}_subsample_{i}_{Sys.Date()}.rds"),
                "n_perms" = nperms,
                "n_cores" = nthreads
            ),
            template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
            log_dir = slurm_logs,
            walltime = "20:00:00",
            memory = "50GB",
            ncpus = nthreads,
            max_jobs = 9999
        )
        
    }
}


res_paths <- list.files(perm_resdir, full.names = TRUE, 
    pattern = "_9999_")

permanova_results <- res_paths %>%
    purrr::set_names(basename(.)) %>%
    purrr::map(~ readRDS(.) %>% broom::tidy()) %>%
    bind_rows(.id = "model_info") %>%
    janitor::clean_names() %>%
    mutate(r2_percent = round(r2 * 100, 3)) %>%
    arrange(p_value) %>%
    glimpse()

p_permanova <- permanova_results %>%
    filter(term %nin% c("Residual", "Total")) %>%
    arrange(p_value) %>%
    mutate(
        significant = p_value < 0.05,
        p_value_formatted = format.pval(p_value, digits = 3)
    ) %>%
    ggplot(aes(y = fct_reorder(term, r2), x = r2*100)) +
    geom_point(aes(color = significant), size = 0.8,
        position = position_jitter(height = 0.2, width = 0), alpha = 0.85) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.5) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    labs(
        title = "PERMANOVA Results: TCR Repertoire Similarity",
        x = "R² % (Effect Size)",
        y = NULL
    ) +
    theme(legend.position = "bottom")

ggsave(
    glue("{imm_plots_dir}/tcr_permanova_results_{Sys.Date()}.png"),
    p_permanova,
    width = 8,
    height = 4
)

```


overlap coefficient (.method = "overlap") - a normalised measure of overlap similarity. It is defined as the size of the intersection divided by the smaller of the size of the two sets.

```{r}
pub_ovov <- readRDS(
    glue("{metrics_dir}/rep-overlap-overlap_2024-12-19.rds")
)
pub_ovov_df <- pub_ovov %>% as.data.frame() %>% 
    glimpse()

pub_ovov_df_long <- pub_ovov_df %>% 
    rownames_to_column("sample_id1") %>%
    pivot_longer(
        cols = -sample_id1,
        names_to = "sample_id2",
        values_to = "overlap_counts"
    ) %>%
    mutate(overlap_counts = replace_na(overlap_counts, 1)) %>%
    glimpse()

# Convert the long format data back to a matrix
pub_ovov_matrix <- pub_ovov_df_long %>%
  pivot_wider(
    id_cols = sample_id1,
    names_from = sample_id2,
    values_from = overlap_counts
  ) %>%
  column_to_rownames("sample_id1") %>%
  as.matrix()

# Display the first few rows and columns
pub_ovov_matrix[1:5, 1:5]
# sid_order_ovov <- seriate_matrix_rows(pub_ovov_matrix, seriate_method = "OLO")
# saveRDS(sid_order_ovov, 
#     glue("{metrics_dir}/repertoire_stats/sid_order_ovov_HC_average_{Sys.Date()}.rds")
#     )
sid_order_ovov <- readRDS( 
    glue("{metrics_dir}/repertoire_stats/sid_order_ovov_HC_average_2025-03-24.rds")
    )

p_ovov_heatmap <- pub_ovov_df_long %>%
    mutate(sample_id1 = factor(sample_id1, levels = sid_order_ovov)) %>%
    mutate(sample_id2 = factor(sample_id2, levels = sid_order_ovov)) %>%
    filter(sample_id1 != sample_id2) %>%
    ggplot(aes(x = sample_id1, y = sample_id2, fill = overlap_counts)) +
    geom_tile() +
    scale_fill_viridis_c(
        option = "mako",
        trans = "log10",
        name = "Overlap Count",
        na.value = "white",
        direction = -1
    ) +
    labs(x = NULL, y = NULL, fill = "Overlap Count") +
    theme_massive_heatmap()

# ggsave(
#     glue("{imm_plots_dir}/public_overlap_overlap_heatmap_v2_{Sys.Date()}.png"),
#     p_ovov_heatmap,
#     width = 10,
#     height = 8,
#     bg = "white"
# )

pub_ovov_df_long %>% glimpse()

pub_ovov_df_long_full <- pub_ovov_df_long %>% 
  # Join metadata for first sample
  left_join(
    key_meta_df %>% 
      select(all_of(essential_meta), sample_id, participant_id) %>%
      rename_with(~paste0(., "1"), everything()),
    by = c("sample_id1")
  ) %>%
  # Join metadata for second sample  
  left_join(
    key_meta_df %>%
      select(all_of(essential_meta), sample_id, participant_id) %>%
      rename_with(~paste0(., "2"), everything()),
    by = c("sample_id2")
  ) %>%
    filter(!is.na(case_control_other_latest1)) %>%
    filter(!is.na(case_control_other_latest2)) %>%
  glimpse()

pub_ovov_df_long_full_annot <- pub_ovov_df_long_full %>% 
    mutate(disease_condition = case_when(
        participant_id1 == participant_id2 & case_control_other_latest1 == "Case" ~ "Within\nCase (Donors)",
        participant_id1 == participant_id2 & case_control_other_latest1 == "Control" ~ "Within\nControl (Donors)",
        case_control_other_latest1 == "Case" & case_control_other_latest2 == "Case" ~ "Between\nCase",
        case_control_other_latest1 == "Control" & case_control_other_latest2 == "Control" ~ "Between\nControl",
        case_control_other_latest1 == "Case" & case_control_other_latest2 == "Control" ~ "Between\nCase-Control",
        case_control_other_latest1 == "Control" & case_control_other_latest2 == "Case" ~ "Between\nCase-Control",
        TRUE ~ "ERROR"
    )) %>%
    mutate(disease_condition = factor(disease_condition, 
        levels = c(
            "Within\nCase (Donors)",
            "Within\nControl (Donors)",
            "Between\nCase",
            "Between\nControl",
            "Between\nCase-Control"
        ))) %>% 
    mutate(HLA_DRB1_risk_grouping = case_when(
        DRB1_risk1 == "risk" & DRB1_risk2 == "risk" ~ "Witihin\nRisk",
        DRB1_risk1 == "protective" & DRB1_risk2 == "protective" ~ "Witihin\nProtective",
        DRB1_risk1 == "mixed" & DRB1_risk2 == "mixed" ~ "Witihin\n Mixed",
        (DRB1_risk1 == "risk" & DRB1_risk2 == "protective") | 
            (DRB1_risk1 == "protective" & DRB1_risk2 == "risk") ~ "Between\nRisk & Protective",
        DRB1_risk1 == "none" & DRB1_risk2 == "none" ~ "Within\nDRX/DRX",
        TRUE ~ "OTHER"
    )) %>%
    glimpse()

pub_ovov_df_long_full_annot %>% glimpse()
# pub_ovov_df_long_full_annot$DRB1_risk1 %>% table()


```

### Hypothesis Testing: Do PD patients share more clones with other PD patients than Controls? More specifically - we want to test the below three hypotheses. Using pairwise overlap scores as iyr data, we utilize mixed effects models (dyadic regression) treating each overlap as the outcome and include crossed random effects for both individuals:

1) Do PD Patients share more receptors amongst their own tiempoints than Ctrl

2) Do PD Patients share more rec with other random PD patients than Healthy Ctrl with one another

3) Do PD patients share more rec with other random PD patients than with Healthy controls

```{r}
pub_ovov_df_long_full_annot$disease_condition %>% table()

pairs_df <- pub_ovov_df_long_full_annot %>% 
    filter(sample_id1 != sample_id2) %>% 
    glimpse()
eps <- 1e-4
pairs_df <- within(pairs_df, {
  logit_ov <- stats::qlogis(pmin(pmax(overlap_counts, eps), 1 - eps))
  s1 <- factor(sample_id1)               # make sure they’re factors
  s2 <- factor(sample_id2)
})
pairs_df$disease_condition %>% table()

# TESTING Question 1)
h1_groups <- c("Within\nCase (Donors)", "Within\nControl (Donors)")
h2_groups <- c("Between\nCase", "Between\nControl")
h3_groups <- c("Between\nCase", "Between\nCase-Control")

mod_within_donors <- nlme::lme(
  fixed  = logit_ov ~ disease_condition,
  random = list(
            s1 = pdIdent(~ 1),          # random intercept for the first sample
            s2 = pdIdent(~ 1)           # random intercept for the second
          ),
        data = pairs_df %>% 
            filter(disease_condition %in% c("Within\nCase (Donors)", "Within\nControl (Donors)")),
  method = "REML",
  control = lmeControl(opt = "optim")
)
summary(mod_within_donors)
broom.mixed::tidy(mod_within_donors)

# TESTING Question 2)
mod_btwn_case_btwn_ctrl <- nlme::lme(
  fixed  = logit_ov ~ disease_condition,
  random = list(
            s1 = pdIdent(~ 1),          # random intercept for the first sample
            s2 = pdIdent(~ 1)           # random intercept for the second
          ),
        data = pairs_df %>% 
            filter(disease_condition %in% h2_group),
  method = "REML",
  control = lmeControl(opt = "optim")
)
summary(mod_btwn_case_btwn_ctrl)
broom.mixed::tidy(mod_btwn_case_btwn_ctrl)

# TESTING Question 3)
mod_btwn_case_btwn_casectrl <- nlme::lme(
  fixed  = logit_ov ~ disease_condition,
  random = list(
            s1 = pdIdent(~ 1),          # random intercept for the first sample
            s2 = pdIdent(~ 1)           # random intercept for the second
          ),
        data = pairs_df %>% 
            filter(disease_condition %in% h3_groups),
  method = "REML",
  control = lmeControl(opt = "optim")
)
summary(mod_btwn_case_btwn_casectrl)
broom.mixed::tidy(mod_btwn_case_btwn_casectrl)

# A less fancy stat test --  Collapse to per-DONOR summaries (breaks the dependence)
# 1.	For every PD DONOR, compute its mean overlap to other PD and mean overlap to controls; take the within-patient difference.
# 2.	Test these paired differences with a Wilcoxon signed-rank (H1).
# 3.	Analogously, compare the per-sample “within-group means” between PD and CTRL with a rank-sum (H2)

pairs_summary_df <- pub_ovov_df_long_full_annot %>% 
    filter(sample_id1 != sample_id2) %>% 
    group_by(participant_id1, disease_condition) %>% 
    dplyr::summarize(mean_overlap_score = mean(overlap_counts)) %>% 
    glimpse()

p_donor_avgs_ovov_boxplot <- pairs_summary_df %>% 
    ggplot(aes(x = disease_condition, y=mean_overlap_score)) +
    scale_y_log10() +
    geom_boxplot() +
    theme_bw() +
    labs(x = NULL, y = "Mean Overlap Score") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    glue("{imm_plots_dir}/public_overlap_overlap_boxplot_donoravgs_{Sys.Date()}.png"),
    p_donor_avgs_ovov_boxplot,
    width = 3.25,
    height = 2.75
)

pairs_summary_df %>% glimpse()

wx1 <- wilcox.test(mean_overlap_score ~ disease_condition, 
    data = pairs_summary_df %>% filter(disease_condition  %in% h1_groups))
wx2 <- wilcox.test(mean_overlap_score ~ disease_condition, 
    data = pairs_summary_df %>% filter(disease_condition  %in% h2_groups))
wx3 <- wilcox.test(mean_overlap_score ~ disease_condition, 
    data = pairs_summary_df %>% filter(disease_condition  %in% h3_groups))


#' No significant differences between distributions with donor summarized testing

#' Next we will test our hypothesis that donors within HLA risk and protective groups share 
pub_ovov_df_long_full_annot %>% glimpse()
pairs_summary_df_hlarisk <- pub_ovov_df_long_full_annot %>% 
    filter(participant_id1 != participant_id2) %>% 
    group_by(participant_id1, DRB1_risk1) %>% 
    dplyr::summarize(mean_overlap_score = mean(overlap_counts)) %>% 
    glimpse()

pairs_summary_df_hlarisk %>% glimpse()
pairs_summary_df_hlarisk$DRB1_risk1 %>% table()

wx_risk_to_none <- wilcox.test(mean_overlap_score ~ DRB1_risk1, 
    data = pairs_summary_df_hlarisk %>% filter(DRB1_risk1 %in% c(
        "risk", "none"
    )))
wx_protective_to_none <- wilcox.test(mean_overlap_score ~ DRB1_risk1, 
    data = pairs_summary_df_hlarisk %>% filter(DRB1_risk1 %in% c(
        "protective", "none"
    )))
wx_protective_to_risk <- wilcox.test(mean_overlap_score ~ DRB1_risk1, 
    data = pairs_summary_df_hlarisk %>% filter(DRB1_risk1 %in% c(
        "protective", "risk"
    )))

p_donor_avgs_ovov_boxplot_hlarisk <- pairs_summary_df_hlarisk %>% 
    ggplot(aes(x = DRB1_risk1, y = mean_overlap_score)) +
    scale_y_log10() +
    geom_boxplot() +
    theme_bw() +
    labs(x = NULL, y = "Mean Overlap Score") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    glue("{imm_plots_dir}/public_overlap_overlap_boxplot_donoravgs_hla-risk_{Sys.Date()}.png"),
    p_donor_avgs_ovov_boxplot_hlarisk,
    width = 4,
    height = 4
)

# What about different subsets of samples between of HLA risk genes?

pub_ovov_df_long_full_annot %>% glimpse()
pub_ovov_df_long_full_annot$DRB1_grouped1 %>% table()

pairs_summary_df_hlariskgenes <- pub_ovov_df_long_full_annot %>% 
    filter(participant_id1 != participant_id2) %>% 
    group_by(participant_id1, DRB1_grouped1) %>% 
    dplyr::summarize(mean_overlap_score = mean(overlap_counts)) %>% 
    glimpse()

wx_risk_dose <- wilcox.test(mean_overlap_score ~ DRB1_grouped1, 
    data = pairs_summary_df_hlariskgenes %>% filter(DRB1_grouped1 %in% c(
        "DRB1*15:01/DRB1*15:01", "DRB1*15:01/DRB1X"
    )))
wx_risk_dose2 <- wilcox.test(mean_overlap_score ~ DRB1_grouped1, 
    data = pairs_summary_df_hlariskgenes %>% filter(DRB1_grouped1 %in% c(
        "DRB1*15:01/DRB1*15:01", "DRB1X/DRB1X"
    )))

wx_protect_dose <- wilcox.test(mean_overlap_score ~ DRB1_grouped1, 
    data = pairs_summary_df_hlariskgenes %>% filter(DRB1_grouped1 %in% c(
        "DRB1*04:01/DRB1*04:01", "DRB1*04:01/DRB1X"
    )))
wx_protect_dose2 <- wilcox.test(mean_overlap_score ~ DRB1_grouped1, 
    data = pairs_summary_df_hlariskgenes %>% filter(DRB1_grouped1 %in% c(
        "DRB1*04:01/DRB1*04:01", "DRB1X/DRB1X"
    )))
wx_protect_dose3 <- wilcox.test(mean_overlap_score ~ DRB1_grouped1, 
    data = pairs_summary_df_hlariskgenes %>% filter(DRB1_grouped1 %in% c(
        "DRB1*04:04/DRB1*04:04", "DRB1*04:04/DRB1X"
    )))

# No difference in any of these tests either






min_overzero <- pub_ovov_df_long_full_annot$overlap_counts %>% unique() %>% sort() %>% .[2]
p_ovov_boxplot <- pub_ovov_df_long_full_annot %>%
    filter(sample_id1 != sample_id2) %>%    
    # filter(participant_id1 != participant_id2) %>%
    mutate(overlap_counts_imputed = (overlap_counts + min_overzero/2)) %>%
    ggplot(aes(x = disease_condition, y = overlap_counts_imputed)) +
    geom_boxplot(outlier.alpha = 0) +
    # geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
    scale_y_log10() +
    theme_bw() +
    labs(x = NULL, y = "Public Overlap") +
    ggpubr::stat_compare_means(
        method = "wilcox.test",
        comparisons = list(
            c("Within\nCase (Donors)", "Within\nControl (Donors)"),
            c("Between\nCase", "Between\nControl"),
            c("Between\nCase", "Between\nCase-Control"),
            c("Between\nControl", "Between\nCase-Control")
        )
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    glue("{imm_plots_dir}/public_overlap_overlap_boxplot_{Sys.Date()}.png"),
    p_ovov_boxplot,
    width = 4,
    height = 4
)

p_ovov_boxplot_nonzero <- pub_ovov_df_long_full_annot %>%
    filter(sample_id1 != sample_id2) %>%    
    filter(participant_id1 != participant_id2) %>%
    filter(overlap_counts > 0) %>%
    ggplot(aes(x = disease_condition, y = overlap_counts)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
    scale_y_log10() +
    theme_bw() +
    labs(x = NULL, y = "Public Overlap") +
    ggpubr::stat_compare_means(
        method = "wilcox.test",
        comparisons = list(c("Within\nCase", "Between\nCase-Control"),
            c("Within\nControl", "Between\nCase-Control")
        )
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    glue("{imm_plots_dir}/public_overlap_overlap_boxplot_nonzero_{Sys.Date()}.png"),
    p_ovov_boxplot_nonzero,
    width = 3,
    height = 4
)


p_ovov_boxplot_hlarisk <- pub_ovov_df_long_full_annot %>%
    filter(sample_id1 != sample_id2) %>%    
    filter(participant_id1 != participant_id2) %>% 
    filter(HLA_DRB1_risk_grouping != "OTHER") %>% 
    filter(overlap_counts > 0) %>%
    ggplot(aes(x = HLA_DRB1_risk_grouping, y = overlap_counts)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
    scale_y_log10() +
    theme_bw() +
    labs(x = NULL, y = "Public Overlap") +
    ggpubr::stat_compare_means(
        method = "wilcox.test",
        comparisons = list(c("Within\nCase", "Between\nCase-Control"),
            c("Within\nControl", "Between\nCase-Control")
        )
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
    glue("{imm_plots_dir}/public_overlap_overlap_boxplot_hlarisk_{Sys.Date()}.png"),
    p_ovov_boxplot_hlarisk,
    width = 3,
    height = 4
)

```



PERMANOVA for public overlap (overlap coefficient)

```{r}
# Convert correlation matrix to distance matrix (1 - correlation)
distance_matrix <- 1 - pub_ovov_matrix

# Define metadata columns for PERMANOVA
key_meta_df %>% glimpse()
full_meta_df %>% glimpse()

target_metadata <- colnames(full_meta_df) %>% 
    discard(. %in% c("sample_id"))
# target_metadata <- c(
#     "study", "raw_total_reads", "raw_total_reads_binned",
#     "case_control_other_latest",
#     "DRB1_grouped", "DRB1_risk",
#     "age_at_baseline", "sex", mds_tests
# )

# Get sample metadata for PERMANOVA
permanova_metadata <- full_meta_df %>%
    filter(!is.na(case_control_other_latest)) %>%
    filter(sample_id %in% rownames(distance_matrix)) %>%
    column_to_rownames("sample_id") %>%
    select(all_of(target_metadata), participant_id) %>%
    glimpse()

subsample_resdir <- glue("{metrics_dir}/permanova_public_overlap_sid_subsample")
subsample_resdir_input <- glue("{subsample_resdir}/model_input")
perm_resdir <- glue("{subsample_resdir}/results")
dir.create(subsample_resdir_input, showWarnings = FALSE, recursive = TRUE)
dir.create(perm_resdir, showWarnings = FALSE, recursive = TRUE)
# Saving subsampled distance matrix and metadata
common_samples <- intersect(
    rownames(distance_matrix), rownames(permanova_metadata)
)

# Saving 50 random subsamples of distance matrix and metadata
for (i in 1:50) {
    message("Processing iter: ", i)
    set.seed(i)
    # already filtered for samples in dist matrix
    i_common_samples <- permanova_metadata %>%
        rownames_to_column(var = "sample_id") %>%
        group_by(participant_id) %>%
        dplyr::slice_sample(n = 1) %>%
        ungroup() %>%
        pull(sample_id)

    distance_matrix_filtered <- as.dist(distance_matrix[i_common_samples, i_common_samples])
    permanova_metadata_filtered <- permanova_metadata[i_common_samples, ] %>% glimpse()

    # Saving distance matrix and metadata
    saveRDS(
        distance_matrix_filtered,
        glue(
            "{subsample_resdir_input}/",
            "distance_matrix_filtered_subsample_{i}.rds"
        )
    )
    saveRDS(
        permanova_metadata_filtered,
        glue(
            "{subsample_resdir_input}/",
            "permanova_metadata_filtered_subsample_{i}.rds"
        )
    )
}

# # Run PERMANOVA for each metadata variable
# slurm_logs <- glue("{wkdir}/.cluster_runs/",
#     "{Sys.Date()}_permanova_public_overlap_sid_subsample")
# dir.create(slurm_logs, showWarnings = FALSE, recursive = TRUE)

# nperms <- 9999
# nthreads <- 2
# for (i in 1:50) {
#     for (var in target_metadata) {
#         message("Running PERMANOVA for: ", var)

#         # Important that distance_matrix is used here below
#         form <- glue("distance_matrix ~ {var}")
#         breg <- easy_batch(
#             cluster_name = glue("pm_{i}_{var}"),
#             fun = run_permanova,
#             params = data.frame(
#                 "dmat_path" = glue(
#                     "{subsample_resdir_input}/",
#                     "distance_matrix_filtered_subsample_{i}.rds"
#                 ),
#                 "meta_path" = glue(
#                     "{subsample_resdir_input}/",
#                     "permanova_metadata_filtered_subsample_{i}.rds"
#                 ),
#                 "model_formula_string" = form,
#                 "outpath" = glue("{perm_resdir}/permanova_{var}_nperms_{nperms}_subsample_{i}_{Sys.Date()}.rds"),
#                 "n_perms" = nperms,
#                 "n_cores" = nthreads
#             ),
#             template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
#             log_dir = slurm_logs,
#             walltime = "20:00:00",
#             memory = "50GB",
#             ncpus = nthreads,
#             max_jobs = 9999
#         )
#     }
# }


batch_params <- expand.grid("var" = target_metadata, "i" = 1:50) %>% 
    mutate(
        dmat_path = glue(
            "{subsample_resdir_input}/",
            "distance_matrix_filtered_subsample_{i}.rds"
        ),
        meta_path = glue(
            "{subsample_resdir_input}/",
            "permanova_metadata_filtered_subsample_{i}.rds"
        ),
        model_formula_string = glue("distance_matrix ~ {var}"),
        outpath = glue("{perm_resdir}/",
            "permanova_{var}_nperms_{nperms}_subsample_{i}_{Sys.Date()}.rds"),
        n_perms = 9999,
        n_cores = 2        
    ) %>%
    dplyr::select(-c(var, i)) %>% 
    glimpse()

# configure registry ----
breg <- breg_setup(cluster_name = "PERM_POVOV", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )

# Submit Jobs ----
jobs <- batchMap(
  fun = run_permanova,
  args = batch_params[21:26800, ],
  reg = breg
)
nrow(batch_params)

jobs[, chunk := chunk(job.id, chunk.size = 10)]
print(jobs[, .N, by = chunk])

submitJobs(jobs,
  resources = list(
    walltime = "2-00:00:00",
    memory = "50GB",
    ncpus = 2,
    max.concurrent.jobs = 9999
  )
)

```

Visualizing Public Overlap PERMANOVA Screen
# TODO FIXME update filters for new outputs

```{r}
subsample_resdir <- glue("{metrics_dir}/permanova_public_overlap_sid_subsample")
perm_resdir <- glue("{subsample_resdir}/results")

res_paths <- list.files(perm_resdir, 
    full.names = TRUE, pattern = "_9999_") %>% 
    keep(grepl("2025-06-06|2025-06-07", .))

future::plan("multisession", workers = 7)
permanova_results <- res_paths %>%
    purrr::set_names(basename(.)) %>%
    furrr::future_map(~ readRDS(.) %>% broom::tidy()) %>%
    bind_rows(.id = "model_info") %>%
    janitor::clean_names() %>%
    mutate(r2_percent = round(r2 * 100, 3)) %>%
    arrange(p_value) %>%
    glimpse()

# Visualize PERMANOVA results
permanova_results %>% glimpse()

# permanova_results %>%
#     filter(term %nin% c("Residual", "Total")) %>%
#     arrange(p_value) %>%
#     slice_min(p_value, n=100) %>% 
#     View

passing_meta <- permanova_results %>%   
    filter(term %nin% c("Residual", "Total")) %>% 
    filter(p_value <= 0.05) %>% 
    pull(term) %>% unique

p_permanova <- permanova_results %>%
    filter(term %nin% c("Residual", "Total")) %>%
    filter(term  %in% passing_meta) %>% 
    arrange(p_value) %>%
    mutate(
        significant = p_value < 0.05,
        p_value_formatted = format.pval(p_value, digits = 3)
    ) %>%
    ggplot(aes(y = fct_reorder(term, r2), x = r2 * 100)) +
    geom_point(aes(color = significant),
        size = 0.8,
        position = position_jitter(height = 0.2, width = 0), alpha = 0.85
    ) +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.5) +
    scale_color_manual(values = c("grey70", "red")) +
    theme_bw() +
    labs(
        title = "PERMANOVA Results: TCR Repertoire Similarity",
        x = "R² % (Effect Size)",
        y = NULL
    ) +
    theme(legend.position = "bottom")

ggsave(
    glue("{imm_plots_dir}/tcr_permanova_results_public_overlap_{Sys.Date()}.png"),
    p_permanova,
    width = 8,
    height = 20
)


# Summarize
perm_summary_df <- permanova_results %>%
  group_by(term) %>%
  summarise(
    mean_log10_p = mean(-log10(p_value), na.rm = TRUE),
    sd_log10_p = sd(-log10(p_value), na.rm = TRUE),
    mean_r2_percent = mean(r2 * 100, na.rm = TRUE),
    sd_r2_percent = sd(r2 * 100, na.rm = TRUE),
    mean_statistic = mean(statistic, na.rm = TRUE),
    sd_statistic = sd(statistic, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  filter(!grepl("guid|GUID", term)) %>% 
  glimpse()

# Flag terms above threshold
pval_thresh <- -log10(0.05)
labeled_points <- perm_summary_df %>%
  filter(mean_log10_p > pval_thresh)

View(labeled_points)

# Plot
p_perm_base <- ggplot(perm_summary_df, aes(x = mean_log10_p, y = mean_r2_percent)) +
  geom_point(alpha = 0.8,) +
  geom_errorbar(aes(ymin = mean_r2_percent - sd_r2_percent,
                    ymax = mean_r2_percent + sd_r2_percent),
                width = 0.01, alpha = 0.5) +
  geom_errorbarh(aes(xmin = mean_log10_p - sd_log10_p,
                     xmax = mean_log10_p + sd_log10_p),
                 height = 0.01, alpha = 0.5) +
    scale_y_log10() +
    geom_vline(xintercept = -log10(0.05)) +
  labs(
    x = expression("Mean -log"[10]*"(p-value)"),
    y = "Mean R² (%)",
    title = "PERMANOVA Summary by Term"
  ) +
  theme_bw()

p_perm_annot <- p_perm_base +
  ggrepel::geom_text_repel(
    data = labeled_points,
    aes(label = term),
    size = 3,
    max.overlaps = Inf,
    force = 50,
    force_pull = 0,
    nudge_x = ifelse(labeled_points$mean_log10_p > 0, 1, -1),
    nudge_y = 1,
    box.padding = 0.8,
    point.padding = 0.5,
    segment.color = "grey",
    segment.size = 0.2,
    segment.alpha = 0.75,
    direction = "both",
    segment.curvature = 0.3,
    segment.ncp = 3,
    segment.angle = 10
  )

ggsave(
    glue("{imm_plots_dir}/tcr_permanova_results_public_overlap_summary_base_{Sys.Date()}.png"),
    p_perm_base,
    width = 10,
    height = 10
)
ggsave(
    glue("{imm_plots_dir}/tcr_permanova_results_public_overlap_summary_annot_{Sys.Date()}.png"),
    p_perm_annot,
    width = 10,
    height = 10
)

```




PERMANOVA on Morisita Horn distance

# TODO FIXME Swap out data with MH index instead and change up code

```{r}
pub_ovov <- readRDS(
    glue("{metrics_dir}/rep-overlap-overlap_2024-12-19.rds")
)
pub_ovov_df <- pub_ovov %>% as.data.frame() %>% 
    glimpse()

pub_ovov_df_long <- pub_ovov_df %>% 
    rownames_to_column("sample_id1") %>%
    pivot_longer(
        cols = -sample_id1,
        names_to = "sample_id2",
        values_to = "overlap_counts"
    ) %>%
    mutate(overlap_counts = replace_na(overlap_counts, 1)) %>%
    glimpse()

# Convert the long format data back to a matrix
pub_ovov_matrix <- pub_ovov_df_long %>%
  pivot_wider(
    id_cols = sample_id1,
    names_from = sample_id2,
    values_from = overlap_counts
  ) %>%
  column_to_rownames("sample_id1") %>%
  as.matrix()

distance_matrix <- 1 - pub_ovov_matrix

# Define metadata columns for PERMANOVA
key_meta_df %>% glimpse()
full_meta_df %>% glimpse()

target_metadata <- colnames(full_meta_df) %>% 
    discard(. %in% c("sample_id"))

# Get sample metadata for PERMANOVA
permanova_metadata <- full_meta_df %>%
    filter(!is.na(case_control_other_latest)) %>%
    filter(sample_id %in% rownames(distance_matrix)) %>%
    column_to_rownames("sample_id") %>%
    select(all_of(target_metadata), participant_id) %>%
    glimpse()

subsample_resdir <- glue("{metrics_dir}/permanova_public_overlap_sid_subsample")
subsample_resdir_input <- glue("{subsample_resdir}/model_input")
perm_resdir <- glue("{subsample_resdir}/results")
dir.create(subsample_resdir_input, showWarnings = FALSE, recursive = TRUE)
dir.create(perm_resdir, showWarnings = FALSE, recursive = TRUE)
# Saving subsampled distance matrix and metadata
common_samples <- intersect(
    rownames(distance_matrix), rownames(permanova_metadata)
)


batch_params <- expand.grid("var" = target_metadata, "i" = 1:50) %>% 
    mutate(
        dmat_path = glue(
            "{subsample_resdir_input}/",
            "distance_matrix_filtered_subsample_{i}.rds"
        ),
        meta_path = glue(
            "{subsample_resdir_input}/",
            "permanova_metadata_filtered_subsample_{i}.rds"
        ),
        model_formula_string = glue("distance_matrix ~ {var}"),
        outpath = glue("{perm_resdir}/",
            "permanova_{var}_nperms_{nperms}_subsample_{i}_{Sys.Date()}.rds"),
        n_perms = 9999,
        n_cores = 2        
    ) %>%
    dplyr::select(-c(var, i)) %>% 
    glimpse()

# configure registry ----
breg <- breg_setup(cluster_name = "PERM_POVOV", 
    log_dir = glue("{wkdir}/.cluster_runs/"),
    template_path = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl")
    )

# Submit Jobs ----
jobs <- batchMap(
  fun = run_permanova,
  args = batch_params[21:26800, ],
  reg = breg
)
nrow(batch_params)

jobs[, chunk := chunk(job.id, chunk.size = 10)]
print(jobs[, .N, by = chunk])

submitJobs(jobs,
  resources = list(
    walltime = "2-00:00:00",
    memory = "50GB",
    ncpus = 2,
    max.concurrent.jobs = 9999
  )
)


```



What fraction of receptors are clones between timepoints within an individual?

overlap coefficient (.method = "overlap") - a normalised measure of overlap similarity. It is defined as the size of the intersection divided by the smaller of the size of the two sets.

```{r}
pub_ovov <- readRDS(
    glue("{metrics_dir}/rep-overlap-overlap_2024-12-19.rds")
)

overlap_df <- pub_ovov %>% as.data.frame() %>% 
    glimpse()

target_metadata <- c(
    "sample_id", 
    "case_control_other_latest", 
    "participant_id", 
    "visit_month", 
    "study"
    )

overlap_df_long <- overlap_df %>%
  rownames_to_column("sample_id1") %>%
  pivot_longer(
    cols = -sample_id1,
    names_to = "sample_id2", 
    values_to = "overlap_score"
  ) %>%
  drop_na(overlap_score) %>%
  # Join metadata for first sample
  left_join(
    key_meta_df %>% 
      select(all_of(target_metadata)) %>%
      rename_with(~paste0(., "1"), everything()),
    by = c("sample_id1")
  ) %>%
  # Join metadata for second sample  
  left_join(
    key_meta_df %>%
      select(all_of(target_metadata)) %>%
      rename_with(~paste0(., "2"), everything()),
    by = c("sample_id2")
  ) %>%
  filter(case_control_other_latest1 != "Other") %>%
  filter(case_control_other_latest2 != "Other") %>%
  filter(participant_id1 == participant_id2) %>%
  filter(sample_id1 != sample_id2) %>%
  mutate(timepoint_delta = visit_month2 - visit_month1) %>%
  mutate(timepoint_delta_cat = case_when(
      timepoint_delta < 0 ~ "before",
      timepoint_delta > 0 ~ "after", 
      TRUE ~ "same"
    )
  )

first_timepoint_ref <- overlap_df_long %>% 
    # filter(participant_id1 == participant_id2) %>% 
    filter(timepoint_delta_cat == "after") %>%
    group_by(participant_id1, visit_month1) %>%
    slice_min(order_by = visit_month1, n = 1) %>%
    glimpse()

# Does distance between timepoints correlate with overlap coefficient?
p_first_timepoint_ref <- first_timepoint_ref %>% 
    ggplot(aes(x = timepoint_delta, y = overlap_score, 
        color = case_control_other_latest1)) +
    geom_point(
        position = position_jitter(width = 1.6, height = 0), 
        alpha = 0.4, size = 0.5) +
    geom_smooth(method = "lm") +
    labs(x = "Timepoint delta [after - before] (months)", 
         y = "Overlap coefficient") +
    ggpubr::stat_regline_equation(
        label.x.npc = 0.02,
        label.y.npc = c(0.95, 0.9),
        aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~")),
        formula = y ~ x
    ) +
    scale_color_d3() +
    theme_bw() +
    guides(color = guide_legend(override.aes = list(text = 0))) +
    labs(color = NULL) +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/public_overlap_first_timepoint_ref_overlap_coef_{Sys.Date()}.png"),
    p_first_timepoint_ref,
    width = 5,
    height = 4
)


# Now if we allow all timedelta comparisons within a participant

overlap_df_long %>% glimpse()
overlap_df_long %>% 
    filter(participant_id1 == "PD-PDAA503EF5") %>%
    View()

cross_timepoint_ref <- overlap_df_long %>% 
    # filter(timepoint_delta_cat == "after") %>%
    glimpse()

# Does distance between timepoints correlate with overlap coefficient?
p_cross_timepoint_ref <- cross_timepoint_ref %>% 
    ggplot(aes(x = timepoint_delta, y = overlap_score, 
        color = case_control_other_latest1)) +
    geom_point(
        position = position_jitter(width = 1.6, height = 0), 
        alpha = 0.4, size = 0.5) +
    geom_smooth(method = "loess", se = TRUE) +
    labs(x = "Timepoint delta [after - before] (months)", 
         y = "Overlap coefficient") +
    scale_color_d3() +
    theme_bw() +
    guides(color = guide_legend(override.aes = list(text = 0))) +
    labs(color = NULL) +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/public_overlap_cross_timepoint_ref_overlap_coef_{Sys.Date()}.png"),
    p_cross_timepoint_ref,
    width = 5,
    height = 4
)



```





```{r}
pub_ov <- readRDS(
    glue("{metrics_dir}/rep-overlap-public_2024-12-19.rds")
)

public_overlap_df <- pub_ov %>% as.data.frame() %>% 
    glimpse()

target_metadata <- c(
    "sample_id", 
    "case_control_other_latest", 
    "participant_id", 
    "visit_month", 
    "study",
    "DRB1_risk"
    )

public_overlap_df_long <- public_overlap_df %>%
  rownames_to_column("sample_id1") %>%
  pivot_longer(
    cols = -sample_id1,
    names_to = "sample_id2",
    values_to = "overlap_score"
  ) %>%
  drop_na(overlap_score) %>%
  # Join metadata for first sample
  left_join(
    key_meta_df %>% 
      select(all_of(target_metadata)) %>%
      rename_with(~paste0(., "1"), everything()),
    by = c("sample_id1")
  ) %>%
  # Join metadata for second sample  
  left_join(
    key_meta_df %>%
      select(all_of(target_metadata)) %>%
      rename_with(~paste0(., "2"), everything()),
    by = c("sample_id2")
  ) %>%
  mutate(
    timepoint_delta = case_when(
      participant_id1 == participant_id2 ~ visit_month2 - visit_month1,
      TRUE ~ NA_real_
    )
  ) %>% 
  mutate(
    timepoint_delta_cat = case_when(
      timepoint_delta < 0 ~ "before",
      timepoint_delta > 0 ~ "after",
      TRUE ~ "same"
    )
  ) %>% 
    mutate(disease_condition = case_when(
        participant_id1 == participant_id2 & case_control_other_latest1 == "Case" ~ "Within\nCase (Donors)",
        participant_id1 == participant_id2 & case_control_other_latest1 == "Control" ~ "Within\nControl (Donors)",
        case_control_other_latest1 == "Case" & case_control_other_latest2 == "Case" ~ "Between\nCase",
        case_control_other_latest1 == "Control" & case_control_other_latest2 == "Control" ~ "Between\nControl",
        case_control_other_latest1 == "Case" & case_control_other_latest2 == "Control" ~ "Between\nCase-Control",
        case_control_other_latest1 == "Control" & case_control_other_latest2 == "Case" ~ "Between\nCase-Control",
        TRUE ~ "ERROR"
    )) %>%
    mutate(disease_condition = factor(disease_condition, 
        levels = c(
            "Within\nCase (Donors)",
            "Within\nControl (Donors)",
            "Between\nCase",
            "Between\nControl",
            "Between\nCase-Control"
        ))) %>% 
    mutate(HLA_DRB1_risk_grouping = case_when(
        DRB1_risk1 == "risk" & DRB1_risk2 == "risk" ~ "Witihin\nRisk",
        DRB1_risk1 == "protective" & DRB1_risk2 == "protective" ~ "Witihin\nProtective",
        DRB1_risk1 == "mixed" & DRB1_risk2 == "mixed" ~ "Witihin\n Mixed",
        (DRB1_risk1 == "risk" & DRB1_risk2 == "protective") | 
            (DRB1_risk1 == "protective" & DRB1_risk2 == "risk") ~ "Between\nRisk & Protective",
        DRB1_risk1 == "none" & DRB1_risk2 == "none" ~ "Within\nDRX/DRX",
        TRUE ~ "OTHER"
    )) %>%
    glimpse()

public_overlap_df_long %>% glimpse()
public_overlap_df_long$overlap_score %>% summary()

within_pid_overlap <- public_overlap_df_long %>% 
    filter(participant_id1 == participant_id2, 
        timepoint_delta_cat == "after") %>%
        group_by(sample_id1, visit_month1) %>%
        slice_min(order_by = timepoint_delta, n = 1) %>%
    glimpse()

# Does distance between timepoints correlate with overlap?
p_public_overlap <- within_pid_overlap %>% 
    ggplot(aes(x = timepoint_delta, y = overlap_score, 
        color = case_control_other_latest1)) +
    geom_point(position = position_jitter(width = 1.6, height = 0), alpha = 0.2, size = 0.5) +
    geom_smooth(method = "loess", se = TRUE) +
    labs(x = "Timepoint delta [after - before] (months)", y = "Shared clones \n within donors") +
    scale_color_d3() +
    ggpubr::stat_regline_equation(
        label.x.npc = 0.02,
        label.y.npc = c(0.9, 0.85),
        aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~")),
        formula = y ~ x
    ) +
    guides(color = guide_legend(override.aes = list(text = 0))) +
    theme_bw() +
    theme(legend.position = "top")

ggsave(
    glue("{imm_plots_dir}/public_overlap_within_pid.png"),
    p_public_overlap,
    width = 5,
    height = 4 
)

  # Keep only one direction of each pair by comparing string values
# public_overlap_df_long %>%
#   filter(stringr::str_compare(sample_id1, sample_id2) < 0) %>%
#   glimpse
    # filter(
    #     (sample_id1 == "BF-1001-SVM0_5T1" & sample_id2 == "BF-1002-SVM0_5T1") | 
    #     (sample_id1 == "BF-1002-SVM0_5T1" & sample_id2 == "BF-1001-SVM0_5T1")
    # )



```


Individual participants should have greater overlap between timepoints than between participants

```{r}
public_overlap_df_long %>% 
    mutate(pair_desc = case_when(
        participant_id1 == participant_id2 ~ "within_pid",
        TRUE ~ "between_pid"
    )) %>%
    group_by(pair_desc) %>%
    summarise(mean_overlap = mean(overlap_score)) %>%
    glimpse()

```


```{r}
# Pulling total unique clones per sample from Hill diversity
unique_clones_per_sample <- readRDS( 
    glue("{metrics_dir}/rep-diversity-hill_2024-12-19_plus_q0.rds")) %>% 
    filter(Q == 0) %>% 
    mutate(sample_id1 = Sample, sample_id2 = Sample) %>% 
    dplyr::rename(overlap_counts = Value) %>% 
    dplyr::select(sample_id1, sample_id2, overlap_counts) %>% 
    glimpse()

public_overlap_df_long_full <- public_overlap_df %>% 
    rownames_to_column("sample_id1") %>%
    pivot_longer(
        cols = -sample_id1,
        names_to = "sample_id2",
        values_to = "overlap_counts"
    ) %>%
    # Create a full join with unique_clones_per_sample
    full_join(unique_clones_per_sample, 
              by = c("sample_id1", "sample_id2"),
              suffix = c("", "_unique")) %>%
    # Give priority to unique_clones_per_sample values
    mutate(overlap_counts = coalesce(overlap_counts_unique, overlap_counts)) %>%
    select(-overlap_counts_unique) %>%
    glimpse()

public_overlap_df_long_full %>% glimpse()


# Convert the long format data back to a matrix
public_overlap_matrix_full <- public_overlap_df_long_full %>%
  pivot_wider(
    id_cols = sample_id1,
    names_from = sample_id2,
    values_from = overlap_counts
  ) %>%
  column_to_rownames("sample_id1") %>%
  as.matrix()

# Display the first few rows and columns
public_overlap_matrix_full[1:5, 1:5]

# # Seriating the matrix
# sid_order_ov_counts <- seriate_matrix_rows(public_overlap_matrix_full, 
#     seriate_method = "HC_average")

# saveRDS(
#     sid_order_ov_counts,
#     glue("{metrics_dir}/repertoire_stats/sid_order_ov_counts_{Sys.Date()}.rds")
# )

sid_order_ov_counts <- readRDS(
    glue("{metrics_dir}/repertoire_stats/sid_order_ov_counts_2025-03-24.rds")
)

p_overlap_heatmap <- public_overlap_df_long_full %>%
    mutate(sample_id1 = factor(sample_id1, levels = sid_order_ov_counts)) %>%
    mutate(sample_id2 = factor(sample_id2, levels = sid_order_ov_counts)) %>%
    filter(sample_id1 != sample_id2) %>%
    ggplot(aes(x = sample_id1, y = sample_id2, fill = overlap_counts)) +
    geom_tile() +
    scale_fill_viridis_c(
        option = "mako",
        trans = "log10",
        name = "Overlap Count",
        na.value = "white"
    ) +
    labs(x = NULL, y = NULL, fill = "Overlap Count") +
    theme_bw() +
    theme(
        axis.text.x = element_blank(),
        axis.text.y = element_blank()
    )

ggsave(
    glue("{imm_plots_dir}/public_overlap_heatmap_{Sys.Date()}.png"),
    p_overlap_heatmap,
    width = 10,
    height = 8
)

```



Testing for overlap in McPas and VDJDB


```{r}
mcpas = dbLoad("https://gitlab.com/immunomind/immunarch/raw/dev-0.5.0/private/McPAS-TCR.csv.gz", "mcpas", .species = "Human", .chain = "TRB")
mcpas
```
