---
title: "Adaptive Immune Receptor Repertoire in Parkinson's Disease: Global Diversity Analysis"
editor: source
author: "Joe Boktor"
date: '2025-07-08'
format: 
  html:
    font-family: helvetica neue
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 4
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
---

Environment setup

```{r setup, message=FALSE, warning=FALSE}
home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))

# Additional libraries
library(data.table)

# Plotting libraries
library(ggsci)
library(ggrepel)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(gt)
library(gtsummary)
library(immunarch)
library(nlme)

metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
imm_plots_dir <- glue("{wkdir}/figures/AIRR/immunarch")
```

Loading metadata and TCRB data

```{r, eval=FALSE}
metadata <- readRDS(
  glue("{wkdir}/data/interim/metadata/compiled_metadata_v4_release_2025-07-21.rds")) %>% 
  dplyr::filter(case_control_other_latest != "Other") %>% 
  glimpse()

key_meta_df <- metadata %>% 
  dplyr::select(
    sample_id, participant_id, case_control_other_latest, 
    age_at_baseline, sex, abis_t_cell, ethnicity, race, visit_month, study,
    aligned_reads, est_tcell_reads_abis, pd_hla_risk_score, pd_hla_risk_score_z
  ) %>% 
  glimpse()

trust4_reports_tcrb <- readRDS(
  glue("{wkdir}/data/interim/airr/trust4_reports_tcrb_clean_df_2025-07-08.rds")
) %>% 
  glimpse()
```

# Collecting Diverstiy Metrics with Immunarch  

### Formatting TRUST4 TCRB data for Immunarch

```{r}
immunarch_formatted <- trust4_reports_tcrb %>%
  dplyr::rename(
    Clones = `#count`,
    Proportion = frequency,
    CDR3.nt = CDR3nt,
    CDR3.aa = CDR3aa,
    V.name = V,
    D.name = D,
    J.name = J
  ) %>%
  dplyr::mutate(
    VJ.ins = NA_character_,
    VD.ins = NA_character_,
    DJ.ins = NA_character_,
    V.end = NA_integer_,
    D.start = NA_integer_,
    D.end = NA_integer_,
    J.start = NA_integer_,
    Sequence = NA_character_
  ) %>%
  glimpse()

# Split data into immunarch format and metadata
immunarch_data <- immunarch_formatted %>%
  dplyr::select(Clones, Proportion, CDR3.nt, CDR3.aa, V.name, D.name, J.name,
                VJ.ins, VD.ins, DJ.ins, Sequence) %>% 
  glimpse()

metadata_immunarch <- immunarch_formatted %>%
  dplyr::select(C, cid, cid_full_length, sample_id) %>% 
  left_join(metadata) %>% 
  glimpse()

# Create list structure expected by immunarch
tictoc::tic("Creating immunarch object")
immdata <- list(
  data = split(immunarch_data, immunarch_formatted$sample_id),
  meta = metadata_immunarch
)
tictoc::toc()

saveRDS(
    immdata,
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_TCRB_{Sys.Date()}.rds"),
    compress = FALSE
)

```


### Executing Immunarch functions

Note that this section takes about 4-6 hours to run.

```{r}
future::plan(future::multisession, workers = 30)
immdata <- readRDS(
    glue("{wkdir}/data/interim/airr/immunarch/immunarch-obj_TCRB_2025-07-08.rds")
)
immdata %>% glimpse()

#' Computes basic statistics, 
#' including clones/distributions of lengths and counts.
#' https://immunarch.com/articles/web_only/v3_basic_analysis.html
clonal_metrics = c("volume", "count", "len", "clones")
for (m in clonal_metrics) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repExplore(immdata$data, .method = m, .col = c("aa")) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-explore-{m}_{Sys.Date()}.rds")
    )
}

# Exploring Clonality
# https://immunarch.com/reference/repClonality.html
clone_mets <- c("clonal.prop", "homeo", "top", "rare")
for (m in clone_mets) {
    message(glue("Calculating {m} clonal metrics"))
    metric_data <- repClonality(immdata$data, .method = m) %>% 
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-clonality-{m}_{Sys.Date()}.rds"
            )
    )
}

# Gene Usage Analysis
# https://immunarch.com/reference/geneUsage.html
gene_metrics <- c("hs.trbv", "HomoSapiens.TRBJ")
for (g in gene_metrics) {
    message(glue("Calculating {g} gene usage metrics"))
    metric_data <- geneUsage(immdata$data, g) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/gene-usage-{g}_{Sys.Date()}.rds"
            )
    )
}

# Analyzing Repertoire Diversity
# https://immunarch.com/articles/web_only/v6_diversity.html
div_metrics <- c(
    # "chao1", "hill", 
    # "div", "gini.simp", "inv.simp", 
    # "gini", 
    "d50" #"raref",
)
for (d in div_metrics) {
    message(glue("Calculating {d} diversity metrics"))
    metric_data <- repDiversity(immdata$data, .method = d) %>%
        glimpse()
    saveRDS(
        metric_data,
        glue("{metrics_dir}/rep-diversity-{d}_{Sys.Date()}.rds"
            )
    )
}

```


# Diversity Analysis of Beta Chain Repertoires

V-gene usage analysis

```{r}
# Visualizing v-gene usage
v_gene_usage <- readRDS(
    glue("{metrics_dir}/gene-usage-hs.trbv_2025-07-09.rds")
)
v_gene_usage_long <- v_gene_usage %>%
    pivot_longer(-Names, names_to = "sample_id", values_to = "counts") %>%
    mutate(counts = ifelse(is.na(counts), 0, counts)) %>%
    left_join(key_meta_df, by = c("sample_id")) %>%
    filter(case_control_other_latest != "Other") %>%
    group_by(sample_id) %>%
    mutate(vgenefreq = counts / sum(counts) * 100) %>%
    glimpse()

p_v_gene_usage <- v_gene_usage_long %>%
    ggplot(aes(
        x = fct_reorder(Names, vgenefreq, .desc = TRUE),
        y = vgenefreq
    )) +
    geom_point(aes(color = case_control_other_latest),
        position = position_jitterdodge(), alpha = 0.1, size = 0.05) +
    geom_boxplot(
        aes(group = interaction(Names, case_control_other_latest)),
        outlier.alpha = 0, # fill.alpha = 0.2,
        position = position_dodge(width = 0.75), alpha = 0.3, color = "black"
    ) +
    scale_color_d3() +
    scale_fill_d3() +
    theme_bw() +
    labs(
        x = "V-Gene", y = "V-Gene Frequency (%)",
        fill = "Disease Status",
        color = "Disease Status"
    ) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top"
    )

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_{Sys.Date()}.png"),
    plot = p_v_gene_usage,
    width = 14,
    height = 6
)

# Grouping by higher level gene id
v_gene_usage_long %>% glimpse()

v_gene_usage_long_broad <- v_gene_usage_long %>%
    mutate(vgenebroad = str_extract(Names, "^[A-Z0-9]+")) %>%
    group_by(sample_id, vgenebroad) %>%
    dplyr::summarise(counts = sum(counts)) %>%
    group_by(sample_id) %>%
    mutate(vgenebroadfreq = counts / sum(counts) * 100) %>%
    left_join(key_meta_df, by = c("sample_id")) %>%
    filter(case_control_other_latest != "Other") %>%
    mutate(case_control_other_latest = factor(case_control_other_latest, 
        levels = c("Control", "Case"))) %>%
    glimpse()

vgenebroad_order <- v_gene_usage_long_broad %>% 
    group_by(vgenebroad) %>% 
    dplyr::summarise(med_freq = median(vgenebroadfreq)) %>% 
    arrange(desc(med_freq)) %>% 
    pull(vgenebroad)

p_v_gene_usage_broad <- v_gene_usage_long_broad %>%
    filter(case_control_other_latest != "Other") %>%
    mutate(vgenebroad = factor(vgenebroad, levels = vgenebroad_order)) %>%
    ggplot(aes(
        x = vgenebroad,
        y = vgenebroadfreq, color = case_control_other_latest
    )) +
    geom_point(aes(color = case_control_other_latest),
        position = position_jitterdodge(), alpha = 0.1, size = 0.05) +
    geom_boxplot(
        aes(group = interaction(vgenebroad, case_control_other_latest)),
        outlier.alpha = 0, # fill.alpha = 0.2,
        position = position_dodge(width = 0.75), alpha = 0.3, color = "black"
    ) +
    scale_color_d3() +
    scale_fill_d3() +
    theme_bw() +
    labs(
        x = "V-Gene", y = "V-Gene Frequency (%)",
        fill = "Disease Status",
        color = "Disease Status"
    ) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_broad_{Sys.Date()}.png"),
    plot = p_v_gene_usage_broad,
    width = 14,
    height = 6
)

vgene_data %>% glimpse()
unique_vgenes <- unique(v_gene_usage_long_broad$vgenebroad)
lmm_results <- list()
# lmm_anova_res <- list()

# Iterate through each vgenebroad
for (vgene in unique_vgenes) {
    # Subset data for current vgene
    vgene_data <- v_gene_usage_long_broad %>%
        filter(vgenebroad == vgene)

    # Skip if not enough data
    if (nrow(vgene_data) < 10) {
        message(glue("Insufficient data for {vgene}"))
        lmm_results[[vgene]] <- NULL
        next
    }

    # Run linear mixed model
    tryCatch(
        {
            lmm_model <- nlme::lme(vgenebroadfreq ~ case_control_other_latest + age_at_baseline + sex + est_tcell_reads_abis + study,
                random = ~ 1 | participant_id,
                data = vgene_data,
                control = lmeControl(opt = "optim")
            )

            # Store model summary
            lmm_results[[vgene]] <- lmm_model

            # Print basic results
            cat("Results for", vgene, ":\n")
            print(anova(lmm_model))
            cat("\n")
        },
        error = function(e) {
            cat("Error in model for", vgene, ":", e$message, "\n\n")
            lmm_results[[vgene]] <- NULL
        }
    )
}

vgene_lmm_res <- lmm_results %>%
    purrr::set_names(names(.)) %>%
    purrr::map(~ broom.mixed::tidy(.)) %>%
    bind_rows(.id = "vgene") %>%
    glimpse()

vgene_lmm_res %>% 
    filter(term == "case_control_other_latestCase") %>% 
    arrange(p.value) 


vgene_lmm_res_better <- tibble()
for (v in names(lmm_results)) {
    d <- lmm_results[[v]] %>% anova() %>% 
        rownames_to_column(var = "term") %>% 
        mutate(vgene = v)
    vgene_lmm_res_better %<>% bind_rows(d)
}
vgene_lmm_res_better %>% glimpse()

vgene_lmm_res_better <- vgene_lmm_res_better %>%
    dplyr::rename(p.value = `p-value`, f.value = `F-value`) %>% 
    filter(term == "case_control_other_latest") %>%
    arrange(p.value) %>% 
    glimpse()

# vgene_lmm_res_better %>% View()

# Create a plot of LMM results
vgene_lmm_plot <- vgene_lmm_res_better %>%
    # filter(term == "case_control_other_latestCase") %>%
    # Match the order of vgenes from the previous plot
    mutate(vgene = factor(vgene, levels = vgenebroad_order)) %>%
    # Add significance flag
    mutate(is_significant = p.value <= 0.05) %>%
    ggplot(aes(x = vgene, y = f.value)) +
    geom_point(aes(color = is_significant)) +
    # geom_pointrange(aes(
    #     ymin = estimate - std.error, 
    #     ymax = estimate + std.error,
    #     color = is_significant
    # ), size = 0.2) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_bw() +
    labs(x = NULL, y = "F-Statistic",
         title = "Linear Mixed Model Results for V Gene Usage") +
    theme(axis.text.x = element_blank(), legend.position = "top") +
    guides(color = "none")

# Stack the plots vertically using aplot
p_v_gene_usage_broad_lmm <- p_v_gene_usage_broad %>%
    aplot::insert_top(vgene_lmm_plot, 
                     height = 0.3)

ggsave(
    filename = glue("{wkdir}/figures/AIRR/immunarch/v_gene_usage_broad_lmm_{Sys.Date()}.png"),
    plot = p_v_gene_usage_broad_lmm,
    width = 12,
    height = 6
)

# vgene_lmm_res %>%
#     filter(term == "case_control_other_latestCase") %>% 
#     DT::datatable()

lmm_results[["TRBV27"]] %>% anova()
lmm_results[["TRBV27"]] %>% broom.mixed::tidy() %>% print
# lmm_results[["TRBV27"]] %>% broom.mixed::tidy() %>% 
#     filter(term == "case_control_other_latestCase") %>% print()

key_meta_df %>% glimpse()

v_gene_usage_long_broad %>% 
    filter(vgenebroad %in% c("TRBV14", "TRBV29", "TRBV3", "TRBV25")) %>% 
    left_join(key_meta_df) %>% 
    ggplot(aes(x = study, y = vgenebroadfreq, color = case_control_other_latest)) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2), alpha = 0.3, size = 0.2) +
    geom_boxplot(
        aes(group = interaction(study, case_control_other_latest)),
        outlier.alpha = 0, # fill.alpha = 0.2,
        position = position_dodge(width = 0.75), alpha = 0.3, color = "black"
    ) +
    scale_color_d3() +
    facet_wrap(~ vgenebroad, scales = "free_y") +
    theme_bw()


```


Since data is not log-scaled and is in percentages, the interpretation of this result is that only one V-gene has a significant difference in frequency between control and case. However, the effect size is about a 0.11% difference in frequency between control and case - which is unlikely to be biologically meaningful.
