---
title: "ClusTCR Stats EDA (Minimal Load)"
editor: source
author: "Joe Boktor"
date: '2026-04-14'
eval: false
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14
---

# ClusTCR Stats EDA (Minimal)

This notebook only loads the core data needed to build the merged clusTCR dataset:
- clusTCR input TCRs
- CDR3 risk scores
- OLGA/SONIA generation probabilities
- clusTCR output + summary

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(glue)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
wkdir <- paste0(pdmbs_dir, "/pdairr")

clustcr_dir <- glue("{wkdir}/data/interim/clusTCR")
tcr_cdr3_riskscore_complied <- glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")
dint <- glue("{wkdir}/data/interim")

risk_score_rds <- Sys.glob(glue("{tcr_cdr3_riskscore_complied}/risk_scores_padj05_*.rds")) %>%
  sort() %>%
  tail(1)
stopifnot(length(risk_score_rds) == 1, file.exists(risk_score_rds))

logreg_lrt_rds <- Sys.glob(glue("{clustcr_dir}/results/logistic_regression_binary_anova_lrt_pval_df_*.rds")) %>%
  sort() %>%
  tail(1)
stopifnot(length(logreg_lrt_rds) == 1, file.exists(logreg_lrt_rds))

logreg_fullstats_rds <- Sys.glob(glue("{clustcr_dir}/results/logistic_regression_binary_full_model_stats_df_*.rds")) %>%
  sort() %>%
  tail(1)
stopifnot(length(logreg_fullstats_rds) == 1, file.exists(logreg_fullstats_rds))

stopifnot(file.exists(glue("{clustcr_dir}/input/amppd_tcrb_2025-08-04.tsv")))
stopifnot(file.exists(glue("{clustcr_dir}/results/clusTCR_output_vgene_restricted_31cpus.csv")))
stopifnot(file.exists(glue("{clustcr_dir}/results/clusTCR_summary_vgene_restricted_31cpus.csv")))
stopifnot(file.exists(glue("{dint}/olga_sonnia/results/tcrb_case_results/tcrb_case_results.csv")))
stopifnot(file.exists(glue("{dint}/olga_sonnia/results/tcrb_ctrl_results/tcrb_ctrl_results.csv")))

```

## Load and join TCR-level annotations

```{r load-tcr-risk-genprob}
clustcr_cdr3b_df <- readr::read_tsv(
  glue("{clustcr_dir}/input/amppd_tcrb_2025-08-04.tsv")
) %>%
  dplyr::rename(case_control_other_latest = condition)

cdr3_risk_score_df <- readRDS(
  risk_score_rds
) %>%
  distinct() %>%
  dplyr::rename(CDR3b_aa = CDR3aa)

genprob_case <- read_csv(
  glue("{dint}/olga_sonnia/results/tcrb_case_results/tcrb_case_results.csv")
) %>%
  mutate(case_control_other_latest = "Case") %>%
  dplyr::rename(CDR3b_aa = amino_acid, TRBV = v_gene, TRBJ = j_gene)

genprob_ctrl <- read_csv(
  glue("{dint}/olga_sonnia/results/tcrb_ctrl_results/tcrb_ctrl_results.csv")
) %>%
  mutate(case_control_other_latest = "Control") %>%
  dplyr::rename(CDR3b_aa = amino_acid, TRBV = v_gene, TRBJ = j_gene)

genprob <- bind_rows(genprob_case, genprob_ctrl)

tcrb_df <- clustcr_cdr3b_df %>%
  left_join(cdr3_risk_score_df, by = "CDR3b_aa") %>%
  left_join(genprob, by = c("CDR3b_aa", "TRBV", "TRBJ", "case_control_other_latest"))

tcrb_df %>% glimpse()

```

## Load clusTCR outputs and build merged dataset

```{r build-compiled-clustcr-results}
clustcr_output <- read_csv(
  glue("{clustcr_dir}/results/clusTCR_output_vgene_restricted_31cpus.csv"),
  col_select = -1
)
clustcr_output %>% glimpse()

clustcr_summary <- read_csv(
  glue("{clustcr_dir}/results/clusTCR_summary_vgene_restricted_31cpus.csv")
) %>%
  dplyr::rename(cluster = ...1)

clustcr_summary$cluster <- unique(clustcr_output$cluster)
clustcr_summary %>% glimpse()

clustcr_results_df <- clustcr_output %>%
  left_join(clustcr_summary, by = "cluster") %>%
  left_join(
    tcrb_df,
    by = c("junction_aa" = "CDR3b_aa", "v_call" = "TRBV"),
    relationship = "many-to-many"
  ) %>%
  group_by(cluster) %>%
  mutate(
    unique_donors_per_clust = n_distinct(participant_id),
    unique_samples_per_clust = n_distinct(sample_id)
  ) %>%
  ungroup() %>%
  mutate(
    motif_length = motif %>%
      gsub("\\[.*?\\]", "X", .) %>%
      gsub("\\.", "", .) %>%
      gsub("[^A-Z]", "", .) %>%
      nchar(),
    cluster_id = glue("{motif}_{v_call}")
  )

clustcr_results_df %>% glimpse()
```

## Load logistic / ANOVA-LRT results and pair to clusters

```{r load-logreg-lrt-and-pair}
# ANOVA-LRT p-values from binary logistic cluster models
anova_lrt_pval_df <- readRDS(
  logreg_lrt_rds
) %>%
  dplyr::rename(anova_lrt_pval = p_value)

# Full model coefficient/stat summaries (includes cluster_id)
logreg_full_model_stats_df <- readRDS(
  logreg_fullstats_rds
)

# Cluster-level table paired with both logistic result objects
cluster_level_w_logreg <- clustcr_results_df %>%
  dplyr::select(
    cluster_id, cluster, motif, v_call, motif_length,
    unique_donors_per_clust, unique_samples_per_clust
  ) %>%
  distinct() %>%
  left_join(anova_lrt_pval_df, by = "cluster_id") %>%
  left_join(logreg_full_model_stats_df, by = "cluster_id")

cluster_level_w_logreg %>% glimpse()
```

## Save EDA-ready dataset

```{r save-eda-dataset}
saveRDS(
  clustcr_results_df,
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_EDA_{Sys.Date()}.rds")
)

saveRDS(
  cluster_level_w_logreg,
  glue("{clustcr_dir}/results/compiled_clustcr_cluster_level_w_logreg_EDA_{Sys.Date()}.rds")
)
```


```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_EDA_{Sys.Date()}.rds")
)

cluster_level_w_logreg <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_cluster_level_w_logreg_EDA_{Sys.Date()}.rds")
)

clustcr_results_df %>% glimpse()
cluster_level_w_logreg %>% glimpse()

```
