---
title: "ClusTCR Distances"
editor: source
author: "Joe Boktor"
date: '2025-09-08'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

# ClusTCR Statistics

Environment setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(styler)  
library(glue)
library(listenv)
library(furrr)
library(tictoc)
library(progress)
library(data.table)
library(seqinr)
library(progress)
library(immunarch)
library(nlme)
library(broom)
library(broom.mixed)
library(parallelDist)
# dimensionality reduction
library(umap)
# plotting
library(ggsci)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)
library(scico)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/_airr-modeling_funcs.R"))
source(paste0(src_dir, "/R_scripts/_airr_analyze-permanova.R"))

scratch_dir <- "/resnick/scratch/jbok/PDMBS"
gliph2_plots_dir <- glue("{wkdir}/figures/AIRR/gliph2/")
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
arcasHLA_interim <- glue("{wkdir}/data/interim/airr/arcasHLA")
gliph2_input_dir <- glue("{wkdir}/data/input/gliph2/param_files")
gliph2_res_dir <- glue("{wkdir}/data/interim/airr/gliph2/processing")
clustcr_dir <- glue("{wkdir}/data/interim/clusTCR")
clustcr_figs <- glue("{wkdir}/figures/clusTCR")
tcr_cdr3_riskscore_complied <- 
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")
cohort_subset_dir <- glue("{clustcr_dir}/results/cohort_subsets")

```


Loading in compiled clusTCR results dataframe
```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds"))
clustcr_results_df %>% glimpse()

clustcr_results_df_filtered <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 9) %>% 
  glimpse()
clustcr_results_df_filtered %>% glimpse()
clustcr_results_df_filtered$cluster_id %>% unique() %>% length()
```

## Calculating euclidean distance between samples based on cluster membership matrices

```{r}
clust_mat_weighted <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_weighted_2025-08-15.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()
clust_mat_binary <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_binary_2025-08-15.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()

dim(clust_mat_weighted)
dim(clust_mat_binary)

colnames(clust_mat_weighted) %>% head()
rownames(clust_mat_weighted) %>% head()
colnames(clust_mat_binary) %>% head()

# Calculating PCA of cluster membership matrices
variance_threshold <- 0.85  # Retain PCs that explain 85% of variance
# For weighted data
pca_w <- prcomp(clust_mat_weighted)
cumulative_var_weighted <- cumsum(pca_w$sdev^2 / sum(pca_w$sdev^2))
n_pcs_weighted <- which(cumulative_var_weighted >= variance_threshold)[1]

# For binary data
pca_b <- prcomp(clust_mat_binary)
cumulative_var_binary <- cumsum(pca_b$sdev^2 / sum(pca_b$sdev^2))
n_pcs_binary <- which(cumulative_var_binary >= variance_threshold)[1]

cat("Final number of PCs - Weighted:", n_pcs_weighted, "\n")
cat("Final number of PCs - Binary:", n_pcs_binary, "\n")

# Extract the selected principal components
pca_scores_weighted <- pca_w$x[, 1:n_pcs_weighted, drop = FALSE]
pca_scores_binary <- pca_b$x[, 1:n_pcs_binary, drop = FALSE]

# Euclidean distance matrix ----
clustcr_mat_w_eucdist <- parallelDist::parDist(
  pca_scores_weighted, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_w_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_{Sys.Date()}.rds"))

clustcr_mat_b_eucdist <- parallelDist::parDist(
  pca_scores_binary, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_b_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_{Sys.Date()}.rds"))

# #--------------------------------
# # Euclidean distance matrix ----
# clustcr_mat_w_eucdist <- parallelDist::parDist(
#   clust_mat_weighted, method = "euclidean", threads = 16)
# saveRDS(clustcr_mat_w_eucdist, 
#   glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_{Sys.Date()}.rds"))

# clustcr_mat_b_eucdist <- parallelDist::parDist(
#   clust_mat_binary, method = "euclidean", threads = 16)
# saveRDS(clustcr_mat_b_eucdist, 
#   glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_{Sys.Date()}.rds"))

```

Visualizing distance matrices

```{r}
clustcr_mat_w_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_2025-08-15.rds"))

clustcr_mat_b_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_2025-08-15.rds"))

w_sid_order <- seriate_dist_obj(clustcr_mat_w_eucdist, seriate_method = "OLO")
b_sid_order <- seriate_dist_obj(clustcr_mat_b_eucdist, seriate_method = "OLO")

# Create long format data frame with seriated order
dist_df_seriated <- clustcr_mat_w_eucdist %>%
  as.matrix() %>% 
  as.data.frame() %>%
  rownames_to_column(var = "row_id") %>% 
  pivot_longer(
    cols = -row_id,
    names_to = "col_id",
    values_to = "distance"
  ) %>%
  mutate(
    row_id = factor(row_id, levels = w_sid_order),
    col_id = factor(col_id, levels = w_sid_order)
  ) %>% 
  arrange(row_id, col_id) %>% 
  filter(row_id %nin% low_vol_samples$sample_id) %>% 
  filter(col_id %nin% low_vol_samples$sample_id) %>% 
  glimpse()

# Create the seriated heatmap
pheat_distmat_w <- dist_df_seriated %>% 
  ggplot(aes(x = col_id, y = row_id, fill = distance)) +
  geom_tile() +
  scale_fill_scico(
    palette = "devon",
    direction = -1
  ) +
  labs(
    title = "Sample Distance Matrix Heatmap",
    subtitle = "Euclidean distances between samples (OLO seriation)",
    x = "Samples",
    y = "Samples",
    fill = "Distance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid = element_blank()
  ) +
  coord_fixed()

tic()
ggsave(
  glue("{clustcr_figs}/clusTCR_EucDist_weighted_heatmap_OLO_{Sys.Date()}.png"),
  pheat_distmat_w,
  width = 10,
  height = 10
)
toc()

# Create long format data frame with seriated order
dist_df_seriated_b <- clustcr_mat_b_eucdist %>%
  as.matrix() %>% 
  as.data.frame() %>%
  rownames_to_column(var = "row_id") %>% 
  pivot_longer(
    cols = -row_id,
    names_to = "col_id",
    values_to = "distance"
  ) %>%
  mutate(
    row_id = factor(row_id, levels = b_sid_order),
    col_id = factor(col_id, levels = b_sid_order)
  ) %>% 
  filter(row_id %nin% low_vol_samples$sample_id) %>% 
  filter(col_id %nin% low_vol_samples$sample_id) %>% 
  glimpse()

# Create the seriated heatmap
pheat_distmat_b <- dist_df_seriated_b %>% 
  ggplot(aes(x = col_id, y = row_id, fill = distance)) +
  geom_tile() +
  scale_fill_scico(
    palette = "devon",
    direction = -1
  ) +
  labs(
    title = "Sample Distance Matrix Heatmap",
    subtitle = "Euclidean distances between samples (OLO seriation)",
    x = "Samples",
    y = "Samples",
    fill = "Distance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid = element_blank()
  ) +
  coord_fixed()

tic()
ggsave(
  glue("{clustcr_figs}/clusTCR_EucDist_binary_heatmap_OLO_{Sys.Date()}.png"),
  pheat_distmat_b,
  width = 10,
  height = 10
)
toc()

```



### Generating cluster membership matrices - V2 on trimmed clusters (top/bottom 10% of Selection scores & top/bottom 10% of CDR3 risk scores)

```{r}
# taking sig clusters from chi2 test
chi2_results_df <- readRDS(
  glue("{clustcr_dir}/results/chi2_results/chi2_results_df_2025-09-08.rds")) %>% 
  glimpse()
sig_cids <- chi2_results_df %>% 
  filter(chi2_p_value <= 0.05) %>% 
  pull(cluster_id) %>% 
  unique()

clustcr_filter_summarized <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 9) %>% 
  dplyr::group_by(sample_id, cluster_id) %>%
  dplyr::summarize(count = sum(count)) %>% 
  glimpse()

# filter names of remaining cluster_sscore
remaining_cids <- clustcr_filter_summarized$cluster_id %>% unique()
```


```{r}
cluster_sscore <- clustcr_results_df %>% 
  dplyr::select(cluster_id, Q_data, risk_score_p2_glrm_strat) %>% 
  group_by(cluster_id) %>% 
  dplyr::summarize(
    avg_risk_score = mean(risk_score_p2_glrm_strat, na.rm = TRUE),
    avg_selection_score = mean(log10(Q_data), na.rm = TRUE)) %>% 
  glimpse()

cids_of_interest <- cluster_sscore %>% 
  filter(cluster_id %in% remaining_cids) %>%
  mutate(quantile_bin_ss = ntile(avg_selection_score, 10)) %>%
  mutate(quantile_bin_rs = ntile(avg_risk_score, 10)) %>%
  filter(quantile_bin_ss == 1 | quantile_bin_ss == 10 | 
    quantile_bin_rs == 1 | quantile_bin_rs == 10 | 
    (cluster_id %in% sig_cids)) %>%
  pull(cluster_id) %>% 
  unique()

sid_count_sum <- clustcr_filter_summarized %>%
  filter(cluster_id %in% cids_of_interest) %>% 
  dplyr::group_by(sample_id) %>% 
  dplyr::summarize(sid_total_count = sum(count)) %>% 
  glimpse()

clust_mat_weighted <- clustcr_filter_summarized %>% 
  filter(cluster_id %in% cids_of_interest) %>% 
  left_join(sid_count_sum, by = "sample_id") %>% 
  mutate(norm_count = count / sid_total_count) %>% 
  dplyr::select(sample_id, cluster_id, norm_count) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = norm_count,
    values_fill = 0
  )

clust_mat_binary <- clustcr_filter_summarized %>% 
  filter(cluster_id %in% cids_of_interest) %>% 
  mutate(present = 1) %>% 
  dplyr::select(sample_id, cluster_id, present) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = present,
    values_fill = 0
  )

dim(clust_mat_weighted)
dim(clust_mat_binary)

# Saving data filtered for extreme Selection Scores or CDR3 risk scores
saveRDS(
  clust_mat_weighted, 
  glue("{clustcr_dir}/results/clust_mat_weighted_SSxRS_{Sys.Date()}.rds")
)
saveRDS(
  clust_mat_binary, 
  glue("{clustcr_dir}/results/clust_mat_binary_SSxRS_{Sys.Date()}.rds")
)

```

```{r}
# latest version is actually 2025-08-26
clust_mat_weighted <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_weighted_SSxRS_2025-08-17.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()
clust_mat_binary <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_binary_SSxRS_2025-08-17.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()

dim(clust_mat_weighted)
dim(clust_mat_binary)

# Calculating PCA of cluster membership matrices
variance_threshold <- 0.85  # Retain PCs that explain 85% of variance
# For weighted data
pca_w <- prcomp(clust_mat_weighted)
cumulative_var_weighted <- cumsum(pca_w$sdev^2 / sum(pca_w$sdev^2))
n_pcs_weighted <- which(cumulative_var_weighted >= variance_threshold)[1]

# For binary data
pca_b <- prcomp(clust_mat_binary)
cumulative_var_binary <- cumsum(pca_b$sdev^2 / sum(pca_b$sdev^2))
n_pcs_binary <- which(cumulative_var_binary >= variance_threshold)[1]

cat("Final number of PCs - Weighted:", n_pcs_weighted, "\n")
cat("Final number of PCs - Binary:", n_pcs_binary, "\n")

# Extract the selected principal components
pca_scores_weighted <- pca_w$x[, 1:n_pcs_weighted, drop = FALSE]
pca_scores_binary <- pca_b$x[, 1:n_pcs_binary, drop = FALSE]

# Euclidean distance matrix ----
clustcr_mat_w_eucdist <- parallelDist::parDist(
  pca_scores_weighted, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_w_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_SSxRS_{Sys.Date()}.rds"))

clustcr_mat_b_eucdist <- parallelDist::parDist(
  pca_scores_binary, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_b_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_SSxRS_{Sys.Date()}.rds"))

saveRDS(
  pca_scores_weighted,
  glue("{clustcr_dir}/results/pca_scores_weighted_SSxRS_{Sys.Date()}.rds")
)
saveRDS(
  pca_scores_binary,
  glue("{clustcr_dir}/results/pca_scores_binary_SSxRS_{Sys.Date()}.rds")
)

```


### Generating cluster membership matrices - V3 only PD altered clusters

```{r}
sid_count_sum <- clustcr_filter_summarized %>%
  # filter(cluster_id %in% sig_cids) %>% 
  dplyr::group_by(sample_id) %>% 
  dplyr::summarize(sid_total_count = sum(count)) %>% 
  glimpse()

clust_mat_weighted <- clustcr_filter_summarized %>% 
  filter(cluster_id %in% sig_cids) %>% 
  left_join(sid_count_sum, by = "sample_id") %>% 
  mutate(norm_count = count / sid_total_count) %>% 
  dplyr::select(sample_id, cluster_id, norm_count) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = norm_count,
    values_fill = 0
  )

clust_mat_binary <- clustcr_filter_summarized %>% 
  filter(cluster_id %in% sig_cids) %>% 
  mutate(present = 1) %>% 
  dplyr::select(sample_id, cluster_id, present) %>% 
  pivot_wider(
    id_cols = sample_id,
    names_from = cluster_id,
    values_from = present,
    values_fill = 0
  )
dim(clust_mat_weighted)
dim(clust_mat_binary)

# Saving data filtered for extreme Selection Scores or CDR3 risk scores
saveRDS(
  clust_mat_weighted, 
  glue("{clustcr_dir}/results/clust_mat_weighted_Chi2SigOnly_{Sys.Date()}.rds")
)
saveRDS(
  clust_mat_binary, 
  glue("{clustcr_dir}/results/clust_mat_binary_Chi2SigOnly_{Sys.Date()}.rds")
)

```


```{r}
clust_mat_weighted <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_weighted_Chi2SigOnly_2025-08-19.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()
clust_mat_binary <- readRDS(
  glue("{clustcr_dir}/results/clust_mat_binary_Chi2SigOnly_2025-08-19.rds")) %>% 
  tibble::column_to_rownames(var = "sample_id") %>% 
  as.matrix()

dim(clust_mat_weighted)
dim(clust_mat_binary)

# Calculating PCA of cluster membership matrices
variance_threshold <- 0.85  # Retain PCs that explain 85% of variance
# For weighted data
pca_w <- prcomp(clust_mat_weighted)
cumulative_var_weighted <- cumsum(pca_w$sdev^2 / sum(pca_w$sdev^2))
n_pcs_weighted <- which(cumulative_var_weighted >= variance_threshold)[1]

# For binary data
pca_b <- prcomp(clust_mat_binary)
cumulative_var_binary <- cumsum(pca_b$sdev^2 / sum(pca_b$sdev^2))
n_pcs_binary <- which(cumulative_var_binary >= variance_threshold)[1]

cat("Final number of PCs - Weighted:", n_pcs_weighted, "\n")
cat("Final number of PCs - Binary:", n_pcs_binary, "\n")

# Extract the selected principal components
pca_scores_weighted <- pca_w$x[, 1:n_pcs_weighted, drop = FALSE]
pca_scores_binary <- pca_b$x[, 1:n_pcs_binary, drop = FALSE]

# Euclidean distance matrix ----
clustcr_mat_w_eucdist <- parallelDist::parDist(
  pca_scores_weighted, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_w_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_Chi2SigOnly_{Sys.Date()}.rds"))

clustcr_mat_b_eucdist <- parallelDist::parDist(
  pca_scores_binary, method = "euclidean", threads = 16)
saveRDS(clustcr_mat_b_eucdist, 
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_Chi2SigOnly_{Sys.Date()}.rds"))

saveRDS(
  pca_scores_weighted,
  glue("{clustcr_dir}/results/pca_scores_weighted_Chi2SigOnly_{Sys.Date()}.rds")
)
saveRDS(
  pca_scores_binary,
  glue("{clustcr_dir}/results/pca_scores_binary_Chi2SigOnly_{Sys.Date()}.rds")
)

```


Visualizing PCA

```{r}
pca_scores_binary <- readRDS(
  glue("{clustcr_dir}/results/pca_scores_binary_SSxRS_2025-08-18.rds")
)
# pca_scores_binary <- readRDS(
#   glue("{clustcr_dir}/results/pca_scores_binary_Chi2SigOnly_2025-08-19.rds")
# )

pca_scores_binary %>% glimpse()

pca_df <- pca_scores_binary %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  left_join(metadata, by = "sample_id") %>%
  filter(sample_id %nin% low_vol_samples$sample_id) %>%
  filter(case_control_other_latest == "Case") %>%
  glimpse()

# Create PCA plot with metadata
var_of_interest <- "moca_total_score"
# pca_plot <- 
pca_df %>% 
  ggplot(aes(x = PC1, y = PC2, color = .data[[var_of_interest]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "PCA of TCR Repertoire Clusters",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  # Add confidence ellipses if you want to show group separation
  # stat_ellipse(aes(color = .data[[var_of_interest]]), type = "norm", level = 0.95) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 11),
    legend.title = element_text(size = 11)
  )

# Display the plot
pca_plot

# Save the plot
ggsave(
  filename = glue("{clustcr_figs}/pca_plot_binary_Chi2SigOnly_{Sys.Date()}.png"),
  plot = pca_plot,
  width = 10,
  height = 8,
  dpi = 300
)

```




Clustering PD samples using DBSCAN

```{r}
clustcr_mat_w_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_SSxRS_2025-08-18.rds"))
clustcr_mat_b_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_SSxRS_2025-08-18.rds"))
clustcr_mat_w_eucdist_input <- clustcr_mat_w_eucdist %>% 
  as.matrix() %>% as.data.frame()
clustcr_mat_b_eucdist_input <- clustcr_mat_b_eucdist %>% 
  as.matrix() %>% as.data.frame()

all(rownames(clustcr_mat_b_eucdist_input) == rownames(clustcr_mat_w_eucdist_input))

# filtering out PD only samples from distance matrices
pd_sids <- metadata %>%
  filter(sample_id %nin% low_vol_samples$sample_id) %>%
  filter(sample_id %in% rownames(clustcr_mat_b_eucdist_input)) %>% 
  filter(case_control_other_latest == "Case") %>% 
  pull(sample_id) %>% 
  unique()

# filtering and converting back into dist object
clustcr_mat_b_eucdist_pd <- clustcr_mat_b_eucdist_input[pd_sids, pd_sids] %>% as.dist()
clustcr_mat_w_eucdist_pd <- clustcr_mat_w_eucdist_input[pd_sids, pd_sids] %>% as.dist()

library(dbscan)
dbscan::kNNdistplot(clustcr_mat_w_eucdist_pd, k =  8)
abline(h = 0.25, lty = 2) # optimal eps looks like 0.25 (at k = 8)

dbscan::kNNdistplot(clustcr_mat_b_eucdist_pd, k =  3)
abline(h = 12, lty = 2) # optimal eps looks like 14 (at k = 8)

# dbscan package
w_db <- dbscan::dbscan(clustcr_mat_w_eucdist_pd, 0.15, 10)
b_db <- dbscan::dbscan(clustcr_mat_b_eucdist_pd, 12, 5)

w_db$cluster %>% table()
b_db$cluster %>% table()



## Trying Distmatrix on only Chi2SigOnly clusters
clustcr_mat_w_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_Chi2SigOnly_2025-08-19.rds"))
clustcr_mat_b_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_Chi2SigOnly_2025-08-19.rds"))
clustcr_mat_w_eucdist_input <- clustcr_mat_w_eucdist %>% 
  as.matrix() %>% as.data.frame()
clustcr_mat_b_eucdist_input <- clustcr_mat_b_eucdist %>% 
  as.matrix() %>% as.data.frame()
all(rownames(clustcr_mat_b_eucdist_input) == rownames(clustcr_mat_w_eucdist_input))

# filtering out PD only samples from distance matrices
pd_sids <- metadata %>%
  filter(sample_id %nin% low_vol_samples$sample_id) %>%
  filter(sample_id %in% rownames(clustcr_mat_b_eucdist_input)) %>% 
  filter(case_control_other_latest == "Case") %>% 
  pull(sample_id) %>% 
  unique()

# filtering and converting back into dist object
clustcr_mat_b_eucdist_pd <- clustcr_mat_b_eucdist_input[pd_sids, pd_sids] %>% as.dist()
clustcr_mat_w_eucdist_pd <- clustcr_mat_w_eucdist_input[pd_sids, pd_sids] %>% as.dist()

library(dbscan)
dbscan::kNNdistplot(clustcr_mat_w_eucdist_pd, k =  8)
abline(h = 0.05, lty = 2) # optimal eps looks like 0.05 (at k = 8)

dbscan::kNNdistplot(clustcr_mat_b_eucdist_pd, k =  8)
abline(h = 8, lty = 2) # optimal eps looks like 8 (at k = 8)

# dbscan package
w_db <- dbscan::dbscan(clustcr_mat_w_eucdist_pd, 0.03, 10)
b_db <- dbscan::dbscan(clustcr_mat_b_eucdist_pd, 7, 10)

w_db$cluster %>% table()
b_db$cluster %>% table()








pub_ovov <- readRDS(
    glue("{metrics_dir}/rep-overlap-overlap_2025-07-08.rds")
)
pub_ovpub <- readRDS(
    glue("{metrics_dir}/rep-overlap-public_2025-07-08.rds")
)
pub_mh <- readRDS(
    glue("{metrics_dir}/rep-overlap-morisita_2025-07-20.rds")
)


# pub_mh %>% glimpse()
pub_mh %>% rownames() %>% head()
pub_mh %>% colnames() %>% head()
pub_mh[1:10, 1:10]

pub_mh_dist <- as.dist(1 - pub_mh)

dbscan::kNNdistplot(pub_mh_dist, k =  5)
abline(h = 0.6, lty = 2) # optimal eps looks like 0.6 (at k = 5)
mh_db <- dbscan::dbscan(pub_mh_dist, 0.6, 10)
mh_db$cluster %>% table()


# # clustcr_mat_w_eucdist
# str(clustcr_mat_b_eucdist)
# rownames(clustcr_mat_b_eucdist)


# Doesn't really seem to produce meaningful clusters

```



Visualizing distance matrices - V2

```{r}
clustcr_mat_w_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_w_eucdist_pca_SSxRS_2025-08-18.rds"))
clustcr_mat_b_eucdist <- readRDS(
  glue("{clustcr_dir}/results/clustcr_mat_b_eucdist_pca_SSxRS_2025-08-18.rds"))

w_sid_order <- seriate_dist_obj(clustcr_mat_w_eucdist, seriate_method = "OLO")
b_sid_order <- seriate_dist_obj(clustcr_mat_b_eucdist, seriate_method = "OLO")

# Create long format data frame with seriated order
dist_df_seriated <- clustcr_mat_w_eucdist %>%
  as.matrix() %>% 
  as.data.frame() %>%
  rownames_to_column(var = "row_id") %>% 
  pivot_longer(
    cols = -row_id,
    names_to = "col_id",
    values_to = "distance"
  ) %>%
  as.data.table() %>% 
  mutate(
    row_id = factor(row_id, levels = w_sid_order),
    col_id = factor(col_id, levels = w_sid_order)
  ) %>% 
  arrange(row_id, col_id) %>% 
  filter(row_id %nin% low_vol_samples$sample_id) %>% 
  filter(col_id %nin% low_vol_samples$sample_id) %>% 
  glimpse()

# Create the seriated heatmap
pheat_distmat_w <- dist_df_seriated %>% 
  ggplot(aes(x = col_id, y = row_id, fill = distance)) +
  geom_tile() +
  scale_fill_scico(
    palette = "devon",
    direction = 1
  ) +
  labs(
    title = "Sample Distance Matrix Heatmap",
    subtitle = "Euclidean distances between samples (OLO seriation)",
    x = "Samples",
    y = "Samples",
    fill = "Distance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid = element_blank()
  ) +
  coord_fixed()

tic()
ggsave(
  glue("{clustcr_figs}/clusTCR_EucDist_weighted_heatmap_SSxRS_OLO_{Sys.Date()}.png"),
  pheat_distmat_w,
  width = 10,
  height = 10
)
toc()

# Create long format data frame with seriated order
dist_df_seriated_b <- clustcr_mat_b_eucdist %>%
  as.matrix() %>% 
  as.data.frame() %>%
  rownames_to_column(var = "row_id") %>% 
  pivot_longer(
    cols = -row_id,
    names_to = "col_id",
    values_to = "distance"
  ) %>%
  as.data.table() %>% 
  mutate(
    row_id = factor(row_id, levels = b_sid_order),
    col_id = factor(col_id, levels = b_sid_order)
  ) %>% 
  filter(row_id %nin% low_vol_samples$sample_id) %>% 
  filter(col_id %nin% low_vol_samples$sample_id) %>% 
  glimpse()

# Create the seriated heatmap
pheat_distmat_b <- dist_df_seriated_b %>% 
  ggplot(aes(x = col_id, y = row_id, fill = distance)) +
  geom_tile() +
  scale_fill_scico(
    palette = "devon",
    direction = 1
  ) +
  labs(
    title = "Sample Distance Matrix Heatmap",
    subtitle = "Euclidean distances between samples (OLO seriation)",
    x = "Samples",
    y = "Samples",
    fill = "Distance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    panel.grid = element_blank()
  ) +
  coord_fixed()

tic()
ggsave(
  glue("{clustcr_figs}/clusTCR_EucDist_binary_heatmap_SSxRS_OLO_{Sys.Date()}.png"),
  pheat_distmat_b,
  width = 10,
  height = 10
)
toc()

```



Visualizing cluster membership matrices as heatmap

```{r}
clustcr_results_df <- readRDS(
  glue("{clustcr_dir}/results/compiled_clustcr_results_df_2025-08-24.rds"))

clustcr_filter_summarized <- clustcr_results_df %>% 
  filter(unique_donors_per_clust >= 10) %>% 
  filter(motif_length >= 8) %>% 
  dplyr::group_by(sample_id, cluster_id) %>%
  dplyr::summarize(count = sum(count)) %>% 
  glimpse()

clust_mat_binary <- clustcr_filter_summarized %>% 
  mutate(present = 1) %>% 
  dplyr::select(sample_id, cluster_id, present) %>% 
  glimpse()
# getting rank for samples and clusters
sid_order <- clustcr_filter_summarized %>%
  dplyr::group_by(sample_id) %>% 
  dplyr::summarize(sid_total_count = sum(count)) %>% 
  pull(sample_id)
cid_order <- clust_mat_binary %>% 
  group_by(cluster_id) %>% 
  dplyr::summarize(cluster_total_count = sum(present)) %>% 
  arrange(desc(cluster_total_count)) %>% 
  pull(cluster_id)
clust_mat_binary %>% glimpse()

clust_mat_binary_sorted <- clust_mat_binary %>% 
  as.data.table() %>% 
  mutate(sample_id = factor(sample_id, levels = sid_order)) %>% 
  mutate(cluster_id = factor(cluster_id, levels = rev(cid_order))) %>% 
  glimpse()

p_cluster_heatmap <- clust_mat_binary_sorted %>% 
  filter(sample_id %nin% low_vol_samples$sample_id) %>% 
  mutate(present = as.character(present)) %>% 
  ggplot(aes(x = sample_id, y = cluster_id, fill = present)) +
  geom_tile() +
  scale_fill_manual(values = c("1" = "black")) +
  # scale_fill_manual(values = c("1" = "#434343"), na.value = "#f1f1f1") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

ggsave(
  glue("{clustcr_figs}/clusTCR_cluster_membership_heatmap_binary_{Sys.Date()}.png"),
  width = 10,
  height = 10,
  dpi = 1200
)

```


