---
title: "cdr3-QTL Single-Cell Localization"
editor: source
author: "Joe Boktor"
date: '2026-04-12'
eval: false
format:
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
    fig-height: 8
    fig-width: 14
---

This notebook localizes CDR3 risk scores (Ishigaki et al., 2022; [GitHub](https://github.com/immunogenomics/cdr3-QTL)) to specific T cell subtypes using external single-cell RNA-seq + TCR-seq datasets from Parkinson's disease studies:

- **Wang et al. 2021** — Blood + CSF T cells (10x 5' GEX + V(D)J); 121K T cells, 21 clusters; [Cell Discovery](https://doi.org/10.1038/s41421-021-00280-3)
- **Ma et al. 2025** — Post-mortem brain (SN + cortex) snRNA-seq + TCR-seq; 207K nuclei; [Nature Communications](https://doi.org/10.1038/s41467-025-62478-3)


## Analysis Setup

```{r setup, include=FALSE}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(glue)
library(furrr)
library(tictoc)
library(progress)
library(entropy)
library(data.table)
library(seqinr)
library(nlme)
library(broom.mixed)
library(Seurat)
# plotting
library(ggsci)
library(scico)
library(plotly)
library(aplot)
library(patchwork)
library(ggpubr)
library(ggrepel)

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
source(paste0(src_dir, "/R_scripts/compute_CDR3-risk-score.R"))
scratch_dir <- "/resnick/scratch/jbok/PDMBS"
metrics_dir <- glue("{wkdir}/data/interim/airr/immunarch/metrics_tcrb")
entropy_dir <- glue("{wkdir}/data/interim/cdr3_entropy")
entropy_plots_dir <- glue("{wkdir}/figures/AIRR/cdr3_entropy")
git_dir <- glue("{home_dir}/git/cdr3-QTL")
hla_plots_dir <- glue("{wkdir}/figures/AIRR/arcasHLA")
hla_dir <- glue("{wkdir}/data/interim/airr/arcasHLA/analysis_cleaned")

c3_pheno_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_phenotypes")
gdata_dir <- glue("{wkdir}/data/interim/cdr3_qtl/cdr3_hla_genotypes")
hla_4d_to_aa_dir <- glue("{wkdir}/data/input/cdr3qtl/hla_4d_to_aa")
tcr_cdr3_riskscore_dir <- glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores")
tcr_cdr3_riskscore_complied <-
    glue("{wkdir}/data/interim/cdr3_qtl/scoring_cdr3-risk-scores_complied")

# External single-cell data directories
ext_sc_dir <- glue("{wkdir}/data/input/external_singlecell")
wang_dir <- glue("{ext_sc_dir}/wang_2021")
ma_dir <- glue("{ext_sc_dir}/ma_2025")
sc_results_dir <- glue("{wkdir}/data/interim/cdr3_qtl/singlecell_localization")
sc_figures_dir <- glue("{wkdir}/figures/CDR3-qtl/singlecell_localization")

dir.create(wang_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(ma_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(sc_results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(sc_figures_dir, showWarnings = FALSE, recursive = TRUE)

future::plan(multisession, workers = 8)
```


# Study 1: Wang et al. 2021 — Blood & CSF T Cells

## Download Wang et al. 2021 Data

Data deposited on [Zenodo record 3993994](https://zenodo.org/record/3993994):

- `counts.rds` (382MB) — raw count matrix
- `normalized.rds` (417MB) — normalized expression matrix
- `metadata.txt` (16MB) — cell annotations with T cell cluster labels + TCR info
- `README.txt`

CSF TCR data originally from Gate et al. (GEO: [GSE134578](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE134578)).

```{r download-wang}
zenodo_base <- "https://zenodo.org/record/3993994/files"
wang_files <- c("counts.rds", "normalized.rds", "metadata.txt", "README.txt")

for (f in wang_files) {
    dest <- glue("{wang_dir}/{f}")
    if (!file.exists(dest)) {
        message(glue("Downloading {f} via curl..."))
        exit_code <- system2("curl",
            args = c("-L", "-o", shQuote(dest),
                     "--retry", "3",
                     "--retry-delay", "5",
                     "-C", "-",
                     shQuote(glue("{zenodo_base}/{f}"))),
            stdout = "", stderr = ""
        )
        if (exit_code != 0) {
            warning(glue("curl failed for {f} (exit code {exit_code})"))
            if (file.exists(dest)) file.remove(dest)
        }
    } else {
        message(glue("{f} already exists, skipping."))
    }
}

# Verify downloads
wang_downloaded <- file.info(list.files(wang_dir, full.names = TRUE))
wang_downloaded$file <- basename(rownames(wang_downloaded))
wang_downloaded %>% dplyr::select(file, size) %>% print()

```


## Load & Inspect Wang et al. 2021 Metadata

```{r load-wang-metadata}
wang_meta <- read.delim(glue("{wang_dir}/metadata.txt"), stringsAsFactors = FALSE) %>%
    as_tibble() %>%
    janitor::clean_names()

message(glue("Wang et al. 2021: {nrow(wang_meta)} cells"))
wang_meta %>% glimpse()

# Inspect available columns — looking for CDR3, cell type, donor, condition
wang_meta %>% colnames() %>% sort() %>% print()
```


## Extract CDR3-Beta Sequences with Cell Type Annotations

```{r wang-extract-cdr3b}
# Inspect TCR-related columns
wang_meta %>%
    dplyr::select(matches("cdr3|CDR3|trb|TRB|tcr|TCR|chain|clono", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

# Inspect cell type / cluster columns
wang_meta %>%
    dplyr::select(matches("cluster|cell.?type|subtype|annotation|ident", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

# Inspect condition/donor columns
wang_meta %>%
    dplyr::select(matches("condition|disease|status|group|donor|patient|sample|subject", ignore.case = TRUE)) %>%
    colnames() %>%
    print()

```


```{r wang-clean-cdr3b}
# NOTE: Column names below need to be adjusted based on actual metadata column names
# after inspecting the output from the chunk above.
# Expected columns (rename as needed):
#   - CDR3 beta AA sequence
#   - Cell type / cluster annotation
#   - Donor ID
#   - Condition (PD vs HC)
#   - Tissue (blood vs CSF)

# Example mapping — UPDATE column names after inspection:
wang_cdr3b <- wang_meta %>%
    # rename(
    #     cdr3b_aa = <CDR3_BETA_COLUMN>,
    #     t_cell_subtype = <CLUSTER_COLUMN>,
    #     donor_id = <DONOR_COLUMN>,
    #     condition = <CONDITION_COLUMN>,
    #     tissue = <TISSUE_COLUMN>
    # ) %>%
    # filter(!is.na(cdr3b_aa) & cdr3b_aa != "") %>%
    # filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
    mutate(
        study = "Wang_2021",
        compartment = "peripheral"  # blood + CSF
    ) %>%
    glimpse()

# CDR3 length distribution
# wang_cdr3b %>%
#     mutate(cdr3_len = nchar(cdr3b_aa)) %>%
#     count(cdr3_len) %>%
#     print()

# T cell subtype breakdown
# wang_cdr3b %>%
#     count(t_cell_subtype, sort = TRUE) %>%
#     print()

# Condition breakdown
# wang_cdr3b %>%
#     count(condition) %>%
#     print()
```


## Download CSF TCR Data (GSE134578)

The CSF T cell TCR data used in Wang et al. 2021 comes from Gate et al. 2020 ([GSE134578](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE134578)).

```{r download-csf-tcr}
csf_dir <- glue("{wang_dir}/csf_GSE134578")
dir.create(csf_dir, showWarnings = FALSE, recursive = TRUE)

# Download supplementary files from GEO
# GEO FTP for supplementary data:
csf_supp_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE134nnn/GSE134578/suppl/"

# List files available — check manually or use:
# system(glue("wget -r -np -nd -P {csf_dir} {csf_supp_url}"))

message("Check GEO supplementary files at: ", csf_supp_url)
message("Download TCR contig/clonotype files with cell barcodes and CDR3 sequences.")
message("Files may include filtered_contig_annotations.csv or similar 10x VDJ output.")
```


```{r load-csf-tcr}
# After downloading, load CSF TCR data
# csf_files <- list.files(csf_dir, pattern = "\\.csv$|\\.tsv$|\\.txt$", full.names = TRUE)
# message("CSF TCR files found: ", length(csf_files))
# csf_files %>% print()

# Load and process CSF TCR data — adjust based on actual file format
# csf_tcr <- read_csv(csf_files[1]) %>%
#     janitor::clean_names() %>%
#     glimpse()

# Extract CDR3-beta with cell annotations
# csf_cdr3b <- csf_tcr %>%
#     filter(chain == "TRB") %>%
#     dplyr::select(barcode, cdr3, ...) %>%
#     rename(cdr3b_aa = cdr3) %>%
#     filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
#     mutate(
#         study = "Wang_2021",
#         tissue = "CSF",
#         compartment = "CSF"
#     )
```


# Study 4: Ma et al. 2025 — Brain T Cells (Substantia Nigra)

## Download Ma et al. 2025 Data

Processed data and supplementary materials deposited on Zenodo:

- [Processed data](https://zenodo.org/records/19454949) — `data.zip` (6.7GB): snRNA-seq processed objects
- [Supplementary data](https://zenodo.org/records/19457731) — `SupplementaryData.zip` (105MB): supplementary tables
- [GitHub](https://github.com/dalhoomist/T-cell_and_glial_pathology_in_PD): analysis code

```{r download-ma}
# -- Processed data (6.7 GB) --
ma_data_zip <- glue("{ma_dir}/data.zip")
if (!file.exists(ma_data_zip)) {
    message("Downloading Ma et al. 2025 processed data (6.7 GB)...")
    system2("curl",
        args = c("-L", "-o", shQuote(ma_data_zip),
                 "--retry", "3", "--retry-delay", "5", "-C", "-",
                 shQuote("https://zenodo.org/records/19454949/files/data.zip")),
        stdout = "", stderr = ""
    )
} else {
    message("data.zip already exists, skipping.")
}

# -- Supplementary data (105 MB) --
ma_supp_zip <- glue("{ma_dir}/SupplementaryData.zip")
if (!file.exists(ma_supp_zip)) {
    message("Downloading Ma et al. 2025 supplementary data (105 MB)...")
    system2("curl",
        args = c("-L", "-o", shQuote(ma_supp_zip),
                 "--retry", "3", "--retry-delay", "5", "-C", "-",
                 shQuote("https://zenodo.org/records/19457731/files/SupplementaryData.zip")),
        stdout = "", stderr = ""
    )
} else {
    message("SupplementaryData.zip already exists, skipping.")
}

# Verify downloads
file.info(c(ma_data_zip, ma_supp_zip)) %>%
    mutate(file = basename(rownames(.))) %>%
    dplyr::select(file, size) %>%
    print()
```


## Extract & Inspect Ma et al. 2025 Archives

```{r extract-ma}
# Extract processed data
ma_data_dir <- glue("{ma_dir}/data")
if (!dir.exists(ma_data_dir)) {
    message("Extracting data.zip...")
    unzip(ma_data_zip, exdir = ma_dir)
}

# Extract supplementary data
ma_supp_dir <- glue("{ma_dir}/SupplementaryData")
if (!dir.exists(ma_supp_dir)) {
    message("Extracting SupplementaryData.zip...")
    unzip(ma_supp_zip, exdir = ma_dir)
}

# List contents
message("=== Processed data contents ===")
list.files(ma_data_dir, recursive = TRUE, full.names = FALSE) %>% print()

message("\n=== Supplementary data contents ===")
list.files(ma_supp_dir, recursive = TRUE, full.names = FALSE) %>% print()
```


## Load Ma et al. 2025 Single-Cell Data

```{r load-ma-data}
# Identify RDS/h5ad/h5seurat files in the extracted data
ma_rds_files <- list.files(ma_data_dir, pattern = "\\.rds$|\\.RDS$",
                           recursive = TRUE, full.names = TRUE)
ma_h5_files <- list.files(ma_data_dir, pattern = "\\.h5ad$|\\.h5seurat$|\\.h5$",
                          recursive = TRUE, full.names = TRUE)

message("RDS files found:")
ma_rds_files %>% print()
message("H5 files found:")
ma_h5_files %>% print()

# Load the primary object — adjust path based on extracted contents
# ma_obj <- readRDS(ma_rds_files[1])
# OR: ma_obj <- SeuratDisk::LoadH5Seurat(ma_h5_files[1])

# Inspect structure
# ma_obj
# head(ma_obj@meta.data) %>% glimpse()
# colnames(ma_obj@meta.data) %>% sort() %>% print()
```


## Extract CDR3-Beta Sequences from Ma et al. 2025

```{r ma-extract-cdr3b}
# Check supplementary tables for TCR clonotype data
ma_supp_files <- list.files(ma_supp_dir, recursive = TRUE, full.names = TRUE)
ma_supp_files %>% print()

# Look for TCR-related files
ma_tcr_files <- ma_supp_files[grepl("tcr|TCR|clono|CDR3|vdj", ma_supp_files, ignore.case = TRUE)]
message("TCR-related supplementary files:")
ma_tcr_files %>% print()

# Load TCR data — adjust based on actual file format
# ma_tcr <- read_csv(ma_tcr_files[1]) %>%
#     janitor::clean_names() %>%
#     glimpse()
```


```{r ma-clean-cdr3b}
# NOTE: Column names need to be adjusted based on actual data inspection.
# The Ma et al. study identified 535 T cells in PD substantia nigra with
# TCR sequences showing homology to alpha-synuclein-reactive clones.

# Extract CDR3-beta sequences with cell annotations
# ma_cdr3b <- ma_tcr %>%
#     filter(chain == "TRB" | grepl("beta", chain, ignore.case = TRUE)) %>%
#     rename(
#         cdr3b_aa = <CDR3_COLUMN>,
#         t_cell_subtype = <CELLTYPE_COLUMN>,
#         donor_id = <DONOR_COLUMN>,
#         condition = <CONDITION_COLUMN>
#     ) %>%
#     filter(!is.na(cdr3b_aa) & cdr3b_aa != "") %>%
#     filter(nchar(cdr3b_aa) >= 12 & nchar(cdr3b_aa) <= 18) %>%
#     mutate(
#         study = "Ma_2025",
#         tissue = "brain",
#         compartment = "CNS"
#     ) %>%
#     glimpse()

# T cell subtype breakdown
# ma_cdr3b %>% count(t_cell_subtype, sort = TRUE) %>% print()
# ma_cdr3b %>% count(condition) %>% print()
```


# Harmonize CDR3-Beta Data Across Studies

```{r harmonize-cdr3b}
# Combine CDR3-beta data from both studies
# Standardize column names to: cell_id, donor_id, condition, tissue,
#   t_cell_subtype, cdr3b_aa, study, compartment

# all_cdr3b <- bind_rows(wang_cdr3b, ma_cdr3b) %>%
#     mutate(
#         # Standardize condition labels
#         condition = case_when(
#             grepl("PD|parkinson|case", condition, ignore.case = TRUE) ~ "PD",
#             grepl("HC|healthy|control|ctrl", condition, ignore.case = TRUE) ~ "HC",
#             TRUE ~ condition
#         ),
#         cdr3_len = nchar(cdr3b_aa)
#     ) %>%
#     glimpse()

# Summary
# message(glue("Total cells with CDR3-beta: {nrow(all_cdr3b)}"))
# all_cdr3b %>% count(study, condition) %>% print()
# all_cdr3b %>% count(study, compartment) %>% print()
# all_cdr3b %>% count(study, t_cell_subtype, sort = TRUE) %>% print()

# CDR3 length distribution by study
# all_cdr3b %>%
#     ggplot(aes(x = cdr3_len, fill = study)) +
#     geom_bar(position = "dodge") +
#     theme_bw() +
#     labs(x = "CDR3-beta length (AA)", y = "Count", fill = "Study") +
#     scale_fill_npg()

# Save harmonized data
# saveRDS(all_cdr3b, glue("{sc_results_dir}/harmonized_cdr3b_singlecell.rds"))
```


# Score CDR3-Beta Sequences with CDR3 Risk Score

Applying the CDR3-QTL HLA risk score beta-coefficients (Ishigaki et al. 2022) to single-cell CDR3-beta sequences, using the same scoring function from [notebook 04](AIRR_cdr3-QTL_04-cdr3riskscore.qmd).

## Load Pre-Computed Beta-Coefficients

```{r load-betacoefs}
resdir <- glue("{wkdir}/data/interim/cdr3_qtl/results_cdr3-risk-score")

all_imgt_pos <- c(
    "P104","P105","P106","P107","P108","P109","P110","P111",
    "P111.1","P112.2","P112.1",
    "P112","P113","P114","P115","P116","P117","P118")

cdr3_betacoef <- list.files(resdir, pattern = ".csv", full.names = TRUE, recursive = TRUE) %>%
    purrr::set_names(basename(.)) %>%
    purrr::map(read_csv) %>%
    bind_rows(.id = "file_name") %>%
    mutate(
        descriptor = basename(file_name) %>% gsub("lm_model_results__", "", .),
        pheno = str_before_first(descriptor, "__"),
        cdr3_pos = descriptor %>% str_after_first("CDR3pos-") %>% str_before_first("__"),
        cdr3_len = descriptor %>% str_after_first("CDR3len-") %>% str_before_first(".csv")
    ) %>%
    janitor::clean_names() %>%
    filter(!is.na(cdr3_aa))

# Filter to significant associations
cdr3_betacoef_sig <- cdr3_betacoef %>%
    group_by(pheno) %>%
    filter(term == "pd_hla_risk_score") %>%
    mutate(adj_p = p.adjust(p_value, method = "BH")) %>%
    filter(adj_p < 0.05) %>%
    arrange(adj_p)

betacoef_list <- cdr3_betacoef_sig %>% split(.$pheno)

message(glue("Loaded {nrow(cdr3_betacoef_sig)} significant beta-coefficients across {length(betacoef_list)} phenotypes"))
names(betacoef_list) %>% print()
```


## Score Single-Cell CDR3-Beta Sequences

```{r score-cdr3b}
# all_cdr3b <- readRDS(glue("{sc_results_dir}/harmonized_cdr3b_singlecell.rds"))

# Get unique CDR3-beta sequences to avoid redundant scoring
# unique_cdr3b <- unique(all_cdr3b$cdr3b_aa)
# message(glue("Scoring {length(unique_cdr3b)} unique CDR3-beta sequences"))

# Score across all phenotype types
# tic("CDR3 risk scoring")
# score_results <- list()
# for (p in names(betacoef_list)) {
#     message(glue("Scoring phenotype: {p}"))
#     score_results[[p]] <- unique_cdr3b %>%
#         purrr::set_names() %>%
#         furrr::future_map_dbl(
#             ~ score_tcrs_with_betacoefs(
#                 x = .,
#                 phenotype_type = p,
#                 betacoef_list = betacoef_list
#             )
#         )
# }
# toc()

# Convert to data frame
# score_df <- score_results %>%
#     purrr::imap(~ tibble(cdr3b_aa = names(.x), !!.y := .x)) %>%
#     purrr::reduce(left_join, by = "cdr3b_aa")

# message(glue("Scored {nrow(score_df)} unique CDR3-beta sequences"))
# score_df %>% glimpse()
```


## Join Scores to Cell-Level Metadata

```{r join-scores}
# scored_cells <- all_cdr3b %>%
#     left_join(score_df, by = "cdr3b_aa")

# message(glue("Scored cells: {nrow(scored_cells)}"))
# scored_cells %>% glimpse()

# Save scored data
# saveRDS(scored_cells, glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))
```


# Analyze CDR3 Risk Scores by T Cell Subtype

## CDR3 Risk Score Distributions by Subtype

```{r plot-scores-by-subtype}
# scored_cells <- readRDS(glue("{sc_results_dir}/scored_cdr3b_singlecell.rds"))

# Select a primary phenotype for visualization
# primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

# -- Per-study, per-subtype boxplot --
# p_subtype_scores <- scored_cells %>%
#     ggplot(aes(x = reorder(t_cell_subtype, .data[[primary_pheno]], FUN = median),
#                y = .data[[primary_pheno]],
#                fill = condition)) +
#     geom_boxplot(outlier.size = 0.3, alpha = 0.8) +
#     facet_wrap(~study, scales = "free_x") +
#     coord_flip() +
#     theme_bw() +
#     labs(
#         x = "T Cell Subtype",
#         y = "CDR3 Risk Score",
#         fill = "Condition",
#         title = "CDR3 Risk Score by T Cell Subtype"
#     ) +
#     scale_fill_npg()

# ggsave(
#     filename = glue("{sc_figures_dir}/cdr3_riskscore_by_subtype.png"),
#     plot = p_subtype_scores,
#     width = 14, height = 10, dpi = 300
# )
# p_subtype_scores
```


## PD vs HC Comparison Within Each Subtype

```{r pd-vs-hc-subtype}
# primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

# Statistical testing: PD vs HC within each subtype
# subtype_tests <- scored_cells %>%
#     group_by(study, t_cell_subtype) %>%
#     filter(n_distinct(condition) == 2) %>%
#     summarize(
#         n_pd = sum(condition == "PD"),
#         n_hc = sum(condition == "HC"),
#         median_pd = median(.data[[primary_pheno]][condition == "PD"]),
#         median_hc = median(.data[[primary_pheno]][condition == "HC"]),
#         wilcox_p = tryCatch(
#             wilcox.test(
#                 .data[[primary_pheno]][condition == "PD"],
#                 .data[[primary_pheno]][condition == "HC"]
#             )$p.value,
#             error = function(e) NA_real_
#         ),
#         .groups = "drop"
#     ) %>%
#     mutate(
#         delta_median = median_pd - median_hc,
#         adj_p = p.adjust(wilcox_p, method = "BH")
#     ) %>%
#     arrange(adj_p)

# subtype_tests %>% print(n = 50)

# saveRDS(subtype_tests, glue("{sc_results_dir}/subtype_wilcox_tests.rds"))
```


## Blood vs Brain Compartment Comparison

```{r compartment-comparison}
# Compare CDR3 risk score profiles across tissue compartments
# primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

# p_compartment <- scored_cells %>%
#     ggplot(aes(x = compartment, y = .data[[primary_pheno]], fill = condition)) +
#     geom_violin(alpha = 0.6, scale = "width") +
#     geom_boxplot(width = 0.2, outlier.size = 0.3, position = position_dodge(0.9)) +
#     theme_bw() +
#     labs(
#         x = "Compartment",
#         y = "CDR3 Risk Score",
#         fill = "Condition",
#         title = "CDR3 Risk Score: Blood vs Brain"
#     ) +
#     scale_fill_npg()

# ggsave(
#     filename = glue("{sc_figures_dir}/cdr3_riskscore_compartment.png"),
#     plot = p_compartment,
#     width = 8, height = 6, dpi = 300
# )
# p_compartment
```


## Summary Heatmap: Mean CDR3 Risk Score by Subtype x Condition

```{r summary-heatmap}
# primary_pheno <- "cdr3_p2_strat_filtered_cv1k"

# heatmap_data <- scored_cells %>%
#     group_by(study, t_cell_subtype, condition) %>%
#     summarize(
#         mean_score = mean(.data[[primary_pheno]]),
#         n_cells = n(),
#         .groups = "drop"
#     )

# p_heatmap <- heatmap_data %>%
#     ggplot(aes(x = condition, y = t_cell_subtype, fill = mean_score)) +
#     geom_tile() +
#     geom_text(aes(label = n_cells), size = 2.5) +
#     facet_wrap(~study, scales = "free_y") +
#     scale_fill_scico(palette = "vik", midpoint = 0) +
#     theme_bw() +
#     labs(
#         x = "Condition", y = "T Cell Subtype",
#         fill = "Mean CDR3\nRisk Score",
#         title = "CDR3 Risk Score Heatmap"
#     )

# ggsave(
#     filename = glue("{sc_figures_dir}/cdr3_riskscore_heatmap.png"),
#     plot = p_heatmap,
#     width = 12, height = 10, dpi = 300
# )
# p_heatmap
```


## Session Info

```{r session-info}
sessionInfo()
```
