---
title: "Distribution Testing"
eval: false
editor: source
author: "Joe Boktor"
date: '2023-10-23'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    theme: cosmo
    column: page
    fig-align: right
    fig-height: 8
    fig-width: 14 
---


```{r}
#| warning: false
#| message: false
home_dir <- "/resnick/groups/MazmanianLab/jboktor"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
scratch_dir <- "/resnick/scratch/jbok/PDMBS"

library(phyloseq)
library(microbiome)
library(ggsci)
library(aplot)
library(usedist)
```

Load phyloseq object

```{r}
# loading metadata (this is a list with both RNASEQ and WGS metadata)
# phymeta <- readRDS("data/interim/metadata/2023-07-14_phyloseq-metadata.rds")

ps <- readRDS(
    glue("{wkdir}/data/processed/phyloseq_objects",
        "/decontaminated/2023-11-01_WGS_KrakenUnique_phyloseq.rds"
    )
)

ps_analysis <- ps %>% 
    microbiome::transform("compositional")
ps_analysis

ps_analysis %>% microbiome::abundances() %>% head()

```


### Kullback-Leibler divergence metric (relative entropy) 
Here we calculate the KL divergence between pairs of samples. This is a measure of the difference between two probability distributions. In this case, we are comparing the distribution of taxa in PD vs Control samples.

$$

D_{KL} (p(x) ||q(x)) = \sum_{x=1}^{n} p_i \ln \frac{p_i}{q_i}

$$


Using `usedist()`, obtain a set of distance pairs of interest

```{r}

kl_metric <- function(p, q) {
     purrr::map2(
        .x = p, .y = q,
        ~ (.x * log(.x/.y))
    )
}

```


```{r}

kl_metric(
    c(0.5, 0.5, 0.2, 0.1),
    c(0.4, 0.5, 0.211, 0.01)
) %>% unlist() %>%sum()



```

For each pair of samples, calculate the KL divergence for all taxa



```{r}

```


