---
title: "Kraken2 Alignment"
eval: false
editor: visual
author: "Joe Boktor"
date: '2022-12-29'
format: 
  html:
    font-family: helvetica neue
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 3
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: center
---

This notebook will execute Kraken2 taxonomic assignment on quality control processed reads.

```{r}
pdmbs_dir <- "/resnick/groups/MazmanianLab/jboktor/PDMBS"
scratch_dir <- "/resnick/scratch/jbok/PDMBS"
wkdir <- paste0(pdmbs_dir, "/pdairr")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(wkdir, "/notebooks/R_scripts/_load-core-pkgs.R"))
source(paste0(wkdir, "/notebooks/R_scripts/_misc_functions.R"))
library(data.table)
library(strex)
library(listenv)
library(phyloseq)

```

#### Executing KrakenUniq Workflow (NEW)

Kraken parameters:
confidence = 0.3 - threshold between [0,1] for confidence in taxonomic assignment

```{r}
sampleIDs <- readRDS("data/interim/metadata/sample-id_WGS.rds")
slurm_out <-  glue("{wkdir}/.cluster_runs/{Sys.Date()}_KrakenUniq_paired")
data_out <- glue("{pdmbs_dir}/workflow/WGS/results/KrakenUniq")
class_reads_out <- glue("{scratch_dir}/WGS/KrakenUniq_classified_reads")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)
dir.create(class_reads_out, recursive = TRUE, showWarnings = FALSE)

for (sampleID in sampleIDs[3]) {
  if (!file.exists(
    glue("{data_out}/{sampleID}_KrakenUniq_report.tsv")
  )) {
    jobname <- glue("krakenUniq_{sampleID}")
    # pause if there are more than 9,999 slurm jobs submitted
    # check_slurm_overload()
    shell_do(
      glue(
        "sbatch",
        " --job-name={jobname}",
        " --output={slurm_out}/{jobname}.out",
        " --error={slurm_out}/{jobname}.err",
        " {src_dir}/shell_scripts/krakenUniq-paired.sh",
        " -s {sampleID}",
        # desired output directory for reports
        " -r {data_out}",
        # directory containing input FQ.gz files
        " -f /resnick/scratch/jbok/PDMBS/WGS/bowtie2_clean_reads",
        # directory for classified reads to be output
        " -c {class_reads_out}"
      )
    )
    # to prevent memory alloc issues
    Sys.sleep(0.02)
  }
}

```


#### Executing Kraken2 Workflow

Kraken parameters:
confidence = 0.3 - threshold between [0,1] for confidence in taxonomic assignment

```{r}
sampleIDs <- readRDS("data/interim/metadata/sample-id_WGS.rds")
slurm_out <-  glue("{wkdir}/.cluster_runs/{Sys.Date()}_Kraken2_paired")
data_out <- glue("{pdmbs_dir}/workflow/WGS/results/kraken2")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

for (sampleID in sampleIDs) {
  if (!file.exists(
    file.path(data_out, glue("UHGG_reports/{sampleID}_UHGG_report.tsv"))
  )) {a
    jobname <- glue("kraken2_{sampleID}")
    # pause if there are more than 9,999 slurm jobs submitted
    check_slurm_overload()
    shell_do(
      glue(
        "sbatch",
        " --job-name={jobname}",
        " --output={slurm_out}/{jobname}.out",
        " --error={slurm_out}/{jobname}.err",
        " {src_dir}/shell_scripts/kraken2-paired.sh",
        " -s {sampleID}",
        " -r /resnick/groups/MazmanianLab/jboktor/PDMBS/workflow/WGS/results/kraken2",
        " -f /resnick/groups/MazmanianLab/jboktor/PDMBS/workflow/WGS/clean_reads"
      )
    )
    # to prevent memory alloc issues
    Sys.sleep(0.02)
  }
}

```

### Kraken2 Workflow - RNASEQ Samples

```{r}
sampleIDs_rna <- readRDS(glue("{wkdir}/data/interim/metadata/sample-id_RNASEQ.rds"))
slurm_out <-  glue("{wkdir}/.cluster_runs/{Sys.Date()}_Kraken2_paired_RNASEQ")
data_out <- glue("{pdmbs_dir}/workflow/RNASEQ/results/kraken2")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)
run_list <- sampleIDs_rna %>%
  discard(
    ~ file.exists(
      file.path(data_out, glue("UHGG_reports/{.}_UHGG_report.tsv"))
    )
  )

for (sampleID in run_list) {
  jobname <- glue("kraken2_{sampleID}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " {src_dir}/shell_scripts/kraken2-paired.sh",
      " -s {sampleID}",
      " -r '/resnick/groups/MazmanianLab/jboktor/PDMBS/workflow/RNASEQ/results/kraken2'",
      " -f '/resnick/groups/MazmanianLab/jboktor/PDMBS/workflow/RNASEQ/clean_reads'"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.02)
}

```


#### Loading Kraken2 reports
Functions for reading in kraken2 tsv files
```{r}
load_kraken_file <- function(reportdir, report) {
  sampleID <- str_before_nth(string = report, pattern = "_", n = -2)
  df_report <- read_tsv(paste0(reportdir, report),
    col_names = FALSE,
    show_col_types = FALSE
  )
  if (nrow(df_report) > 0) {
    df_report %<>%
      as.data.table() %>%
      dplyr::rename(
        relative_abundance = X1,
        clade_counts = X2,
        taxon_counts = X3,
        taxon_rank_code = X4,
        NCBI_taxon_ID = X5,
        feature = X6
      ) %>%
      dplyr::mutate(sample_id = sampleID)
  }
}

aggregate_kraken_files <- function(input_files) {
  require(purrr)
  require(data.table)
  require(stringr)
  require(dplyr)
  require(readr)
  purrr::map(
    input_files,
    ~ load_kraken_file(report = ., reportdir = reportdir)
  ) %>% bind_rows()
}
```

```{r}
seq_methods <- c("WGS", "RNASEQ")
refDBlist <- c("RefSeqPlusPF", "UHGG")
outputdir <- glue("{wkdir}/data/interim/kraken_results")
dir.create(outputdir, showWarnings = FALSE)

# INITATE FUTURE-SLURM BACKEND FOR  PARALLEL PROCESSING
future::plan(
  future.batchtools::batchtools_slurm,
  template = 
    glue("{wkdir}/batchtools_templates/batchtools-workflow-submission.slurm.tmpl"),
  resources = list(
    name = "aggreg-krak",
    memory = "10G",
    ncpus = 2
  )
)

#' Loop through chunks of file names, RefDBs, and data-types
#' and send jobs to slurm
tic.clearlog()
for (seq_method in seq_methods) {
  for (refDB in refDBlist) {
    message(glue("Aggregating {seq_method} {refDB} Kraken2 files ..."))
    reportdir <- glue(
      "{pdmbs_dir}/workflow/{seq_method}",
      "/results/kraken2/{refDB}_reports/"
    )
    filepaths <- list.files(path = reportdir)
    print(length(filepaths))
    n_jobs <- ceiling(length(filepaths) / 200) # ~200 files/job
    kraken2_data <- listenv()

    tic(msg = glue("{seq_method} {refDB} Kraken2 files aggregating ..."))
    for (job in 1:n_jobs) {
      input_chunk <- chunk_func(filepaths, n_jobs)[[job]]
      kraken2_data[[job]] %<-% aggregate_kraken_files(input_chunk)
    }
    kraken2_table <- kraken2_data %>%
      as.list() %>%
      data.table::rbindlist()
    saveRDS(
      kraken2_table,
      glue(
        "{outputdir}/",
        "{Sys.Date()}_{seq_method}_{refDB}_kraken2_datatable.rds"
      )
    )
    toc()
  }
}

```

Create Phyloseq Profiles using RefSeqPlusPF and UHGG Kraken2 profiles

```{r}
outputdir <- glue("{wkdir}/data/interim/kraken_results")
rank_code_map <- list(
  "U" = "U",
  "R" = "R",
  "R1" = "D",
  "R2" = "P",
  "R3" = "C",
  "R4" = "O",
  "R5" = "F",
  "R6" = "G",
  "R7" = "S"
)
uhgg_to_refseq_rankcode <- function(rank) {
  rank_code_map[[rank]]
}
uhgg_to_refseq_rankcode <- Vectorize(uhgg_to_refseq_rankcode)
taxa_subsets <- c(
  "All" = "",
  "Species" = "S",
  "Genus" = "G",
  "Phylum" = "P"
)
meta_static <-
  readRDS(
    glue(
      "{wkdir}/data/interim/metadata/2023-06-06_static_metdata.rds"
      )) %>%
  mutate(sample_id = participant_id)
meta_long <-
  readRDS(
    glue("{wkdir}/data/interim/metadata/2023-06-06_longitudinal_metadata.rds"
    ))
rna_samples <- read.csv(
  glue(
    "{wkdir}/data/input/metadata/",
    "2021_v2-5release_0510/rna_sample_inventory.csv"
  ),
  header = TRUE
)
core_meta <- readRDS(
  glue("{wkdir}/data/interim/metadata/2023-06-06_core-metadata.rds")
)
rna_samples <- read.csv(
  glue(
    "{wkdir}/data/input/metadata/",
    "2021_v2-5release_0510/rna_sample_inventory.csv"
  ),
  header = TRUE
)
meta_long %<>%
  dplyr::select(-(colnames(core_meta)[-1]))
meta_rnaseq <- rna_samples %>%
  left_join(core_meta, by = join_by(participant_id)) %>%
  left_join(meta_long, by = c("participant_id", "visit_month"))

sample_metadata <- list(
  "WGS" = meta_static,
  "RNASEQ" = meta_rnaseq
)
saveRDS(
  sample_metadata,
  glue("{wkdir}/data/interim/metadata/{Sys.Date()}_phyloseq-metadata.rds")
)

for (seq_method in c("RNASEQ", "WGS")){
  for (refDB in c("UHGG", "RefSeqPlusPF")){
    for (taxa_rank_filter in names(taxa_subsets)) {
      message(glue("{seq_method} {refDB} {taxa_rank_filter}"))
      kraken_long <- readRDS(
        glue(
          "{outputdir}/",
          "2023-07-13_{seq_method}_{refDB}_kraken2_datatable.rds"
        )
      )
      kraken_wide  <- kraken_long %>%
      dplyr::select(
        feature, NCBI_taxon_ID, taxon_rank_code, sample_id, taxon_counts
        ) %>%
      pivot_wider(
        names_from = "sample_id",
        values_from = taxon_counts,
        values_fill = 0
      )
      if (refDB == "UHGG") {
        kraken_wide %<>%
          mutate(taxon_rank_code = uhgg_to_refseq_rankcode(taxon_rank_code))
      }

      # Construct Phyloseq Objects ---
      count_table <-
        kraken_wide %>%
        filter(grepl(taxa_subsets[[taxa_rank_filter]], taxon_rank_code)) %>%
        mutate(
          feature = gsub(" ", "__", feature),
          feature = glue("{taxon_rank_code}_{feature}")) %>%
        column_to_rownames(var = "feature") %>%
          select(-c(NCBI_taxon_ID, taxon_rank_code))

      physeq_count_table <- otu_table(count_table, taxa_are_rows = TRUE)
      physeq_meta <- sample_metadata[[seq_method]] %>%
        filter(sample_id %in% colnames(count_table)) %>%
        sample_data()
      rownames(physeq_meta) <- physeq_meta$sample_id

      cat("\nMetadata dimensions:", dim(physeq_meta), "\n")
      cat("\nCount table dimensions:", dim(count_table), "\n")
      ps <- phyloseq(physeq_count_table, physeq_meta)
      ps %>% print()
      saveRDS(ps, file = glue(
        "{wkdir}/data/processed/phyloseq_objects/raw/",
        "{Sys.Date()}_{seq_method}_{refDB}_{taxa_rank_filter}_phyloseq.rds"
      ))
    }
  }
}

```

