---
title: "WGS Read Processing"
eval: false
editor: source
author: "Joe Boktor"
date: '2022-12-23'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    column: page
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

A Cromwell [workflow](https://app.terra.bio/#workspaces/amp-pd-gcp-joeb/AMP-PD/job_history/bc0291d7-7f4e-44b7-a1ca-608cb565bd22) was employed in Google Cloud to select unmapped reads from GATK (4.0.1.2) processed CRAM files and return BAM files containing sequence metadata, reads, and sequencing flagstat metrics. 

#### The goals of this notebook are to: 
1. Download flagstat metric for GATK CRAM files and unmapped BAM file reads
2. Extract key sample-level sequence metadata from BAM files such as sequencing facility, date, flow cell, machine, etc.
3. Download and process unmapped bam files as follows
    a. Remove PCR and optical duplicate reads, secondary/supplementary alignments.
    b. Align BAM files with problematic Y-chromosome sequences and remove aligned reads.
    c. Attempt read alignment across sample pairs and record success.
    d. Conduct quality control filtering and read trimming for all reads, returning singular FASTQ files for each sample.


```{mermaid}
graph TD
    B[CRAM/BAM \n all reads]
    C[BAM unmapped]
    D[BAM unmapped \ncleaned]
    E[R1/R2 FASTQ]
    F[R1/R2 FASTQ QC'd]
    G[<b>R1/R2 FASTQ \nFINAL</b>]
    H[<b>R1/R2 FASTQ \nFINAL</b>]
    I[KrakenUniq Profiles]
    J[(<b>KrakenUniq MicrobialDB</b> \n human, archaea, bacteria, \nviral, UniVec_Core, \nEukaryotic \npathogen genomes)]
    K{KrakenUniq}

    B --> |Filter unmapped reads from\n BWA hg38 alignment| C
    C --> |Filter PCR/optical duplicates \n& secondary alignments| D
    D --> |Select paired reads| E
    E --> |minimum read length - 50bp \n phred theshold - 10 \n remove sex-chromosome alignments | F
    F --> |Bowtie2 T2T-CHM13+Y \nalignment with \n updated Human Genome| G

    H --> K
    J --> K
    K --> I

  linkStyle 0 stroke-width:4px,stroke:red
  linkStyle 3 stroke-width:4px,stroke:red
  linkStyle 4 stroke-width:4px,stroke:red
  linkStyle 7 stroke-width:4px,stroke:red

```



#### Downloading samtools flagstat metrics from Google Cloud Storage

```{r}
#| warning: false
#| message: false

library(glue)
library(purrr)
library(tibble)
library(magrittr)
library(dplyr)
library(stringr)
library(ggplot2)
library(stringr)
library(ggsci)
library(aplot)

home_dir <- "/central/groups/MazmanianLab/joeB"
pdmbs_dir <- paste0(home_dir, "/PDMBS")
ref_dir <- paste0(home_dir, "/Downloads/RefDBs")
wkdir <- paste0(pdmbs_dir, "/parkinsons-microbial-blood-signatures")
src_dir <- paste0(wkdir, "/notebooks")
source(paste0(src_dir, "/R_scripts/_load-core-pkgs.R"))
source(paste0(src_dir, "/R_scripts/_misc_functions.R"))
load_dot_env(glue("{wkdir}/.env"))
scratch_dir <- "/central/scratch/jbok/PDMBS"

```

### WGS Data Processing
Get a list of all google cloud storage file locations generated from workflow.
```{r}
tic()
bam_urls <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_WGS')}/**/*_unmapped.bam"),
  stdout_path = TRUE
)
flagstat_urls <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_WGS')}/**/*_flagstat.tsv"),
  stdout_path = TRUE
)
toc()

flagstat_cram_urls <- flagstat_urls %>%
  discard(grepl("unmapped-f4_flagstat.tsv", .))
flagstat_bam_urls <- flagstat_urls %>%
  keep(grepl("unmapped-f4_flagstat.tsv", .))

saveRDS(
  bam_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_unmapped-bam.rds")
)
saveRDS(
  flagstat_cram_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_flagstat-cram.rds")
)
saveRDS(
  flagstat_bam_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_URLs_flagstat-unmapped-f4.rds")
)

```


Next, we will do some quality control and ensure the proper samples are contained within our lists.

```{r}
# load in data
bam_urls <-
  readRDS(
    glue("{wkdir}/data/interim/gcs_locations/2022-12-23_URLs_unmapped-bam.rds")
  )
flagstat_cram_urls <-
  readRDS(
    glue("{wkdir}/data/interim/gcs_locations/2022-12-23_URLs_flagstat-cram.rds")
  )
flagstat_bam_urls <-
  readRDS(
    glue("{wkdir}/data/interim/gcs_locations/2022-12-23_URLs_flagstat-unmapped-f4.rds")
  )

# qc sample ids -  there should be 9833 samples
flagstat_cram_urls %>% length()
flagstat_bam_urls %>% length()
bam_urls %>% length()

#' Each list appears to contain one additional file (9844), so we check 
#' our data for duplicate samples
flagstat_cram_urls[duplicated(basename(flagstat_cram_urls))]

#' a duplicate file is caused by a sample which required a secondary run,
#' where the first run failed. So, we will selectively remove 
#' all primary data for samples which have a recorded secondary attempt.

```

Function to select url paths with the highest attempt number for each sample.
```{r}
gc_dedup_urls <- function(urls) {
  data.frame(
    url = urls
  ) %>%
    mutate(file_name = strex::str_after_last(url, "/")) %>%
    mutate(
      attempt_num = case_when(
        grepl("attempt", url) ~ str_before_last(url, "/") %>%
          str_after_last("/") %>%
          str_remove("attempt-"),
        TRUE ~ "1"
      ),
      attempt_num = as.numeric(attempt_num)
    ) %>%
    group_by(file_name) %>%
    slice_max(order_by = attempt_num, n = 1) %>%
    pull(url)
}

```

```{r}
bam_urls %<>% gc_dedup_urls
flagstat_cram_urls %<>% gc_dedup_urls
flagstat_bam_urls %<>% gc_dedup_urls
# double check
flagstat_cram_urls %>% length()
flagstat_bam_urls %>% length()
bam_urls %>% length()

sampleIDs <- stringr::str_remove(basename(bam_urls), "_unmapped.bam")
saveRDS(sampleIDs, "data/interim/metadata/sample-id_WGS.rds")
```

Downloading 01-CRAM flagstats

```{r}
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_download_01_CRAM-flagstat")
data_out <- glue("{pdmbs_dir}/workflow/WGS/flagstats/01_CRAM")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

for (url in flagstat_cram_urls) {
  if (file.exists(file.path(data_out, basename(url)))) {
    next
  }
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu=50MB",
      " --time=10:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.3)
}
```


Downloading 02-BAM-unmapped flagstats

```{r}
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_download_02_BAM-flagstat")
data_out <- glue("{pdmbs_dir}/workflow/WGS/flagstats/02_BAM-unmapped")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

for (url in flagstat_bam_urls) {
  if (file.exists(file.path(data_out, basename(url)))) {
    next
  }
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu=50MB",
      " --time=10:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.05)
}

```


Downloading unmapped BAM files

```{r}
slurm_out <-  glue("{wkdir}/.cluster_runs/gcs_download_BAM-unmapped")
data_out <- glue("/central/scratch/jbok/PDMBS/WGS/BAM-unmapped-raw")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

for (url in bam_urls) {
  if (file.exists(file.path(data_out, basename(url)))) {
    next
  }
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # pause if there are more than 9,999 slurm jobs submitted
  check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=2",
      " --nodes=1",
      " --mem-per-cpu=20GB",
      " --time=20:00:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.025)
}

```

Examining workflow success.

```{r}
# function to iterate through std-error files and report samples with issues
run_status <- list()
slurm_out <-  glue("{wkdir}/.cluster_runs/gcs_download_BAM-unmapped")
stderr_paths <- list.files(slurm_out) %>% keep(grepl(".err", .))
error_list <- c(
  "/etc/bashrc: fork: Resource temporarily unavailable",
  "error"
)
success <- c("Operation completed")
pb <- progress_bar$new(total = length(stderr_paths))

for (url in stderr_paths) {
   pb$tick()
  stderr <- read.table(file.path(slurm_out, url), sep = "\t", header = FALSE)
  if (any(str_detect(stderr$V1, success))) {
    run_status[[url]] <- "SUCCESS"
  } else if (any(stderr$V1 %in% error_list)) {
    run_status[[url]] <- "ERROR"
  } else {
    run_status[[url]] <- "UNKNOWN ERROR"
  }
}

cat(
  "Workflow status:",
  "\n SUCCESS:",
  run_status %>% keep(grepl("SUCCESS", .)) %>% length(),
  "\n ERROR:",
  run_status %>% keep(grepl("ERROR", .)) %>% length(),
  "\n UNKNOWN ERROR:", 
  run_status %>% keep(grepl("UNKNOWN ERROR", .)) %>% length(),
  "\n"
)
```


#### Read Quality Control 
##### Workflow:
1. Trim PCR & optical duplicates / secondary and supplementary alignments.
2. Collect sample sequencing metadata from BAM file
3. Run flagstat metrics

```{r}
sampleIDs <- readRDS("data/interim/metadata/sample-id_WGS.rds")
slurm_out <-  glue("{wkdir}/.cluster_runs/2022-12-27_BAM-QC")
data_out <- glue("{pdmbs_dir}/workflow/WGS/sequence_metadata")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)

for (sampleID in sampleIDs) {
  if (file.exists(
    file.path(data_out, glue("{sampleID}_sequencing_params.tsv"))
  )) {
    next
  }
  jobname <- glue("bam-qc_{sampleID}")
  # pause if there are more than 9,999 slurm jobs submitted
  check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/filter_bam_reads.sh",
      " -s {sampleID}",
      " -i /central/scratch/jbok/PDMBS/WGS/BAM-unmapped-raw",
      " -o /central/groups/MazmanianLab/joeB/PDMBS/workflow/WGS"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.02)
}

```


##### Workflow:
1. Convert Bam file into Fastq
2. Trim low quality reads and reads aligning to sex chromosome fragments using bbduk parameters as follows:
    + minlen     = 50 - discard reads below this length (post-trimming)
    + k          = 25 - kmer length  
    + mink       = 8 - allow a minimum kmer length of 8 bp at the end of a sequence
    + ref = (x&y_chromosome_fragments)- sex-chromosome fragment sequences that can mismap to microbial genes
    + ktrim      = r -  trim sequences that map to reference from the right (arbitrary selection)
    + hdist      = 1 -  max hamming distance for reference kmers - num mistmatch allowed 
    + qtrim      = rl -  trim R and L ends  at trimq (after kmer search) 
    + trimq      = 10 -  trim regions with Phred quality below 20 (q=2 ~ 1% error)

[confidence scoring algorithm](https://github.com/DerrickWood/kraken2/blob/master/docs/MANUAL.markdown#confidence-scoring)

[sex-chromosome fragments](https://github.com/briannachrisman/blood_microbiome/tree/main/public_data)


```{r}
sampleIDs <- readRDS("data/interim/metadata/sample-id_WGS.rds")
slurm_out <-  glue("{wkdir}/.cluster_runs/{Sys.Date()}_Fastq-QC_bbduk")
data_out <- glue("{pdmbs_dir}/workflow/WGS/clean_reads")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)

for (sampleID in sampleIDs)  {
  if (!file.exists(
    file.path(data_out, glue("{sampleID}_R2.fastq.gz"))
  )) {
    jobname <- glue("readqc-WGS_{sampleID}")
    # pause if there are more than 9,999 slurm jobs submitted
    check_slurm_overload()
    shell_do(
      glue(
        "sbatch",
        " --job-name={jobname}",
        " --output={slurm_out}/{jobname}.out",
        " --error={slurm_out}/{jobname}.err",
        " notebooks/shell_scripts/readQC.sh",
        " -s {sampleID}",
        " -w /central/scratch/jbok/PDMBS/WGS",
        " -o /central/groups/MazmanianLab/joeB/PDMBS/workflow/WGS"
      )
    )
    # to prevent memory alloc issues
    Sys.sleep(0.02)
  }
}

```


### RNA-SEQ Data Processing

```{r}
tic()
bam_urls_b1 <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_RNA_B1')}/**/*_unmapped.bam"),
  stdout_path = TRUE
)
flagstat_urls_b1 <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_RNA_B1')}/**/*_flagstat.tsv"),
  stdout_path = TRUE
)
bam_urls_b2 <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_RNA_B2')}/**/*_unmapped.bam"),
  stdout_path = TRUE
)
flagstat_urls_b2 <- shell_do(
  glue("gsutil ls {Sys.getenv('GCS_RNA_B2')}/**/*_flagstat.tsv"),
  stdout_path = TRUE
)
toc()

bam_urls <- gc_dedup_urls(c(bam_urls_b1, bam_urls_b2))
flagstat_bam_urls <- c(flagstat_urls_b1, flagstat_urls_b2) %>%
  discard(grepl("unmapped-f4_flagstat.tsv", .)) %>% 
  gc_dedup_urls()
flagstat_bamunmapped_urls <- c(flagstat_urls_b1, flagstat_urls_b2) %>%
  keep(grepl("unmapped-f4_flagstat.tsv", .)) %>%
  gc_dedup_urls()

length(bam_urls)
length(flagstat_bam_urls)
length(flagstat_bamunmapped_urls)

saveRDS(
  bam_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_RNASEQ_URLs_unmapped-bam.rds")
)
saveRDS(
  flagstat_bam_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_RNASEQ_URLs_flagstat-bam.rds")
)
saveRDS(
  flagstat_bamunmapped_urls,
  glue("{wkdir}/data/interim/gcs_locations/{Sys.Date()}_RNASEQ_URLs_flagstat-unmapped-f4.rds")
)
sampleIDs_rna <- stringr::str_remove(basename(bam_urls), "_unmapped.bam")
saveRDS(sampleIDs_rna, 
  glue("{wkdir}/data/interim/metadata/sample-id_RNASEQ.rds")
)
```

Downloading 01-RNASEQ-BAM-flagstat

```{r}
rnaseq_wkdir <- glue("{pdmbs_dir}/workflow/RNASEQ")
bam_urls <- readRDS(
  glue(
    "{wkdir}/data/interim/gcs_locations/",
    "2023-07-12_RNASEQ_URLs_unmapped-bam.rds"
  )
)
flagstat_bam_urls <- readRDS(
  glue(
    "{wkdir}/data/interim/gcs_locations/",
    "2023-07-12_RNASEQ_URLs_flagstat-bam.rds"
  )
)
flagstat_bamunmapped_urls <- readRDS(
  glue(
    "{wkdir}/data/interim/gcs_locations/",
    "2023-07-12_RNASEQ_URLs_flagstat-unmapped-f4.rds"
  )
)
```

```{r}
slurm_out <- glue("{wkdir}/.cluster_runs/{get_time()}_gcs_download_01-RNASEQ-BAM-flagstat")
data_out <- glue("{rnaseq_wkdir}/flagstats/01_BAM")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)
download_list <- flagstat_bam_urls %>% 
  discard(~file.exists(file.path(data_out, basename(.))))

for (url in download_list) {
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --mem=50M",
      " --time=30:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " {src_dir}/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.05)
}
```


Downloading 02-BAM-unmapped flagstats

```{r}
slurm_out <- glue("{wkdir}/.cluster_runs/{get_time()}_gcs_download_02-RNASEQ-unmapped-BAM-flagstat")
data_out <- glue("{rnaseq_wkdir}/flagstats/02_BAM-unmapped")
dir.create(slurm_out, recursive = TRUE)
dir.create(data_out, recursive = TRUE)
download_list <- flagstat_bamunmapped_urls %>% 
  discard(~file.exists(file.path(data_out, basename(.))))

for (url in download_list) {
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --mem-per-cpu=500M",
      " --time=20:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " {src_dir}/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.05)
}

```


Downloading unmapped BAM files

```{r}
slurm_out <-  glue("{wkdir}/.cluster_runs/{get_time()}_gcs_download_RNASEQ_BAM-unmapped")
data_out <- glue("/central/scratch/jbok/PDMBS/RNASEQ/BAM-unmapped-raw")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)
rnaseq_bam_urls <- bam_urls %>%
  discard(~ file.exists(
    file.path(
      glue("{pdmbs_dir}/workflow/RNASEQ/bam_unmapped"),
      gsub("_unmapped", "", basename(.))
    )
  )) %>%
  discard(~ file.exists(
    file.path(
      "/central/scratch/jbok/PDMBS/RNASEQ/BAM-unmapped-raw/",
      basename(.)
    )
    )
  ) 
  
for (url in rnaseq_bam_urls) {
  # filesize_gb <- 
  #   as.numeric(shell_do(glue("gsutil du {url} | cut -d' ' -f1"), stdout_path = TRUE))/1e9
  jobname <- glue("gcs_{tools::file_path_sans_ext(basename(url))}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --cpus-per-task=2",
      " --mem=10G",
      # " --mem={ceiling(filesize_gb)}G",
      " --time=10:00:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " {src_dir}/shell_scripts/download_gcs_url.sh",
      " -i {url}",
      " -o {data_out}/"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.01)
}

```



#### Read Quality Control 
##### Workflow:
1. Trim PCR & optical duplicates / secondary and supplementary alignments.
2. Collect sample sequencing metadata from BAM file
3. Run flagstat metrics

```{r}
sampleIDs_rna <- readRDS(glue("{wkdir}/data/interim/metadata/sample-id_RNASEQ.rds"))
slurm_out <-  glue("{wkdir}/.cluster_runs/{get_time()}_BAM-QC_RNASEQ")
data_out <- "/central/scratch/jbok/PDMBS/RNASEQ/sequence_metadata"
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out, recursive = TRUE, showWarnings = FALSE)
dir.create(
  "/central/scratch/jbok/PDMBS/RNASEQ/bam_unmapped", 
  recursive = TRUE, showWarnings = FALSE)
dir.create(
  glue("{pdmbs_dir}/workflow/RNASEQ/flagstats/03_BAM-unmapped-3328-filtered"), 
  recursive = TRUE, showWarnings = FALSE)
# run_list <- sampleIDs_rna %>%
#   discard(
#     ~ file.exists(
#       file.path(data_out, glue("{.}_sequencing_params.tsv"))
#     )
#   )
run_list <- sampleIDs_rna %>%
  discard(
    ~ file.exists(
      file.path("/central/scratch/jbok/PDMBS/RNASEQ/bam_unmapped", glue("{.}.bam"))
    )
  )
for (sampleID in run_list) {
  jobname <- glue("bam-qc_{sampleID}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " {src_dir}/shell_scripts/filter_bam_reads.sh",
      " -s {sampleID}",
      " -i /central/scratch/jbok/PDMBS/RNASEQ/BAM-unmapped-raw",
      " -o /central/scratch/jbok/PDMBS/RNASEQ"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.02)
}
```


##### Workflow:
1. Convert Bam file into Fastq
2. Trim low quality reads and reads aligning to sex chromosome fragments using bbduk parameters as follows:
    + minlen     = 50 - discard reads below this length (post-trimming)
    + k          = 25 - kmer length  
    + mink       = 8 - allow a minimum kmer length of 8 bp at the end of a sequence
    + ref = (x&y_chromosome_fragments)- sex-chromosome fragment sequences that can mismap to microbial genes
    + ktrim      = r -  trim sequences that map to reference from the right (arbitrary selection)
    + hdist      = 1 -  max hamming distance for reference kmers - num mistmatch allowed 
    + qtrim      = rl -  trim R and L ends  at trimq (after kmer search) 
    + trimq      = 20 -  trim regions with Phred quality below 20 (q=2 ~ 1% error)

[confidence scoring algorithm](https://github.com/DerrickWood/kraken2/blob/master/docs/MANUAL.markdown#confidence-scoring)

[sex-chromosome fragments](https://github.com/briannachrisman/blood_microbiome/tree/main/public_data)

```{r}
sampleIDs_rna <- readRDS(glue("{wkdir}/data/interim/metadata/sample-id_RNASEQ.rds"))
slurm_out_bbduk <-  glue("{wkdir}/.cluster_runs/{get_time()}_Fastq-QC_bbduk_RNASEQ")
data_out_bbduk <- glue("{pdmbs_dir}/workflow/RNASEQ/clean_reads")
stats_out_bbduk <- glue("{pdmbs_dir}/workflow/RNASEQ/stats_clean_reads")
dir.create(slurm_out_bbduk, recursive = TRUE, showWarnings = FALSE)
dir.create(data_out_bbduk, recursive = TRUE, showWarnings = FALSE)
dir.create(stats_out_bbduk, recursive = TRUE, showWarnings = FALSE)
dir.create("/central/scratch/jbok/PDMBS/RNASEQ/raw_reads", 
  recursive = TRUE, showWarnings = FALSE)
run_list <- sampleIDs_rna %>%
  discard(
    ~ file.exists(
      file.path(data_out_bbduk, glue("{.}_R1.fastq.gz"))
    )
  )

for (sampleID in run_list)  {
  jobname <- glue("readqc_{sampleID}")
  # check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --output={slurm_out_bbduk}/{jobname}.out",
      " --error={slurm_out_bbduk}/{jobname}.err",
      " {src_dir}/shell_scripts/readQC.sh",
      " -s {sampleID}",
      " -w /central/scratch/jbok/PDMBS/RNASEQ",
      " -o /central/groups/MazmanianLab/joeB/PDMBS/workflow/RNASEQ"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(0.02)
}

```

### WGS/RNA Bowtie2 Alignment against T2T-CH13+newY

Downloading CHM13v2

```{r, eval = FALSE}
shell_do("cd /central/groups/MazmanianLab/joeB/Downloads/RefDBs")
shell_do('curl -OJX GET "https://api.ncbi.nlm.nih.gov/datasets/v2alpha/genome/accession/GCF_009914755.1/download?include_annotation_type=GENOME_FASTA,GENOME_GFF,RNA_FASTA,CDS_FASTA,PROT_FASTA,SEQUENCE_REPORT&filename=GCF_009914755.1.zip" -H "Accept: application/zip"')

# ref_genomes/T2T-CHM13v2.0/GCF_009914755.1/GCF_009914755.1_T2T-CHM13v2.0_genomic.fna

```

Downloading updated [T2T+Y](https://github.com/marbl/CHM13)

```{r, eval = FALSE}
t2t_genome_dir <- glue("{home_dir}/Downloads/RefDBs/ref_genomes/T2T-CHM13v2.0_plusY")
shell_do(
  glue(
    "wget -P {t2t_genome_dir}",
    " https://s3-us-west-2.amazonaws.com/human-pangenomics/T2T/CHM13/assemblies/analysis_set/chm13v2.0.fa.gz"
  )
)
shell_do(
  glue(
    "gunzip",
    " {t2t_genome_dir}/chm13v2.0.fa.gz"
  )
)

```

Building a bowtie2 index for CHM13v2 (with updated Y chromosome)

```{r, eval = FALSE}
bowtiew_ind_dir <- glue("{ref_dir}/bowtie2_indices/T2T-CHM13v2.0_plusY_genomic")
shell_do("mkdir -p {bowtiew_ind_dir}")
slurm_shell_do(
  glue(
    "bowtie2-build",
    " {ref_dir}/ref_genomes/T2T-CHM13v2.0_plusY/chm13v2.0.fa",
    " {bowtiew_ind_dir}/T2T-CHM13v2pY"
  ),
   memory = "100G",
   walltime = 10800
)

```


Executing Bowtie2 Alignment.

```{r, eval = FALSE}

bowtiew_ind <-
  glue(
    "{home_dir}/Downloads/RefDBs/bowtie2_indices/",
    "T2T-CHM13v2.0_plusY_genomic/T2T-CHM13v2pY"
  )

seq_method <- "WGS"
bowtie_temp <- glue("{scratch_dir}/{seq_method}/bowtie2_temp")
bowtie_cleanreads_v2 <- glue(
  "{scratch_dir}/{seq_method}/",
  "bowtie2_clean_reads"
)
bowtie_stats <- glue(
  "{pdmbs_dir}/workflow/{seq_method}/",
  "stats_clean_reads_bowtie2_CHM13"
)
dir.create(bowtie_temp, recursive = TRUE, showWarnings = FALSE)
dir.create(bowtie_cleanreads_v2, recursive = TRUE, showWarnings = FALSE)
dir.create(bowtie_stats, recursive = TRUE, showWarnings = FALSE)

fq_paths <- list.files(glue("{pdmbs_dir}/workflow/{seq_method}/clean_reads"),
  full.names = TRUE
)

batchtools_params_bowtie2 <- data.frame(
  "paths" = fq_paths
  ) %>%
  mutate(sample_id = basename(paths) %>% str_remove("_R[1-2].fastq.gz")) %>%
  mutate(paired_end = case_when(
    grepl("_R1.fastq.gz", paths) ~ "inpath_f",
    grepl("_R2.fastq.gz", paths) ~ "inpath_r",
    TRUE ~ "ERROR"
  )) %>%
  pivot_wider(names_from = "paired_end", values_from = "paths") %>%
    mutate(
      outpath_sam = glue("{bowtie_temp}/{sample_id}.sam"),
      outpath_fqs = glue("{bowtie_cleanreads_v2}/{sample_id}_R%.fastq.gz"),
      outpath_metrics = glue("{bowtie_stats}/{sample_id}_aln_metrics.txt"),
      outpath_stderr = glue("{bowtie_stats}/{sample_id}_stderr.txt"),
      bowtiew_ind = bowtiew_ind,
      threads = 8,
      wkdir = wkdir
    ) %>%
    # USE TO RERUN FAILURES ONLY (See below)
    filter(sample_id %in% samples_to_rerun) %>%
    select(-sample_id)

batchtools_params_bowtie2 %>% glimpse()


# TO RERUN FAILURES
# 1) look for temp SAM files that failed to be removed
# 2) look for stderr files that have error messages

# configure registry ----
cluster_run <- glue("{get_time()}_Bowtie2_CHM13_{seq_method}")
message("\n\nRUNNING:  ", cluster_run, "\n")
breg <- makeRegistry(
  file.dir = glue(
    "{wkdir}/.cluster_runs/",
    cluster_run
  ),
  seed = 42
)
breg$cluster.functions <- batchtools::makeClusterFunctionsSlurm(
  template = glue("{wkdir}/batchtools_templates/batchtools.slurm.tmpl"),
  scheduler.latency = 0.1,
  fs.latency = 1
)
# Submit Jobs ----
jobs <- batchMap(
  fun = shell_do_bowtie2,
  args = batchtools_params_bowtie2,
  reg = breg
)
jobs[, chunk := chunk(job.id, chunk.size = 3)]
print(jobs[, .N, by = chunk])

submitJobs(jobs,
  resources = list(
    walltime = 172800,
    memory = 10000,
    ncpus = 8,
    max.concurrent.jobs = 9999
  )
)

```


Checking log files for any errors

```{r}

stderr_files <- list.files(
  glue("{pdmbs_dir}/workflow/{seq_method}/stats_clean_reads_bowtie2_CHM13"),
  pattern = "_stderr.txt",
  full.names = TRUE
)

future::plan("multisession", workers = 8)
stderr_failures <- stderr_files %>%
  purrr::set_names() %>%
  furrr::future_map(~ suppressWarnings(has_error_message(.)))
future::plan("sequential")

stderr_failures %>% unlist() %>% table()
stderr_failures_paths <- stderr_failures %>%
  unlist() %>%
  keep(. == TRUE) %>%
  names()

stderr_failures_names <- stderr_failures_paths %>%
  basename() %>%
  strex::str_before_last("_stderr.txt")
temp_remove_failures <- list.files(bowtie_temp) %>%
  fs::path_ext_remove()

samples_to_rerun <- c(stderr_failures_names, temp_remove_failures)

```

```{r}
sessionInfo()
```
