---
title: "SourceTracker analysis data processing"
editor: source
eval: false
author: "Joe Boktor"
date: '2022-12-29'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    theme: cosmo
    column: page
    fig-align: right
    fig-height: 8
    fig-width: 14 
---


Samples to include in human body site deconvolution:
PMID:

```{r}
pdmbs_dir <- "/central/groups/MazmanianLab/joeB/PDMBS"
wkdir <- paste0(pdmbs_dir, "/parkinsons-microbial-blood-signatures")
source(paste0(wkdir, "/notebooks/R_scripts/_misc_functions.R"))
source(paste0(wkdir, "/notebooks/R_scripts/_load-core-pkgs.R"))
library(janitor)
library(progress)

```


```{r}
# read in metadata
hmp_manifest <-
  readr::read_tsv("data/input/metadata/HMP/hmp_manifest_19484f718e.tsv",
    col_names = T
  )
# keep only HMP download paths
hmp_manifest_processed <- hmp_manifest %>%
  mutate(
    urls = stringr::word(urls, 1, sep = ","),
    sample_name = gsub(".tar.bz2", "", basename(urls))
  )

unwanted_dups <- 
  c(
    "91319e642fdd8a6e3b059cfb05ab9545",
    "91319e642fdd8a6e3b059cfb05ab9c37",
    "91319e642fdd8a6e3b059cfb05ab9eeb",
    "91319e642fdd8a6e3b059cfb05ab77fd"
    )

# N. of unique sample names
hmp_manifest_processed %>% dim()
hmp_manifest_processed$sample_name %>% unique() %>% length()

hmp_manifest_processed %<>% 
# removing HMP V1 files
  filter(!grepl("hmwgsqc/v1", urls), 
  # removing duplicate samples
  file_id %nin% unwanted_dups
  )
  
hmp_manifest_processed %>% dim()
hmp_manifest_processed$sample_name %>% unique() %>% length()

# format for TERRA workflow
hmp_manifest_formated <- hmp_manifest_processed %>% 
  relocate(sample_name) %>% 
  rename(
    sample_id2 = sample_id,
    "entity:sample_id" = sample_name
    )
write.table(hmp_manifest_formated, 
  file = "data/input/metadata/HMP/hmp_manifest_filtered-TERRA-formatted.tsv", 
  sep="\t", quote = FALSE, row.names = FALSE
  )
write.table(hmp_manifest_formated$urls, 
  file = "data/input/metadata/HMP/hmp_manifest_filtered-URLS-ONLY.tsv", 
  sep="\t", quote = FALSE, row.names = FALSE, col.names = FALSE
  )

```

Download files into google storage bucket

```{r, eval=FALSE}
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_upload_HMP2")
dir.create(slurm_out, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(hmp_manifest_processed))) {
  url <- hmp_manifest_processed[i, ]$urls
  filesize_mb <- ceiling(hmp_manifest_processed[i, ]$size / 1e6)
  file_exists <- shell_do(
    glue(
      "gsutil ls gs://shotgun-metagenomic-data/HMP2/{basename(url)} | wc -l"
    ),
    stdout_path = TRUE
  )
  if (file_exists == "1") {
    next
  }
  jobname <- glue("gcs_{gsub('.tar.bz2', '', basename(url) )}")
  cat("\n Processing: ", basename(url), "\n")
  # pause if there are more than 9,999 slurm jobs submitted
  check_slurm_overload()
  shell_do(
    glue(
      "sbatch",
      " --job-name={jobname}",
      " --ntasks=1",
      " --nodes=1",
      " --mem-per-cpu={filesize_mb*3}MB",
      " --time=3-12:00:00",
      " --output={slurm_out}/{jobname}.out",
      " --error={slurm_out}/{jobname}.err",
      " notebooks/shell_scripts/upload_gcs_HMP.sh",
      " -p {url}",
      " -f {basename(url)}"
    )
  )
  # to prevent memory alloc issues
  Sys.sleep(10)
}
```


Extract downloads with errors and re-run

```{r}

run_status <- list()
slurm_out <- glue("{wkdir}/.cluster_runs/gcs_upload_HMP2")
stderr_paths <- list.files(slurm_out, full.names = TRUE) %>% keep(grepl(".err", .))
pb <- progress_bar$new(total = length(stderr_paths))

for (url in stderr_paths) {
  pb$tick()
  print(url)
  stderr <- read.table(url, sep = "\t")
  if (stderr$V1 %>% grepl("Operation completed over 1 objects", .) %>% any()) {
    run_status[[url]] <- "SUCCESS"
  } else {
    run_status[[url]] <- "ERROR"
  }
}

cat(
  "Workflow status:",
  "\n SUCCESS:",
  run_status %>% keep(grepl("SUCCESS", .)) %>% length(),
  "\n ERROR:",
  run_status %>% keep(grepl("ERROR", .)) %>% length(),
  "\n"
)

```


Required data for analysis
1. Sample Biomass prior to mixing and library prep
2. Cohort 
3. Sequencing Run
4. Library Prep Lot
5. Sequencing Location
6. Flow Cell


Read in sample sequencing metadata, previously extracted from BAM file headers. 

```{r}





as.integer(file_exists) == 0

```